### 2017-06-12 paper/paper.tex
Start of the paper.

\documentclass[prd,aps,letter,twocolumn,floatfix,notitlepage]{revtex4-1}

%% packages
\usepackage{graphicx,psfrag}
\usepackage{mathrsfs,amsmath,amsfonts,amssymb, bm}
\usepackage{multirow}
\usepackage{comment,hyperref}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algpseudocode} % in place of algorithmic. Conflicts with revtex4 unless [H] option passed
\usepackage{setspace} % for pleasent spacing in algorithm

%% macros
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\bsube}{\begin{subequations}}
\newcommand{\esube}{\end{subequations}}

\def\l{\ell}
\def\blambda{{\bm \lambda}}
\def\bLambda{{\bm \Lambda}}
\def\blam{\bar{\lambda}}
\def\Lam{\Lambda}
\def\Lamt{\tilde{\Lambda}}
\def\dLamt{\tilde{\delta{\Lambda}}}
\def\bmu{{\bm \mu}}
\def\btheta{{\bm \theta}}

%% macros for comments
\usepackage{color}
%% colors
\definecolor{cyan}{rgb}{0,0.9,0.9}
\definecolor{orange}{rgb}{0.9,0.5,0}
\definecolor{magenta}{rgb}{1,0,1}
\definecolor{purple}{rgb}{0.8,0.4,0.8}
\definecolor{gray}{rgb}{0.8242,0.8242,0.8242}
%%
\newcommand{\bs}[1]{{\textcolor{green}{\texttt{SB: #1}} }}
\newcommand{\bl}[1]{{\textcolor{blue}{\texttt{BL: #1}} }}
\newcommand{\cg}[1]{{\textcolor{orange}{{CG: #1}} }}
\newcommand{\cvdb}[1]{{\textcolor{purple}{\texttt{CVDB: #1}} }}
\newcommand{\red}[1]{\textcolor{red}{#1}}
%% +++++++++++++++++++++++++++++++++++++++++++++++++++++++
\begin{document}

\title{Surrogate model for aligned-spin binary neutron star waveforms using gaussian process regression}

\author{
Benjamin D. Lackey$^1$, 
Michael P\"{u}rrer$^1$
Andrea Taracchini$^1$
}
\affiliation{
$^1$Max Planck Institute for Gravitational Physics, Albert Einstein Institute, D-14476 Golm, Germany
}

\date{\today}

\begin{abstract}

Construct 5d hierarchical surrogate model using reduced bases and GPR. Includes $\ell=2,3,4$, dynamical tides, 
and tapering fit to NR.

Provide framework for building high dimensional surrogate from NR.

\end{abstract}

\pacs{
  % 04.25.D-,   % numerical relativity
  % 
  04.30.Db,   % gravitational wave generation and sources
  % 04.40.Dg,   % Relativistic stars: structure, stability, and oscillations
  % 04.70.Bw,   % classical black holes
  95.30.Sf,     % relativity and gravitation
  % 
  % 95.30.Lz,   % Hydrodynamics
  % 97.60.Jd    % Neutron stars
  % 97.60.Lf    % black holes (astrophysics)
  % 98.62.Mw    % Infall, accretion, and accretion disks
}

%%
\maketitle

%% Template for figures
%% \begin{figure}[t]
%%   \centering 
%%     \includegraphics[width=0.49\textwidth]{../fig/ }
%%     \caption{ }
%%  \label{fig: }
%% \end{figure}

%% Template for tables
%% \begin{table}[t]
%%   \centering    
%%   \caption{ }
%%   \begin{tabular}{ccc}        
%%     \hline
%%     col$1$ & col$2$ & col$3$ \\
%%     \hline
%%     &  &  \\
%%     \hline
%%   \end{tabular}
%%  \label{tab: }
%% \end{table}

\red{Change arrow vectors to bold vectors. ${\bm x}^i$ (bold) for example $i$. $x_j$ (not bold) for coordinate/parameter/feature $j$. 
$x_j^i$ (not bold) for feature $j$ of example $i$.}

%% ________________________________________________________
\section{Introduction}

We organize the paper as follows. In Section...

\textit{Conventions:} Unless explicitly stated, we use units where $G=c=1$. To agree with the machine learning literature, we will refer to the waveform parameters (features) as $\vec x$ and the hyperparameters in the Gaussian Process Regression model as $\vec\theta$.

%% ________________________________________________________
\section{Aligned-spin, dynamical tides EOB model}

--Key features of dynamical tides EOB model

--Total mass is scaled out. This is fine, even up to zero amplitude point. Estimate the error in this approximation.

--Tapering fit to NR.

--Universal relations to relate all matter parameters to quadrupolar $\Lambda$.

--Results in 5 waveform parameters $\vec x=\{q, S_1, S_2, \Lambda_1, \Lambda_2\}$.

%% ________________________________________________________
\section{Evaluating waveform model accuracy}

In the sections below, we will use several criteria to choose waveforms samples and evaluate the accuracy of the 
surrogate waveform model. We briefly discuss them here.

--Phase error. (needs to be significantly smaller than tidal effects) Use this for order of magnitude estimate of required
accuracy.

--distinguishability

--Mismatch

As another common measure of the surrogate model accuracy, we examine the mismatch between 
the surrogate model and the original EOB waveform.
The mismatch represents the loss in signal-to-noise ratio that would result 
from using the surrogate model instead of the original EOB waveform. 
It is defined by the deviation from a perfect overlap after aligning the two waveforms
using the time and phase free parameters $t_0$ and $\phi_0$:
\begin{equation}
\mathcal{M} = 1 - \max_{t_0, \phi_0} \frac{(h_{\rm EOB}, h_{\rm Sur})} {\sqrt{(h_{\rm EOB}, h_{\rm EOB}) (h_{\rm Sur}, h_{\rm Sur})}}.
\end{equation}
The inner product here  is the integral of the Fourier transformed waveforms $\tilde h(f)$ weighted by the noise power spectral 
density (PSD) $S_n(f)$ of the detector:
\begin{equation}
(h_1, h_2) = 4 \Re \int_{f_{\rm low}}^{f_{\rm high}} \frac{\tilde h_1(f) \tilde h^*_2(f)} {S_n(f)} df.
\end{equation}

--Fisher matrix systematic error. (See spin NSBH paper.)

%% ________________________________________________________
\section{Surrogate model}

The problem of constructing a surrogate model for a computer simulation that takes parameters 
$\vec x$ and outputs a vector of data $\vec y$ is the following: construct approximate functions $f_j(x)$
for each quantity $y_j$ using a training set of precomputed examples $\{\vec x^i, \vec y^i\}$. For the problem here, 
our parameters are the intrinsic waveform parameters $\vec x$ and the data is the waveform frequency series $\tilde h(f_j)$.

--Frequency-domain so Fourier transform can be done offline and all issues related to tapering, etc. are taken care of.

We aim to construct a sufficiently accurate waveform surrogate using as few waveform evaluations as possible.
To do this we perform the following steps:
\begin{enumerate}
\item Select the initial training set parameters.

\item Precondition an initial training set of waveforms, and decompose them into functions that smoothly vary
as a function of waveform parameters $\vec x$.

\item Compress the waveform output using a greedy method for generating a reduced orthonormal basis and 
generate empirical interpolating functions and nodes form this reduced space, to minimize the number of interpolations
required.

\item Interpolate the waveform at the empirical nodes as a function of the waveform parameters $\vec x$ using GPR.

\item With the resulting surrogate model and GPR uncertainty estimate, perform uncertainty sampling to choose the 
next waveforms to add to the training set.

\item Repeat steps 2--5 to reconstruct the surrogate model with the updated training set. This can be repeated
if necessary.
\end{enumerate}

\subsection{Initial training set design}

In constructing a surrogate model, we will need to interpolate several quantities as a function of the five
waveform parameters ${\bm x}$.
Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) 
require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality; 
the number of samples grows exponentially with the dimension $d$ ($N^d$ for $N$ samples per dimension).
In the 5-dimensional problem discussed here, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
At 10 minutes--1 hour per EOB waveform on a standard CPU, this is at the limit of what is reasonable.

However, there are other methods such as Gaussian Process Regression (GPR), discussed below, that allow nonuniform designs.
One such standard design is a Latin Hypercube Design (LHD). An LHD with $N$ samples divides each dimension uniformly into $N$
grid points per dimension for a total of $N^d$ possible locations. However, for each dimension each of the $N$ values is sampled 
exactly once. This has the convenient property that the samples are non-collapsing; if one of the parameters is much less 
significant than other parameters, samples are not wasted on that parameter, as they would be with a uniform grid. And, a projection
onto a subspace is still a LHD. [[Also, each dimension is uniformly sampled which is not the case for MC sampling.]]

As an example of the usefulness of noncollapsing designs, the tidal parameters have a minimal impact on the waveform
at low frequencies, so it wouldn't make sense to densely sample a grid of tidal parameters at low frequencies. However, near
the merger the tidal parameters can be a important as the spin parameters. One therefore might want more samples for the tidal
parameters and less samples for the spin. However, due to the expense of the wavefroms we must use the same grid of waveforms 
for low and high frequencies. A LHD, with its noncollapsing property, bypasses this problem.

For an LHD there are $(N!)^d$ possible designs, and a standard requirement is that the LHD be space-filling, meaning that the 
points are as far apart as possible from each other. (See [[cite]] for a review of methods for optimizing the placement of samples.) 
The measure used here is that we maximize the minimum Euclidean distance between any two samples [[cite]]. We do this by 
sampling $xxx$ random designs and choosing the one with the maximum minimum distance.

For a smooth approximately monotonic function in $d$-dimensions, standard lore is that $10d$ samples are needed to interpolate 
a function with GPR [[cite]] [[To what accuracy]]. We therefore construct our initial training set with 50 waveform samples, and use 
these waveforms to construct the reduced basis and perform interpolation with GPR. Once an initial surrogate model is constructed,
we will use the GPR error estimate to choose new waveforms to add to the training set, and update the surrogate model.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{trainingset3d.png}\\
\includegraphics[width=0.49\textwidth]{trainingset2d.png}
\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, S_1, S_2\}$ (top)
and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom). The 160 black circles were used for the initial
training set (32 corner points + 128 LHD points) that generated the initial surrogate. The 200 red squares were generated
using uncertainty sampling from the GPR error estimate of the RMS phase error at the empirical nodes. 
}
\label{fig:LHD}
\end{figure}

\subsection{Conditioning and decomposing the training set waveforms}

--Messy details of preconditioning lalsimulation waveforms to make them smooth functions of parameters ${\bm x}$.
(Rescale with total mass to convert to geometric units, time shift to align at max amplitude or specific frequency, 
taper beginning and end, pad end with zeros, Fourier transform, truncation, resampling, zeroing initial phase, ...)

--Show plot (?) to validate that waveform is properly conditioned by plotting amplitude/phase as function of waveform parameters
at specific frequencies and show that they're smooth functions.

\begin{equation}
\tilde h(f, x) = A(f, x) e^{i\Phi(f, x)}
\end{equation}

Because the waveform $h$ is an oscillatory function of $f$ and ${\bm x}$ that is difficult to interpolate, we instead construct 
surrogates for the amplitude $A(f, x)$ and phase $\Phi(f, x)$. These are smooth functions. However, for aligned-spin waveforms
these functions are already approximately known from the 3.5PN TaylorF2 waveform 
$\tilde h_{\rm F2}(f, x) = A_{\rm F2}(f, x) e^{i\Phi_{\rm F2} (f, x)}$, where
\begin{equation}
A_{\rm F2} = simple equation,
\end{equation}
and
\begin{equation}
\Phi_{\rm F2} = not simple equation.
\end{equation}
We therefore only have to construct surrogates for the difference between EOB and TaylorF2:
\begin{align}
\Delta\ln(A) &= \ln(A/A_{\rm F2})\\
\Delta\Phi & = \Phi - \Phi_{\rm F2}
\end{align}
These quantities are shown in Fig.~\ref{fig:dh} for the training set wavefroms. We note that for small differences 
between the EOB and TaylorF2 waveform, $\Delta\ln A$ is approximately the fractional difference $A/A_{\rm F2} - 1$
between the EOB and TaylorF2 waveform. We use the log-ratio instead of the ratio, because it guarantees that any
error in interpolating this quantity doesn't result in a negative amplitude. Also, because waveforms tend to deviate from 
each other near the end of the inspiral exponentially, depending on each waveform's merger frequency the log captures
this behavior. [[Is this true for EOB? We'll find out.]]

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
\caption{Training set}
\label{fig:dh}
\end{figure}

We show the range of values of the phase $\Phi$ over the training set $\Phi_{\rm range} = \Phi_{\rm max} - \Phi_{\rm min}$ 
as well as the range of values of the phase residual $\Delta\Phi$ over the training set 
$\Delta\Phi_{\rm range} = \Delta\Phi_{\rm max}-\Delta\Phi_{\rm min}$ in Fig.~\ref{fig:phaserange}. 
The range in $\Phi$ can be larger than 1000~radians, meaning
that if we require 0.1~radian accuracy, we need a fractional error of $<10^{-4}$. The range of $\Delta\Phi$ is 1--2 orders 
of magnitude smaller, and as a result, the interpolation accuracy requirements are 1--2 orders of magnitude smaller as well.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{phaserange.png}
\caption{Range of phase values for training set waveforms. The range is a factor of $\gtrsim 100$ larger for $\Phi$ than for $\Delta\Phi$, 
meaning that the interpolation accuracy requirements are at least $\sim 2$ orders of magnitude less for $\Delta\Phi$.}
\label{fig:phaserange}
\end{figure}

\subsection{Compression of waveform data with reduced bases and empirical interpolation method}

For many surrogate problems a function $f({\bm x})$ is sought to interpolate a scalar quantity $y$. For the
problem here, the outputs are the quantities $\Delta\ln A$ and $\Delta\Phi$ sampled at the frequencies $f_j$. 
One could interpolate $\Delta\ln A$ and $\Delta\Phi$ as a function of ${\bm x}$ at each sample $f_j$, 
however, this would be prohibitively expensive. Instead we use the standard method of constructing an
orthonormal reduced basis to compress each training set waveform into a small set of coefficients, 
and then using the empirical interpolation method
to optimize the placement of a corresponding number of empirical nodes $F^A_j$ and $F^\Phi_j$. One then interpolates
$\Delta\ln A$ and $\Delta\Phi$ as a function of ${\bm x}$ at these nodes. The full waveform can then be reconstructed by 
multiplying the value at these nodes by the the corresponding empirical interpolating functions $B_j^A(f)$ and $B_j^\Phi(f)$. 
The methods used here are presented in Refs.[[cite]], and we use the exact implementation described in 
Ref.~[[LackeyBernuzziGalley2017]]. We briefly describe the details.

In order to construct a reduced basis from the training set waveforms we use the greedy method~[[LackeyBernuzziGalley2017]
to choose a subset of the training set waveforms and orthonormalize them with the Grahm-Schmidt procedure. The end result
is similar to singular value decomposition, but one can make a one-to-one mapping between each chosen waveform and a 
corresponding basis function. The quantities $\Delta\ln A$ and $\Delta\Phi$ can be approximately described by
\begin{subequations}
\label{eq:basis1}
\begin{align}
\Delta\ln A(f; \btheta) \approx {} & \sum_{i=1}^{N_A} e^A_i(f) c^A_i(\btheta)\\
\Delta\Phi(f; \btheta) \approx {} & \sum_{i=1}^{N_\Phi} e^\Phi_i(f) c^\Phi_i(\btheta),
\end{align}
\end{subequations}
where the coefficients are
\begin{subequations}
\label{eq:basiscoeffs1}
\begin{align}
c_i^A(\btheta) & = \langle e_i^A(\cdot) , \Delta\ln A(\cdot; \btheta) \rangle \\
c_i^\Phi(\btheta) & = \langle e_i^\Phi(\cdot) , \Delta\Phi(\cdot; \btheta) \rangle  .
\end{align}
\end{subequations}

In the empirical interpolation method, one chooses a set of frequencies $F_j$ [[to ... minimize impact of interpolation error]].
The reduced basis can then be inverted to construct empirical interpolating functions $B_j(f)$ [[details...]].
\begin{subequations}
\label{eq:empinterpFinal}
\begin{align}
\Delta\ln A_S(f; \btheta ) & = \sum_{j=1}^{N_A} B_j^A(f) \Delta\ln A(F^A_j; \btheta) \\
\Delta\Phi_S(f; \btheta ) & = \sum_{j=1}^{N_\Phi} B_j^\Phi(f) \Delta\Phi(F^\Phi_j; \btheta)
\end{align}
\end{subequations}
The interpolating functions for $\Delta\ln A$ and $\Delta\Phi$ are shown in Fig.~\ref{fig:B}.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{Bamp.png}
\includegraphics[width=0.49\textwidth]{Bphase.png}
\caption{Top: $B^{\rm Amp}_j(Mf)$. Bottom: $B^{\rm Phase}_j(Mf)$. Black dots indicate the frequency of the empirical nodes $MF_j$.}
\label{fig:B}
\end{figure}

\subsection{Interpolating over waveform parameter space}

We use Gaussian process regression to fit $\Delta\ln A$ and $\Delta\Phi$ at the empirical nodes as a function of
the waveform parameters. In GPR a function $f(\vec x)$ is described in terms of a mean $m(\vec x)$ and 
covariance $k(\vec x, \vec x')$ between points [[cite RassmussenWilliamsChapter2, IntroArxivPaper]].
\begin{equation}
f(\vec x) \approx \mathcal{GP}(m(\vec x), k(\vec x, \vec x')).
\end{equation}
The mean $m(x)$ can be some parameterized function which is the standard version of regression, 
while $k(\vec x, \vec x')$ is a kernel with tunable hyperparameters that describes the covariance between
the point $\vec x$ and the sampled point $\vec x'$. This covariance can be used to describe features of the function 
not captured by the parameterized mean as well as genuine noise in the data ${\bm y}$. 
For the problem here, we have already subtracted the 
TaylorF2 waveform from EOB waveform, so we set $m=0$ and model the remainder terms purely in terms 
of the covariance $k$.

In a zero-mean Gaussian process, the data and function $({\bm y}, f_*)$ at the training set and new points $({\bm x}, x_*)$ are
drawn from a multivariate normal distribution
\begin{equation}
\label{eq:gaussian}
\begin{bmatrix}
{\bm y} \\
f_* \\
\end{bmatrix}
\sim \mathcal{N}
\left({\bm 0}, 
\begin{bmatrix}
K & K_*^T \\
K_* & K_{**} \\
\end{bmatrix}
\right)
\end{equation}
where $K = K({\bm x}^i, {\bm x}^j)$ is a matrix, $K_* = K({\bm x}^i, {\bm x}_*)$ is a vector, and $K_{**} = K({\bm x}_*, {\bm x}_*)$ is a scalar.

The conditional probability for $f_*$ given the training set examples ${\bm y}$ with hyperparameters $\theta$ is also a Gaussian
\begin{equation}
f_* | {\bm x}, x_*, {\bm y}, {\bm \theta} \sim \mathcal{N}(\bar f_*, {\rm var}(f_*))
\end{equation}
where the mean and variance are
\begin{align}
\label{eq:mean}
\bar f_* &= K_*^T K^{-1} y \\
\label{eq:var}
{\rm var}(f_*) &= K_{**} - K_*^T K^{-1} K_*.
\end{align}
Eq.~\eqref{eq:mean} is the estimate of the function, and Eq.~\eqref{eq:var} is the estimate of the uncertainty, where we note that the variance does not depend on the data $\bm{y}$.

We use a radial kernel which expresses the covariance in terms of a distance $r$ between points
\begin{equation}
r^2 = (x-x')^T M (x-x'),
\end{equation}
where we choose the matrix M to be diagonal
\begin{equation}
M = {\rm diag}(\ell_1^{-2}, \ell_2^{-2}, \dots, \ell_d^{-2}).
\end{equation}
The tunable hyperparameters $\ell_i$ represent the length scale over which the function $f(x)$ varies in 
each coordinate $x_i$. Specifically, we use the Mat\'{e}rn class of kernels
\begin{equation}
k_{\rm Matern}(r) = \frac{2^{1-\nu}}{\Gamma(\nu)} \left(\sqrt{2\nu} r\right)^\nu K_\nu \left(\sqrt{2\nu} r \right).
\end{equation}
where $K_\nu(x)$ is a modified Bessel function. The value of $\nu$ parameterizes the smoothness of the 
Gaussian Process $f(x)$, and is $k$ times mean-square differentiable if $\nu>k$ [[cite]]. For half-integer values of $\nu$, this 
kernel has a computationally cheap form without special functions, and we have had good results with $\nu=5/2$, 
resulting in a twice-differentiable function. The $\nu=5/2$ kernel is
\begin{equation}
k_{\rm Matern}^{\nu=5/2}(r) = \left(1+\sqrt{5}r + \frac{5r^2}{3}\right) \exp\left(-\sqrt{5}r\right).
\end{equation}

Our final kernel takes the form
\begin{equation}
k(x_p, x_q) = \sigma_f^2 k_{\rm Matern}^{\nu=5/2}(r) + \sigma_n^2 \delta_{pq}
\end{equation}
where $\sigma_f$ is a scale factor that describes the range of values that $f(x)$ takes over the domain, 
and $\sigma_n$ is a noise parameter. The white noise kernel $\sigma_n^2 \delta_{pq}$ (also called a nugget), which is
$\sigma_n^2$ when $x_p=x_q$ and zero otherwise, is generically used for numerical stability, but also parameterizes
noise in the data ${\bm y}$. In our case, the training set waveforms which are evaluated, tapered, and 
Fourier transformed numerically have numerical noise that we will estimate by optimizing the hyperparameters.
The hyperparameters now are $\theta = \{\sigma_f, \ell_q, \ell_{S_1}, \ell_{S_2}, \ell_{\Lambda_1}, \ell_{\Lambda_2}, \sigma_n\}$.

[[In principle, one can select the optimal class of kernels using k-fold or leave-one-out cross validation. We have also tried the 
more common infinitely differentiable squared exponential kernel. However, we found it difficult to reliably tune the 
hyperparameters for the problem here.]]

In order to estimate the hyperparameters, we use the above assumption (Eq.~\eqref{eq:gaussian}) that the joint distribution of 
the data ${\bm y}$ is a multivariate Gaussian which has the following distribution
\begin{equation}
\ln p({\bm y} | {\bm x}, {\bm \theta}) = -\frac{1}{2}{\bm y}^T K^{-1} {\bm y} - \frac{1}{2} \ln |K| - \frac{d}{2} \ln 2\pi.
\end{equation}
This is the log-likelihood for ${\bm y}$ given the hyperparameters ${\bm \theta}$, and we can find the posterior for ${\bm \theta}$
given ${\bm y}$ using Bayes' theorem
\begin{equation}
p({\bm \theta} | {\bm x}, {\bm y}) \propto p({\bm \theta}) p({\bm y} | {\bm x}, {\bm \theta}).
\end{equation}
The prior $p({\bm \theta})$ is typically uniform and used to set the bounds on ${\bm \theta}$. If interested in the distribution of the 
hyperparamaters, one can sample this posterior with, for example, Markov chain Monte Carlo. However, for the problem here, we are
simply interested in finding the maximum of the posterior. To do this, we use the \texttt{gaussian\_process} module in the
\texttt{scikit-learn} package [[cite]]. Finally, with the optimized hyperparameters ${\bm \theta}$, Eq.~\eqref{eq:var} is the 
interpolating function. We note that evaluating Eq.~\eqref{eq:var}, is an $\mathcal{O}(N)$ operation if the vector $K^{-1}{\bm y}$ is
precomputed.

\subsection{Choosing new training set waveforms}

Once we have an initial surrogate model, we can choose new training set samples by iteratively searching the parameter space
for new points ${\bm x}$ that maximize some error criterion. The quantity that we use is the distinguishability between the
EOB waveform and its surrogate [[cite]]
\begin{equation}
|\delta h | = \sqrt{(h - h_{\rm Sur} | h - h_{\rm Sur})}.
\end{equation}
For an initial training set with N samples, this function has $\sim N$ local maxima. We use Monte Carlo sampling with ~100,000
samples and choose the point with the largest distinguishability [[for now. Could also use multistart optimization.]] to find the 
approximate global maximum. Because $|\delta h |$ is an expensive function that requires an EOB or NR evaluation, 
we approximate it as follows:

[[Your inexpensive way of approximating $|\delta h |$ that does not depend on $h_{\rm EOB}$ here.]]
\begin{equation}
\label{eq:distinguish}
|\delta h | \approx blah.
\end{equation}

With this new sample, one could evaluate the waveform at the new point and reconstruct the surrogate model, then iterate.

Reconstructing the surrogate for each new sample is inexpensive compared to the time needed to simulate a new NR waveform. 
However, even for NR waveforms, one often wants to perform $N_{\rm new}$ new waveform simulations in parallel. 
We therefore need a method for optimally choosing $N_{\rm new}$ new samples. One could simply choose the 
$N_{\rm new}$ local maxima with the largest $|\delta h |$. However, there is a more optimal way of iteratively finding new points.
If we hold the hyperparameters ${\bm \theta}$ fixed, we note that the GPR error estimate (Eq.~\eqref{eq:var}) does not depend
on the data ${\bm y}$, just on the samples ${\bm x}$. Therefore, the approximate $|\delta h |$ (Eq.~\eqref{eq:distinguish}) does not depend
on the new samples either. The method uses the idea of uncertainty sampling [[cite]] but without updating the hyperparameters
after each new point is chosen. The algorithm for choosing the new points is as follows.

While $|\delta h |_{\rm max} > \epsilon$:
\begin{enumerate}
\item Construct the GPR error functions (Eq.~\eqref{eq:var}) at each of the empirical nodes $F^A_j$ and $F^\Phi_j$.
In practice this can be done by specifying the samples ${\bm x}$, the hyperparameters ${\bm \theta}$, and dummy 
data ${\bm y}={\bm 1}$. These GPR error functions are the input to Eq.~\eqref{eq:distinguish}.

\item Find $({\bm x}_{\rm max}, |\delta h |_{\rm max})$ that maximizes Eq.~\eqref{eq:distinguish} over the parameter space. 
\item Add ${\bm x}_{\rm max}$ to the list of samples: ${\bm x} = [{\bm x}, {\bm x}_{\rm max}]$.
\end{enumerate}
These new waveforms can now be evaluated in parallel.

[[The methods used here fall under the terms active learning and Bayesian optimization [[cite]]. The specific method being used is
uncertainty sampling [[cite]]. The parallelization method I am using I have not seen in the literature. Is it actually new? Ask Niki.]]

[[Another method you could try is k-means clustering of the mismatch for your validation set (using mismatch above some threshold), 
where you have $k=N$, because there will
be $\sim N$ local maxima of the mismatch. Within each cluster, choose the waveform with the largest mismatch and add it to the
training set. This way you only use one new point per local maximum region. You could do the same thing with the Fisher matrix
systematic error calculation instead of the mismatch if that's a better quantity to use.]]

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{uncertaintysampling.png}
\caption{Estimated RMS phase error as function of number of sampled points}
\label{fig:phaserange}
\end{figure}

\subsection{Surrogate waveform evaluation}

--Constructing the final model with units, inclination angle, time and phase shift, etc.

--Implementation details

%% ________________________________________________________
\section{Results}

\subsection{Accuracy}

--Phase error. (needs to be significantly smaller than tidal effects) Use this for order of magnitude estimate of required
accuracy.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{maxamphist.png}
\includegraphics[width=0.49\textwidth]{maxphasehist.png}
\caption{Amplitude and phase error.}
\label{fig:phaserange}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{dhmaxerror.png}
\caption{Waveform with largest phase error. It's parameters are....}
\label{fig:phaserange}
\end{figure}

In Fig.~\ref{fig:mismatch}, we show the distribution of mismatch $\mathcal{M}$ between our surrogate
and the $10^4$ randomly sampled EOB waveforms. We use the design sensitivity aLIGO PSD~\cite{Aasi:2013wya} and
a sampling rate of 4096~Hz. Our integration bounds are $f_{\rm low} = x$~Hz and the Nyquist frequency
$f_{\rm high} = 2048$~Hz. Because the surrogate can be rescaled with mass,
we show results for the smaller mass $M_B$ fixed at $1M_\odot$ or fixed at $2M_\odot$.
We also show the mismatch as a function of parameters in Fig.~\ref{fig:mismatchtriangle}.
It is generally largest for as the mass ratio decreases toward $1/3$.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{mismatch.png}
\caption{Mismatch. \red{[[You should do mismatch with TaylorF2 as a reference for the improvement in 
going through all the effort in this paper.]]}
}
\label{fig:mismatch}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{mismatchtrianglem10.png}
\caption{Mismatch for systems with $m_2=1M_\odot$.}
\label{fig:mismatchtriangle}
\end{figure}

--distinguishability

--Fisher matrix systematic error.

\subsection{Timing}

Although optimizing the hyperparameters scales with number of samples $n$ as $\mathcal{O}(n^3)$
due to the required matrix inversion, the evaluation time of a stored GPR scales as $\mathcal{O}(n)$.
The evaluation time for the waveform functions will therefore be $\mathcal{O}(n(N_A+N_\Phi))$.

The surrogate has a fixed time of $\sim 50$~ms to evaluate the GPR at each node then evaluate the amplitude
and phase at the log-spaced frequencies. Then, there is an additional cost to resample the amplitude and phase
at uniformly-spaced frequencies with interpolation, and finally evaluating $\tilde h_+$ and $\tilde h_\times$. 
Below 50~Hz the surrogate is faster than all models except TaylorF2 [[you didn't try PhenomP etc.]], and below
18~Hz it is faster than TaylorF2.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{timing.png}
\caption{Waveform evaluation time as a function of the waveform starting frequency. The time-domain waveforms
were sampled at 4096~Hz then Fourier transformed. The frequency-domain waveforms used the
same frequency samples as the Fourier transformed time-domain waveforms.}
\label{fig:timing}
\end{figure}

--Frequency domain allows for evaluating likelihood with non-uniform sampling.

--Can construct a ROQ of the model if needed.

\subsection{Parameter estimation}

--Can only include this section if there's a LAL version. Should you bother?

\section{Discussion}

\subsection{Review}

\subsection{Future work}

Hierarchical method and uncertainty sampling provides way to construct surrogate from minimal number of waveforms.

--Could be used to correct the very end of inspiral (and post-merger) with NR simulations where only O(100) simulations are feasable.

%% ________________________________________________________
\begin{acknowledgments}

BL thanks... BL was supported by...

\end{acknowledgments}

%% ________________________________________________________
%\bibliographystyle{revtex}     
\bibliography{paper}

\end{document}

### 2017-06-13 src/TEOB_generation/Generate_TEOB_TD_MPI.py
Script to generate TEOB waveforms in parallel
Currently, raw TD data is output in .npy files.

@@ -1,0 +1,402 @@
+#!/usr/bin/env python
+
+# coding: utf-8
+
+import lal
+import lalsimulation as LS
+
+import numpy as np
+import shutil, os, sys
+import traceback
+from subprocess import check_call
+import scipy
+import scipy.interpolate as ip
+from scipy import optimize
+
+import os.path
+import h5py
+import glob
+
+from mpi4py import MPI
+import warnings
+import json
+from multiprocessing import Process, Queue
+
+import romspline
+
+from tqdm import *
+
+# Q: Andrea, do we want 'TEOBv2' or 'TEOBv4'?
+# Q: Is the head of TEOBBNS fine?
+
+# Ben
+# Q: which distance, inclination do you want me to use?
+# Q: which total mass do you want me to use?
+# Q: Do you want both hp, hc; since the wf is aligned, there is no need to store both.
+# Q: why is tstart needed? shouldn't we just align at the peak of h22?
+
+# Q: Is a common time grid needed? I think not and will return 
+    # either a romspline grid for each cfg (but that may be too slow to generate)
+    # or data interpolated onto a common romspline grid, but using zeros instead of extrapolation)
+    # or the TD amplitude, phase data on a suitable grid
+    # or the raw data
+
+
+
+#------------------------------------------------------------------------------
+def spin_tidal_eob(m1, m2, s1z, s2z, lambda1, lambda2,
+                    f_min, 
+                    delta_t=1.0/16384.0, distance=1.0, inclination=0.0,
+                    approximant='TEOBv4', verbose=True):
+    """EOB waveform with aligned spin and tidal interactions. 
+    l=3 tidal interaction and l=2,3 f-mode calculated with universal relations.
+    
+    Parameters
+    ----------
+    approximant : 'TEOBv2' or 'TEOBv4'
+        Based on the inspiral model given by 'SEOBNRv2' or 'SEOBNRv4'.
+    
+    Returns
+    -------
+    Waveform object
+    """
+    # print m1, m2, s1z, s2z, lambda1, lambda2, f_min, delta_t, distance, inclination
+    f_ref = 0.
+    phiRef = 0.
+    
+    # Must have aligned spin
+    s1x, s1y, s2x, s2y = 0., 0., 0., 0.
+    
+    # Eccentricity is not part of the model
+    longAscNodes = 0.
+    eccentricity = 0.
+    meanPerAno = 0.
+    
+    # Set the EOB approximant
[…]

### 2017-06-14 src/TEOB_generation/Generate_TEOB_TD_MPI.py
Add data compression for output to .npy

@@ -26,21 +26,25 @@ import romspline
 
 from tqdm import *
 
-# Q: Andrea, do we want 'TEOBv2' or 'TEOBv4'?
-# Q: Is the head of TEOBBNS fine?
+# Q: Andrea, do we want 'TEOBv2' or 'TEOBv4'? Both, but start with TEOBv4
+# Q: Is the head of TEOBBNS fine? Yes
+# Q: I get nans when calling SimUniversalRelationlambda3TidalVSlambda2Tidal and similar functions with lambda = 0; Andrea knows about this; needs to fix domain of validity of universal relations
+
 
 # Ben
-# Q: which distance, inclination do you want me to use?
-# Q: which total mass do you want me to use?
+# Q: which distance, inclination do you want me to use?: d=1, iota=0
+# Q: which total mass do you want me to use? M=2Msun DOn't
 # Q: Do you want both hp, hc; since the wf is aligned, there is no need to store both.
 # Q: why is tstart needed? shouldn't we just align at the peak of h22?
 
-# Q: Is a common time grid needed? I think not and will return 
-    # either a romspline grid for each cfg (but that may be too slow to generate)
-    # or data interpolated onto a common romspline grid, but using zeros instead of extrapolation)
-    # or the TD amplitude, phase data on a suitable grid
-    # or the raw data
+# Q: Is a common time grid needed? Yes
+    # output the TD amplitude and phase data on a suitable grid (either romspline, or const phase grid); use using zeros instead of extrapolation
+    
 
+# Could fix mass of smaller object to 1Msun and keep f_min_Hz fixed; this would mean that
+# the wf starts at a higher Mf and would require hybridization; this would be computationally cheaper
+# Ben would like to avoid this
+# So, for the time being I can just fix Mtot=2Msun and f_min_Hz=8Hz
 
 
 #------------------------------------------------------------------------------
@@ -192,8 +196,40 @@ def compute_corners_from_domain(d):
                                                 for lambda2 in d.get_lambda2()]
 
 #------------------------------------------------------------------------------
+def time_grid_for_longest_waveform(Mtot, f_min, outfile, deg=3, abstol=5e-5):
+    # positive aligned spins and equal mass-ratio increase the length of the waveform in time
+    # high lambda also makes the wf shorter, so use lambda ~ 0 here.
+    
[…]

### 2017-06-14 src/TEOB_generation/Generate_TEOB_TD_MPI.py
Add data consolidation step

@@ -437,9 +437,58 @@ def main():
     comm.Barrier()
     if rank == 0: print '=============================================================='
 
-    # FIXME: add consolidation of data step
-    # load up .npy and save as hdf5
-    
+    # Load all .npy and save as hdf5
+    if rank == 0:
+        print 'Loading npy files for all configurations'
+
+        # FIXME: add git branch and hash
+        # how to find build directory? then can use git rev-parse HEAD
+        generation_str = \
+        '''
+        approximant = %s
+        M = %g
+        fs = %g
+        iota = %g
+        f_min = %g
+        distance = %g
+
+        Using lalsimulation version %s
+        ''' % (approximant, M, fs, iota, f_min, distance, LS.LALSIMULATION_VERSION)
+
+        print 'Outputting interpolated data as HDF5'
+        outfname = 'TEOB_dataI.hdf5'
+        fh5 = h5py.File(outfname, 'w')
+        fh5.create_dataset('configurations', data=cfgs)
+        fh5.create_dataset('configurations_keys', data=['q', 'chi1', 'chi2', 'lambda1', 'lambda2'])
+        fh5.create_dataset('grid', data=grid)
+        fh5.attrs['Description'] = np.string_('TEOB TD data')
+        fh5.attrs['GenerationSettings'] = generation_str
+
+        data_keys_name = ['amp', 'phi']
+        basename = '%s/TEOB_TD'%(tmpdir)
+        fh5.create_dataset('data_keys_name', data=data_keys_name)
+        for i in tqdm(np.arange(n)):
+            if verbose:
+                print 'Loading data for waveform %d of %d with index %d.' %(i,n,cfgs[i])
+            # Data has already been interpolated onto target grid
+            f = '%s_%d.npy'%(basename, i)
+            try:
+                data = np.load(f)
+                grp = fh5.create_group('h_'+str(i))
+                for j in range(len(data_keys_name)):
+                # create one data set for each quantity
+                # each data set includes data as a function of
+                # time and physical parameter(s)
+                    if np.isnan(data[j]).any():
+                        print '*'*80
+                        print '*** Warning: dataset %d for quantity %s contains one or more nans! ***' %(i, data_keys_name[j])
+                        print data[j]
+                        print '*'*80
+                    grp.create_dataset(data_keys_name[j], data=data[j])
+            except Exception as e:
+                print 'Loading %s failed. Skipping configuration.' % (f)
+
+        fh5.close()
     
     print '=============================================================='
     print 'All Done!'

### 2017-06-14 src/TEOB_generation/Generate_TEOB_TD_MPI.py
Set up test run on vulcan

@@ -490,9 +490,9 @@ def main():
 
         fh5.close()
     
-    print '=============================================================='
-    print 'All Done!'
-    print '=============================================================='
+        print '=============================================================='
+        print 'All Done!'
+        print '=============================================================='
 
 if __name__ == "__main__":
     main()

### 2017-06-14 src/TEOB_generation/runs/test.sh

@@ -1,0 +1,14 @@
+#!/bin/bash
+
+source /home/mpuer/lsc/TEOBBNS/etc/lalsuiterc
+
+export PYTHONPATH=$PYTHONPATH:/home/mpuer/projects/
+
+SDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/
+RDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/runs
+CFG=${RDIR}/test.json
+LOG=${RDIR}/test.log
+
+cd ${SDIR}
+mpirun -n 24 Generate_TEOB_TD_MPI.py -o ${CFG} > ${LOG} 2>&1 &
+

### 2017-06-26 src/TEOB_generation/Generate_TEOB_TD_MPI.py
Add option to read configurations from text file; assume that q <= 1

@@ -334,7 +334,7 @@ def TEOB_process_array_TD(i, M,
         return
     try:
         print 'Generate wf:', args
-
+        print q
         m1 = M * q/(1.0+q)
         m2 = M * 1.0/(1.0+q)
         t, hp, hc = spin_tidal_eob(m1, m2, chi1, chi2, lambda1, lambda2,
@@ -468,25 +468,22 @@ def main():
     d.set_lambda2(lambda2_min, lambda2_max)
     cfgs = compute_corners_from_domain(d) # 2^5 = 32 corner points
 
-    # FIXME: write function that reads list of configs from txt or npy
-
-
-    # Adhoc 1-parameter family for tests of smoothness
-    q = 0.6
-    spin1z_arr = np.linspace(-0.7, 0.7, 51)
-    spin2z = 0.2
-    lambda1 = 2000.0
-    lambda2 = 1000.0
-    cfgs = [(q, spin1z, spin2z, lambda1, lambda2) for spin1z in spin1z_arr]
-
-    # we are not using a fixed grid anymore
-    # g_filename = 'time_grid_M%.fMsun_fmin%.fHz.npy' % (M, f_min)
-    # if rank == 0 and not os.path.isfile(g_filename):
-    #     # Generate sparse time grid
-    #     print 'Computing target sparse time grid, chosen based on total mass and f_min.'
-    #     time_grid_for_longest_waveform(M, f_min, outdir+g_filename, deg=3, abstol=5e-5)
-    # comm.Barrier()
-    # grid = np.load(g_filename) # Now read it back in for all tasks
+    if opts['cfgfile']:
+        print 'Loading configurations from textfile', cfgfile
+        cfgs = np.loadtxt(cfgfile)
+        if cfgs.shape[1] != 5:
+            raise ValueError("Expected configurations in file %s to contain 5 columns "
+            +"(q, spin1z, spin2z, lambda1, lambda2)."%cfgfile)
+        if np.min(cfgs.T[0]) > 1.0:
+            raise ValueError("Expected in file %s mass-ratio <= 1"%cfgfile)
+    else:
+        print 'Using adhoc 1-parameter family for tests of smoothness'
+        q = 0.6
+        spin1z_arr = np.linspace(-0.7, 0.7, 51)
+        spin2z = 0.2
+        lambda1 = 2000.0
+        lambda2 = 1000.0
+        cfgs = [(q, spin1z, spin2z, lambda1, lambda2) for spin1z in spin1z_arr]
 
     n = len(cfgs)
[…]

### 2017-06-26 src/TEOB_generation/Generate_TEOB_TD_MPI.py
Make cfgfile the default

@@ -457,17 +457,6 @@ def main():
     comm.Barrier()
     os.chdir(outdir)
 
-    # Compute corners of specified 5D domain
-    q_min, chi1_min, chi2_min, lambda1_min, lambda2_min = opts['params']['min_vals']
-    q_max, chi1_max, chi2_max, lambda1_max, lambda2_max = opts['params']['max_vals']    
-    d = domain()
-    d.set_q(q_min, q_max)
-    d.set_chi1(chi1_min, chi1_max)
-    d.set_chi2(chi2_min, chi2_max)
-    d.set_lambda1(lambda1_min, lambda1_max)
-    d.set_lambda2(lambda2_min, lambda2_max)
-    cfgs = compute_corners_from_domain(d) # 2^5 = 32 corner points
-
     if opts['cfgfile']:
         print 'Loading configurations from textfile', cfgfile
         cfgs = np.loadtxt(cfgfile)
@@ -476,6 +465,17 @@ def main():
             +"(q, spin1z, spin2z, lambda1, lambda2)."%cfgfile)
         if np.min(cfgs.T[0]) > 1.0:
             raise ValueError("Expected in file %s mass-ratio <= 1"%cfgfile)
+    elif opts['corners']:
+        print 'Computing corners of specified 5D domain'
+        q_min, chi1_min, chi2_min, lambda1_min, lambda2_min = opts['params']['min_vals']
+        q_max, chi1_max, chi2_max, lambda1_max, lambda2_max = opts['params']['max_vals']    
+        d = domain()
+        d.set_q(q_min, q_max)
+        d.set_chi1(chi1_min, chi1_max)
+        d.set_chi2(chi2_min, chi2_max)
+        d.set_lambda1(lambda1_min, lambda1_max)
+        d.set_lambda2(lambda2_min, lambda2_max)
+        cfgs = compute_corners_from_domain(d) # 2^5 = 32 corner points
     else:
         print 'Using adhoc 1-parameter family for tests of smoothness'
         q = 0.6

### 2017-06-26 src/TEOB_generation/Generate_TEOB_TD_MPI.py
fix minor bug

@@ -458,6 +458,7 @@ def main():
     os.chdir(outdir)
 
     if opts['cfgfile']:
+        cfgfile = opts['cfgfile']
         print 'Loading configurations from textfile', cfgfile
         cfgs = np.loadtxt(cfgfile)
         if cfgs.shape[1] != 5:

### 2017-06-26 src/TEOB_generation/runs/set1/test_train_40Hz.sh
Add config files and runscripts for first set, 40Hz

@@ -1,0 +1,14 @@
+#!/bin/bash
+
+source /home/mpuer/lsc/TEOBBNS/etc/lalsuiterc
+
+export PYTHONPATH=$PYTHONPATH:/home/mpuer/projects/
+
+SDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/
+RDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/runs/set1
+CFG=${RDIR}/test_train_40Hz.json
+LOG=${RDIR}/test_train_40Hz.log
+
+cd ${SDIR}
+mpirun -n 24 Generate_TEOB_TD_MPI.py -o ${CFG} > ${LOG} 2>&1 &
+

### 2017-06-26 src/TEOB_generation/runs/set1/test_validation_40Hz.sh

@@ -1,0 +1,14 @@
+#!/bin/bash
+
+source /home/mpuer/lsc/TEOBBNS/etc/lalsuiterc
+
+export PYTHONPATH=$PYTHONPATH:/home/mpuer/projects/
+
+SDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/
+RDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/runs/set1
+CFG=${RDIR}/test_validation_40Hz.json
+LOG=${RDIR}/test_validation_40Hz.log
+
+cd ${SDIR}
+mpirun -n 24 Generate_TEOB_TD_MPI.py -o ${CFG} > ${LOG} 2>&1 &
+

### 2017-06-27 src/diagnostics.py
EOB waveform parameters that cause problems.

@@ -1,6 +1,7 @@
 import numpy as np
 import matplotlib.pyplot as plt
 
+########################## 2d error plots ##############################
 def max_error_2d_projection_plot(axes, x, y, error, threshold=None,
                                  x_label='x', y_label='y', colorbar=None, colorbarlabel='error'):
     """2d plot of the errors.
@@ -15,7 +16,7 @@ def max_error_2d_projection_plot(axes, x, y, error, threshold=None,
     # Sort errors so largest errors are plotted on top of smaller errors
     error_params = np.array([error, x, y]).T
     error_params_sort = error_params[error_params[:, 0].argsort()]
-    
+
     # Remove points below threshold
     if threshold!=None:
         ep = error_params_sort[error_params_sort[:, 0]>=threshold]
@@ -35,13 +36,13 @@ def max_error_2d_projection_plot(axes, x, y, error, threshold=None,
         cb = plt.colorbar(mappable=sc, ax=axes)
         cb.set_label(label=colorbarlabel, fontsize=18)
         cb.ax.tick_params(labelsize=14)
-    
+
     # buffers for plot
     xmin, xmax = x.min(), x.max()
     ymin, ymax = y.min(), y.max()
     bufx = 0.05*(xmax - xmin)
     bufy = 0.05*(ymax - ymin)
-    
+
     axes.set_xlim([xmin-bufx, xmax+bufx])
     axes.set_ylim([ymin-bufy, ymax+bufy])
     axes.set_xlabel(x_label)
@@ -91,3 +92,87 @@ def error_2d_triangle_plot(params, error, labels, threshold=None, figsize=(10, 1
     fig.subplots_adjust(hspace=0.0, wspace=0.0)
 
     return fig, ax
+
+#################### 1d cross section plots ###########################
+
+def plot_1d_amp_phase_variation(ax_amp, ax_phase, sur, params, xi, xlow, xhigh, nx, mfs, legend=True):
+    """Plot amplitude(1parameter) and phase(1parameter) as the single parameter is varied.
+    Do this for all the frequencies mfs.
+
+    Parameters
+    ----------
+    params : 1d array
+        Point in full parameter space where the 1d slice should pass through
+    xi : int
+        Index of the parameter you want to vary
+    xlow, xhigh, nx : bounds and number of samples for the feature
+    mfs : 1d array
[…]

### 2017-06-28 paper/paper.tex
Start section on EOB

@@
 \usepackage{setspace} % for pleasent spacing in algorithm
+\usepackage{times}
 
@@
 \section{Aligned-spin, dynamical tides EOB model}
+Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects.
+State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries.
+The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}.
+This research program has been extended to accommodate matter effects~\cite{}, which become relevant XXX
 
---Key features of dynamical tides EOB model
+References~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides.
+Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star.
+Because of binary motion, the forcing tidal field varies at the orbital frequency.
+Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode.
+The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field.
+Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.
+Here we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
+\begin{equation}
+\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
+\end{equation}
+where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides.
+We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below.
+Let $A$ and $B$ be the body labels.
+Let $M^{A,B}$ be the mass of either body and $M=M^A+M^B$ be the total mass of the binary.
+Explicitly, including quadrupolar and octupolar tidal effects, we have
+\begin{widetext}
+\begin{align}
+ \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)\,u^6\frac{X_B}{X_A} \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
+ &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)\,u^8 \frac{X_B}{X_A}\left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
+\end{align}
+\end{widetext}
+where $X_{A,B}=M^{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate, $\Lambda_{2,\textrm{dyn}}^{A,B}(u)$ are the dimensionless quadrupolar dynamical tidal polarizabilities, and $\Lambda_{3,\textrm{dyn}}^{A,B}(u)$ are the dimensionless octupolar dynamical tidal polarizabilities.
+Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies.
+In particular, the dimensionless dynamical polarizabilities read
+\begin{equation}
+\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\frac{2}{(2\ell-1)!!}k^{A,B}_{\ell}\hat{k}^{A,B}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\left(\frac{R_{A,B}}{M}\right)^{2\ell+1}\,,
+\end{equation}
+where $k^{A,B}_{\ell}$ is the tidal Love number and $R_{A,B}$ is the neutron star radius.
+Here $\hat{k}^{A,B}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode angular frequency of the $2^\ell$-th multipole, $\omega_{0\ell}^{A,B}$.
+
+Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
+\begin{equation}
+h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{insp-plunge, pm}} + h_{\ell m}^{\textrm{tidal}}\,,
+\end{equation} 
+where $h_{\ell m}^{\textrm{insp-plunge, pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs.
+(A14)-(A17) of Ref.~\cite{Damour:2012yf}.
+Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.\red{[Tanja should confirm for 21]}
 
@@
 %\bibliographystyle{revtex}     
-\bibliography{paper}
+\bibliography{paper,refs}
 
 \end{document}
+
+\begin{widetext}
+\begin{align}
+\Lambda_{2,\textrm{dyn}}^{\textrm{(A,B)}}(u) = &\frac{\lambda_{2}^{\textrm{(1,2)}}}{M^5} \left\{\frac{1}{4} + \frac{3}{4} \left(\frac{\omega^{(1,2)}_{f,2}}{2\Omega}\right)^2\left[\frac{4\Omega^2}{(\omega^{(1,2)}_{f,2})^2 - 4\Omega^2} + \frac{10}{3Q}\right.\right.
+\nonumber\\
+&\left.\left.+\sqrt{\frac{\pi}{3}}\frac{1}{\epsilon_{1,2}}\left[(1 + 2\mathcal{S}{(y_{1,2})})\cos{\left(\frac{\pi}{2}y^2_{1,2}\right)} - (1 + 2\mathcal{C}{(y_{1,2})})\sin{\left(\frac{\pi}{2}y_{1,2}^2\right)}\right]\right]\right\}\,,\\
+\Lambda_{3,\textrm{eff}}^{\textrm{(1,2)}} (u)= &\frac{\lambda_{3}^{\textrm{(1,2)}}}{M^7} \left\{\frac{3}{8} + \frac{(\omega^{(1,2)}_{f,3})^2}{\Omega^2}\left(\frac{25}{48 O_{1,2}} + \frac{5}{72}\frac{9\Omega^2}{(\omega^{(1,2)}_{f,3})^2-9\Omega^2}\right)\right.
+\nonumber\\
+&\left.+ O_{1,2}\left[\cos{(w^2)}\left(\frac{1}{2} +\mathcal{S}(sqrt2pi*w)\right) - \sin{(w^2)}\left(\frac{1}{2} + \mathcal{C}(sqrt2pi*w)\right)\right]\right\}\,,
+\end{align}
+\end{widetext}
+where $\lambda_{2}^{\textrm{(1,2)}}$ are the (constant) adiabatic values of the quadrupolar tidal polarizabilities, $\lambda_{3}^{\textrm{(1,2)}}$ are the (constant) adiabatic values of the octupolar tidal polarizabilities, $\omega^{(1,2)}_{f,2}$ are the quadrupolar $f$-mode frequencies, $\omega^{(1,2)}_{f,3}$ are the octupolar $f$-mode frequencies, $\mathcal{S}$ is the Fresnel sine integral, $\mathcal{C}$ is the Fresnel cosine integral, $\epsilon_{1,2} = 2^{19/3}\nu (M\omega^{(1,2)}_{f,2})^{5/3}/5$,   $y_{1,2}=\sqrt{3/(\pi \epsilon)} Q_{1,2}/5$, and $Q_{1,2} = 4- 2^{1/3}(M\omega^{(1,2)}_{f,2})^{5/3}u^{-5/2}$.
+, $w_{1,2}=O_{1,2}/(4\,3^{2/3}\sqrt{10}(M\omega^{(1,2)}_{f,3})^{5/6}\sqrt{\nu})$, $O_{1,2}=5\sqrt{5\pi}/u^3(M\omega^{(1,2)}_{f,3})^{7/6}/(192\,3^{2/3}\sqrt{\nu})$.

### 2017-06-28 src/TEOB_generation/runs/test_9Hz.sh
1D test for f_min=9Hz
expect about 1 waveform per two days

@@ -1,0 +1,14 @@
+#!/bin/bash
+
+source /home/mpuer/lsc/TEOBBNS/etc/lalsuiterc
+
+export PYTHONPATH=$PYTHONPATH:/home/mpuer/projects/
+
+SDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/
+RDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/runs
+CFG=${RDIR}/test_9Hz.json
+LOG=${RDIR}/test_9Hz.log
+
+cd ${SDIR}
+mpirun -n 24 Generate_TEOB_TD_MPI.py -o ${CFG} > ${LOG} 2>&1 &
+

### 2017-06-28 src/TEOB_generation/runs/set1/test_train_uncsamp_40Hz.sh
Setup for new configurations

@@ -1,0 +1,14 @@
+#!/bin/bash
+
+source /home/mpuer/lsc/TEOBBNS/etc/lalsuiterc
+
+export PYTHONPATH=$PYTHONPATH:/home/mpuer/projects/
+
+SDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/
+RDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/runs/set1
+CFG=${RDIR}/test_train_uncsamp_40Hz.json
+LOG=${RDIR}/test_train_uncsamp_40Hz.log
+
+cd ${SDIR}
+mpirun -n 24 Generate_TEOB_TD_MPI.py -o ${CFG} > ${LOG} 2>&1 &
+

### 2017-06-28 src/TEOB_generation/runs/set1/test_validation2_40Hz.sh

@@ -1,0 +1,14 @@
+#!/bin/bash
+
+source /home/mpuer/lsc/TEOBBNS/etc/lalsuiterc
+
+export PYTHONPATH=$PYTHONPATH:/home/mpuer/projects/
+
+SDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/
+RDIR=/home/mpuer/projects/gpsurrogate/src/TEOB_generation/runs/set1
+CFG=${RDIR}/test_validation2_40Hz.json
+LOG=${RDIR}/test_validation2_40Hz.log
+
+cd ${SDIR}
+mpirun -n 24 Generate_TEOB_TD_MPI.py -o ${CFG} > ${LOG} 2>&1 &
+

### 2017-06-29 paper/paper.tex
More work on EOB section

@@
 
+The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO), Virgo, and KAGRA which are now making routine detections of binary black hole (BBH) systems or will begin in the next few years.
+Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]].
+From the nondetection of BNS events in aLIGO's first observing run (O1) the 90\% upper limit is still consistent with the upper end of this range [[O1UpperLimits]].
+However, continued nondetection will begin to narrow down this range.
+
+The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters.
+These parameters are measured by matched-filtering of the GW data with waveform models, and require highly accurate waveform models to reliably measure the parameters with minimal systematic errors.
+[[Review of systematic error BNS papers...]] Previous work on estimating the tidal parameters, for example, has shown that even phase uncertainties of several radians, as is present in current 3.5-post-Newtonian waveform models, leads to systematic errors that are as large as the statistical errors [[cite]] and this also biases the recovered NS equation of state as well [[cite]].
+Uncertainties in the PN waveform model also lead to significant biases in the recovered masses and spins as well [[cite]].
+[[What about sky position?
+Does it matter?]]
+
+The effective one body (EOB) formalism and numerical relativity (NR) simulations, on the other hand, have shown significantly smaller uncertainties for BNS systems in their respective ranges of validity than PN waveforms.
+For the inspiral, EOB waveforms have total phase uncertainties of $\sim 1$~rad as compared to PN waveforms with phase uncertainties of $\sim 10$~rad as determined by the variation from freedom within each formalism [[explain better.
+show your plot for this.]].
+NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]].
+
+Unfortunately, available codes for generating EOB waveforms for BNS systems can take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]].
+
+These quantities are measured with Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling.
+
+Previous work on surrogate models has focused mainly on compressing the waveform parameter space into a linear combination of a small number of basis functions using either greedy methods [[cite]] or singular value decomposition [[cite]].
+Significantly less work has been performed on optimally sampling waveforms so that the coefficients of these linear combinations of waveforms can be chosen through fitting or interpolation with the fewest number of waveform evaluations.
+Standard techniques such as tensor spline or Chebyshev interpolation require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
+For the 5-dimensional problem here, this is somewhat unfeasable.
+Recent work on optimally choosing waveforms for both EOB [[cite GPRpaper]] and NR [[cite CaltechPrecessing]] waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.
+
+In this paper, we will develop methods to construct surrogate models from EOB and NR waveform codes that accurately reproduce waveforms from this high-dimensional parameter space, using as few waveform evaluations as possible.
+We will construct frequency-domain surrogates of the amplitude and phase.
+
+--Overview of what was done in nonspinning BNS surrogate paper, and the improvements used for constructing aligned-spin surrogate.
+
+--Advantages of using frequency domain surrogate (and why you didn't use it in last paper).
+
+[[A notable effect that has not been included is the spin induced quadrupole moment of each NS.]]
+
 We organize the paper as follows. In Section...
@@
 (A14)-(A17) of Ref.~\cite{Damour:2012yf}.
-Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.\red{[Tanja should confirm for 21]}
+Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
+
+The inspiral-plunge orbital dynamics is obtained by numerically solving Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s.
 

### 2017-07-06 paper/paper.tex
Debugging and filtering EOB training set waveforms. Initial analysis of EOB surrogate.

@@
 (A14)-(A17) of Ref.~\cite{Damour:2012yf}.
-Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
-
-The inspiral-plunge orbital dynamics is obtained by numerically solving Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s.
+Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.\red{[Tanja should confirm for 21]}
 

### 2017-07-06 src/trainingseteob.py

@@ -1,0 +1,533 @@
+import numpy as np
+import h5py
+
+import scipy.interpolate as interpolate
+import scipy.optimize as optimize
+
+import waveform as wave
+import waveformset as ws
+import window
+import taylorf2
+import trainingset as train
+
+
+def time_at_max_amp(time, amp):
+    """Find the time corresponding to the maximum amplitude.
+    This function interpolates between data points using a 2nd-order spline,
+    before finding the maximum.
+    """
+    nsamp = len(amp)
+
+    # Find index of maximum amplitude
+    imax = np.argmax(amp)
+    # Associated time used as initial guess in numerical maximization
+    tmax = time[imax]
+
+    if imax==nsamp-1:
+        raise Exception, "Maximum amplitude is at last data point. Can't calculate global maximum."
+
+    # Range of parameters to interpolate
+    # (Use no more than the 7 points surrounding the max point.)
+    istart = imax-3
+    iend = min(len(amp)-1, imax+4)
+
+    # Just interpolate a small number of points surrounding imax
+    tlist = time[istart:iend]
+    alist = amp[istart:iend]
+
+    # Use 2nd order interpolation
+    # because it's safer and gives pretty much the same results as higher order interpolation.
+    # Note: A spline is not just a local quadratic fit of 3 points.
+    # It matches derivatives between local quadratics.
+    # So, the answer changes as you add more points to tlist and alist.
+    negampoft = interpolate.UnivariateSpline(tlist, -1.0*alist, k=2, s=0)
+
+    # Find the minimum, starting the search at tmax
+    result = optimize.minimize(negampoft, tmax)
+    return result.x[0]
+
+
+class ConditionedWaveform(object):
+    """Methods for conditioning a time-domain Waveform object
+    to be used in a frequency-domain training set.
+    """
+
+    def __init__(self, h):
+        """
+        """
+        # Original waveform not to be manipulated
+        self.h_orig = h
+        # Copy of the waveform for manipulation
+        self.h = h.copy()
+
+    def resample_uniform(self, delta_t, order=2):
+        """Resample waveform with spacing delta_t.
+        Includes first sample, but does not necessarily include last sample.
+        """
[…]

### 2017-07-06 src/waveform.py

@@ -322,7 +322,7 @@ def fourier_transform_uniform_sampled_waveform(h):
 
 ##################### Functions for plotting waveforms ####################
 
-def plot_waveforms(waveforms, xi=-np.inf, xf=np.inf, npoints=1000, amp=True, hp=True, hc=False):
+def plot_waveforms(waveforms, xi=-np.inf, xf=np.inf, npoints=10000, amp=True, hp=True, hc=False):
     """
     """
     fig, axes = plt.subplots(1, figsize=(16, 3))
@@ -378,7 +378,7 @@ def plot_waveforms_fd(waveforms, exp=False):
     return fig, (ax1, ax2)
 
 
-def plot_waveforms_fd_resample(waveforms, xi=None, xf=None, npoints=1000, exp=False):
+def plot_waveforms_fd_resample(waveforms, xi=None, xf=None, npoints=10000, exp=False):
     """Plot amplitudes and phases of the waveforms.
     """
     fig, (ax1, ax2) = plt.subplots(2, sharex=True, figsize=(16, 6))
@@ -408,7 +408,7 @@ def plot_waveforms_fd_resample(waveforms, xi=None, xf=None, npoints=1000, exp=Fa
     return fig, (ax1, ax2)
 
 
-def plot_waveform_difference_fd(h1, h2, xi=None, xf=None, npoints=1000):
+def plot_waveform_difference_fd(h1, h2, xi=None, xf=None, npoints=10000):
     """Plots of A_1/A_2 - 1 and Phi_1 - Phi_2.
     """
     fig, (ax1, ax2) = plt.subplots(2, sharex=True, figsize=(16, 6))

### 2017-07-18 src/TEOB_generation/Generate_TEOB_TD_MPI.py
Fix issue with phase monotonicity

@@ -27,29 +27,7 @@ import romspline
 
 from tqdm import *
 
-# MP June 2017
-
-#------------------------------------------------------------------------------
-# Choices
-
-# Q: Andrea, do we want 'TEOBv2' or 'TEOBv4'? Both, but start with TEOBv4
-# Q: Is the head of TEOBBNS fine? Yes
-# Q: I get nans when calling SimUniversalRelationlambda3TidalVSlambda2Tidal and similar functions with lambda = 0; Andrea knows about this; needs to fix domain of validity of universal relations
-
-
-# Ben
-# Q: which distance, inclination do you want me to use?: d=1, iota=0
-# Q: which total mass do you want me to use? M=2Msun
-# Q: Do you want both hp, hc; since the wf is aligned, there is no need to store both.
-# Q: why is tstart needed? shouldn't we just align at the peak of h22?
-
-# Q: Is a common time grid needed? No
-
-# Could fix mass of smaller object to 1Msun and keep f_min_Hz fixed; this would mean that
-# the wf starts at a higher Mf and would require hybridization; this would be computationally cheaper
-# Ben would like to avoid this
-# So, for the time being I can just fix Mtot=2Msun and f_min_Hz=8Hz
-
+# MP 06-07, 2017
 
 #------------------------------------------------------------------------------
 def spin_tidal_eob(m1, m2, s1z, s2z, lambda1, lambda2,
@@ -141,12 +119,9 @@ def spin_tidal_eob(m1, m2, s1z, s2z, lambda1, lambda2,
     
     # Extract time array from lalsimulation's structures
     tstart = hp.epoch.gpsSeconds + hp.epoch.gpsNanoSeconds*1.0e-9 
-    # Q: why is this needed? shouldn't we just align at the peak?
     ts = tstart + hp.deltaT*np.arange(hp.data.length)
     
     return ts, hp.data.data, hc.data.data
-    #return wave.Waveform.from_hp_hc(ts, hp.data.data, hc.data.data)
-
 
 #------------------------------------------------------------------------------
 class domain:
@@ -262,6 +237,12 @@ def Generate_phase_grid(t, phi):
     # Compute time values from spline of original data
     t_grid = spline(phim[idx_ok], tm[idx_ok])(phase_grid)
 
+    # Add points beyond where ceased to be monotonic
+    if len(tm) < len(t):
+        idx_nm = np.where(t > tm[-1])[0]
+        t_grid = np.concatenate([t_grid, t[idx_nm]])
[…]

### 2017-07-18 src/TEOB_generation/Generate_TEOB_TD_MPI.py
Fix phase monotonicty via extrapolation

@@ -246,6 +246,44 @@ def Generate_phase_grid(t, phi):
     return t_grid, phase_grid
 
 #------------------------------------------------------------------------------
+def Generate_phase_grid_extrapolate(t, phi):
+    """
+    Using raw time and phase data, construct a grid that has
+    a specific number of points per cycle.
+    """
+    # Make sure that the phase data is monotonically increasing
+    # This is required for the spline interpolation below
+    tm, phim = monotonically_increasing_timeseries(t, -phi)
+
+    i = 0
+    phi_max = max(phim)
+    phi_g = [phim[i]]
+
+    while phi_g[i] < phi_max:
+        p = pts_per_cycle_function(phi_g[i] - phi_max, n=1.0, a=10.0, sigma=0.01)
+        phi_new = phi_g[i] + 2*np.pi / p
+        phi_g.append(phi_new)
+        i += 1
+    phase_grid = np.array(phi_g[:-1])
+
+    # Remove points where the phase does not increase monotonically
+    idx_ok = np.where(np.diff(phim) > 0)
+
+    # Compute time values from spline of original data
+    t_grid = spline(phim[idx_ok], tm[idx_ok])(phase_grid)
+
+    # Add points beyond where phase ceased to be monotonic
+    if len(tm) < len(t):
+        # extrapolate phase
+        # cubic extrapolation is more accurate; linear extrapolation would be safer
+        phiI = spline(tm[idx_ok], phim[idx_ok], k=3, ext=0)
+        idx_nm = np.where(t > tm[-1])[0]
+        t_grid = np.concatenate([t_grid, t[idx_nm]])
+        phase_grid = np.concatenate([phase_grid, phiI(t[idx_nm])])
+    
+    return t_grid, phase_grid
+
+#------------------------------------------------------------------------------
 def time_grid_for_longest_waveform(Mtot, f_min, outfile, deg=3, abstol=5e-5):
     # Note: This function is no longer used
     # 1) this is very slow for long waveforms (O(day))
@@ -331,7 +369,7 @@ def TEOB_process_array_TD(i, M,
         ampI = ip.InterpolatedUnivariateSpline(t, amp, k=3, ext='zeros')
  
         # Compute phase grid with variable number of points per cycle
-        t_grid, phase_grid = Generate_phase_grid(t, phi)
+        t_grid, phase_grid = Generate_phase_grid_extrapolate(t, phi)
 
         # use only non-zero amplitude data for constructing spline
         idx = np.where(amp > 0.0)

### 2017-07-18 paper/paper.tex
Tweak EOB formulas in paper

--- 2017-06-29 paper/paper.tex
@@
 Let $A$ and $B$ be the body labels.
-Let $M^{A,B}$ be the mass of either body and $M=M^A+M^B$ be the total mass of the binary.
+Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.
 Explicitly, including quadrupolar and octupolar tidal effects, we have
@@
 \begin{align}
- \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)\,u^6\frac{X_B}{X_A} \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
- &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)\,u^8 \frac{X_B}{X_A}\left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
+ \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
+ &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
 \end{align}
 \end{widetext}
-where $X_{A,B}=M^{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate, $\Lambda_{2,\textrm{dyn}}^{A,B}(u)$ are the dimensionless quadrupolar dynamical tidal polarizabilities, and $\Lambda_{3,\textrm{dyn}}^{A,B}(u)$ are the dimensionless octupolar dynamical tidal polarizabilities.
+where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate, $\Lambda_{2,\textrm{dyn}}^{A,B}(u)$ are the dimensionless quadrupolar dynamical tidal polarizabilities, and $\Lambda_{3,\textrm{dyn}}^{A,B}(u)$ are the dimensionless octupolar dynamical tidal polarizabilities.
 Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies.
@@
 \begin{equation}
-\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\frac{2}{(2\ell-1)!!}k^{A,B}_{\ell}\hat{k}^{A,B}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\left(\frac{R_{A,B}}{M}\right)^{2\ell+1}\,,
+\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\,,
 \end{equation}
-where $k^{A,B}_{\ell}$ is the tidal Love number and $R_{A,B}$ is the neutron star radius.
-Here $\hat{k}^{A,B}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode angular frequency of the $2^\ell$-th multipole, $\omega_{0\ell}^{A,B}$.
+where $k^{A,B}_{\ell}$ is the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ is the neutron star compactness, with $R_{A,B}$ the neutron star radius.
+Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole angular frequency, $\omega_{0\ell}^{A,B}$.
 

### 2017-07-19 paper/paper.tex
More work on EOB section

@@
 The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}.
-This research program has been extended to accommodate matter effects~\cite{}, which become relevant XXX
-
-References~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides.
+This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter --- typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs).
+The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case.
+In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides.
 Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star.
@@
 Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.
-Here we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
+
+In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
 \begin{equation}
@@
 \end{widetext}
-where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate, $\Lambda_{2,\textrm{dyn}}^{A,B}(u)$ are the dimensionless quadrupolar dynamical tidal polarizabilities, and $\Lambda_{3,\textrm{dyn}}^{A,B}(u)$ are the dimensionless octupolar dynamical tidal polarizabilities.
+where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
 Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies.
-In particular, the dimensionless dynamical polarizabilities read
+In particular, the dimensionless dynamical tidal polarizability reads
 \begin{equation}
-\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\,,
+\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\,,
 \end{equation}
-where $k^{A,B}_{\ell}$ is the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ is the neutron star compactness, with $R_{A,B}$ the neutron star radius.
+where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
+\begin{equation}
+\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
+\end{equation}
+with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star.
 Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole angular frequency, $\omega_{0\ell}^{A,B}$.
@@
 \begin{equation}
-h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{insp-plunge, pm}} + h_{\ell m}^{\textrm{tidal}}\,,
+h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{insp-plunge, pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
 \end{equation} 
@@
 
-The inspiral-plunge orbital dynamics is obtained by numerically solving Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s.
+The inspiral-plunge orbital dynamics is obtained by numerically integrating Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s: see Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}.
+The numerical integration starts from quasicircular initial conditions that are built following the prescription of Ref.~\cite{Buonanno:2005xu}.
+We exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics.
+When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case.
+
+Let us now discuss the construction of the full waveform, from the inspiral to the post-merger stage.
+In this work we only consider the dominant mode, $(\ell,m)=(2,2)$.
+The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$, that is the time when either the amplitude or the GW frequency of $h_{22}^{\textrm{insp-plunge}}$ reaches a global maximum, whichever occurs first.
+We model the post-merger GW emission --- i.e., at $t>t_{\textrm{match}}^{22}$ --- by a signal that tapers down in amplitude and has a constant GW frequency.
+In particular, we employ the amplitude of the merger-ringdown of the BBH model (as given by the magnitude of Eq.~(2.10) of Ref.~\cite{Bohe:2016gbl}), using as inputs the mass ratio and spins of the BNS/NSBH of interest.
+This choice is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}.
+The constant value of the GW frequency is taken to be that of $h_{22}^{\textrm{insp-plunge}}$ at time $t=t_{\textrm{match}}^{22}$.
+A more sophisticated, NR-informed, time-domain model of the post-merger emission will be part of future investigations.
+
+As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless spin components along the angular momentum of the binary $-1 \leq \chi_{A,B} \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
+We note that this list does not include the total mass $M$ of the system, since the point-mass part of the model is scale-invariant and tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$, and $\Lambda_2^{A,B}$, which is inherent to the PN analysis of tidal effects~\cite{Flanagan:2007ix}.
+To reduce the dimensionality of the intrinsic parameter space, we exploit universal (i.e., equation-of-state-independent) relations that are available in the literature.
+In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} with the coefficients from the first row of Table I therein expresses $\Lambda_3^{A,B}$ as a function of $\Lambda_2^{A,B}$ with an accuracy to within about 10\%, depending on the equation of state and the value of $\Lambda_2^{A,B}$.
+Moreover, Eq.
+(3.5) of Ref.~\cite{Chan:2014kua} with the coefficients from the first column of Table I therein gives $\omega_{02}^{A,B}$ as a function of $\Lambda_2^{A,B}$ to within a few percent, while the same equation but with the coefficients from the second column of Table I therein gives $\omega_{03}^{A,B}$ as a function of $\Lambda_3^{A,B}$ to within a few percent.
+Thanks to these relations, the waveform only depends on 5 intrinsic parameters, which we denote as $\boldsymbol{x}=\{q, \chi_A,\chi_B, \Lambda_2^A,\Lambda_2^B\}$.
 
 --Total mass is scaled out. This is fine, even up to zero amplitude point. Estimate the error in this approximation.
-
---Tapering fit to NR.
-
---Universal relations to relate all matter parameters to quadrupolar $\Lambda$.
-
---Results in 5 waveform parameters $\vec x=\{q, S_1, S_2, \Lambda_1, \Lambda_2\}$.
 

### 2017-07-20 paper/paper.tex
Add comment on comparisons to NR

@@
 A more sophisticated, NR-informed, time-domain model of the post-merger emission will be part of future investigations.
+
+Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state, finding that the model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state.
+Similar results hold for the tidal EOB model of Ref.~\cite{Bernuzzi:2014owa}, whose radial potential includes tidal effects in a form that is inspired by first-order gravitational-self-force (GSF) calculations of tidal invariants in the small-mass-ratio limit.
+Both the EOB model with dynamical tides and the GSF-inspired one are significantly more faithful to NR than the adiabatic tidal EOB model of Refs.~\cite{Damour:2009wj,Bini:2012gu}.
 

### 2017-07-20 paper/paper.tex
Spin effects in tidal EOB

@@
 Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects.
-State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries.
+State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries.
 The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}.
@@
 
+We note that, to date, this is the only tidal EOB model that includes the effect of spin.
+More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.
+The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning.
+This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled.
+The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.
+
 As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless spin components along the angular momentum of the binary $-1 \leq \chi_{A,B} \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
-We note that this list does not include the total mass $M$ of the system, since the point-mass part of the model is scale-invariant and tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$, and $\Lambda_2^{A,B}$, which is inherent to the PN analysis of tidal effects~\cite{Flanagan:2007ix}.
 To reduce the dimensionality of the intrinsic parameter space, we exploit universal (i.e., equation-of-state-independent) relations that are available in the literature.
@@
 Thanks to these relations, the waveform only depends on 5 intrinsic parameters, which we denote as $\boldsymbol{x}=\{q, \chi_A,\chi_B, \Lambda_2^A,\Lambda_2^B\}$.
-
---Total mass is scaled out. This is fine, even up to zero amplitude point. Estimate the error in this approximation.
+We note that this list does not include the total mass $M$ of the system.
+Of course, the point-mass part of the model is scale-invariant.
+As to the inspiral-plunge phase, in this model tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$, and $\Lambda_2^{A,B}$, which is inherent to the post-Newtonian analysis of tidal effects~\cite{Flanagan:2007ix}.
+Finally, the simple model of the post-merger signal that we employ does not depend on $M$ either, although a more sophisticated description will have to include such dependence to account for the phenomenology that is observed in NR simulations of this phase.
 

### 2017-07-20 paper/paper.tex
Include more explicit formulas in the EOB section

@@
 
-Let us now discuss the construction of the full waveform, from the inspiral to the post-merger stage.
+Let us now discuss the construction of the full waveform.
 In this work we only consider the dominant mode, $(\ell,m)=(2,2)$.
-The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$, that is the time when either the amplitude or the GW frequency of $h_{22}^{\textrm{insp-plunge}}$ reaches a global maximum, whichever occurs first.
-We model the post-merger GW emission --- i.e., at $t>t_{\textrm{match}}^{22}$ --- by a signal that tapers down in amplitude and has a constant GW frequency.
-In particular, we employ the amplitude of the merger-ringdown of the BBH model (as given by the magnitude of Eq.~(2.10) of Ref.~\cite{Bohe:2016gbl}), using as inputs the mass ratio and spins of the BNS/NSBH of interest.
+The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$.
+Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when $|h_{22}^{\textrm{insp-plunge}}|$ is maximum.
+Let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum.
+Then we put
+\begin{equation}
+t_{\textrm{match}}^{22} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.
+\end{equation}
+We model the post-merger GW emission by a signal that tapers down in amplitude and has a constant GW frequency.
+In particular, we employ the amplitude of the merger-ringdown of the underlying BBH model~\cite{Bohe:2016gbl}, using as inputs the mass ratio and spins of the BNS/NSBH of interest.
+Explicitly,
+\begin{equation}
+|h_{22}^{\textrm{post-mrg}}(t)|=\nu \tilde{A}_{22}(t)\exp{\left[-(t-t^{22}_{\textrm{match}})/\tau_{220}\right]}\,,
+\end{equation}
+where $\nu=M_A M_B/M^2$ is the symmetric mass ratio, $\tilde{A}_{22}(t)$ is a monotonic and bounded function given in Eq.~(2.12) of Ref.~\cite{Bohe:2016gbl}, and $\tau_{220}$ is the decay time of the least-damped quasinormal mode of the remnant BH that forms after the BBH merger.
 This choice is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}.
-The constant value of the GW frequency is taken to be that of $h_{22}^{\textrm{insp-plunge}}$ at time $t=t_{\textrm{match}}^{22}$.
+The constant value of the GW frequency is taken to be that of $h_{22}^{\textrm{insp-plunge}}$ at time $t=t_{\textrm{match}}^{22}$, so that the post-merger phase evolution reads
+\begin{align}
+\phi_{22}^{\textrm{post-mrg}}(t)&=\omega_{22}^{\textrm{insp-plunge}}(t_{\textrm{match}}^{22})(t-t_{\textrm{match}}^{22})\nonumber\\
+&+\phi_{22}^{\textrm{inspiral-plunge}}(t_{\textrm{match}}^{22})\,
+\end{align}
+where $\phi_{22}^{\textrm{inspiral-plunge}}$ is the phase of $h_{22}^{\textrm{insp-plunge}}$.
 A more sophisticated, NR-informed, time-domain model of the post-merger emission will be part of future investigations.
+Finally, the full waveform reads
+\begin{align}
+h_{22}(t) &= h_{22}^{\textrm{insp-plunge}}(t)\, \theta\left(t_{\textrm{match}}^{22}-t\right)\nonumber\\
+&+h_{22}^{\textrm{post-mrg}}(t)\, \theta\left(t-t_{\textrm{match}}^{22}\right)\,.
+\end{align}
 
@@
 
-As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless spin components along the angular momentum of the binary $-1 \leq \chi_{A,B} \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
+As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless components of the spins $\boldsymbol{S}_{A,B}$ along the Newtonian angular momentum ${\boldsymbol{L}}_N$, $-1 \leq \chi_{A,B}=(\boldsymbol{S}_{A,B}\cdot \hat{\boldsymbol{L}}_N)/M_{A,B}^2 \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
 To reduce the dimensionality of the intrinsic parameter space, we exploit universal (i.e., equation-of-state-independent) relations that are available in the literature.
 In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} with the coefficients from the first row of Table I therein expresses $\Lambda_3^{A,B}$ as a function of $\Lambda_2^{A,B}$ with an accuracy to within about 10\%, depending on the equation of state and the value of $\Lambda_2^{A,B}$.
-Moreover, Eq.
+While this relation has been fitted to NR data only for $1 \leq \Lambda_2^{A,B}\leq10^6$, we use it also outside such range.
+However, this formula is divergent at small values of $\Lambda_2^{A,B}$, thus in such regime we replace it with a polynomial function that vanishes at 0, equals the universal relation at $\Lambda_2^{A,B}=10^{-2}$, and best-fits it for $10^{-5}\leq\Lambda_2^{A,B}\leq10^{-2}$ Explicitly, putting $\zeta_{A,B}=\frac{\log\Lambda^{A,B}_2}{10}$,
+\begin{widetext}
+\begin{equation}
+\Lambda_3^{A,B}=\left\{\begin{array}{ll}
+\Lambda_2^{A,B}\left[0.441 - 34.6\,\Lambda_2^{A,B} + 1760(\Lambda_2^{A,B})^2\right]\,,&\textrm{if}\quad0\leq\Lambda_2^{A,B}\leq10^{-2}\\
+&\\
+\exp{\left[-1.15 +11.8\,\zeta_{A,B}+2.51\,\zeta_{A,B}^2-1.31\,\zeta_{A,B}^3+0.252\,\zeta_{A,B}^4\right]}\,,&\textrm{otherwise}\,.
+\end{array}\right.
+\end{equation}
+\end{widetext}
+Eq.
 (3.5) of Ref.~\cite{Chan:2014kua} with the coefficients from the first column of Table I therein gives $\omega_{02}^{A,B}$ as a function of $\Lambda_2^{A,B}$ to within a few percent, while the same equation but with the coefficients from the second column of Table I therein gives $\omega_{03}^{A,B}$ as a function of $\Lambda_3^{A,B}$ to within a few percent.
+Putting $\xi_{A,B}=\log\Lambda_2^{A,B}$, over the fitting range used by Ref.~\cite{Chan:2014kua} (i.e., $0\leq \xi_{A,B} \leq 9$), we set
+\begin{align}
+M_{A,B}\omega_{02}^{A,B}&=
+0.182 -6.836\times10^{-3}\xi_{A,B}\nonumber\\
+&-4.196\times10^{-3}\xi_{A,B}^2+5.215\times10^{-4}\xi_{A,B}^3\nonumber\\
+&-1.857\times10^{-5}\xi_{A,B}^4\,;\label{URomega2}
+\end{align}
+outside this range, we set $M_{A,B}\omega_{02}^{A,B}=0.182$ if $\xi_{A,B} < 1$ and $M_{A,B}\omega_{02}^{A,B}=0.161$ if $\xi_{A,B} >9$, that is Eq.~(\ref{URomega2}) evaluated for $\xi_{A,B}=1$ and $\xi_{A,B}=9$, respectively.
+Moreover, putting $\Upsilon_{A,B}=\log\Lambda_3^{A,B}$, over the fitting range used by Ref.~\cite{Chan:2014kua} (i.e., $-1 \leq \Upsilon_{A,B} \leq 10$), we set
+\begin{align}
+M_{A,B}\omega_{03}^{A,B}&=
+\,0.2245 -1.5\times10^{-2}\Upsilon_{A,B}\nonumber\\
+&-1.412\times10^{-3}\Upsilon_{A,B}^2+1.832\times10^{-4}\Upsilon_{A,B}^3\nonumber\\
+&-5.561\times10^{-6}\Upsilon_{A,B}^4\,;\label{URomega3}
+\end{align}
+outside this range, we set $M_{A,B}\omega_{03}^{A,B}=0.2379$ if $\Upsilon_{A,B} < -1$ and $M_{A,B}\omega_{03}^{A,B}=0.1165$ if $\Upsilon_{A,B} >10$, that is Eq.~(\ref{URomega3}) evaluated for $\Upsilon_{A,B}=-1$ and $\Upsilon_{A,B}=10$, respectively.
 Thanks to these relations, the waveform only depends on 5 intrinsic parameters, which we denote as $\boldsymbol{x}=\{q, \chi_A,\chi_B, \Lambda_2^A,\Lambda_2^B\}$.

### 2017-07-24 paper/paper.tex
Initial version of abstract and introduction without references.

--- 2017-07-20 paper/paper.tex
@@
 
-\title{Surrogate model for aligned-spin binary neutron star waveforms using gaussian process regression}
+\title{Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression}
 
@@
 
-Construct 5d hierarchical surrogate model using reduced bases and GPR. Includes $\ell=2,3,4$, dynamical tides, 
-and tapering fit to NR.
+Fast and accurate waveform models are necessary for measuring properties of inspiraling binary neutron star systems such as the masses, spins, and tidal parameters.
+The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as corrections for dynamical tides, that are accurate enough to avoid significant biases in the measured parameters.
+However, current implementations take at least 10 minutes to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations.
+We construct a frequency-domain surrogate model that requires less than 0.2~s to evaluate for a waveform with a starting frequency of 10~Hz (0.05~s for a starting frequency of 30~Hz).
+We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter.
+We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with approximately 500 waveforms instead of the $>10^5$ waveforms that would be required with standard, grid-based interpolation methods.
+For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.
 
-Provide framework for building high dimensional surrogate from NR.
+%Finally, I will show how this same method can be used to optimally choose parameters for numerical relativity simulations to correct the the EOB surrogate during the last several orbits and merger.
 
@@
 
-The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO), Virgo, and KAGRA which are now making routine detections of binary black hole (BBH) systems or will begin in the next few years.
+%The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO), Virgo, and KAGRA which are now making routine detections of binary black hole (BBH) systems or will begin in the next few years.
 Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]].
@@
 
-The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters.
-These parameters are measured by matched-filtering of the GW data with waveform models, and require highly accurate waveform models to reliably measure the parameters with minimal systematic errors.
+The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO) [[cite]], now making routine detections of binary black hole (BBH) systems [[cite]], as well as Virgo [[cite]], KAGRA [[cite]], and LIGO-India [[cite]].
+Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]], and the nondetection of BNS events in aLIGO's first observing run (O1) is still consistent with the upper end of this range to 90\% confidence [[O1UpperLimits]].
+However, continued nondetection in runs now underway will narrow down this range.
+
+%The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters.
+These parameters are measured with Bayesian parameter estimation that compares the GW data with waveform templates generated on-the-fly, and require highly accurate waveform models to reliably measure the parameters with minimal systematic errors.
 [[Review of systematic error BNS papers...]] Previous work on estimating the tidal parameters, for example, has shown that even phase uncertainties of several radians, as is present in current 3.5-post-Newtonian waveform models, leads to systematic errors that are as large as the statistical errors [[cite]] and this also biases the recovered NS equation of state as well [[cite]].
@@
 
-The effective one body (EOB) formalism and numerical relativity (NR) simulations, on the other hand, have shown significantly smaller uncertainties for BNS systems in their respective ranges of validity than PN waveforms.
-For the inspiral, EOB waveforms have total phase uncertainties of $\sim 1$~rad as compared to PN waveforms with phase uncertainties of $\sim 10$~rad as determined by the variation from freedom within each formalism [[explain better.
-show your plot for this.]].
-NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]].
+The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters.
+These parameters, measured with Bayesian parameter estimation techniques, require highly accurate waveform models to avoid systematic errors.
+For example, post-Newtonian (PN) waveform models which are truncated at the currently known 3.5PN order, lead to a uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz.
+This is approximately the same size as the contribution to the phase from tidal effects, and as a result the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]], and this also biases the recovered NS equation of state as well [[cite]].
+Uncertainties in the PN waveform model also lead to significant biases in the recovered masses and spins as well [[cite]].
 
-Unfortunately, available codes for generating EOB waveforms for BNS systems can take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]].
+The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms.
+Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveforms uncertainties of $\sim 10$~rad.
+NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger.
+Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]].
+Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable.
+We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.
 
-These quantities are measured with Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling.
+Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\mathbf{x})$ given its waveform parameters $\mathbf{x}$.
+The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce the original training set of waveforms to some desired accuracy.
+This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions.
+In previous work, this reduced basis has been constructed either with singular value decomposition (SVD) [[Michael's papers]] or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization [[All the other papers]].
 
-Previous work on surrogate models has focused mainly on compressing the waveform parameter space into a linear combination of a small number of basis functions using either greedy methods [[cite]] or singular value decomposition [[cite]].
-Significantly less work has been performed on optimally sampling waveforms so that the coefficients of these linear combinations of waveforms can be chosen through fitting or interpolation with the fewest number of waveform evaluations.
-Standard techniques such as tensor spline or Chebyshev interpolation require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
+Once this reduced basis is built, one can interpolate each of the coefficients $c_i(\mathbf{x})$ as a function of the waveform parameters $\mathbf{x}$.
+This method has been used for frequency-domain surrogates of aligned spin BBH waveforms [[Michael's papers]].
+Alternatively, one can use the empirical interpolation method [[cite]] which re-expresses the linear combination of orthonormal basis functions in terms of a sum of empirical interpolating functions $B_j(t)$ times the value of the waveform $h(T_j; \mathbf{x})$ at empirical nodes $T_j$.
+One then interpolates the waveform $h(T_j; \mathbf{x})$ as a function of $\mathbf{x}$ at each node $T_j$.
+This method has been used for time domain-waveforms for nonspinning [[cite]], aligned-spin [[cite]], and precessing [[cite]] BBH waveforms as well as nonspinning BNS waveforms with tidal interactions [[cite]], as well as for frequency-domain waveforms [[cite]].
+
+Previous work on surrogate models has focused mainly on parameter spaces $\mathbf{x}$ with $<=3$~dimensions.
+This allows one to approximate the coefficients $c_i(\mathbf{x})$ or waveform values $h(T_j; \mathbf{x})$ as functions of $\mathbf{x}$ using simple fitting functions [[cite]], or standard interpolation techniques such as tensor spline [[Michael, others]] or Chebyshev interpolation [[Ben]].
+These interpolation techniques typically require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
 For the 5-dimensional problem here, this is somewhat unfeasable.
@@
 
-In this paper, we will develop methods to construct surrogate models from EOB and NR waveform codes that accurately reproduce waveforms from this high-dimensional parameter space, using as few waveform evaluations as possible.
-We will construct frequency-domain surrogates of the amplitude and phase.
+In this paper we focus on breaking this curse of dimensionality by using an interpolation method known as Gaussian process regression (GPR) that does not require a regular grid.
+Importantly, GPR also provides a convenient estimate of its own uncertainty.
+This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set.
+A similar approach has recently been used to optimally sample waveforms for a 2-dimensional parameter space using GPR [[DoctorEtAl]].
+GPR has also been used recently in GW data analysis to marginalize over waveform uncertainties in parameter estimation [[ChrisMoore2papers?]].
 
---Overview of what was done in nonspinning BNS surrogate paper, and the improvements used for constructing aligned-spin surrogate.
+In this paper, we will develop methods to construct a surrogate model for the inspiral through merger of an aligned spin BNS waveform with tidal interactions using the EOB tidal description described in Hinderer et al.~[[cite1, cite2]].
+This work is a continuation of the surrogate developed in [[cite]] for a nonspinning BNS waveform a using slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~[[cite, ...]].
+Whereas in [[Ben]] we constructed a time-domain surrogate for the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation.
 
---Advantages of using frequency domain surrogate (and why you didn't use it in last paper).
-
-[[A notable effect that has not been included is the spin induced quadrupole moment of each NS.]]
-
-We organize the paper as follows. In Section...
+We organize the paper as follows.
+In Section II we describe the EOB waveform model used for constructing the surrogate and the approximations we use to reduce the dimensionality of the parameter space.
+In section III, we describe the details of building the surrogate model, choosing training set waveforms, and evaluating the final model.
+We then compare the accuracy and speed of the surrogate to the original model in section IV.
+Then, finally discuss future improvements and applications in Section V.
 
 \textit{Conventions:} Unless explicitly stated, we use units where $G=c=1$.
-To agree with the machine learning literature, we will refer to the waveform parameters (features) as $\vec x$ and the hyperparameters in the Gaussian Process Regression model as $\vec\theta$.
+To agree with the machine learning literature, we will refer to the waveform parameters (features) as $\mathbf{x}$ and the hyperparameters in the Gaussian Process Regression model as $\mathbf{\theta}$.
 
@@
 
-%% ________________________________________________________
-\section{Evaluating waveform model accuracy}
-
-In the sections below, we will use several criteria to choose waveforms samples and evaluate the accuracy of the 
-surrogate waveform model. We briefly discuss them here.
-
---Phase error. (needs to be significantly smaller than tidal effects) Use this for order of magnitude estimate of required
-accuracy.
-
---distinguishability
-
---Mismatch
-
-As another common measure of the surrogate model accuracy, we examine the mismatch between 
-the surrogate model and the original EOB waveform.
-The mismatch represents the loss in signal-to-noise ratio that would result 
-from using the surrogate model instead of the original EOB waveform. 
-It is defined by the deviation from a perfect overlap after aligning the two waveforms
-using the time and phase free parameters $t_0$ and $\phi_0$:
-\begin{equation}
-\mathcal{M} = 1 - \max_{t_0, \phi_0} \frac{(h_{\rm EOB}, h_{\rm Sur})} {\sqrt{(h_{\rm EOB}, h_{\rm EOB}) (h_{\rm Sur}, h_{\rm Sur})}}.
-\end{equation}
-The inner product here  is the integral of the Fourier transformed waveforms $\tilde h(f)$ weighted by the noise power spectral 
-density (PSD) $S_n(f)$ of the detector:
-\begin{equation}
-(h_1, h_2) = 4 \Re \int_{f_{\rm low}}^{f_{\rm high}} \frac{\tilde h_1(f) \tilde h^*_2(f)} {S_n(f)} df.
-\end{equation}
-
---Fisher matrix systematic error. (See spin NSBH paper.)
+\red{A notable effect that has not been included is the spin induced quadrupole moment of each NS [[Poisson etc.]]. The impact on detection and parameter estimation is discussed in [[HarryHinderer]].}
 
@@
 %% ________________________________________________________
+\section{Evaluating waveform model accuracy}
+
+In the sections below, we will use several criteria to choose waveforms samples and evaluate the accuracy of the 
+surrogate waveform model. We briefly discuss them here.
+
+--Phase error. (needs to be significantly smaller than tidal effects) Use this for order of magnitude estimate of required
+accuracy.
+
+--distinguishability
+
+--Mismatch
+
+As another common measure of the surrogate model accuracy, we examine the mismatch between 
+the surrogate model and the original EOB waveform.
+The mismatch represents the loss in signal-to-noise ratio that would result 
+from using the surrogate model instead of the original EOB waveform. 
+It is defined by the deviation from a perfect overlap after aligning the two waveforms
+using the time and phase free parameters $t_0$ and $\phi_0$:
+\begin{equation}
+\mathcal{M} = 1 - \max_{t_0, \phi_0} \frac{(h_{\rm EOB}, h_{\rm Sur})} {\sqrt{(h_{\rm EOB}, h_{\rm EOB}) (h_{\rm Sur}, h_{\rm Sur})}}.
+\end{equation}
+The inner product here  is the integral of the Fourier transformed waveforms $\tilde h(f)$ weighted by the noise power spectral 
+density (PSD) $S_n(f)$ of the detector:
+\begin{equation}
+(h_1, h_2) = 4 \Re \int_{f_{\rm low}}^{f_{\rm high}} \frac{\tilde h_1(f) \tilde h^*_2(f)} {S_n(f)} df.
+\end{equation}
+
+--Fisher matrix systematic error. (See spin NSBH paper.)
+
+%% ________________________________________________________
 \begin{acknowledgments}

### 2017-07-25 paper/paper.tex
Cleaned up universal relations subsection of EOB section.

@@
 \section{Aligned-spin, dynamical tides EOB model}
+
+\subsection{Inspiral-plunge waveform}
+
 Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects.
@@
 The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}.
-This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter --- typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs).
+This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter---typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs).
 The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case.
 In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides.
-Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star.
+Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion---be it a black hole or another neutron star.
 Because of binary motion, the forcing tidal field varies at the orbital frequency.
@@
 The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field.
+
 Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.
@@
 
+Let us now discuss the construction of the full waveform.
+The numerical integration starts from quasicircular initial conditions that are built following the prescription of Ref.~\cite{Buonanno:2005xu}.
+
 The inspiral-plunge orbital dynamics is obtained by numerically integrating Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s: see Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}.
-The numerical integration starts from quasicircular initial conditions that are built following the prescription of Ref.~\cite{Buonanno:2005xu}.
-We exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics.
-When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case.
 
-Let us now discuss the construction of the full waveform.
-In this work we only consider the dominant mode, $(\ell,m)=(2,2)$.
 The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$.
@@
 \end{equation}
+In this work we only consider the dominant mode, $(\ell,m)=(2,2)$.
+
+%We exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics.
+When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case.
+
+\subsection{Post-merger waveform}
+
+\red{[[Mention that this is just a tapering and does not include post-merger oscillations seen in numerical simulations.]]}
+
 We model the post-merger GW emission by a signal that tapers down in amplitude and has a constant GW frequency.
@@
 
-As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless components of the spins $\boldsymbol{S}_{A,B}$ along the Newtonian angular momentum ${\boldsymbol{L}}_N$, $-1 \leq \chi_{A,B}=(\boldsymbol{S}_{A,B}\cdot \hat{\boldsymbol{L}}_N)/M_{A,B}^2 \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
+\subsection{Approximation of higher-order tidal effects}
+
+As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless components of the spins along the Newtonian orbital angular momentum $-1 \leq S_{A,B} \leq 1$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
+To reduce the dimensionality of the intrinsic parameter space, we exploit nearly EOS-independent fits for the higher-order effects in terms of $\Lambda_2$ usually referred to as universal relations.
+In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} provides a fit for $\Lambda_3$ as a function of $\Lambda_2$ with an accuracy of about 10\%, depending on the EOS.
+This relation has been fit for a sample of EOSs and for tidal parameters in the range $1 \leq \Lambda_2\leq10^6$.
+This formula, however, diverges at small values of $\Lambda_2$, so we replace it with a polynomial function that vanishes at $\Lambda_2 = 0$.
+We require this extension to be continuous and smooth at $\Lambda_2 = 10^{-2}$, and to fit the universal relation in the range $10^{-5}\leq\Lambda_2^{A,B}\leq10^{-2}$.
+The exact relation we use is
+\begin{equation}
+\Lambda_3 = \left\{\begin{array}{ll}
+\Lambda_2 (f + g \Lambda_2 + h \Lambda_2^2)\, , & 0 \leq \Lambda_2 \leq 10^{-2} \\
+e^{a + b \xi + c \xi^2 + d \xi^3 + e \xi^4}\, , & \Lambda_2>10^{-2}
+\end{array}\right.,
+\end{equation}
+where $\xi = \ln\Lambda_2$, $\{a, b, c, d, e\} = \{-1.15, 1.18, 2.51\times 10^{-2}, -1.31\times 10^{-3}, 2.52\times 10^{-5}\}$ from Ref.~\cite{Yagi:2013sva}, and $\{f, g, h\}=\{0.441, -34.6, 1760\}$.
+
+Eq.
+(3.5) of Ref.~\cite{Chan:2014kua} gives relations for $\omega_{02}$ as a function of $\Lambda_2$ and $\omega_{03}$ as a function of $\Lambda_3$ to within a few percent error.
+For $\omega_{02}$, the fitting range used by Ref.~\cite{Chan:2014kua} was $1\leq \xi \leq 9$, and outside this range we require continuity.
+The relation we use is then
+\begin{equation}
+\omega_{02} = \left\{\begin{array}{ll}
+f\, , & \xi < 1 \\
+a + b\xi + c\xi^2 + d\xi^3 + e\xi^4\, , & 1 \le \xi \le 9 \\
+g\, , & \xi > 9
+\end{array}\right.,
+\end{equation}
+where $\{a, b, c, d, e\} = \{0.182, -6.836\times10^{-3}, -4.196\times10^{-3}, 5.215\times10^{-4}, -1.857\times10^{-5}\}$ from Ref.~\cite{Chan:2014kua}, and $\{f, g\}=\{0.182, 0.161\}$.
+For $\omega_{03}$, the fit is given in terms of $\Upsilon=\ln\Lambda_3$ in the range $-1\leq \Upsilon \leq 10$, and outside this range we require continuity.
+The relation we use is then
+\begin{equation}
+\omega_{03} = \left\{\begin{array}{ll}
+f\, , & \Upsilon < -1 \\
+a + b\Upsilon + c\Upsilon^2 + d\Upsilon^3 + e\Upsilon^4\, , & -1 \le \Upsilon \le 10 \\
+g\, , & \Upsilon > 10
+\end{array}\right.,
+\end{equation}
+where $\{a, b, c, d, e\} = \{0.2245, -1.5\times10^{-2}, -1.412\times10^{-3}, 1.832\times10^{-4}, -5.561\times10^{-6}\}$ and $\{f, g\}=\{0.2379, 0.1165\}$.
+
+With these relations the waveform only depends on 5 intrinsic parameters $\mathbf{x}=\{q, S_A, S_B, \Lambda_2^A,\Lambda_2^B\}$.
+We note that this list does not include the total mass $M$ of the system.
+The point-mass part of the model is scale-invariant, and the tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$ and $\Lambda_2^{A,B}$.
+Finally, the simple model of the post-merger signal that we employ does not depend on $M$ either.
+It is not clear at what point in the transition from the inspiral to the post-merger this approximation breaks down.
+For example, the frequency of post-merger oscillations depends sensitively on the total mass of the hypermassive remnant.
+A set of NR simulations with a fixed mass ratio and tidal parameters but different total masses (and therefore different EOSs in order to keep the tidal parameters fixed) will be needed to determine the time when this approximation breaks down.
+
+%%% ________________________________________________________
+%\section{Aligned-spin, dynamical tides EOB model}
+%Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects.
+State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries.
+The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}.
+This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter --- typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs).
+The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case.
+In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides.
+Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star.
+Because of binary motion, the forcing tidal field varies at the orbital frequency.
+Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode.
+The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field.
+Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.
+%
+%In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
+%\begin{equation}
+%\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
+%\end{equation}
+%where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides.
+We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below.
+Let $A$ and $B$ be the body labels.
+Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.
+Explicitly, including quadrupolar and octupolar tidal effects, we have
+%\begin{widetext}
+%\begin{align}
+% \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
+% &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
+%\end{align}
+%\end{widetext}
+%where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
+Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies.
+In particular, the dimensionless dynamical tidal polarizability reads
+%\begin{equation}
+%\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\,,
+%\end{equation}
+%where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
+%\begin{equation}
+%\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
+%\end{equation}
+%with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star.
+Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole angular frequency, $\omega_{0\ell}^{A,B}$.
+%
+%Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
+%\begin{equation}
+%h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{insp-plunge, pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
+%\end{equation} 
+%where $h_{\ell m}^{\textrm{insp-plunge, pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs.
+(A14)-(A17) of Ref.~\cite{Damour:2012yf}.
+Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
+%
+%The inspiral-plunge orbital dynamics is obtained by numerically integrating Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s: see Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}.
+The numerical integration starts from quasicircular initial conditions that are built following the prescription of Ref.~\cite{Buonanno:2005xu}.
+We exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics.
+When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case.
+%
+%Let us now discuss the construction of the full waveform.
+In this work we only consider the dominant mode, $(\ell,m)=(2,2)$.
+The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$.
+Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when $|h_{22}^{\textrm{insp-plunge}}|$ is maximum.
+Let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum.
+Then we put
+%\begin{equation}
+%t_{\textrm{match}}^{22} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.
+%\end{equation}
+%We model the post-merger GW emission by a signal that tapers down in amplitude and has a constant GW frequency.
+In particular, we employ the amplitude of the merger-ringdown of the underlying BBH model~\cite{Bohe:2016gbl}, using as inputs the mass ratio and spins of the BNS/NSBH of interest.
+Explicitly,
+%\begin{equation}
+%|h_{22}^{\textrm{post-mrg}}(t)|=\nu \tilde{A}_{22}(t)\exp{\left[-(t-t^{22}_{\textrm{match}})/\tau_{220}\right]}\,,
+%\end{equation}
+%where $\nu=M_A M_B/M^2$ is the symmetric mass ratio, $\tilde{A}_{22}(t)$ is a monotonic and bounded function given in Eq.~(2.12) of Ref.~\cite{Bohe:2016gbl}, and $\tau_{220}$ is the decay time of the least-damped quasinormal mode of the remnant BH that forms after the BBH merger.
+This choice is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}.
+The constant value of the GW frequency is taken to be that of $h_{22}^{\textrm{insp-plunge}}$ at time $t=t_{\textrm{match}}^{22}$, so that the post-merger phase evolution reads
+%\begin{align}
+%\phi_{22}^{\textrm{post-mrg}}(t)&=\omega_{22}^{\textrm{insp-plunge}}(t_{\textrm{match}}^{22})(t-t_{\textrm{match}}^{22})\nonumber\\
+%&+\phi_{22}^{\textrm{inspiral-plunge}}(t_{\textrm{match}}^{22})\,
+%\end{align}
+%where $\phi_{22}^{\textrm{inspiral-plunge}}$ is the phase of $h_{22}^{\textrm{insp-plunge}}$.
+A more sophisticated, NR-informed, time-domain model of the post-merger emission will be part of future investigations.
+Finally, the full waveform reads
+%\begin{align}
+%h_{22}(t) &= h_{22}^{\textrm{insp-plunge}}(t)\, \theta\left(t_{\textrm{match}}^{22}-t\right)\nonumber\\
+%&+h_{22}^{\textrm{post-mrg}}(t)\, \theta\left(t-t_{\textrm{match}}^{22}\right)\,.
+%\end{align}
+%
+%Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state, finding that the model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state.
+Similar results hold for the tidal EOB model of Ref.~\cite{Bernuzzi:2014owa}, whose radial potential includes tidal effects in a form that is inspired by first-order gravitational-self-force (GSF) calculations of tidal invariants in the small-mass-ratio limit.
+Both the EOB model with dynamical tides and the GSF-inspired one are significantly more faithful to NR than the adiabatic tidal EOB model of Refs.~\cite{Damour:2009wj,Bini:2012gu}.
+%
+%We note that, to date, this is the only tidal EOB model that includes the effect of spin.
+More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.
+The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning.
+This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled.
+The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.
+%
+%As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless components of the spins $\boldsymbol{S}_{A,B}$ along the Newtonian angular momentum ${\boldsymbol{L}}_N$, $-1 \leq \chi_{A,B}=(\boldsymbol{S}_{A,B}\cdot \hat{\boldsymbol{L}}_N)/M_{A,B}^2 \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
 To reduce the dimensionality of the intrinsic parameter space, we exploit universal (i.e., equation-of-state-independent) relations that are available in the literature.
@@
 However, this formula is divergent at small values of $\Lambda_2^{A,B}$, thus in such regime we replace it with a polynomial function that vanishes at 0, equals the universal relation at $\Lambda_2^{A,B}=10^{-2}$, and best-fits it for $10^{-5}\leq\Lambda_2^{A,B}\leq10^{-2}$ Explicitly, putting $\zeta_{A,B}=\frac{\log\Lambda^{A,B}_2}{10}$,
-\begin{widetext}
-\begin{equation}
-\Lambda_3^{A,B}=\left\{\begin{array}{ll}
-\Lambda_2^{A,B}\left[0.441 - 34.6\,\Lambda_2^{A,B} + 1760(\Lambda_2^{A,B})^2\right]\,,&\textrm{if}\quad0\leq\Lambda_2^{A,B}\leq10^{-2}\\
-&\\
-\exp{\left[-1.15 +11.8\,\zeta_{A,B}+2.51\,\zeta_{A,B}^2-1.31\,\zeta_{A,B}^3+0.252\,\zeta_{A,B}^4\right]}\,,&\textrm{otherwise}\,.
-\end{array}\right.
-\end{equation}
-\end{widetext}
-Eq.
+%\begin{widetext}
+%\begin{equation}
+%\Lambda_3^{A,B}=\left\{\begin{array}{ll}
+%\Lambda_2^{A,B}\left[0.441 - 34.6\,\Lambda_2^{A,B} + 1760(\Lambda_2^{A,B})^2\right]\,,&\textrm{if}\quad0\leq\Lambda_2^{A,B}\leq10^{-2}\\
+%&\\
+%\exp{\left[-1.15 +11.8\,\zeta_{A,B}+2.51\,\zeta_{A,B}^2-1.31\,\zeta_{A,B}^3+0.252\,\zeta_{A,B}^4\right]}\,,&\textrm{otherwise}\,.
+%\end{array}\right.
+%\end{equation}
+%\end{widetext}
+%Eq.
 (3.5) of Ref.~\cite{Chan:2014kua} with the coefficients from the first column of Table I therein gives $\omega_{02}^{A,B}$ as a function of $\Lambda_2^{A,B}$ to within a few percent, while the same equation but with the coefficients from the second column of Table I therein gives $\omega_{03}^{A,B}$ as a function of $\Lambda_3^{A,B}$ to within a few percent.
 Putting $\xi_{A,B}=\log\Lambda_2^{A,B}$, over the fitting range used by Ref.~\cite{Chan:2014kua} (i.e., $0\leq \xi_{A,B} \leq 9$), we set
-\begin{align}
-M_{A,B}\omega_{02}^{A,B}&=
-0.182 -6.836\times10^{-3}\xi_{A,B}\nonumber\\
-&-4.196\times10^{-3}\xi_{A,B}^2+5.215\times10^{-4}\xi_{A,B}^3\nonumber\\
-&-1.857\times10^{-5}\xi_{A,B}^4\,;\label{URomega2}
-\end{align}
-outside this range, we set $M_{A,B}\omega_{02}^{A,B}=0.182$ if $\xi_{A,B} < 1$ and $M_{A,B}\omega_{02}^{A,B}=0.161$ if $\xi_{A,B} >9$, that is Eq.~(\ref{URomega2}) evaluated for $\xi_{A,B}=1$ and $\xi_{A,B}=9$, respectively.
+%\begin{align}
+%M_{A,B}\omega_{02}^{A,B}&=
+%0.182 -6.836\times10^{-3}\xi_{A,B}\nonumber\\
+%&-4.196\times10^{-3}\xi_{A,B}^2+5.215\times10^{-4}\xi_{A,B}^3\nonumber\\
+%&-1.857\times10^{-5}\xi_{A,B}^4\,;\label{URomega2}
+%\end{align}
+%outside this range, we set $M_{A,B}\omega_{02}^{A,B}=0.182$ if $\xi_{A,B} < 1$ and $M_{A,B}\omega_{02}^{A,B}=0.161$ if $\xi_{A,B} >9$, that is Eq.~(\ref{URomega2}) evaluated for $\xi_{A,B}=1$ and $\xi_{A,B}=9$, respectively.
 Moreover, putting $\Upsilon_{A,B}=\log\Lambda_3^{A,B}$, over the fitting range used by Ref.~\cite{Chan:2014kua} (i.e., $-1 \leq \Upsilon_{A,B} \leq 10$), we set
-\begin{align}
-M_{A,B}\omega_{03}^{A,B}&=
-\,0.2245 -1.5\times10^{-2}\Upsilon_{A,B}\nonumber\\
-&-1.412\times10^{-3}\Upsilon_{A,B}^2+1.832\times10^{-4}\Upsilon_{A,B}^3\nonumber\\
-&-5.561\times10^{-6}\Upsilon_{A,B}^4\,;\label{URomega3}
-\end{align}
-outside this range, we set $M_{A,B}\omega_{03}^{A,B}=0.2379$ if $\Upsilon_{A,B} < -1$ and $M_{A,B}\omega_{03}^{A,B}=0.1165$ if $\Upsilon_{A,B} >10$, that is Eq.~(\ref{URomega3}) evaluated for $\Upsilon_{A,B}=-1$ and $\Upsilon_{A,B}=10$, respectively.
-Thanks to these relations, the waveform only depends on 5 intrinsic parameters, which we denote as $\boldsymbol{x}=\{q, \chi_A,\chi_B, \Lambda_2^A,\Lambda_2^B\}$.
+%\begin{align}
+%M_{A,B}\omega_{03}^{A,B}&=
+%\,0.2245 -1.5\times10^{-2}\Upsilon_{A,B}\nonumber\\
+%&-1.412\times10^{-3}\Upsilon_{A,B}^2+1.832\times10^{-4}\Upsilon_{A,B}^3\nonumber\\
+%&-5.561\times10^{-6}\Upsilon_{A,B}^4\,;\label{URomega3}
+%\end{align}
+%outside this range, we set $M_{A,B}\omega_{03}^{A,B}=0.2379$ if $\Upsilon_{A,B} < -1$ and $M_{A,B}\omega_{03}^{A,B}=0.1165$ if $\Upsilon_{A,B} >10$, that is Eq.~(\ref{URomega3}) evaluated for $\Upsilon_{A,B}=-1$ and $\Upsilon_{A,B}=10$, respectively.
+%Thanks to these relations, the waveform only depends on 5 intrinsic parameters, which we denote as $\boldsymbol{x}=\{q, \chi_A,\chi_B, \Lambda_2^A,\Lambda_2^B\}$.
 We note that this list does not include the total mass $M$ of the system.
@@
 Finally, the simple model of the post-merger signal that we employ does not depend on $M$ either, although a more sophisticated description will have to include such dependence to account for the phenomenology that is observed in NR simulations of this phase.
-
-\red{A notable effect that has not been included is the spin induced quadrupole moment of each NS [[Poisson etc.]]. The impact on detection and parameter estimation is discussed in [[HarryHinderer]].}
+%
+%\red{A notable effect that has not been included is the spin induced quadrupole moment of each NS [[Poisson etc.]].
+The impact on detection and parameter estimation is discussed in [[HarryHinderer]].}
 

### 2017-07-27 paper/paper.tex
More reorganization/suggestions for the EOB section.

@@
 
-\def\l{\ell}
-\def\blambda{{\bm \lambda}}
-\def\bLambda{{\bm \Lambda}}
-\def\blam{\bar{\lambda}}
-\def\Lam{\Lambda}
-\def\Lamt{\tilde{\Lambda}}
-\def\dLamt{\tilde{\delta{\Lambda}}}
-\def\bmu{{\bm \mu}}
-\def\btheta{{\bm \theta}}
+%\def\l{\ell}
+%\def\blambda{{\bm \lambda}}
+%\def\bLambda{{\bm \Lambda}}
+%\def\blam{\bar{\lambda}}
+%\def\Lam{\Lambda}
+%\def\Lamt{\tilde{\Lambda}}
+%\def\dLamt{\tilde{\delta{\Lambda}}}
+%\def\bmu{{\bm \mu}}
+
+\def\bx{\mathbf{x}}
+\def\btheta{\boldsymbol{\theta}}
 
@@
 We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter.
-We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with approximately 500 waveforms instead of the $>10^5$ waveforms that would be required with standard, grid-based interpolation methods.
+We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with approximately 500 waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods.
 For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.
@@
 
-\red{Change arrow vectors to bold vectors. ${\bm x}^i$ (bold) for example $i$. $x_j$ (not bold) for coordinate/parameter/feature $j$. 
-$x_j^i$ (not bold) for feature $j$ of example $i$.}
+\red{[[Change arrow vectors to bold vectors. ${\bm x}^i$ (bold) for example $i$. $x_j$ (not bold) for coordinate/parameter/feature $j$. 
+$x_j^i$ (not bold) for feature $j$ of example $i$.]]}
 
@@
 
-%The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO), Virgo, and KAGRA which are now making routine detections of binary black hole (BBH) systems or will begin in the next few years.
-Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]].
-From the nondetection of BNS events in aLIGO's first observing run (O1) the 90\% upper limit is still consistent with the upper end of this range [[O1UpperLimits]].
-However, continued nondetection will begin to narrow down this range.
-
-The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO) [[cite]], now making routine detections of binary black hole (BBH) systems [[cite]], as well as Virgo [[cite]], KAGRA [[cite]], and LIGO-India [[cite]].
+The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO) [[cite]] now making routine detections of binary black hole (BBH) systems [[cite]], as well as Virgo [[cite]], KAGRA [[cite]], and LIGO-India [[cite]].
 Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]], and the nondetection of BNS events in aLIGO's first observing run (O1) is still consistent with the upper end of this range to 90\% confidence [[O1UpperLimits]].
-However, continued nondetection in runs now underway will narrow down this range.
-
-%The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters.
-These parameters are measured with Bayesian parameter estimation that compares the GW data with waveform templates generated on-the-fly, and require highly accurate waveform models to reliably measure the parameters with minimal systematic errors.
-[[Review of systematic error BNS papers...]] Previous work on estimating the tidal parameters, for example, has shown that even phase uncertainties of several radians, as is present in current 3.5-post-Newtonian waveform models, leads to systematic errors that are as large as the statistical errors [[cite]] and this also biases the recovered NS equation of state as well [[cite]].
-Uncertainties in the PN waveform model also lead to significant biases in the recovered masses and spins as well [[cite]].
-[[What about sky position?
-Does it matter?]]
+However, continued nondetection in runs now underway will decrease the upper limit of this range.
 
@@
 These parameters, measured with Bayesian parameter estimation techniques, require highly accurate waveform models to avoid systematic errors.
-For example, post-Newtonian (PN) waveform models which are truncated at the currently known 3.5PN order, lead to a uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz.
-This is approximately the same size as the contribution to the phase from tidal effects, and as a result the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]], and this also biases the recovered NS equation of state as well [[cite]].
-Uncertainties in the PN waveform model also lead to significant biases in the recovered masses and spins as well [[cite]].
+For example, post-Newtonian (PN) waveform models, which are truncated at the currently known 3.5PN order, lead to uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz [[cite]].
+This is approximately the same size as the contribution to the phase from tidal effects, and as a result, the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]].
+This also biases the recovered NS equation of state and radius [[cite]], as well as the masses and spins [[cite]].
 
 The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms.
-Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveforms uncertainties of $\sim 10$~rad.
-NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger.
+Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad.
+Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]].
 Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]].
 Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable.
+Significant work has been done to optimize existing implementations of EOB for BBH systems [[EtienneMcWilliams]].
+However, it is not clear that these optimizations can be readily adapted for the BNS model discussed here with dynamical tides.
 We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.
 
-Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\mathbf{x})$ given its waveform parameters $\mathbf{x}$.
-The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce the original training set of waveforms to some desired accuracy.
+Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$.
+The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce any waveform in the parameter space to some desired accuracy.
 This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions.
@@
 
-Once this reduced basis is built, one can interpolate each of the coefficients $c_i(\mathbf{x})$ as a function of the waveform parameters $\mathbf{x}$.
+Once this reduced basis is built, one can interpolate each of the coefficients $c_i(\bx)$ as a function of the waveform parameters $\bx$.
 This method has been used for frequency-domain surrogates of aligned spin BBH waveforms [[Michael's papers]].
-Alternatively, one can use the empirical interpolation method [[cite]] which re-expresses the linear combination of orthonormal basis functions in terms of a sum of empirical interpolating functions $B_j(t)$ times the value of the waveform $h(T_j; \mathbf{x})$ at empirical nodes $T_j$.
-One then interpolates the waveform $h(T_j; \mathbf{x})$ as a function of $\mathbf{x}$ at each node $T_j$.
-This method has been used for time domain-waveforms for nonspinning [[cite]], aligned-spin [[cite]], and precessing [[cite]] BBH waveforms as well as nonspinning BNS waveforms with tidal interactions [[cite]], as well as for frequency-domain waveforms [[cite]].
+Alternatively, one can use the empirical interpolation method (EIM) [[cite]] which re-expresses the linear combination of orthonormal basis functions in terms of a sum of empirical interpolating functions $B_j(t)$ times the value of the waveform $h(T_j; \bx)$ at empirical nodes $T_j$.
+One then interpolates the waveform $h(T_j; \bx)$ as a function of $\bx$ at each node $T_j$.
+This method has been used for time domain-waveforms for nonspinning [[cite]], aligned-spin [[cite]], and precessing [[cite]] BBH waveforms as well as nonspinning BNS waveforms with tidal interactions [[cite]], and frequency-domain, precessing waveforms [[cite]].
 
-Previous work on surrogate models has focused mainly on parameter spaces $\mathbf{x}$ with $<=3$~dimensions.
-This allows one to approximate the coefficients $c_i(\mathbf{x})$ or waveform values $h(T_j; \mathbf{x})$ as functions of $\mathbf{x}$ using simple fitting functions [[cite]], or standard interpolation techniques such as tensor spline [[Michael, others]] or Chebyshev interpolation [[Ben]].
+Previous work on surrogate models has focused mainly on parameter spaces $\bx$ with $\le 3$~dimensions.
+This allows one to approximate the coefficients $c_i(\bx)$ or waveform values $h(T_j; \bx)$ as functions of $\bx$ using simple fitting functions [[cite]] or standard interpolation techniques such as tensor spline [[Michael, others]] or Chebyshev interpolation [[Ben]].
 These interpolation techniques typically require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
@@
 In this paper, we will develop methods to construct a surrogate model for the inspiral through merger of an aligned spin BNS waveform with tidal interactions using the EOB tidal description described in Hinderer et al.~[[cite1, cite2]].
-This work is a continuation of the surrogate developed in [[cite]] for a nonspinning BNS waveform a using slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~[[cite, ...]].
-Whereas in [[Ben]] we constructed a time-domain surrogate for the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation.
+This work is a continuation of the surrogate developed in [[cite]] for a nonspinning BNS waveform using a slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~[[cite, ...]].
+Whereas in [[Ben]] we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation.
+It also enables additional techniques such as reduced order quadrature [[cite]] to accelerate likelihood evaluation.
 
 We organize the paper as follows.
-In Section II we describe the EOB waveform model used for constructing the surrogate and the approximations we use to reduce the dimensionality of the parameter space.
-In section III, we describe the details of building the surrogate model, choosing training set waveforms, and evaluating the final model.
+In Section II we provide an overview of the EOB waveform model used for constructing the surrogate and the approximations we use to reduce the dimensionality of the parameter space.
+In section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model.
 We then compare the accuracy and speed of the surrogate to the original model in section IV.
@@
 \textit{Conventions:} Unless explicitly stated, we use units where $G=c=1$.
-To agree with the machine learning literature, we will refer to the waveform parameters (features) as $\mathbf{x}$ and the hyperparameters in the Gaussian Process Regression model as $\mathbf{\theta}$.
+To agree with the GPR literature, we will refer to the waveform parameters as $\bx$ and the hyperparameters in the GPR model as $\btheta$.
+
+\red{[[In the introduction you should emphasize the parts of this paper that are new to the GW field: Applying GPR to a high dimensional parameter space, showing that you can choose several new training set samples then calculate them in parallel without reconstructing the reduced basis.
+Building a surrogate for the residual to reduce the required interpolation accuracy.
+The first BNS waveform that realistically tapers to zero amplitude.
+Active learning method that readily extends to higher dimensions.
+By attaching to TaylorF2, the surrogate is valid for all frequencies.]]}
+
+\red{[[We aim to construct a sufficiently accurate waveform surrogate using as few waveform evaluations as possible, to show how one could use NR simulations/go to higher dimensions.]]}
 
@@
 
-\subsection{Inspiral-plunge waveform}
+\red{[[Say somewhere that this model is based off of what is known as SEOBNRv4 (and/or 2?) in lalsuite.]]}
 
-Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects.
-State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries.
+\red{[[The inspiral part is exactly as described in previous papers. The post-merger part is new to this paper. If this is true, we should write that somewhere.]]}
+
+The effective-one-body (EOB) approach to the general-relativistic 2-body problem, first described in Ref.~\cite{Buonanno:1998gg}, has proven successful in modeling the dynamics and GW emission of binaries of compact objects.
+State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of aligned-spin binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal \red{[[quantify this]]} for equal-mass) binaries.
 The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}.
-This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter---typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs).
+This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter---BNS or neutron-star-black-hole (NSBH) binaries.
 The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case.
@@
 Because of binary motion, the forcing tidal field varies at the orbital frequency.
-Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode.
+Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of the f-mode.
 The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field.
 
+We note that, to date, this is the only tidal EOB model that includes the effect of spin \red{[[Sebastiano also has a model that includes spin and tides.]]}.
+More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.
+The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning.
+This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled.
+The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.
+
+\subsection{Inspiral-plunge waveform}
+
 Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.
-
+\red{[[Show how $\Delta_u$ enters the Hamiltonian.]]}.
 In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
@@
 \end{widetext}
-where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
-Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies.
+where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$ and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
+Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\omega_{0\ell}^{A,B}$.
 In particular, the dimensionless dynamical tidal polarizability reads
@@
 Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole angular frequency, $\omega_{0\ell}^{A,B}$.
+\red{[[Does it need to be made clear that the $f$-mode is rescaled by the individual mass not the total mass?]]} This enhancement factor is 1 for low orbital frequencies, and increases by about \red{x\%} as $\ell$ times the orbital frequency approaches the $f$-mode frequency.
 
@@
 \begin{equation}
-h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{insp-plunge, pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
+h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
 \end{equation} 
-where $h_{\ell m}^{\textrm{insp-plunge, pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs.
+where the point-mass piece $h_{\ell m}^{\textrm{pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and the tidal piece $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs.
 (A14)-(A17) of Ref.~\cite{Damour:2012yf}.
-Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
+Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
+The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the GW flux (Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived.
 
 Let us now discuss the construction of the full waveform.
-The numerical integration starts from quasicircular initial conditions that are built following the prescription of Ref.~\cite{Buonanno:2005xu}.
-
-The inspiral-plunge orbital dynamics is obtained by numerically integrating Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s: see Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}.
-
-The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$.
-Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when $|h_{22}^{\textrm{insp-plunge}}|$ is maximum.
-Let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum.
-Then we put
+The Hamiltonian equations of motion with the above radiation-reaction force are numerically integrated beginning with the quasicircular initial conditions described in Ref.~\cite{Buonanno:2005xu}.
+The inspiral-plunge waveform is obtained by evaluating Eq.~(\ref{hlminspplunge}) using the solution to the orbital dynamics up to some end time $t=t_{\textrm{match}}^{22}$ that we now define.
+Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when the amplitude of the $(\ell,m)=(2,2)$ mode $|h_{22}^{\textrm{insp-plunge}}|$ is maximum, and let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum.
+We then choose the stopping condition to be the earliest of these two times
 \begin{equation}
@@
 \end{equation}
-In this work we only consider the dominant mode, $(\ell,m)=(2,2)$.
+\red{[[Explain/motivate/justify this choice in a sentence or two.
+Why don't you just pick one of the conditions?]]} In our final waveform model we only use the dominant $(\ell,m)=(2,2)$ mode in constructing the final waveform, although we use all modes in evaluating the radiation reaction force.
 
@@
 
-\red{[[Mention that this is just a tapering and does not include post-merger oscillations seen in numerical simulations.]]}
-
-We model the post-merger GW emission by a signal that tapers down in amplitude and has a constant GW frequency.
-In particular, we employ the amplitude of the merger-ringdown of the underlying BBH model~\cite{Bohe:2016gbl}, using as inputs the mass ratio and spins of the BNS/NSBH of interest.
+We model the post-merger GW emission that tapers to zero amplitude in terms of analytic functions for the amplitude $A(t)$ and phase $\Phi(t)$ such that $h(t) = A(t) e^{i\Phi(t)}$.
+For the amplitude, we use the merger-ringdown prescription of the underlying SEOBNRv4 BBH model~\cite{Bohe:2016gbl}.
 Explicitly,
 \begin{equation}
-|h_{22}^{\textrm{post-mrg}}(t)|=\nu \tilde{A}_{22}(t)\exp{\left[-(t-t^{22}_{\textrm{match}})/\tau_{220}\right]}\,,
+|A^{\textrm{post-mrg}}(t)|=\nu \tilde{A}_{22}(t; \nu, S_1, S_2)\exp{\left[-(t-t^{22}_{\textrm{match}})/\tau_{220}\right]}\,,
 \end{equation}
-where $\nu=M_A M_B/M^2$ is the symmetric mass ratio, $\tilde{A}_{22}(t)$ is a monotonic and bounded function given in Eq.~(2.12) of Ref.~\cite{Bohe:2016gbl}, and $\tau_{220}$ is the decay time of the least-damped quasinormal mode of the remnant BH that forms after the BBH merger.
-This choice is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}.
-The constant value of the GW frequency is taken to be that of $h_{22}^{\textrm{insp-plunge}}$ at time $t=t_{\textrm{match}}^{22}$, so that the post-merger phase evolution reads
+where $\nu=M_A M_B/M^2$ is the symmetric mass ratio. $\tilde{A}_{22}(t; \nu, S_1, S_2)$ is the analytic function given in Ref.~\cite{Bohe:2016gbl} (Eq.~(2.12))
+\begin{equation}
+\tilde{A}_{22}(t; \nu, S_1, S_2) = c_1^2\tanh[c_1^f(\nu, S_1, S_2)(t-t_{\textrm{match}}^{22})+c_2^f(\nu, S_1, S_2)]+c_2^c.
+\end{equation}
+The coefficients $c_1^f(\nu, S_1, S_2)$ and $c_2^f(\nu, S_1, S_2)$ are given by the fitting functions Eqs.~(2.13) and~(2.14) of Ref.~\cite{Bohe:2016gbl}, and the coefficients $c_1^c$ and $c_2^c$ are fixed by requiring that the waveform amplitude be continuous and smooth at $t^{22}_{\textrm{match}}$.
+$\tau_{220}$ is the decay time of the least-damped quasinormal mode of the remnant BH that forms after the BBH merger, and is given in terms of the total mass and angular momentum of the system at $t^{22}_{\rm match}$ using the fitting formula Eq.~{E2) of Ref.~\cite{BertiCordosoWill2006}.
+
+The phase $\Phi(t)$ is defined by requiring that the GW frequency have the constant value determined by the final inspiral frequency at $t_{\textrm{match}}^{22}$.
+Explicitly, this linear phase evolution reads
 \begin{align}
@@
 where $\phi_{22}^{\textrm{inspiral-plunge}}$ is the phase of $h_{22}^{\textrm{insp-plunge}}$.
-A more sophisticated, NR-informed, time-domain model of the post-merger emission will be part of future investigations.
-Finally, the full waveform reads
-\begin{align}
-h_{22}(t) &= h_{22}^{\textrm{insp-plunge}}(t)\, \theta\left(t_{\textrm{match}}^{22}-t\right)\nonumber\\
-&+h_{22}^{\textrm{post-mrg}}(t)\, \theta\left(t-t_{\textrm{match}}^{22}\right)\,.
-\end{align}
 
-Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state, finding that the model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state.
+Although this post-merger model has not been fit to NR BNS simulations, it is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}.
+\red{[[Maybe we want plots of these two waveforms if there's enough space?]]} A more sophisticated, NR-informed model of the post-merger emission will be part of future investigations.
+
+\red{[[This paragraph is talking about the inspiral right?
+In this case it should go in the subsection above.]]} Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state, finding that the model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state.
 Similar results hold for the tidal EOB model of Ref.~\cite{Bernuzzi:2014owa}, whose radial potential includes tidal effects in a form that is inspired by first-order gravitational-self-force (GSF) calculations of tidal invariants in the small-mass-ratio limit.
@@
 
-We note that, to date, this is the only tidal EOB model that includes the effect of spin.
-More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.
-The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning.
-This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled.
-The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.
-
 \subsection{Approximation of higher-order tidal effects}
 
-As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless components of the spins along the Newtonian orbital angular momentum $-1 \leq S_{A,B} \leq 1$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
+As described above, an aligned-spin EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless components of the spins along the Newtonian orbital angular momentum $-1 \leq S_{A,B} \leq 1$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
 To reduce the dimensionality of the intrinsic parameter space, we exploit nearly EOS-independent fits for the higher-order effects in terms of $\Lambda_2$ usually referred to as universal relations.
@@
 This formula, however, diverges at small values of $\Lambda_2$, so we replace it with a polynomial function that vanishes at $\Lambda_2 = 0$.
-We require this extension to be continuous and smooth at $\Lambda_2 = 10^{-2}$, and to fit the universal relation in the range $10^{-5}\leq\Lambda_2^{A,B}\leq10^{-2}$.
+We require this extension to be continuous and smooth at $\Lambda_2 = 10^{-2}$, and to fit the universal relation in the range $10^{-5}\leq\Lambda_2\leq10^{-2}$.
 The exact relation we use is
@@
 
-With these relations the waveform only depends on 5 intrinsic parameters $\mathbf{x}=\{q, S_A, S_B, \Lambda_2^A,\Lambda_2^B\}$.
+With these relations the waveform only depends on 5 intrinsic parameters $\bx=\{q, S_A, S_B, \Lambda_2^A,\Lambda_2^B\}$.
 We note that this list does not include the total mass $M$ of the system.
@@
 
-The problem of constructing a surrogate model for a computer simulation that takes parameters 
-$\vec x$ and outputs a vector of data $\vec y$ is the following: construct approximate functions $f_j(x)$
-for each quantity $y_j$ using a training set of precomputed examples $\{\vec x^i, \vec y^i\}$. For the problem here, 
-our parameters are the intrinsic waveform parameters $\vec x$ and the data is the waveform frequency series $\tilde h(f_j)$.
+As discussed above in Sec.~II, the current implementation of the EOB model takes at least 10~minutes to evaluate.
+In this section we will describe in detail the method for training and evaluating our surrogate model.
+The first decision in building a waveform surrogate is choosing between building a surrogate of the time-domain waveform $h(t;\bx)$ or the frequency-domain waveform $\tilde h(f;\bx)$.
+A time-domain surrogate has the advantage that significantly less processing of the original waveform is needed, resulting in a surrogate that typically more accurately reproduces the original waveform model.
+A frequency-domain surrogate, on the other hand, requires windowing of the Fourier transformed waveform to avoid Gibbs oscillations.
+For the beginning of the waveform, one can simply begin at a lower frequency, window the waveform, then truncate the Fourier-transformed waveform above some desired starting frequency.
+At the end of the waveform, the BNS amplitude drops rapidly over $\sim 1$~GW cycle, resulting in Gibbs oscillations at the high-frequency end.
+One could window the end of the waveform over several cycles, however, this would remove available information about tidal interactions.
+As we will show in Sec.~[[]], these high-frequency Gibbs oscillations in the amplitude of the Fourier transformed waveform are a genuine feature of BNS waveforms unlike BBH waveforms which typically taper over a larger number of GW cycles during their ringdown.
+These oscillations make it difficult to construct a surrogate.
+However, we will show that smoothing these oscillations in the Fourier transformed waveform with a moving average filter leads to a surrogate that is sufficiently faithful to the original time-domain model.
+Another important advantage is that a surrogate of $\tilde h(f;\bx)$ bypasses the need to evaluate the Fourier transform at each waveform evaluation for parameter estimation.
+And, as we shall see, a frequency-domain surrogate can be trivially attached to analytic models that are accurate at low-frequencies, such as TaylorF2, making analysis for third-generation detectors which are sensitive at lower frequencies easier.
+Finally, methods for speeding up the likelihood evaluation in parameter estimation such as reduced order quadrature typically require a fast frequency-domain model for calculating the appropriate weights online [[cite]].
+For these reasons we will construct a surrogate of $\tilde h(f;\bx)$.
 
---Frequency-domain so Fourier transform can be done offline and all issues related to tapering, etc. are taken care of.
+%The problem of constructing a surrogate model for a computer simulation that takes parameters 
+%$\vec x$ and outputs a vector of data $\vec y$ is the following: construct approximate functions $f_j(x)$
+%for each quantity $y_j$ using a training set of precomputed examples $\{\vec x^i, \vec y^i\}$. For the problem here, 
+%our parameters are the intrinsic waveform parameters $\vec x$ and the data is the waveform frequency series $\tilde h(f_j)$.
 
-We aim to construct a sufficiently accurate waveform surrogate using as few waveform evaluations as possible.
-To do this we perform the following steps:
+In the subsections below we will build our surrogate with the aim of using as few waveform evaluations as possible.
+We do this because EOB models of BNS systems are already expensive, and we would like a method that extends to higher dimensions and more expensive simulations such as numerical relativity simulations.
+A general overview of the process we have used is as follows:
 \begin{enumerate}
-\item Select the initial training set parameters.
+\item Select an initial, sparse set of waveform parameters $\bx$ that cover the parameter space using a Latin Hypercube Design (LHD).
 
-\item Precondition an initial training set of waveforms, and decompose them into functions that smoothly vary
-as a function of waveform parameters $\vec x$.
+\item Evaluate the initial training set of EOB waveforms $h(t;\bx)$ and condition them to produce smoothed, frequency-domain waveforms $\tilde h(f;\bx)$.
 
-\item Compress the waveform output using a greedy method for generating a reduced orthonormal basis and 
-generate empirical interpolating functions and nodes form this reduced space, to minimize the number of interpolations
-required.
+\item Construct separate orthonormal, reduced bases for the amplitude $A(f;\bx)$ and phase $\Phi(f;\bx)$ of the training set waveforms.
+Then, use the empirical interpolation method (EIM) to generate separate sets of frequency-dependent empirical interpolating functions for the amplitude $B^A_j(f)$ and phase $B^\Phi_j(f)$.
 
-\item Interpolate the waveform at the empirical nodes as a function of the waveform parameters $\vec x$ using GPR.
+\item For each empirical interpolating function, interpolate the amplitude $A(F^A_j;\bx)$ or phase $\Phi(F^\Phi_j;\bx)$ at the associated frequency nodes $F^A_j$ or $F^\Phi_j$, respectively, as a function of the waveform parameters $\bx$ using GPR.
 
-\item With the resulting surrogate model and GPR uncertainty estimate, perform uncertainty sampling to choose the 
-next waveforms to add to the training set.
+\item With the resulting surrogate model and GPR uncertainty estimates, iteratively choose the points in parameter space with the largest estimated waveform errors.
+Then, generate these waveforms and add them to the initial training set.
 
-\item Repeat steps 2--5 to reconstruct the surrogate model with the updated training set. This can be repeated
-if necessary.
+\item Reconstruct the surrogate model with the updated training set. This can be repeated as necessary with additional training set waveforms.
+
 \end{enumerate}
@@
 \begin{equation}
-A_{\rm F2} = simple equation,
+A_{\rm F2} = \sqrt{\frac{5\pi\eta}{6}} x^{-7/4},
 \end{equation}
 and
-\begin{equation}
-\Phi_{\rm F2} = not simple equation.
-\end{equation}
+\begin{align}
+\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
+                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, S_1, S_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
+\end{align}
+Where $x=(\pi f)^{2/3}$ is the standard PN parameter.
+We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~[[]] of Ref.~[[]], the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order in Eqs.~[[]] of Ref.~[[]] and Eqs.~[[]] of Ref.~[[]], and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order in Eqs.~[[]] of Ref.~[[]].
 We therefore only have to construct surrogates for the difference between EOB and TaylorF2:

### 2017-08-31 paper/paper.tex

@@
-\documentclass[prd,aps,letter,twocolumn,floatfix,notitlepage]{revtex4-1}
+\documentclass[prd,aps,letter,twocolumn,floatfix,notitlepage,nofootinbib]{revtex4-1}
 
@@
 \newcommand{\red}[1]{\textcolor{red}{#1}}
+\newcommand{\blue}[1]{\textcolor{blue}{#1}}
 %% +++++++++++++++++++++++++++++++++++++++++++++++++++++++
@@
 The effective-one-body (EOB) approach to the general-relativistic 2-body problem, first described in Ref.~\cite{Buonanno:1998gg}, has proven successful in modeling the dynamics and GW emission of binaries of compact objects.
-State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of aligned-spin binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal \red{[[quantify this]]} for equal-mass) binaries.
+State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of aligned-spin binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to 0.98 for equal-mass) binaries.
 The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}.
@@
 
-We note that, to date, this is the only tidal EOB model that includes the effect of spin \red{[[Sebastiano also has a model that includes spin and tides.]]}.
-More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.
-The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning.
+We note that, to date, this is the only tidal EOB model that includes the effect of spin \red{[[Sebastiano also has a model that includes spin and tides.]]}\blue{[[Which is the Reference?]]}.
+More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.\footnote{This model is implemented in the LIGO Algorithm Library as SEOBNRv4.} The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning.
 This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled.
@@
 
-Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.
-\red{[[Show how $\Delta_u$ enters the Hamiltonian.]]}.
+Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.
+\red{[[Show how $\Delta_u$ enters the Hamiltonian.]]}\blue{[[There's no easy/illuminating way to explain how this potential enters the Hamiltonian]]}.
 In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
@@
 where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$ and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
-Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\omega_{0\ell}^{A,B}$.
+Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$.
 In particular, the dimensionless dynamical tidal polarizability reads
 \begin{equation}
-\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\,,
+\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
 \end{equation}
@@
 with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star.
-Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole angular frequency, $\omega_{0\ell}^{A,B}$.
-\red{[[Does it need to be made clear that the $f$-mode is rescaled by the individual mass not the total mass?]]} This enhancement factor is 1 for low orbital frequencies, and increases by about \red{x\%} as $\ell$ times the orbital frequency approaches the $f$-mode frequency.
+Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.\footnote{Here $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.} This enhancement factor is 1 for low orbital frequencies, and increases to about 2 as $\ell$ times the orbital frequency approaches the $f$-mode frequency.
 
@@
 Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when the amplitude of the $(\ell,m)=(2,2)$ mode $|h_{22}^{\textrm{insp-plunge}}|$ is maximum, and let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum.
-We then choose the stopping condition to be the earliest of these two times
+We then choose the stopping condition to be the earliest of these two times,
 \begin{equation}
-t_{\textrm{match}}^{22} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.
+t_{\textrm{match}}^{22} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.\label{tmatch}
 \end{equation}
-\red{[[Explain/motivate/justify this choice in a sentence or two.
-Why don't you just pick one of the conditions?]]} In our final waveform model we only use the dominant $(\ell,m)=(2,2)$ mode in constructing the final waveform, although we use all modes in evaluating the radiation reaction force.
+In the BBH model, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks at a prescribed time and has prescribed amplitude and frequency, hence $t_{\textrm{match}}^{22}=t_{\textrm{peak}}^{\textrm{amp}}$.
+Here, however, we do not impose such corrections because of limited input from NR BNS simulations, therefore we have little control over the behavior of Eq.~(\ref{hlminspplunge}) in the late inspiral.
+This motivates our choice of $t_{\textrm{match}}^{22}$ in Eq.~(\ref{tmatch}), with the goal of reducing to the BBH prescription in the non-tidal limit and avoiding non-monotonic growth of the GW frequency in the late inspiral.
+To help bracket the stopping time, we exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics.
+%When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case.
+In our model we only include the dominant $(\ell,m)=(2,2)$ mode in constructing the final waveform, although we use all modes when evaluating the radiation reaction force.
 
-%We exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics.
-When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case.
+Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state, finding that the model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state.
+Similar results hold for the tidal EOB model of Ref.~\cite{Bernuzzi:2014owa}, whose radial potential includes tidal effects in a form that is inspired by first-order gravitational-self-force (GSF) calculations of tidal invariants in the small-mass-ratio limit.
+Both the EOB model with dynamical tides and the GSF-inspired one are significantly more faithful to NR than the adiabatic tidal EOB model of Refs.~\cite{Damour:2009wj,Bini:2012gu}.
 
@@
 
-We model the post-merger GW emission that tapers to zero amplitude in terms of analytic functions for the amplitude $A(t)$ and phase $\Phi(t)$ such that $h(t) = A(t) e^{i\Phi(t)}$.
+We model the post-merger GW emission that tapers to zero amplitude in terms of analytic functions for the amplitude $A_{22}^{\textrm{post-mrg}}(t)$ and phase $\phi^{\textrm{post-mrg}}(t)$ such that $h_{22}^{\textrm{post-mrg}}(t) = A_{22}^{\textrm{post-mrg}}(t) \exp{[i\phi_{22}^{\textrm{post-mrg}}(t)]}$.
 For the amplitude, we use the merger-ringdown prescription of the underlying SEOBNRv4 BBH model~\cite{Bohe:2016gbl}.
@@
 \begin{equation}
-|A^{\textrm{post-mrg}}(t)|=\nu \tilde{A}_{22}(t; \nu, S_1, S_2)\exp{\left[-(t-t^{22}_{\textrm{match}})/\tau_{220}\right]}\,,
+A_{22}^{\textrm{post-mrg}}(t)=\nu \tilde{A}_{22}(t; \nu, S_1, S_2)e^{-(t-t^{22}_{\textrm{match}})/\tau_{220}}\,,
 \end{equation}
@@
 \begin{equation}
-\tilde{A}_{22}(t; \nu, S_1, S_2) = c_1^2\tanh[c_1^f(\nu, S_1, S_2)(t-t_{\textrm{match}}^{22})+c_2^f(\nu, S_1, S_2)]+c_2^c.
+\tilde{A}_{22}(t; \nu, S_1, S_2) = c_1^2\tanh[c_1^f(t-t_{\textrm{match}}^{22})+c_2^f]+c_2^c.
 \end{equation}
-The coefficients $c_1^f(\nu, S_1, S_2)$ and $c_2^f(\nu, S_1, S_2)$ are given by the fitting functions Eqs.~(2.13) and~(2.14) of Ref.~\cite{Bohe:2016gbl}, and the coefficients $c_1^c$ and $c_2^c$ are fixed by requiring that the waveform amplitude be continuous and smooth at $t^{22}_{\textrm{match}}$.
-$\tau_{220}$ is the decay time of the least-damped quasinormal mode of the remnant BH that forms after the BBH merger, and is given in terms of the total mass and angular momentum of the system at $t^{22}_{\rm match}$ using the fitting formula Eq.~{E2) of Ref.~\cite{BertiCordosoWill2006}.
+The coefficients $c_{1,2}^f=c_{1,2}^f(\nu, S_1, S_2)$ are given by the fitting functions Eqs.~(2.13) and~(2.14) of Ref.~\cite{Bohe:2016gbl}, and the coefficients $c_1^c$ and $c_2^c$ are fixed by requiring that the waveform amplitude be continuous and smooth at $t^{22}_{\textrm{match}}$.
+$\tau_{220}$ is the decay time of the least-damped quasinormal mode of the remnant BH that forms after the BBH merger, and is given in terms of the total mass and angular momentum of the system at $t^{22}_{\rm match}$ using the fitting formula Eq.~(E2) of Ref.~\cite{Berti:2005ys}.
 
-The phase $\Phi(t)$ is defined by requiring that the GW frequency have the constant value determined by the final inspiral frequency at $t_{\textrm{match}}^{22}$.
-Explicitly, this linear phase evolution reads
+The phase is defined by requiring that the GW frequency have the constant value determined by the final inspiral frequency at $t_{\textrm{match}}^{22}$. Explicitly, this linear phase evolution reads
 \begin{align}
@@
 \red{[[Maybe we want plots of these two waveforms if there's enough space?]]} A more sophisticated, NR-informed model of the post-merger emission will be part of future investigations.
-
-\red{[[This paragraph is talking about the inspiral right?
-In this case it should go in the subsection above.]]} Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state, finding that the model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state.
-Similar results hold for the tidal EOB model of Ref.~\cite{Bernuzzi:2014owa}, whose radial potential includes tidal effects in a form that is inspired by first-order gravitational-self-force (GSF) calculations of tidal invariants in the small-mass-ratio limit.
-Both the EOB model with dynamical tides and the GSF-inspired one are significantly more faithful to NR than the adiabatic tidal EOB model of Refs.~\cite{Damour:2009wj,Bini:2012gu}.
 

### 2017-09-28 paper/paper.tex
Update model section to reflect new tapering

@@
 The Hamiltonian equations of motion with the above radiation-reaction force are numerically integrated beginning with the quasicircular initial conditions described in Ref.~\cite{Buonanno:2005xu}.
-The inspiral-plunge waveform is obtained by evaluating Eq.~(\ref{hlminspplunge}) using the solution to the orbital dynamics up to some end time $t=t_{\textrm{match}}^{22}$ that we now define.
-Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when the amplitude of the $(\ell,m)=(2,2)$ mode $|h_{22}^{\textrm{insp-plunge}}|$ is maximum, and let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum.
-We then choose the stopping condition to be the earliest of these two times,
+The inspiral-plunge waveform is obtained by evaluating Eq.~(\ref{hlminspplunge}) using the solution to the orbital dynamics.
+Let $\Omega$ be the instantaneous orbital frequency, $r$ be orbital separation, $p_r$ be the radial momentum, $f_s$ be the waveform sampling frequency in Hz, $\omega_{\textrm{peak}}^{\textrm{NR}}$ be the NR fit for the (2,2)-mode amplitude peak of a BNS merger defined in Eq.~(2) of Ref.~\cite{Bernuzzi:2015rla}~\footnote{Note that this fit was tuned to NR simulations only for $0 \leq \kappa_2^T \leq 500$, where $\kappa_2^T$ is defined by Eq.~(1) of Ref.~\cite{Bernuzzi:2015rla}.
+In our model, for $\kappa_2^T >500$ we use the value of this fit at $\kappa_2^T=500$.}.
+Then the numerical integration of the orbital dynamics stops when any of the following events happens: (i) $\Omega$ peaks, (ii) $p_r$ becomes positive, (iii) $\dot{r}$ becomes positive, (iv) $\dot{p}_r$ becomes positive, (v) $r$ does not evolve more than a part in $10^7$, (vi) $2\Omega$ reaches $\omega_{\textrm{peak}}^{\textrm{NR}}$, (vii) $2M\Omega$ reaches $\pi M f_s$.
+In order to avoid triggering these stopping conditions at low frequency, where residual orbital eccentricity due to imperfect quasicircular initial conditions could be present, we exploit the $\omega_{\textrm{peak}}^{\textrm{NR}}$ fit to define a radius below which the stopping conditions can be triggered; such threshold value is taken to be $1.5 (\omega_{\textrm{peak}}^{\textrm{NR}}/2)^{-2/3}$.
+
+Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the earliest time when the amplitude of the $(\ell,m)=(2,2)$ mode $|h_{22}^{\textrm{insp-plunge}}|$ peaks, and let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum.
+Then
 \begin{equation}
@@
 \end{equation}
+If $|h_{22}^{\textrm{insp-plunge}}|$  does not peak, then we choose $t_{\textrm{peak}}^{\textrm{amp}}$ to be the earliest time when $\partial_t |h_{22}^{\textrm{insp-plunge}}|$ reaches a minimum after having reached a peak.
+
 In the BBH model, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks at a prescribed time and has prescribed amplitude and frequency, hence $t_{\textrm{match}}^{22}=t_{\textrm{peak}}^{\textrm{amp}}$.
@@
 We model the post-merger GW emission that tapers to zero amplitude in terms of analytic functions for the amplitude $A_{22}^{\textrm{post-mrg}}(t)$ and phase $\phi^{\textrm{post-mrg}}(t)$ such that $h_{22}^{\textrm{post-mrg}}(t) = A_{22}^{\textrm{post-mrg}}(t) \exp{[i\phi_{22}^{\textrm{post-mrg}}(t)]}$.
-For the amplitude, we use the merger-ringdown prescription of the underlying SEOBNRv4 BBH model~\cite{Bohe:2016gbl}.
-Explicitly,
-\begin{equation}
-A_{22}^{\textrm{post-mrg}}(t)=\nu \tilde{A}_{22}(t; \nu, S_1, S_2)e^{-(t-t^{22}_{\textrm{match}})/\tau_{220}}\,,
+For the amplitude, we define the following linear extension of the inspiral-plunge amplitude
+\begin{align}
+\hat{A}(t)=\left\{\begin{array}{ll}
+|h_{22}^{\textrm{insp-plunge}}(t)|, & t \leq t^{22}_{\textrm{match}}\,, \\
+a + b (t-t^{22}_{\textrm{match}}) , & t > t^{22}_{\textrm{match}}\,,
+\end{array}\right.
+\end{align}
+where $a=|h_{22}^{\textrm{insp-plunge}}(t^{22}_{\textrm{match}})|$ and $b=\partial_t |h_{22}^{\textrm{insp-plunge}}(t^{22}_{\textrm{match}})|$.
+Then we introduce the tapering function \begin{equation}
+W(t) = \frac{1}{1+\exp{[-(t- t^{22}_{\textrm{match}}-15M)/\tau]}}\,,
 \end{equation}
-where $\nu=M_A M_B/M^2$ is the symmetric mass ratio. $\tilde{A}_{22}(t; \nu, S_1, S_2)$ is the analytic function given in Ref.~\cite{Bohe:2016gbl} (Eq.~(2.12))
-\begin{equation}
-\tilde{A}_{22}(t; \nu, S_1, S_2) = c_1^2\tanh[c_1^f(t-t_{\textrm{match}}^{22})+c_2^f]+c_2^c.
-\end{equation}
-The coefficients $c_{1,2}^f=c_{1,2}^f(\nu, S_1, S_2)$ are given by the fitting functions Eqs.~(2.13) and~(2.14) of Ref.~\cite{Bohe:2016gbl}, and the coefficients $c_1^c$ and $c_2^c$ are fixed by requiring that the waveform amplitude be continuous and smooth at $t^{22}_{\textrm{match}}$.
-$\tau_{220}$ is the decay time of the least-damped quasinormal mode of the remnant BH that forms after the BBH merger, and is given in terms of the total mass and angular momentum of the system at $t^{22}_{\rm match}$ using the fitting formula Eq.~(E2) of Ref.~\cite{Berti:2005ys}.
+with $\tau=\pi/(2\omega_{\textrm{match}})$ for $\omega_{\textrm{match}} =\omega_{22}^{\textrm{insp-plunge}}(t^{22}_{\textrm{insp-plunge}})$.
+The amplitude at all times is then given by $|h_{22}(t)|=\hat{A}(t)W(t)$.
 
-The phase is defined by requiring that the GW frequency have the constant value determined by the final inspiral frequency at $t_{\textrm{match}}^{22}$. Explicitly, this linear phase evolution reads
+Let $\phi_{22}^{\textrm{inspiral-plunge}}$ be the phase of $h_{22}^{\textrm{insp-plunge}}$. We define a time $t_{\textrm{freq}}=t^{22}_{\textrm{insp-plunge}}-20M$. The phase at all times is given by
 \begin{align}
-\phi_{22}^{\textrm{post-mrg}}(t)&=\omega_{22}^{\textrm{insp-plunge}}(t_{\textrm{match}}^{22})(t-t_{\textrm{match}}^{22})\nonumber\\
-&+\phi_{22}^{\textrm{inspiral-plunge}}(t_{\textrm{match}}^{22})\,
+\phi_{22}(t)=\left\{\begin{array}{ll}
+\phi_{22}^{\textrm{inspiral-plunge}}(t), & t \leq t_{\textrm{freq}}\,, \\
+\phi_{\textrm{freq}} + \omega_{\textrm{match}}(t-t_{\textrm{freq}})  & \\
++\frac{\Delta\omega^2}{\dot{\omega}_{\textrm{freq}}} [\exp{(-\dot{\omega}_{\textrm{freq}}(t-t_{\textrm{freq}})/\Delta\omega)}-1]\,,& t > t_{\textrm{freq}}\,,
+\end{array}\right.
 \end{align}
-where $\phi_{22}^{\textrm{inspiral-plunge}}$ is the phase of $h_{22}^{\textrm{insp-plunge}}$.
+where $\phi_{\textrm{freq}}=_{22}^{\textrm{inspiral-plunge}}(t_{\textrm{freq}})$, $\Delta\omega=\omega_{\textrm{match}} - \omega_{\textrm{freq}}$, $\omega_{\textrm{freq}}=\omega_{22}^{\textrm{insp-plunge}}(t_{\textrm{freq}})$, and $\dot{\omega}_{\textrm{freq}}=\partial_t\omega_{22}^{\textrm{insp-plunge}}(t_{\textrm{freq}})$.
 
 Although this post-merger model has not been fit to NR BNS simulations, it is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}.
-\red{[[Maybe we want plots of these two waveforms if there's enough space?]]} A more sophisticated, NR-informed model of the post-merger emission will be part of future investigations.
+A more sophisticated, NR-informed model of the post-merger emission will be part of future investigations.
 
@@
 This formula, however, diverges at small values of $\Lambda_2$, so we replace it with a polynomial function that vanishes at $\Lambda_2 = 0$.
-We require this extension to be continuous and smooth at $\Lambda_2 = 10^{-2}$, and to fit the universal relation in the range $10^{-5}\leq\Lambda_2\leq10^{-2}$.
+We require this extension to be continuous at $\Lambda_2 = 10^{-2}$, and to fit the universal relation in the range $10^{-5}\leq\Lambda_2\leq10^{-2}$.
 The exact relation we use is
@@
 \end{equation}
-where $\xi = \ln\Lambda_2$, $\{a, b, c, d, e\} = \{-1.15, 1.18, 2.51\times 10^{-2}, -1.31\times 10^{-3}, 2.52\times 10^{-5}\}$ from Ref.~\cite{Yagi:2013sva}, and $\{f, g, h\}=\{0.441, -34.6, 1760\}$.
+where $\xi = \ln\Lambda_2$, $\{a, b, c, d, e\} = \{-1.15, 1.18, 2.51\times 10^{-2}, -1.31\times 10^{-3}, 2.52\times 10^{-5}\}$ from Ref.~\cite{Yagi:2013sva}, and $\{f, g, h\}=\{0.4406491912035266, -34.63232296075433,1762.112913125107\}$.
 
@@
 (3.5) of Ref.~\cite{Chan:2014kua} gives relations for $\omega_{02}$ as a function of $\Lambda_2$ and $\omega_{03}$ as a function of $\Lambda_3$ to within a few percent error.
-For $\omega_{02}$, the fitting range used by Ref.~\cite{Chan:2014kua} was $1\leq \xi \leq 9$, and outside this range we require continuity.
+For $\omega_{02}$, the fitting range used by Ref.~\cite{Chan:2014kua} was $0\leq \xi \leq 9$, and outside this range we require continuity.
 The relation we use is then
@@
 \omega_{02} = \left\{\begin{array}{ll}
-f\, , & \xi < 1 \\
+f\, , & \xi < 0 \\
 a + b\xi + c\xi^2 + d\xi^3 + e\xi^4\, , & 1 \le \xi \le 9 \\

### 2017-10-03 src/trainingseteob.py
Changed traning set to match to TaylorF2 at low frequencies.

@@ -3,6 +3,7 @@ import h5py
 
 import scipy.interpolate as interpolate
 import scipy.optimize as optimize
+import scipy.stats as stats
 
 import waveform as wave
 import waveformset as ws
@@ -153,97 +154,6 @@ class ConditionedWaveform(object):
         return h_ref, dh
 
 
-# def condition_eob_waveform(
-#     h, params, delta_t,
-#     winon_i, winon_f,
-#     n_ext,
-#     trunc_i, trunc_f, npoints=10000,
-#     win='planck', plots=False):
-#     """Generate a conditioned Frequency-domain waveform from an EOB waveform with arbitrary time samples.
-#     0. Resample the waveform.
-#     1. Window the beginning and end.
-#     2. Pad the end with zeros so all the waveforms in the training set have the exact same time samples.
-#     3. Fourier transform the waveform.
-#     4. Time shift waveform so t=0 corresponds to maximum amplitude.
-#     5. Resample the waveform, truncating the beginning and end to remove the windowing effect.
-#
-#     Parameters
-#     ----------
-#     h : Waveform
-#         Uniformly sampled waveform.
-#     winon_i : Initial frequency of on window.
-#     winon_f : Final frequency of on window.
-#     winoff_i : Initial frequency of off window.
-#     winoff_f : Final frequency of off window. Should be less than the ending frequency of the waveform h.
-#     n_ext : int
-#         Number of samples for the extended (padded) waveform.
-#         All training set waveforms should have the same time samples.
-#         A power of 2 will make the Fourier transform efficient.
-#         If the phase of the Fourier-transformed waveform doesn't look correctly unwrapped,
-#         increase n_ext so that delta_f will be smaller.
-#     trunc_i : Initial frequency of the truncated waveform after Fourier transforming.
-#     trunc_f : Final frequency of the truncated waveform after Fourier transforming.
-#     npoints : int
-#         number of logarithmically-spaced samples for the final conditioned waveform.
-#     win : 'hann' or 'planck'
-#         Type of window to use.
-#
-#     Returns
-#     -------
-#     h_tilde : Waveform
[…]

### 2017-10-03 paper/paper.tex

@@
 In our model, for $\kappa_2^T >500$ we use the value of this fit at $\kappa_2^T=500$.}.
-Then the numerical integration of the orbital dynamics stops when any of the following events happens: (i) $\Omega$ peaks, (ii) $p_r$ becomes positive, (iii) $\dot{r}$ becomes positive, (iv) $\dot{p}_r$ becomes positive, (v) $r$ does not evolve more than a part in $10^7$, (vi) $2\Omega$ reaches $\omega_{\textrm{peak}}^{\textrm{NR}}$, (vii) $2M\Omega$ reaches $\pi M f_s$.
+Then the numerical integration of the orbital dynamics stops when any of the following events happens: (i) $\Omega$ peaks, (ii) $p_r$ becomes positive, (iii) $\dot{r}$ becomes positive, (iv) $\dot{p}_r$ becomes positive, (v) $r$ does not evolve more than a 1000th of the leading-order estimate for $\dot{r}$, that is $\propto 64/5 \nu/r^4$, (vi) $2\Omega$ reaches $\omega_{\textrm{peak}}^{\textrm{NR}}$, (vii) $2M\Omega$ reaches $\pi M f_s$.
 In order to avoid triggering these stopping conditions at low frequency, where residual orbital eccentricity due to imperfect quasicircular initial conditions could be present, we exploit the $\omega_{\textrm{peak}}^{\textrm{NR}}$ fit to define a radius below which the stopping conditions can be triggered; such threshold value is taken to be $1.5 (\omega_{\textrm{peak}}^{\textrm{NR}}/2)^{-2/3}$.

### 2017-11-17 paper/paper.tex
First draft of paper with mostly written sections.

@@
 \usepackage{times}
+\usepackage{ulem}
 
@@
 Fast and accurate waveform models are necessary for measuring properties of inspiraling binary neutron star systems such as the masses, spins, and tidal parameters.
-The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as corrections for dynamical tides, that are accurate enough to avoid significant biases in the measured parameters.
-However, current implementations take at least 10 minutes to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations.
-We construct a frequency-domain surrogate model that requires less than 0.2~s to evaluate for a waveform with a starting frequency of 10~Hz (0.05~s for a starting frequency of 30~Hz).
+The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as the effect of dynamical tides, and the formalism is accurate enough to avoid significant biases in the measured parameters.
+Unfortunately, current implementations take several minutes to a day to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations.
+We construct a frequency-domain surrogate of this model that requires less than 0.3~s to evaluate for a waveform with a starting frequency of 10~Hz (0.04~s for a starting frequency of 30~Hz).
 We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter.
-We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with approximately 500 waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods.
+We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with $\sim 1000$ waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods.
+
 For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.
-
-%Finally, I will show how this same method can be used to optimally choose parameters for numerical relativity simulations to correct the the EOB surrogate during the last several orbits and merger.
 
@@
 
-%% Template for figures
-%% \begin{figure}[t]
-%%   \centering 
-%%     \includegraphics[width=0.49\textwidth]{../fig/ }
-%%     \caption{ }
-%%  \label{fig: }
-%% \end{figure}
-
-%% Template for tables
-%% \begin{table}[t]
-%%   \centering    
-%%   \caption{ }
-%%   \begin{tabular}{ccc}        
-%%     \hline
-%%     col$1$ & col$2$ & col$3$ \\
-%%     \hline
-%%     &  &  \\
-%%     \hline
-%%   \end{tabular}
-%%  \label{tab: }
-%% \end{table}
-
-\red{[[Change arrow vectors to bold vectors. ${\bm x}^i$ (bold) for example $i$. $x_j$ (not bold) for coordinate/parameter/feature $j$. 
-$x_j^i$ (not bold) for feature $j$ of example $i$.]]}
-
 %% ________________________________________________________
@@
 
-The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO) [[cite]] now making routine detections of binary black hole (BBH) systems [[cite]], as well as Virgo [[cite]], KAGRA [[cite]], and LIGO-India [[cite]].
-Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]], and the nondetection of BNS events in aLIGO's first observing run (O1) is still consistent with the upper end of this range to 90\% confidence [[O1UpperLimits]].
-However, continued nondetection in runs now underway will decrease the upper limit of this range.
-
-The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters.
+\red{[[What is possible with BNS systems:]]}
+
+The detection of the binary neutron star (BNS) system GW170817 [[cite]] by the Advanced LIGO [[cite]] and Advanced Virgo [[cite]] gravitational-wave detectors, has demonstrated that the sky position, distance, masses, spins, and tidal parameters can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant [[cite]], and understanding the structure of neutron stars (NSs) through tidal interactions [[cite]].
+Measuring these parameters accurately will also be necessary to properly interpret the gamma ray burst [[cite]] and optical afterglow [[cite]] that was found in conjunction.
+Furthermore, an estimate of the BNS merger rate of x--y events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval) [[GW170817]] indicates that, when second generation detectors reach design sensitivity, including KAGRA [[cite]] and possibly LIGO-India [[cite]], we could eventually observe a population of tens or hundreds of events after a couple of years of observation [[2010Rates?]].
+This will allow us to estimate the NS mass distribution [[cite]], spin distribution [[cite]], and universal equation of state (EOS) that is common to all NSs [[DelPozzo, Lackey, ...]].
+
+\red{[[Why we need more accurate waveform models to do these things:]]}
+
+These results are obtained using a Bayesian analysis that compares the data to parameterized waveform templates.
+To avoid systematic errors highly accurate waveform templates are needed.
+In addition, the Bayesian parameter estimation alogrithms (Markov chain Monte Carlo and Nested Sampling [[parallel algorithms are becoming available]]) typically perform several million sequential waveform evaluations, requiring waveform evaluation times to be less than $\mathcal{O}(1s)$ in order to run in less than a month for each event.
+
+%The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters.
 These parameters, measured with Bayesian parameter estimation techniques, require highly accurate waveform models to avoid systematic errors.
@@
 
+\red{[[The problems with the current waveform models/templates.]]}
+
+To date, there are a limited number BNS waveform models that include spin and tides, and that are fast enough for standard parameter estimation.  
+TaylorF2: (not so good point particle. okay tidal terms)
+For example, post-Newtonian (PN) waveform models, which are truncated at the currently known 3.5PN order, lead to uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz [[cite]].
+This is approximately the same size as the contribution to the phase from tidal effects, and as a result, the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]].
+This also biases the recovered NS equation of state and radius [[cite]], as well as the masses and spins [[cite]].
+
+\*\_NRTides: (better/good point particle. better tidal terms)
+Initial parameter estimation results for the first BNS event, GW170817 using a variety of waveform templates (TaylorF2, SEOBNRv4\_ROM\_NRTides, and IMRPhenomD\_NRTides) showed that there are noticeable differences in the measured parameters between the different waveform templates [[cite]], and in particular the tidal parameter.
+[[NR\_Tides is a fit to NR waveforms at high frequency and has only been fit to a limited number of low spin, low tidal parameter NR simulations.]]
+
+TEOBResumS:
+
+\red{[[What is good about the new EOB model:]]}
+
 The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms.
+Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad.
+Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]].
+
+Another model
+
+\red{[[Problem with EOB (speed). Attempts to speed it up with optimization. Alternative using surrogates.]]}
+
+Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]].
+Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable.
+Significant work has been done to optimize existing implementations of EOB for BBH systems [[EtienneMcWilliams]].
+However, it is not clear that these optimizations can be readily adapted for the BNS model discussed here with dynamical tides.
+We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.
+
+Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$.
+The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce any waveform in the parameter space to some desired accuracy.
+This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions.
+In previous work, this reduced basis has been constructed either with singular value decomposition (SVD) [[Michael's papers]] or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization [[All the other papers]].
+
+\red{[[Overview of surrogates. Typical methods that have been used. Their problems. What you are doing that is new.]]}
+
+%The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms.
 Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad.
@@
 
-Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$.
+%Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$.
 The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce any waveform in the parameter space to some desired accuracy.
@@
 It also enables additional techniques such as reduced order quadrature [[cite]] to accelerate likelihood evaluation.
+
+\red{[[Obligitory summary:]]}
 
@@
 
-\subsection{Approximation of higher-order tidal effects}
+\subsection{Approximation of higher-order tidal parameters}
 
@@
 
-As discussed above in Sec.~II, the current implementation of the EOB model takes at least 10~minutes to evaluate.
+%As discussed above in Sec.~II, the current implementation of the EOB model takes at least 10~minutes to evaluate.
 In this section we will describe in detail the method for training and evaluating our surrogate model.
@@
 
-%The problem of constructing a surrogate model for a computer simulation that takes parameters 
-%$\vec x$ and outputs a vector of data $\vec y$ is the following: construct approximate functions $f_j(x)$
-%for each quantity $y_j$ using a training set of precomputed examples $\{\vec x^i, \vec y^i\}$. For the problem here, 
-%our parameters are the intrinsic waveform parameters $\vec x$ and the data is the waveform frequency series $\tilde h(f_j)$.
-
-In the subsections below we will build our surrogate with the aim of using as few waveform evaluations as possible.
-We do this because EOB models of BNS systems are already expensive, and we would like a method that extends to higher dimensions and more expensive simulations such as numerical relativity simulations.
-A general overview of the process we have used is as follows:
-\begin{enumerate}
-\item Select an initial, sparse set of waveform parameters $\bx$ that cover the parameter space using a Latin Hypercube Design (LHD).
-
-\item Evaluate the initial training set of EOB waveforms $h(t;\bx)$ and condition them to produce smoothed, frequency-domain waveforms $\tilde h(f;\bx)$.
-
-\item Construct separate orthonormal, reduced bases for the amplitude $A(f;\bx)$ and phase $\Phi(f;\bx)$ of the training set waveforms.
-Then, use the empirical interpolation method (EIM) to generate separate sets of frequency-dependent empirical interpolating functions for the amplitude $B^A_j(f)$ and phase $B^\Phi_j(f)$.
-
-\item For each empirical interpolating function, interpolate the amplitude $A(F^A_j;\bx)$ or phase $\Phi(F^\Phi_j;\bx)$ at the associated frequency nodes $F^A_j$ or $F^\Phi_j$, respectively, as a function of the waveform parameters $\bx$ using GPR.
-
-\item With the resulting surrogate model and GPR uncertainty estimates, iteratively choose the points in parameter space with the largest estimated waveform errors.
-Then, generate these waveforms and add them to the initial training set.
-
-\item Reconstruct the surrogate model with the updated training set. This can be repeated as necessary with additional training set waveforms.
-
-\end{enumerate}
-
-\subsection{Initial training set design}
-
-In constructing a surrogate model, we will need to interpolate several quantities as a function of the five
-waveform parameters ${\bm x}$.
-Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) 
-require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality; 
-the number of samples grows exponentially with the dimension $d$ ($N^d$ for $N$ samples per dimension).
-In the 5-dimensional problem discussed here, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
-At 10 minutes--1 hour per EOB waveform on a standard CPU, this is at the limit of what is reasonable.
-
-However, there are other methods such as Gaussian Process Regression (GPR), discussed below, that allow nonuniform designs.
-One such standard design is a Latin Hypercube Design (LHD). An LHD with $N$ samples divides each dimension uniformly into $N$
-grid points per dimension for a total of $N^d$ possible locations. However, for each dimension each of the $N$ values is sampled 
-exactly once. This has the convenient property that the samples are non-collapsing; if one of the parameters is much less 
-significant than other parameters, samples are not wasted on that parameter, as they would be with a uniform grid. And, a projection
-onto a subspace is still a LHD. [[Also, each dimension is uniformly sampled which is not the case for MC sampling.]]
-
-As an example of the usefulness of noncollapsing designs, the tidal parameters have a minimal impact on the waveform
-at low frequencies, so it wouldn't make sense to densely sample a grid of tidal parameters at low frequencies. However, near
-the merger the tidal parameters can be a important as the spin parameters. One therefore might want more samples for the tidal
-parameters and less samples for the spin. However, due to the expense of the wavefroms we must use the same grid of waveforms 
-for low and high frequencies. A LHD, with its noncollapsing property, bypasses this problem.
-
-For an LHD there are $(N!)^d$ possible designs, and a standard requirement is that the LHD be space-filling, meaning that the 
-points are as far apart as possible from each other. (See [[cite]] for a review of methods for optimizing the placement of samples.) 
-The measure used here is that we maximize the minimum Euclidean distance between any two samples [[cite]]. We do this by 
-sampling $xxx$ random designs and choosing the one with the maximum minimum distance.
-
-For a smooth approximately monotonic function in $d$-dimensions, standard lore is that $10d$ samples are needed to interpolate 
-a function with GPR [[cite]] [[To what accuracy]]. We therefore construct our initial training set with 50 waveform samples, and use 
-these waveforms to construct the reduced basis and perform interpolation with GPR. Once an initial surrogate model is constructed,
-we will use the GPR error estimate to choose new waveforms to add to the training set, and update the surrogate model.
+\subsection{Decomposition of the waveform}
+
+In this section we will describe the method for decomposing the waveform into smooth, slowly-varying functions and training a surrogate model for these functions.
+As discussed above, we work with the frequency domain waveform $\tilde h(f; \bx)$.
+This is favorable for data analysis which is usually done in the frequency domain.
+However, Fourier transforming a finite length, discretely sampled waveform, typically leads to a surrogate with more noise.
+We will show below, that sufficient filtering can solve most of these problems.
+Because the waveform $\tilde h(f; \bx)$ is an oscillatory function of $f$ and $\bx$, the waveform is usually decomposed into an amplitude and phase $\tilde h(f; \bx) = A(f; \bx) e^{i\Phi(f; \bx)}$, which are smoother, mostly monotonic functions of $f$ and $\bx$.
+The amplitude and phase of these waveforms are shown in Fig.~\ref{fig:h}.
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{trainingset3d.png}\\
-\includegraphics[width=0.49\textwidth]{trainingset2d.png}
-\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, S_1, S_2\}$ (top)
-and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom). The 160 black circles were used for the initial
-training set (32 corner points + 128 LHD points) that generated the initial surrogate. The 200 red squares were generated
-using uncertainty sampling from the GPR error estimate of the RMS phase error at the empirical nodes. 
-}
-\label{fig:LHD}
-\end{figure}
-
-\subsection{Conditioning and decomposing the training set waveforms}
-
---Messy details of preconditioning lalsimulation waveforms to make them smooth functions of parameters ${\bm x}$.
-(Rescale with total mass to convert to geometric units, time shift to align at max amplitude or specific frequency, 
-taper beginning and end, pad end with zeros, Fourier transform, truncation, resampling, zeroing initial phase, ...)
-
---Show plot (?) to validate that waveform is properly conditioned by plotting amplitude/phase as function of waveform parameters
-at specific frequencies and show that they're smooth functions.
-
-\begin{equation}
-\tilde h(f, x) = A(f, x) e^{i\Phi(f, x)}
-\end{equation}
-
-Because the waveform $h$ is an oscillatory function of $f$ and ${\bm x}$ that is difficult to interpolate, we instead construct 
-surrogates for the amplitude $A(f, x)$ and phase $\Phi(f, x)$. These are smooth functions. However, for aligned-spin waveforms
-these functions are already approximately known from the 3.5PN TaylorF2 waveform 
-$\tilde h_{\rm F2}(f, x) = A_{\rm F2}(f, x) e^{i\Phi_{\rm F2} (f, x)}$, where
-\begin{equation}
-A_{\rm F2} = \sqrt{\frac{5\pi\eta}{6}} x^{-7/4},
-\end{equation}
-and
-\begin{align}
-\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
-                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, S_1, S_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
-\end{align}
-Where $x=(\pi f)^{2/3}$ is the standard PN parameter.
-We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~[[]] of Ref.~[[]], the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order in Eqs.~[[]] of Ref.~[[]] and Eqs.~[[]] of Ref.~[[]], and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order in Eqs.~[[]] of Ref.~[[]].
-We therefore only have to construct surrogates for the difference between EOB and TaylorF2:
-\begin{align}
-\Delta\ln(A) &= \ln(A/A_{\rm F2})\\
-\Delta\Phi & = \Phi - \Phi_{\rm F2}
-\end{align}
-These quantities are shown in Fig.~\ref{fig:dh} for the training set wavefroms. We note that for small differences 
-between the EOB and TaylorF2 waveform, $\Delta\ln A$ is approximately the fractional difference $A/A_{\rm F2} - 1$
-between the EOB and TaylorF2 waveform. We use the log-ratio instead of the ratio, because it guarantees that any
-error in interpolating this quantity doesn't result in a negative amplitude. Also, because waveforms tend to deviate from 
-each other near the end of the inspiral exponentially, depending on each waveform's merger frequency the log captures
-this behavior. [[Is this true for EOB? We'll find out.]]
-
-\begin{figure}[htb]
-\centering
+\includegraphics[width=0.49\textwidth]{h_trainingset.png}\\
 \includegraphics[width=0.49\textwidth]{dh_trainingset.png}
 \caption{Training set}
-\label{fig:dh}
+\label{fig:h}
 \end{figure}
 
-We show the range of values of the phase $\Phi$ over the training set $\Phi_{\rm range} = \Phi_{\rm max} - \Phi_{\rm min}$ 
-as well as the range of values of the phase residual $\Delta\Phi$ over the training set 
-$\Delta\Phi_{\rm range} = \Delta\Phi_{\rm max}-\Delta\Phi_{\rm min}$ in Fig.~\ref{fig:phaserange}. 
-The range in $\Phi$ can be larger than 1000~radians, meaning
-that if we require 0.1~radian accuracy, we need a fractional error of $<10^{-4}$. The range of $\Delta\Phi$ is 1--2 orders 
-of magnitude smaller, and as a result, the interpolation accuracy requirements are 1--2 orders of magnitude smaller as well.
+Although the amplitude and phase are relatively smooth functions, they still span a wide range of values.
+The phase, for example, spans about $10^4$~rad.
+Unfortunately, to avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad.
+This is a difficult requirement to achieve in the 5-dimensional parameter space here.
+However, for aligned-spin waveforms these functions are approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2 $\tilde h_{\rm F2}(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2} (f; \bx)}$ [[cite]].
+We can therefore evaluate the EOB waveform in terms of the residuals, $\Delta\ln(A)$ and $\Delta\Phi$, between EOB and TaylorF2
+\begin{align}
+\tilde h(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2}(f; \bx)} e^{\Delta\ln(A)(f; \bx) + i  \Delta\Phi(f; \bx)},
+\label{eq:hdecomp}
+\end{align}
+where
+\begin{align}
+\Delta\ln(A)(f; \bx) &= \ln\left(\frac{ A(f; \bx) }{ A_{\rm F2}(f; \bx)}\right),\\
+\Delta\Phi(f; \bx) & = \Phi(f; \bx) - \Phi_{\rm F2}(f; \bx).
+\end{align}
+These residuals are shown in Fig.~\ref{fig:h}.
+For the amplitude, we use the log-ratio instead of the ratio because it guarantees that the amplitude of the reconstructed waveform in Eq.~\eqref{eq:hdecomp} is positive even if interpolation error leads to a negative $\Delta\ln(A)(f; \bx)$.
+In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency of each waveform, the log-ratio captures this behavior.
+
+To demonstrate the usefulness of this approach, in Fig.~\ref{fig:phaserange}, we show the range of values of the phase $\Phi$ over the training set $\Phi_{\rm range} = \Phi_{\rm max} - \Phi_{\rm min}$ as well as the range of values of the phase residual $\Delta\Phi$ over the training set $\Delta\Phi_{\rm range} = \Delta\Phi_{\rm max}-\Delta\Phi_{\rm min}$.
+The range in $\Phi$ can be larger than 1000~radians.
+The range of $\Delta\Phi$ is 1--2 orders of magnitude smaller, and as a result, the interpolation accuracy requirements are 1--2 orders of magnitude smaller as well.
 
@@
 
-\subsection{Compression of waveform data with reduced bases and empirical interpolation method}
-
-For many surrogate problems a function $f({\bm x})$ is sought to interpolate a scalar quantity $y$. For the
-problem here, the outputs are the quantities $\Delta\ln A$ and $\Delta\Phi$ sampled at the frequencies $f_j$. 
-One could interpolate $\Delta\ln A$ and $\Delta\Phi$ as a function of ${\bm x}$ at each sample $f_j$, 
-however, this would be prohibitively expensive. Instead we use the standard method of constructing an
-orthonormal reduced basis to compress each training set waveform into a small set of coefficients, 
-and then using the empirical interpolation method
-to optimize the placement of a corresponding number of empirical nodes $F^A_j$ and $F^\Phi_j$. One then interpolates
-$\Delta\ln A$ and $\Delta\Phi$ as a function of ${\bm x}$ at these nodes. The full waveform can then be reconstructed by 
-multiplying the value at these nodes by the the corresponding empirical interpolating functions $B_j^A(f)$ and $B_j^\Phi(f)$. 
-The methods used here are presented in Refs.[[cite]], and we use the exact implementation described in 
-Ref.~[[LackeyBernuzziGalley2017]]. We briefly describe the details.
-
-In order to construct a reduced basis from the training set waveforms we use the greedy method~[[LackeyBernuzziGalley2017]
-to choose a subset of the training set waveforms and orthonormalize them with the Grahm-Schmidt procedure. The end result
-is similar to singular value decomposition, but one can make a one-to-one mapping between each chosen waveform and a 
-corresponding basis function. The quantities $\Delta\ln A$ and $\Delta\Phi$ can be approximately described by
-\begin{subequations}
-\label{eq:basis1}
+The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
+\begin{equation}
+A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} x^{-7/4} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)x \right],
+\end{equation}
+where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{lalsuite}. The phase has the schematic form
 \begin{align}
-\Delta\ln A(f; \btheta) \approx {} & \sum_{i=1}^{N_A} e^A_i(f) c^A_i(\btheta)\\
-\Delta\Phi(f; \btheta) \approx {} & \sum_{i=1}^{N_\Phi} e^\Phi_i(f) c^\Phi_i(\btheta),
+\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
+                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, S_1, S_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
 \end{align}
-\end{subequations}
-where the coefficients are
-\begin{subequations}
-\label{eq:basiscoeffs1}
-\begin{align}
-c_i^A(\btheta) & = \langle e_i^A(\cdot) , \Delta\ln A(\cdot; \btheta) \rangle \\
-c_i^\Phi(\btheta) & = \langle e_i^\Phi(\cdot) , \Delta\Phi(\cdot; \btheta) \rangle  .
-\end{align}
-\end{subequations}
-
-In the empirical interpolation method, one chooses a set of frequencies $F_j$ [[to ... minimize impact of interpolation error]].
-The reduced basis can then be inverted to construct empirical interpolating functions $B_j(f)$ [[details...]].
-\begin{subequations}
-\label{eq:empinterpFinal}
-\begin{align}
-\Delta\ln A_S(f; \btheta ) & = \sum_{j=1}^{N_A} B_j^A(f) \Delta\ln A(F^A_j; \btheta) \\
-\Delta\Phi_S(f; \btheta ) & = \sum_{j=1}^{N_\Phi} B_j^\Phi(f) \Delta\Phi(F^\Phi_j; \btheta)
-\end{align}
-\end{subequations}
-The interpolating functions for $\Delta\ln A$ and $\Delta\Phi$ are shown in Fig.~\ref{fig:B}.
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.49\textwidth]{Bamp.png}
-\includegraphics[width=0.49\textwidth]{Bphase.png}
-\caption{Top: $B^{\rm Amp}_j(Mf)$. Bottom: $B^{\rm Phase}_j(Mf)$. Black dots indicate the frequency of the empirical nodes $MF_j$.}
-\label{fig:B}
-\end{figure}
-
-\subsection{Interpolating over waveform parameter space}
-
-We use Gaussian process regression to fit $\Delta\ln A$ and $\Delta\Phi$ at the empirical nodes as a function of
-the waveform parameters. In GPR a function $f(\vec x)$ is described in terms of a mean $m(\vec x)$ and 
+We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~[[]] of Ref.~[[]], the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order in Eqs.~[[]] of Ref.~[[]] and Eqs.~[[]] of Ref.~[[]], and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order in Eqs.~[[]] of Ref.~[[]].
+These terms for the phase are exactly as used in the \texttt{lalsuite} waveform \texttt{TaylorF2} [[cite]].
+
+\subsection{Preconditioning the training set waveforms}
+
+We evaluate the EOB waveform with a starting frequency equivalent to 20Hz for a total mass of $M=2M_\odot$ ($Mfwi=0.000197$).
+In order to take a discrete Fourier transform, we window the start of the waveform with a Planck window [[cite]] in the interval $(Mfwi, Mfwf) = (0.000197, 0.00021)$ to avoid Gibbs oscillations.
+We then resample the waveform with a spacing $\Delta t/M = 5$, and pad the end of the waveform with zeros such that all waveforms in the training set have the exact same time samples.
+
+After evaluating the discrete Fourier transform, evaluate the residuals $\Delta\ln(A)$ and $\Delta\Phi$ between the EOB and TaylorF2 waveforms.
+Because waveforms have free time and phase parameters, subtract a linear fit to $\Delta\Phi$ at the beginning of the waveform in the window $(Mf_{fit,i}, Mf_{fit,f}) = 0.00021(1, 1.05)$.
+The resulting $\Delta\Phi$ is zero and has zero slope at $Mf_{fit,i}$, guaranteeing that the surrogate smoothly matches to TaylorF2 at lower frequencies.
+
+As a result of Gibbs oscillations in the Fourier transformed waveform that weren't completely removed by the Planck waveform as well as from the end of the waveform where the amplitude rapidly drops to zero amplitude after the peak amplitude over $\sim 1$ cycle, small amplitude, high frequency oscillations remain in the residuals $\Delta\ln(A)$ and $\Delta\Phi$.
+We remove these oscillations using a moving average filter centered on $Mf$ with width $[Mf(1-\Delta Mf), Mf(1+\Delta Mf)]$.
+We use a width of $\Delta Mf=0.1$ for the amplitude and $\Delta Mf=0.05$ for the phase.
+
+Finally, we truncate the data to the interval $(Mf_{trunc,i}, Mf_{trunc,f}) = (0.00021, 0.07)$.
+We note that the gravitational wave frequency at the Schwarzschild ISCO is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$.
+Although second generation detectors are only marginally sensitive to signals above this frequency, we use the high-frequency cutoff of $Mf_{trunc,f}=0.07$ because some of the EOB waveforms still have enough amplitude at these high frequencies to affect the structure of the surrogate waveform if inverse Fourier transformed back into the time domain.
+
+\red{[[Show plot (?) to validate that waveform is properly conditioned by plotting amplitude/phase as function of waveform parameters at specific frequencies and show that they're smooth functions?]]}
+
+\subsection{Spline interpolation for frequency $f$}
+
+We now seek to interpolate the conditioned and filtered residuals $\Delta\ln(A)(f; \bx)$ and $\Delta\Phi(f; \bx)$.
+We begin by interpolating as a function of $f$ for fixed $\bx$.
+The majority of previous papers on gravitational waveform surrogates have used an orthonormal basis of global functions $\hat e_i(f)$ for frequency (or time) interpolation [[cite]].
+This type of surrogate is referred to as a reduced order model.
+For a generic function $g(f; \bx)$, this can be written as $g(f; \bx) \approx \sum_{i=1}^N c_i(\bx) \hat e_i(f)$.
+The coefficients $c_i(\bx)$ are then interpolated as functions of $\bx$ [[Puerrer]].
+Alternatively, using the empirical interpolation method [[cite]], one can re-express these basis functions $\hat e_i(f)$ in terms of empirical interpolating functions $B_j(f)$ and the value of $g(f; \bx)$ at empirical nodes $F_j$ as $g(f; \bx) \approx \sum_{j=1}^N B_j(f) g(F_j; \bx)$.
+The location of these nodes $F_j$ can be optimized to minimize interpolation errors.
+This method has been used in Refs.~[[cite]].
+
+For the problem here, we have found that global basis functions do not work well.
+The residuals are small and smooth at low frequencies and large and noisy at high frequencies.
+Errors in evaluating $g(F_j;\bx)$ for high frequency can propagate to large errors at low frequencies.
+Instead, we find that spline interpolation works significantly better.
+We use 20 frequency nodes $F_j$ log-spaced the interval $[f_{trunc,i}, f_{trunc,f}]$, and evaluate the residuals $\Delta\ln(A)(F_j; \bx)$ and $\Delta\Phi(F_j; \bx)$ at these nodes using GPR discussed below.
+We then choose third-order spline interpolation to interpolate between these points.
+The local, third-order polynomials, that are only connected by the requirement of smoothness [[check this]] do not propagate high-frequency errors down to low-frequency errors as significantly.
+
+\subsection{Gaussian process regression for parameters $\bx$}
+
+We use Gaussian process regression to fit the residuals $\Delta\ln(A)(F_j; \bx)$ and $\Delta\Phi(F_j; \bx)$ at the empirical nodes $F_j$ as a function of the waveform parameters.
+In GPR a function $f(\vec x)$ is described in terms of a mean $m(\vec x)$ and
 covariance $k(\vec x, \vec x')$ between points [[cite RassmussenWilliamsChapter2, IntroArxivPaper]].
@@
 
-\subsection{Choosing new training set waveforms}
+\subsection{Iterative construction of training set}
+
+In the subsections below we will build our surrogate with the aim of using as few waveform evaluations as possible.
+We do this because EOB models of BNS systems are already expensive, and we would like a method that extends to higher dimensions and more expensive simulations such as numerical relativity simulations.
+A general overview of the process we have used is as follows:
+\begin{enumerate}
+\item Select an initial, sparse set of waveform parameters $\bx$ that cover the parameter space using a Latin Hypercube Design (LHD).
+
+\item Evaluate the initial training set of EOB waveforms $h(t;\bx)$ and condition them to produce smoothed, frequency-domain waveforms $\tilde h(f;\bx)$.
+
+%\item Construct separate orthonormal, reduced bases for the amplitude $A(f;\bx)$ and phase $\Phi(f;\bx)$ of the training set waveforms.
+Then, use the empirical interpolation method (EIM) to generate separate sets of frequency-dependent empirical interpolating functions for the amplitude $B^A_j(f)$ and phase $B^\Phi_j(f)$.
+
+\item Choose interpolation nodes.
+
+%\item For each empirical interpolating function, interpolate the amplitude $A(F^A_j;\bx)$ or phase $\Phi(F^\Phi_j;\bx)$ at the associated frequency nodes $F^A_j$ or $F^\Phi_j$, respectively, as a function of the waveform parameters $\bx$ using GPR.
+
+\item GPR
+
+\item With the resulting surrogate model and GPR uncertainty estimates, iteratively choose the points in parameter space with the largest estimated waveform errors.
+Then, generate these waveforms and add them to the initial training set.
+
+\item Reconstruct the surrogate model with the updated training set. This can be repeated as necessary with additional training set waveforms.
+
+\end{enumerate}
+
+In constructing a surrogate model, we will need to interpolate several quantities as a function of the five
+waveform parameters ${\bm x}$.
+Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) 
+require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality; 
+the number of samples grows exponentially with the dimension $d$ ($N^d$ for $N$ samples per dimension).
+In the 5-dimensional problem discussed here, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
+At 10 minutes--1 hour per EOB waveform on a standard CPU, this is at the limit of what is reasonable.
+
+However, there are other methods such as Gaussian Process Regression (GPR), discussed below, that allow nonuniform designs.
+One such standard design is a Latin Hypercube Design (LHD). An LHD with $N$ samples divides each dimension uniformly into $N$
+grid points per dimension for a total of $N^d$ possible locations. However, for each dimension each of the $N$ values is sampled 
+exactly once. This has the convenient property that the samples are non-collapsing; if one of the parameters is much less 
+significant than other parameters, samples are not wasted on that parameter, as they would be with a uniform grid. And, a projection
+onto a subspace is still a LHD. [[Also, each dimension is uniformly sampled which is not the case for MC sampling.]]
+
+As an example of the usefulness of noncollapsing designs, the tidal parameters have a minimal impact on the waveform
+at low frequencies, so it wouldn't make sense to densely sample a grid of tidal parameters at low frequencies. However, near
+the merger the tidal parameters can be a important as the spin parameters. One therefore might want more samples for the tidal
+parameters and less samples for the spin. However, due to the expense of the wavefroms we must use the same grid of waveforms 
+for low and high frequencies. A LHD, with its noncollapsing property, bypasses this problem.
+
+For an LHD there are $(N!)^d$ possible designs, and a standard requirement is that the LHD be space-filling, meaning that the 
+points are as far apart as possible from each other. (See [[cite]] for a review of methods for optimizing the placement of samples.) 
+The measure used here is that we maximize the minimum Euclidean distance between any two samples [[cite]]. We do this by 
+sampling $xxx$ random designs and choosing the one with the maximum minimum distance.
+
+For a smooth approximately monotonic function in $d$-dimensions, standard lore is that $10d$ samples are needed to interpolate 
+a function with GPR [[cite]] [[To what accuracy]]. We therefore construct our initial training set with 50 waveform samples, and use 
+these waveforms to construct the reduced basis and perform interpolation with GPR. Once an initial surrogate model is constructed,
+we will use the GPR error estimate to choose new waveforms to add to the training set, and update the surrogate model.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.49\textwidth]{trainingset3d.png}\\
+\includegraphics[width=0.49\textwidth]{trainingset2d.png}
+\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, S_1, S_2\}$ (top)
+and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom). The 160 black circles were used for the initial
+training set (32 corner points + 128 LHD points) that generated the initial surrogate. The 200 red squares were generated
+using uncertainty sampling from the GPR error estimate of the RMS phase error at the empirical nodes. 
+}
+\label{fig:LHD}
+\end{figure}
 
@@
 \caption{Estimated RMS phase error as function of number of sampled points}
-\label{fig:phaserange}
+\label{fig:uncsamp}
 \end{figure}
@@
 
---Constructing the final model with units, inclination angle, time and phase shift, etc.
-
---Implementation details
+Given the interpolating functions for the amplitude and phase residuals $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ and $\{\mathcal{I}_{\rm GPR, j}[\Delta\Phi_j](\bx)\}$ evaluated at the frequency nodes $\{F_j\}$, we can now reconstruct the frequency-domain waveform.
+The surrogates for the residuals $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$ are constructed by interpolating between the nodes $\{F_j\}$ with cubic splines:
+\begin{align}
+\Delta\ln A_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}](Mf),\\
+\Delta\Phi_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\mathcal{I}_{\rm GPR}[\{\Delta\Phi_j](\bx)\}](Mf).
+\end{align}
+We set these functions to 0 below the first frequency node $MF_0 =  Mf_{trunc,i}$ so that the waveform transitions to TaylorF2 at lower frequencies.
+With the analytic expressions for $A_{\rm F2}(Mf;\bx)$ and $\Phi_{\rm F2}(Mf;\bx)$ and the interpolated expressions $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$, the final surrogates for the amplitude and phase are
+\begin{align}
+A_S(Mf;\bx) &= A_{\rm F2}(Mf;\bx) \exp\left[ \Delta\ln A_S(Mf;\bx) \right] \\
+\Phi_S(Mf;\bx) &= \Phi_{\rm F2}(Mf;\bx) + \Delta\Phi_S(Mf;\bx).
+\end{align}
+In physical units the $+$ and $\times$ polarizations of the waveform are then 
+\begin{align}
+\tilde h_+(f; \bx) &= \frac{1}{2}(1+\cos^2\iota) \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
+& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right)\right] \\
+\tilde h_\times(f; \bx) &= \cos\iota \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
+& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right) + i \frac{\pi}{2}\right]
+\end{align}
+where $\iota$ is the inclination angle.
 
@@
 
---Phase error. (needs to be significantly smaller than tidal effects) Use this for order of magnitude estimate of required
-accuracy.
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.49\textwidth]{maxamphist.png}
-\includegraphics[width=0.49\textwidth]{maxphasehist.png}
-\caption{Amplitude and phase error.}
-\label{fig:phaserange}
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.49\textwidth]{dhmaxerror.png}
-\caption{Waveform with largest phase error. It's parameters are....}
-\label{fig:phaserange}
-\end{figure}
+The accuracy of the surrogate can be assessed by comparing it to a test set of waveforms with randomly sampled parameters.
+We evaluate an additional 1000 waveforms randomly sampled in parameter space using the same distribution discussed above to generate the last 1000 waveforms for the training set.
+
+A common measure of the surrogate model accuracy is the mismatch between 
+the surrogate model and the original EOB waveform.
+The mismatch represents the loss in signal-to-noise ratio that would result 
+from using the surrogate model instead of the original EOB waveform. 
+It is defined by the deviation from a perfect overlap after aligning the two waveforms
+using the time and phase free parameters $t_0$ and $\phi_0$:
+\begin{equation}
+\mathcal{M} = 1 - \max_{t_0, \phi_0} \frac{(h_{\rm EOB}, h_{\rm Sur})} {\sqrt{(h_{\rm EOB}, h_{\rm EOB}) (h_{\rm Sur}, h_{\rm Sur})}}.
+\end{equation}
+The inner product here  is the integral of the Fourier transformed waveforms $\tilde h(f)$ weighted by the noise power spectral 
+density (PSD) $S_n(f)$ of the detector:
+\begin{equation}
+(h_1, h_2) = 4 \Re \int_{f_{\rm low}}^{f_{\rm high}} \frac{\tilde h_1(f) \tilde h^*_2(f)} {S_n(f)} df.
+\end{equation}
 
 In Fig.~\ref{fig:mismatch}, we show the distribution of mismatch $\mathcal{M}$ between our surrogate
-and the $10^4$ randomly sampled EOB waveforms. We use the design sensitivity aLIGO PSD~\cite{Aasi:2013wya} and
+and the $10^3$ randomly sampled EOB waveforms. We use the design sensitivity aLIGO PSD~\cite{Aasi:2013wya} and
 a sampling rate of 4096~Hz. Our integration bounds are $f_{\rm low} = x$~Hz and the Nyquist frequency
@@
 we show results for the smaller mass $M_B$ fixed at $1M_\odot$ or fixed at $2M_\odot$.
-We also show the mismatch as a function of parameters in Fig.~\ref{fig:mismatchtriangle}.
-It is generally largest for as the mass ratio decreases toward $1/3$.
+[[This is all wrong. Fix it.]]
+We compare this mismatch to that of th
 
@@
 \includegraphics[width=0.49\textwidth]{mismatch.png}
-\caption{Mismatch. \red{[[You should do mismatch with TaylorF2 as a reference for the improvement in 
-going through all the effort in this paper.]]}
-}
+\caption{Mismatch}
 \label{fig:mismatch}
@@
 \centering
-\includegraphics[width=0.49\textwidth]{mismatchtrianglem10.png}
-\caption{Mismatch for systems with $m_2=1M_\odot$.}
-\label{fig:mismatchtriangle}
+\includegraphics[width=0.49\textwidth]{htildemaxerror.png}
+\caption{Amplitude and phase errors as functions of frequency for the waveform parameters $\bx$ that gave the largest mismatch with the test-set waveforms.}
+\label{fig:maxmismatch}
 \end{figure}
 
---distinguishability
-
---Fisher matrix systematic error.
+To demonstrate that we can correctly recover the behavior of the original unfiltered time-domain EOB waveform, we inverse Fourier transform the surrogate and compare it to the EOB waveform.
+In Fig.~\ref{fig:maxmismatchtd} we compare the surrogate to the test set waveform that had the largest mismatch.
+We align the two waveforms in time and phase by maximizing the overlap at early times in the interval $-200,000 < t/M < -100,000$.
+The two waveforms are nearly identical except during the last $\sim 2$ cycles where the surrogate peaks in amplitude before the EOB waveform.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.49\textwidth]{hmaxerror.png}
+\caption{Inverse Fourier transformed surrogate compared to the original EOB waveform.}
+\label{fig:maxmismatchtd}
+\end{figure}
+
+%\begin{figure}[htb]
+%\centering
+%\includegraphics[width=0.49\textwidth]{mismatchtrianglem10.png}
+%\caption{Mismatch for systems with $m_2=1M_\odot$.}
+%\label{fig:mismatchtriangle}
+%\end{figure}
+
+%--Fisher matrix systematic error. (See spin NSBH paper.)
+%
+%--Phase error. (needs to be significantly smaller than tidal effects) Use this for order of magnitude estimate of required
+%accuracy.
+%
+%\begin{figure}[htb]
+%\centering
+%\includegraphics[width=0.49\textwidth]{maxamphist.png}
+%\includegraphics[width=0.49\textwidth]{maxphasehist.png}
+%\caption{Amplitude and phase error.}
+%\label{fig:ampphaseerr}
+%\end{figure}
+%
+
+%--distinguishability
+%
+%--Fisher matrix systematic error.
 
@@
 
-\subsection{Parameter estimation}
-
---Can only include this section if there's a LAL version. Should you bother?
-
-\section{Discussion}
-
-\subsection{Review}
-
-\subsection{Future work}
-
-Hierarchical method and uncertainty sampling provides way to construct surrogate from minimal number of waveforms.
-
---Could be used to correct the very end of inspiral (and post-merger) with NR simulations where only O(100) simulations are feasable.
-
-%% ________________________________________________________
-\section{Evaluating waveform model accuracy}
-
-In the sections below, we will use several criteria to choose waveforms samples and evaluate the accuracy of the 
-surrogate waveform model. We briefly discuss them here.
-
---Phase error. (needs to be significantly smaller than tidal effects) Use this for order of magnitude estimate of required
-accuracy.
-
---distinguishability
-
---Mismatch
-
-As another common measure of the surrogate model accuracy, we examine the mismatch between 
-the surrogate model and the original EOB waveform.
-The mismatch represents the loss in signal-to-noise ratio that would result 
-from using the surrogate model instead of the original EOB waveform. 
-It is defined by the deviation from a perfect overlap after aligning the two waveforms
-using the time and phase free parameters $t_0$ and $\phi_0$:
-\begin{equation}
-\mathcal{M} = 1 - \max_{t_0, \phi_0} \frac{(h_{\rm EOB}, h_{\rm Sur})} {\sqrt{(h_{\rm EOB}, h_{\rm EOB}) (h_{\rm Sur}, h_{\rm Sur})}}.
-\end{equation}
-The inner product here  is the integral of the Fourier transformed waveforms $\tilde h(f)$ weighted by the noise power spectral 
-density (PSD) $S_n(f)$ of the detector:
-\begin{equation}
-(h_1, h_2) = 4 \Re \int_{f_{\rm low}}^{f_{\rm high}} \frac{\tilde h_1(f) \tilde h^*_2(f)} {S_n(f)} df.
-\end{equation}
-
---Fisher matrix systematic error. (See spin NSBH paper.)
+%\subsection{Parameter estimation}
+%
+%--Can only include this section if there's a LAL version. Should you bother?
+
+\section{Discussion and future work}
+
+We have constructed a fast frequency-domain surrogate of the most accurate BNS waveform currently available for data analysis.
+This aligned-spin model, TEOBv4, incorporates the tidally induced $\ell=2$ and 3 multipole moments as well as the effect of dynamical tides as the excitation approaches the $\ell=2$ and 3 f-mode frequencies.
+We have achieved mismatches of no more than $7 \times 10^{-4}$ and phase errors of $\lesssim 1$~rad up to the merger frequency.
+These are sufficient to not bias results in any of the parameters.
+
+The evaluation time has a flat cost of $\sim 0.02$~s to perform the GPR interpolation at each node.
+The rest of the time is spent resampling the waveform with spline interpolation.
+For a starting frequency of 10~Hz, this takes 0.3~s when the waveform is matched with data uniformly sampled at 4096Hz.
+For a starting frequency of 30~Hz, the total evaluation time is 0.04~s.
+For current parameter estimation codes, this is sufficient.
+However, one could further improve run times using reduced order quadrature [[cite]] which requires frequency-domain waveforms, or adaptively sampled likelihood evaluation [[cite]].
+
+The hierarchical method presented here, where we begin with the analytic TaylorF2 reference model then make a surrogate of the residual, can be used to further improve the waveform model.
+For example, one could build a surrogate for numerical BNS simulations using TEOBv4\_surrogate as the base model, and constructing a surrogate of the residual.
+Such a model would have the accuracy of EOB below $\sim 400$~Hz and the accuracy of numerical simulations for the last several cycles before merger (once numerical simulations can be proven to be more accurate than EOB [[cite Tim?
+Kyohei?]]).
+Using GPR and uncertainty sampling discussed above, one could optimally choose the waveform parameters for the numerical simulations, and run them in parallel to build the training set.
+Because the difference between TEOBv4 and numerical BNS simulations is exceptionally small, one would not need a high fractional accuracy for a surrogate of the difference, and 10--100 waveforms may be sufficient.
+
+This model notably does not include two effects that can be important for systems with large spin.
+The first is the spin-induced quadrupole moment $Q$ that effects the waveform phase at 2PN order [[cite Poisson]].
+The easiest way to correct for this effect is to simply add this 2PN term to the TaylorF2 reference waveform.
+One can use the universal relation between the quadrupole moment $Q$ and the tidal parameter $\Lambda$ as was done with the other NS matter parameters.
+Once this effect is incorporated into the EOB model, one could easily rebuild the surrogate presented in this paper to incorporate this effect.
+
+The second effect is precession for non-aligned spins.
+Although none of the EOB models currently available include both tidal effects and precession, there are ways to boost this waveform model to a precessing frame, and approximately incorporate precession.
+For example, Chatziioanno et al.
+have analytically solved the 2PN-accurate precession equations for generic spins.
+This only works for systems that do not include transitional precession, as is unlikely for low mass ratio BNS systems [[cite]].
+Using stationary uniform asymptotics, they can also analytically Fourier-transform the solution.
+Importantly, one is free to specify the frequency-domain amplitude and phase evolution, such as the TEOBv4\_surrogate here, for the waveform in the co-precessing frame.
+This approach would provide a fast, accurate model for BNS systems with tides and generic spins.
 
@@
 
-BL thanks... BL was supported by...
+BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the Institute for Nuclear Theory where many of the methods used here were discussed.
+BL, MP, and AT were supported by...
 
@@
 \end{document}
+
+%\section{Old Junk text}
+%
+%
+%
+%\sout{A time-domain surrogate has the advantage that significantly less processing of the original waveform is needed, resulting in a surrogate that typically more accurately reproduces the original waveform model.}
+%%A frequency-domain surrogate, on the other hand, requires windowing of the Fourier transformed waveform to avoid Gibbs oscillations.
+For the beginning of the waveform, one can simply begin at a lower frequency, window the waveform, then truncate the Fourier-transformed waveform above some desired starting frequency.
+%
+%
+%\sout{As we will show in Sec.~[[]], these high-frequency Gibbs oscillations in the amplitude of the Fourier transformed waveform are a genuine feature of BNS waveforms unlike BBH waveforms which typically taper over a larger number of GW cycles during their ringdown.
+These oscillations make it difficult to construct a surrogate.
+However, we will show that smoothing these oscillations in the Fourier transformed waveform with a moving average filter leads to a surrogate that is sufficiently faithful to the original time-domain model.
+Another important advantage is that a surrogate of $\tilde h(f;\bx)$ bypasses the need to evaluate the Fourier transform at each waveform evaluation for parameter estimation.
+And, as we shall see, a frequency-domain surrogate can be trivially attached to analytic models that are accurate at low-frequencies, such as TaylorF2, making analysis for third-generation detectors which are sensitive at lower frequencies easier.
+Finally, methods for speeding up the likelihood evaluation in parameter estimation such as reduced order quadrature typically require a fast frequency-domain model for calculating the appropriate weights online [[cite]].
+For these reasons we will construct a surrogate of $\tilde h(f;\bx)$.}
 

### 2017-11-17 src/surrogate.py

@@ -512,7 +512,7 @@ class GPSplineSurrogate(object):
         self.order = order
 
     @classmethod
-    def load(cls, nodes_filename, damp_gp_filename, dphase_gp_filename, order=2, npoints=10000):
+    def load(cls, nodes_filename, damp_gp_filename, dphase_gp_filename, order=3, npoints=10000):
         """Load surrogate model from 4 hdf5 data files.
         """
         mf_amp, mf_phase = load_amp_phase_nodes(nodes_filename)

### 2017-11-17 paper/paper.tex
Incremental improvement in section III.E.

@@
 
-We use Gaussian process regression to fit the residuals $\Delta\ln(A)(F_j; \bx)$ and $\Delta\Phi(F_j; \bx)$ at the empirical nodes $F_j$ as a function of the waveform parameters.
-In GPR a function $f(\vec x)$ is described in terms of a mean $m(\vec x)$ and
-covariance $k(\vec x, \vec x')$ between points [[cite RassmussenWilliamsChapter2, IntroArxivPaper]].
+We now need a method to interpolate the residuals $\Delta\ln(A)(F_j; \bx)$ and $\Delta\Phi(F_j; \bx)$ at the frequency nodes $F_j$ as a function of the five waveform parameters $\bx$.
+Most multivariate interpolation techniques (e.g.
+tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality; the number of samples grows exponentially with the dimension $d$ ($N^d$ for $N$ samples per dimension).
+For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
+At 10 minutes--1 hour per EOB waveform on a standard CPU, this is at the limit of what is reasonable.
+However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs.
+The method we choose is Gaussian Process Regression (GPR).
+
+In GPR a function $f(\vec x)$ is described in terms of a mean $m(\vec x)$ and  covariance $k(\vec x, \vec x')$ between points [[cite RassmussenWilliamsChapter2, IntroArxivPaper]].
 \begin{equation}
@@
 \end{equation}
-The mean $m(x)$ can be some parameterized function which is the standard version of regression, 
-while $k(\vec x, \vec x')$ is a kernel with tunable hyperparameters that describes the covariance between
-the point $\vec x$ and the sampled point $\vec x'$. This covariance can be used to describe features of the function 
-not captured by the parameterized mean as well as genuine noise in the data ${\bm y}$. 
-For the problem here, we have already subtracted the 
-TaylorF2 waveform from EOB waveform, so we set $m=0$ and model the remainder terms purely in terms 
-of the covariance $k$.
-
-In a zero-mean Gaussian process, the data and function $({\bm y}, f_*)$ at the training set and new points $({\bm x}, x_*)$ are
-drawn from a multivariate normal distribution
+The mean $m(x)$ can be some parameterized function which is the standard version of regression, while $k(\vec x, \vec x')$ is a kernel with tunable hyperparameters that describes the covariance between the point $\vec x$ and the sampled point $\vec x'$.
+This covariance can be used to describe features of the function not captured by the parameterized mean as well as genuine noise in the data ${\bm y}$.
+For the problem here, we have already subtracted the TaylorF2 waveform from EOB waveform, so we set $m=0$ and model the remainder terms purely in terms of the covariance $k$.
+
+In a zero-mean Gaussian process, the data and function $({\bm y}, f_*)$ at the training set and new points $({\bm x}, x_*)$ are drawn from a multivariate normal distribution
 \begin{equation}
@@
 
-\subsection{Iterative construction of training set}
-
-In the subsections below we will build our surrogate with the aim of using as few waveform evaluations as possible.
-We do this because EOB models of BNS systems are already expensive, and we would like a method that extends to higher dimensions and more expensive simulations such as numerical relativity simulations.
-A general overview of the process we have used is as follows:
-\begin{enumerate}
-\item Select an initial, sparse set of waveform parameters $\bx$ that cover the parameter space using a Latin Hypercube Design (LHD).
-
-\item Evaluate the initial training set of EOB waveforms $h(t;\bx)$ and condition them to produce smoothed, frequency-domain waveforms $\tilde h(f;\bx)$.
-
-%\item Construct separate orthonormal, reduced bases for the amplitude $A(f;\bx)$ and phase $\Phi(f;\bx)$ of the training set waveforms.
-Then, use the empirical interpolation method (EIM) to generate separate sets of frequency-dependent empirical interpolating functions for the amplitude $B^A_j(f)$ and phase $B^\Phi_j(f)$.
-
-\item Choose interpolation nodes.
-
-%\item For each empirical interpolating function, interpolate the amplitude $A(F^A_j;\bx)$ or phase $\Phi(F^\Phi_j;\bx)$ at the associated frequency nodes $F^A_j$ or $F^\Phi_j$, respectively, as a function of the waveform parameters $\bx$ using GPR.
-
-\item GPR
-
-\item With the resulting surrogate model and GPR uncertainty estimates, iteratively choose the points in parameter space with the largest estimated waveform errors.
-Then, generate these waveforms and add them to the initial training set.
-
-\item Reconstruct the surrogate model with the updated training set. This can be repeated as necessary with additional training set waveforms.
-
-\end{enumerate}
-
-In constructing a surrogate model, we will need to interpolate several quantities as a function of the five
-waveform parameters ${\bm x}$.
-Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) 
-require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality; 
-the number of samples grows exponentially with the dimension $d$ ($N^d$ for $N$ samples per dimension).
-In the 5-dimensional problem discussed here, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
-At 10 minutes--1 hour per EOB waveform on a standard CPU, this is at the limit of what is reasonable.
-
-However, there are other methods such as Gaussian Process Regression (GPR), discussed below, that allow nonuniform designs.
-One such standard design is a Latin Hypercube Design (LHD). An LHD with $N$ samples divides each dimension uniformly into $N$
-grid points per dimension for a total of $N^d$ possible locations. However, for each dimension each of the $N$ values is sampled 
-exactly once. This has the convenient property that the samples are non-collapsing; if one of the parameters is much less 
-significant than other parameters, samples are not wasted on that parameter, as they would be with a uniform grid. And, a projection
-onto a subspace is still a LHD. [[Also, each dimension is uniformly sampled which is not the case for MC sampling.]]
-
-As an example of the usefulness of noncollapsing designs, the tidal parameters have a minimal impact on the waveform
-at low frequencies, so it wouldn't make sense to densely sample a grid of tidal parameters at low frequencies. However, near
-the merger the tidal parameters can be a important as the spin parameters. One therefore might want more samples for the tidal
-parameters and less samples for the spin. However, due to the expense of the wavefroms we must use the same grid of waveforms 
-for low and high frequencies. A LHD, with its noncollapsing property, bypasses this problem.
-
-For an LHD there are $(N!)^d$ possible designs, and a standard requirement is that the LHD be space-filling, meaning that the 
-points are as far apart as possible from each other. (See [[cite]] for a review of methods for optimizing the placement of samples.) 
-The measure used here is that we maximize the minimum Euclidean distance between any two samples [[cite]]. We do this by 
-sampling $xxx$ random designs and choosing the one with the maximum minimum distance.
-
-For a smooth approximately monotonic function in $d$-dimensions, standard lore is that $10d$ samples are needed to interpolate 
-a function with GPR [[cite]] [[To what accuracy]]. We therefore construct our initial training set with 50 waveform samples, and use 
-these waveforms to construct the reduced basis and perform interpolation with GPR. Once an initial surrogate model is constructed,
-we will use the GPR error estimate to choose new waveforms to add to the training set, and update the surrogate model.
+\subsection{Iterative construction of training set and surrogate}
+
+One of the main aims of this paper is to build a surrogate with as few waveform evaluations as possible.
+EOB models are expensive, and we would like a method that extends to higher dimensions and more expensive simulations such as numerical relativity simulations.
+Using the freedom provided by GPR to sample waveforms at arbitrary locations in parameter space, we try a set of more optimal designs.
+
+One such design is a Latin Hypercube Design (LHD) [[cite]].
+An LHD with $N$ samples divides each of $d$ dimension uniformly into $N$ grid points, for a total of $N^d$ grid points.
+However, for each dimension, the $N$ values are sampled exactly once.
+For an LHD there are $(N!)^d$ possible designs, and we choose one randomly~\footnote{There are ways to better choose an LHD.
+A standard requirement is that the LHD be space-filling, meaning that the points are as far apart as possible from each other.
+(See [[cite]] for a review of methods for optimizing the placement of samples.) One such definition of space filling is that the chosen locations maximize the minimum Euclidean distance between any two samples [[cite]].
+We have not studied this here.}.
+
+An LHD has the property that the samples are non-collapsing; a projection onto a subspace is still a uniformly sampled LHD.
+This avoids wasting samples when one of the parameters has much less of an effect than the other parameters.
+As an example of the usefulness of non-collapsing designs, the tidal parameters have a minimal impact on the waveform at low frequencies, so it wouldn't make sense to densely sample a grid of tidal parameters at low frequencies.
+However, near the merger the tidal parameters can be as important as the spin parameters.
+One therefore might want more samples for the tidal parameters and less samples for the spin.
+However, due to the expense of the waveforms we must use the same set of waveforms for low and high frequencies.
+An LHD with its noncollapsing property bypasses this problem.
+
+We build an initial training set with 128 waveforms sampled with an LHD in our 5-dimensional parameter space given by $q\in[1/3, 1]$, $S_1, S_2 \in [-0.5, 0.5]$ and $\Lambda_1, \Lambda_2 \in [0, 5000]$.
+We also find empirically that the GPR uncertainty estimates are largest at the corners of the parameter space, so we also sample the 32 corners.
+We construct our initial surrogate with these 160 waveforms shown in Fig.~\ref{fig:LHD}.
 
@@
 
-Once we have an initial surrogate model, we can choose new training set samples by iteratively searching the parameter space
-for new points ${\bm x}$ that maximize some error criterion. The quantity that we use is the distinguishability between the
-EOB waveform and its surrogate [[cite]]
-\begin{equation}
-|\delta h | = \sqrt{(h - h_{\rm Sur} | h - h_{\rm Sur})}.
-\end{equation}
-For an initial training set with N samples, this function has $\sim N$ local maxima. We use Monte Carlo sampling with ~100,000
-samples and choose the point with the largest distinguishability [[for now. Could also use multistart optimization.]] to find the 
-approximate global maximum. Because $|\delta h |$ is an expensive function that requires an EOB or NR evaluation, 
-we approximate it as follows:
-
-[[Your inexpensive way of approximating $|\delta h |$ that does not depend on $h_{\rm EOB}$ here.]]
-\begin{equation}
-\label{eq:distinguish}
-|\delta h | \approx blah.
-\end{equation}
+With this initial surrogate model, we can choose new training set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion.
+The quantity that we use is the root-mean-square error in the phase at frequencies below $Mf=0.03$ [[justify this]].
+
+[[equation here]]
+
+For an initial training set with $N$ samples, this function has $\sim N$ local maxima. To find the global maximum, we use a basin jumping method [[cite]] to avoid getting stuck in local maxima.
+
+%The quantity that we use is the distinguishability between the
+%EOB waveform and its surrogate [[cite]]
+%\begin{equation}
+%|\delta h | = \sqrt{(h - h_{\rm Sur} | h - h_{\rm Sur})}.
+%\end{equation}
+%For an initial training set with N samples, this function has $\sim N$ local maxima. We use Monte Carlo sampling with ~100,000
+%samples and choose the point with the largest distinguishability [[for now. Could also use multistart optimization.]] to find the 
+%approximate global maximum. Because $|\delta h |$ is an expensive function that requires an EOB or NR evaluation, 
+%we approximate it as follows:
+%
+%[[Your inexpensive way of approximating $|\delta h |$ that does not depend on $h_{\rm EOB}$ here.]]
+%\begin{equation}
+%\label{eq:distinguish}
+%|\delta h | \approx blah.
+%\end{equation}
 
@@
 \end{figure}
+
+%In constructing a surrogate model, we will need to interpolate several quantities as a function of the five
+%waveform parameters ${\bm x}$.
+%Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) 
+%require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality; 
+%the number of samples grows exponentially with the dimension $d$ ($N^d$ for $N$ samples per dimension).
+%In the 5-dimensional problem discussed here, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
+%At 10 minutes--1 hour per EOB waveform on a standard CPU, this is at the limit of what is reasonable.
+
+%For a smooth approximately monotonic function in $d$-dimensions, standard lore is that $10d$ samples are needed to interpolate 
+%a function with GPR [[cite]] [[To what accuracy]]. We therefore construct our initial training set with 50 waveform samples, and use 
+%these waveforms to construct the reduced basis and perform interpolation with GPR. Once an initial surrogate model is constructed,
+%we will use the GPR error estimate to choose new waveforms to add to the training set, and update the surrogate model.
+
+%A general overview of the process we have used is as follows:
+%\begin{enumerate}
+%\item Select an initial, sparse set of waveform parameters $\bx$ that cover the parameter space using a Latin Hypercube Design (LHD).
+%
+%\item Evaluate the initial training set of EOB waveforms $h(t;\bx)$ and condition them to produce smoothed, frequency-domain waveforms $\tilde h(f;\bx)$.
+%
+%%\item Construct separate orthonormal, reduced bases for the amplitude $A(f;\bx)$ and phase $\Phi(f;\bx)$ of the training set waveforms.
+Then, use the empirical interpolation method (EIM) to generate separate sets of frequency-dependent empirical interpolating functions for the amplitude $B^A_j(f)$ and phase $B^\Phi_j(f)$.
+%
+%\item Choose interpolation nodes.
+%
+%%\item For each empirical interpolating function, interpolate the amplitude $A(F^A_j;\bx)$ or phase $\Phi(F^\Phi_j;\bx)$ at the associated frequency nodes $F^A_j$ or $F^\Phi_j$, respectively, as a function of the waveform parameters $\bx$ using GPR.
+%
+%\item GPR
+%
+%
+%\item With the resulting surrogate model and GPR uncertainty estimates, iteratively choose the points in parameter space with the largest estimated waveform errors.
+Then, generate these waveforms and add them to the initial training set.
+%
+%\item Reconstruct the surrogate model with the updated training set. This can be repeated as necessary with additional training set waveforms.
+%
+%\end{enumerate}
 
@@
 BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the Institute for Nuclear Theory where many of the methods used here were discussed.
+We thank Sylvain Marsat for fixing issues that caused the original TEOBv4 model to be numerically unstable.
 BL, MP, and AT were supported by...

### 2017-11-17 src/TEOB-LAL-implementation/LALSimIMRTEOBv4ROM.c
LAL code agrees with Python surrogate code
Comparison against FFTed TEOBv4 is reasonable. Phase convention should be correct now.
Timeshift should be tweaked a bit further.

@@ -1053,9 +1053,12 @@ static int TaylorF2Phasing(
   double q,
   double chi1,
   double chi2,
-  double lambda1,
-  double lambda2,
-  gsl_vector *Mf
+  double lambda1,   // tidal deformability of body 1
+  double lambda2,   // tidal deformability of body 2
+  double dquadmon1, // self-spin deformation of body 1
+  double dquadmon2, // self-spin deformation of body 2
+  gsl_vector *Mfs,
+  gsl_vector **PNphase // Output: TaylorF2 phase at frequencies Mfs
 );
 
 static int TaylorF2Phasing(
@@ -1063,10 +1066,17 @@ static int TaylorF2Phasing(
   double q,
   double chi1,
   double chi2,
-  double lambda1,
-  double lambda2,
-  gsl_vector *Mfs
+  double lambda1,   // tidal deformability of body 1
+  double lambda2,   // tidal deformability of body 2
+  double dquadmon1, // self-spin deformation of body 1
+  double dquadmon2, // self-spin deformation of body 2
+  gsl_vector *Mfs,
+  gsl_vector **PNphase // Output: TaylorF2 phase at frequencies Mfs
 ) {
+  XLAL_CHECK(PNphase != NULL, XLAL_EFAULT);
+  XLAL_CHECK(*PNphase == NULL, XLAL_EFAULT);
+  *PNphase = gsl_vector_alloc(Mfs->size);
+
   PNPhasingSeries *pn = NULL;
   LALDict *extraParams = XLALCreateDict();
   XLALSimInspiralWaveformParamsInsertPNSpinOrder(extraParams, LAL_SIM_INSPIRAL_SPIN_ORDER_35PN);
@@ -1078,8 +1088,11 @@ static int TaylorF2Phasing(
 
   // FIXME: We should be able to switch on quadrupole-monopole terms (self-spin deformation of the two bodies)
   // from TaylorF2; later these can be replaced with EOB as well. Just add a switch for now.
-  // XLALSimInspiralWaveformParamsInsertdQuadMon1(extraParams, dquadmon1);
-  // XLALSimInspiralWaveformParamsInsertdQuadMon2(extraParams, dquadmon2);
[…]

### 2017-11-17 src/TEOB-LAL-implementation/LALSimIMR.h
Major cleanup of C code
This changed the timeshift in the phase again and also the overall amount of zeros in the polarizations.
Otherwise the waveform is unchanged!

Still to be fixed: TF2 discrepancy between Ben's code and LAL

@@ -1,0 +1,298 @@
+/*
+ * Copyright (C) 2011 N. Fotopoulos <nickolas.fotopoulos@ligo.org>
+ *
+ *  This program is free software; you can redistribute it and/or modify
+ *  it under the terms of the GNU General Public License as published by
+ *  the Free Software Foundation; either version 2 of the License, or
+ *  (at your option) any later version.
+ *
+ *  This program is distributed in the hope that it will be useful,
+ *  but WITHOUT ANY WARRANTY; without even the implied warranty of
+ *  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+ *  GNU General Public License for more details.
+ *
+ *  You should have received a copy of the GNU General Public License
+ *  along with with program; see the file COPYING. If not, write to the
+ *  Free Software Foundation, Inc., 59 Temple Place, Suite 330, Boston,
+ *  MA  02111-1307  USA
+ */
+
+#ifndef _LALSIMIMR_H
+#define _LALSIMIMR_H
+
+#include <lal/LALDatatypes.h>
+#include <lal/LALSimInspiral.h>
+
+#ifdef LAL_HDF5_ENABLED
+#include <lal/H5FileIO.h>
+#endif
+
+#if defined(__cplusplus)
+extern "C" {
+#elif 0
+} /* so that editors will match preceding brace */
+#endif
+
+/**
+ * @defgroup LALSimIMR_h Header LALSimIMR.h
+ * @ingroup lalsimulation_inspiral
+ *
+ * @brief Routines for generating inspiral-merger-ringdown waveforms.
+ *
+ * @{
+ * @defgroup LALSimIMRPhenom_c                   LALSimIMRPhenom.c
+ * @defgroup LALSimIMREOBNRv2_c                  LALSimIMREOBNRv2.c
+ * @defgroup LALSimIMRSpinAlignedEOB_c           LALSimIMRSpinAlignedEOB.c
+ * @defgroup LALSimIMRSpinPrecEOB_c              LALSimIMRSpinPrecEOB.c
+ * @defgroup LALSimIMRSEOBNRROM_c                LALSimIMRSEOBNRvxROMXXX.c
+ * @defgroup LALSimIMRSEOBNRv2ChirpTime_c        LALSimIMRSEOBNRv2ChirpTime.c
[…]

### 2017-11-17 src/TEOB-LAL-implementation/LALSimIMRTEOBv4ROM.c

@@ -56,15 +56,20 @@
 #ifdef LAL_HDF5_ENABLED
 #include <lal/H5FileIO.h>
 // FIXME: change these to our datafile and version number 1.0.0
-static const char ROMDataHDF5[] = "SEOBNRv4ROM_v2.0.hdf5";
-static const INT4 ROMDataHDF5_VERSION_MAJOR = 2;
-static const INT4 ROMDataHDF5_VERSION_MINOR = 0;
-static const INT4 ROMDataHDF5_VERSION_MICRO = 0;
+//static const char ROMDataHDF5[] = "SEOBNRv4ROM_v2.0.hdf5";
+//static const char ROMDataHDF5[] = "/Users/mpuer/Documents/gpsurrogate/src/TEOB-LAL-implementation/TEOBv4_surrogate.hdf5";
+// FIXME: missing attributes in HDF5 file: Email, Description (rather than description), version_major, version_minor, version_micro
+static const char ROMDataHDF5[] = "TEOBv4_surrogate.hdf5";
+// FIXME: uncomment after adding attributes
+// static const INT4 ROMDataHDF5_VERSION_MAJOR = 1;
+// static const INT4 ROMDataHDF5_VERSION_MINOR = 0;
+// static const INT4 ROMDataHDF5_VERSION_MICRO = 0;
 #endif
 
 #include <lal/LALSimInspiral.h>
 #include <lal/LALSimIMR.h>
 
+#include "LALSimIMREOBNRv2.h"
 #include "LALSimInspiralPNCoefficients.c"
 #include "LALSimIMRSEOBNRROMUtilities.c"
 
@@ -75,16 +80,10 @@ static const INT4 ROMDataHDF5_VERSION_MICRO = 0;
 
 
 // TODO:
-// D add lambda1,2 args
-// D add prototypes
-// * remove the glueing functions -- will be easier to copy logic from earlier code
-// D test reading of HDF5 test data
-// D implement code to compute kernel and gpr
-// * compute and check amp,phi at EIM nodes
-// * assemble waveform and check against python output
 // * add waveform approximant and glueing code in LALSimInspiral
-// remove all debugging code
-// add more checks and tests
+// * remove all debugging code: fprintf statements
+// * add more checks and tests
+// * take care of all FIXMEs
 
 
 #ifdef LAL_PTHREAD_LOCK
[…]

### 2017-11-20 paper/paper.tex
Mostly complete descriptions in all subsections.

@@
 \def\bx{\mathbf{x}}
+\def\by{\mathbf{y}}
 \def\btheta{\boldsymbol{\theta}}
@@
 One such design is a Latin Hypercube Design (LHD) [[cite]].
-An LHD with $N$ samples divides each of $d$ dimension uniformly into $N$ grid points, for a total of $N^d$ grid points.
+An LHD with $N$ samples divides each of $d$ dimensions uniformly into $N$ grid points, for a total of $N^d$ grid points.
 However, for each dimension, the $N$ values are sampled exactly once.
-For an LHD there are $(N!)^d$ possible designs, and we choose one randomly~\footnote{There are ways to better choose an LHD.
+For an LHD there are $(N!)^d$ possible designs, and we choose one randomly\footnote{There are ways to better choose an LHD.
 A standard requirement is that the LHD be space-filling, meaning that the points are as far apart as possible from each other.
@@
 We have not studied this here.}.
-
-An LHD has the property that the samples are non-collapsing; a projection onto a subspace is still a uniformly sampled LHD.
+An LHD has the property that the samples are non-collapsing; a projection onto a subspace is still an LHD.
 This avoids wasting samples when one of the parameters has much less of an effect than the other parameters.
 As an example of the usefulness of non-collapsing designs, the tidal parameters have a minimal impact on the waveform at low frequencies, so it wouldn't make sense to densely sample a grid of tidal parameters at low frequencies.
-However, near the merger the tidal parameters can be as important as the spin parameters.
-One therefore might want more samples for the tidal parameters and less samples for the spin.
+However, during the merger the tidal parameters can be more important than the other parameters.
+One therefore might want more samples for the tidal parameters and less samples for the other parameters.
 However, due to the expense of the waveforms we must use the same set of waveforms for low and high frequencies.
@@
 We build an initial training set with 128 waveforms sampled with an LHD in our 5-dimensional parameter space given by $q\in[1/3, 1]$, $S_1, S_2 \in [-0.5, 0.5]$ and $\Lambda_1, \Lambda_2 \in [0, 5000]$.
-We also find empirically that the GPR uncertainty estimates are largest at the corners of the parameter space, so we also sample the 32 corners.
+In addition, we find empirically that the GPR uncertainty estimates are largest at the corners of the parameter space, so we also sample the 32 corners.
 We construct our initial surrogate with these 160 waveforms shown in Fig.~\ref{fig:LHD}.
@@
 \includegraphics[width=0.49\textwidth]{trainingset2d.png}
-\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, S_1, S_2\}$ (top)
-and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom). The 160 black circles were used for the initial
-training set (32 corner points + 128 LHD points) that generated the initial surrogate. The 200 red squares were generated
-using uncertainty sampling from the GPR error estimate of the RMS phase error at the empirical nodes. 
+\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, S_1, S_2\}$ (top) and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom).
+The 160 black circles were used for the initial training set (32 corner points + 128 LHD points) that generated the initial surrogate.
+The 200 red squares were generated using uncertainty sampling from the GPR error estimate of the RMS phase error at the interpolation nodes.
+The final 1000 waveforms are drawn uniformly from parameter space.
 }
@@
 With this initial surrogate model, we can choose new training set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion.
-The quantity that we use is the root-mean-square error in the phase at frequencies below $Mf=0.03$ [[justify this]].
-
-[[equation here]]
-
-For an initial training set with $N$ samples, this function has $\sim N$ local maxima. To find the global maximum, we use a basin jumping method [[cite]] to avoid getting stuck in local maxima.
-
-%The quantity that we use is the distinguishability between the
-%EOB waveform and its surrogate [[cite]]
-%\begin{equation}
-%|\delta h | = \sqrt{(h - h_{\rm Sur} | h - h_{\rm Sur})}.
-%\end{equation}
-%For an initial training set with N samples, this function has $\sim N$ local maxima. We use Monte Carlo sampling with ~100,000
-%samples and choose the point with the largest distinguishability [[for now. Could also use multistart optimization.]] to find the 
-%approximate global maximum. Because $|\delta h |$ is an expensive function that requires an EOB or NR evaluation, 
-%we approximate it as follows:
-%
-%[[Your inexpensive way of approximating $|\delta h |$ that does not depend on $h_{\rm EOB}$ here.]]
-%\begin{equation}
-%\label{eq:distinguish}
-%|\delta h | \approx blah.
-%\end{equation}
-
-With this new sample, one could evaluate the waveform at the new point and reconstruct the surrogate model, then iterate.
-
-Reconstructing the surrogate for each new sample is inexpensive compared to the time needed to simulate a new NR waveform. 
-However, even for NR waveforms, one often wants to perform $N_{\rm new}$ new waveform simulations in parallel. 
-We therefore need a method for optimally choosing $N_{\rm new}$ new samples. One could simply choose the 
-$N_{\rm new}$ local maxima with the largest $|\delta h |$. However, there is a more optimal way of iteratively finding new points.
-If we hold the hyperparameters ${\bm \theta}$ fixed, we note that the GPR error estimate (Eq.~\eqref{eq:var}) does not depend
-on the data ${\bm y}$, just on the samples ${\bm x}$. Therefore, the approximate $|\delta h |$ (Eq.~\eqref{eq:distinguish}) does not depend
-on the new samples either. The method uses the idea of uncertainty sampling [[cite]] but without updating the hyperparameters
-after each new point is chosen. The algorithm for choosing the new points is as follows.
-
-While $|\delta h |_{\rm max} > \epsilon$:
+The quantity that we use is the root-mean-square error in the phase at frequencies below $Mf=0.03$:
+\begin{equation}
+\epsilon_{\rm RMS} = \sqrt{\frac{1}{N} \sum_i (\sigma_{\Delta\Phi_i})^2},
+\label{eq:errorrms}
+\end{equation}
+where $\sigma_{\Delta\Phi_i}$ is the GPR estimate of the uncertainty in $\Delta\Phi(MF_i, \bx)$.
+[[justify the definition of error]].
+For an initial training set with $N$ samples, this function has $\sim N$ local maxima located in the voids between samples $\bx$.
+To find the global maximum, we use a basin jumping method [[cite]] to avoid getting stuck in local maxima.
+
+For low starting frequencies, reconstructing the surrogate for each new EOB waveform is smaller than the time needed to simulate a new EOB waveform.
+However, EOB waveforms are expensive enough that we would like a method to efficinetly choose $N_{\rm new}$ new waveforms in parallel.
+Reconstructing the surrogate with each new EOB waveform allows us to retune the hyperparameters $\btheta$ with each new waveform.
+However, if we hold the hyperparameters $\btheta$ fixed, we note that the GPR error estimate (Eq.~\eqref{eq:var}) does only depends on the samples $\bx$ and not on the data $\by$.
+Therefore, the approximate $\epsilon_{\rm rms}$ (Eq.~\eqref{eq:distinguish}) only depends on the parameters of the new waveform, not on the waveform data.
+The method uses the idea of uncertainty sampling [[cite]] but without updating the hyperparameters after each new point is chosen.
+The algorithm for choosing the $N_{\rm new}$ new points is as follows.
+For $i = 1, \dots, N_{\rm new}$:
 \begin{enumerate}
-\item Construct the GPR error functions (Eq.~\eqref{eq:var}) at each of the empirical nodes $F^A_j$ and $F^\Phi_j$.
-In practice this can be done by specifying the samples ${\bm x}$, the hyperparameters ${\bm \theta}$, and dummy 
-data ${\bm y}={\bm 1}$. These GPR error functions are the input to Eq.~\eqref{eq:distinguish}.
-
-\item Find $({\bm x}_{\rm max}, |\delta h |_{\rm max})$ that maximizes Eq.~\eqref{eq:distinguish} over the parameter space. 
+
+\item Construct the GPR error estimator (Eq.~\eqref{eq:errorrms}).
+In practice this can be done by specifying the samples $\bx$, the hyperparameters $\btheta$, and dummy data for $\by$ since Eq.~\eqref{eq:errorrms} does not depend on $\by$.
+
+\item Find $({\bm x}_{\rm max}, \epsilon_{\rm rms})$ that maximizes Eq.~\eqref{eq:errorrms} over the parameter space.
+
 \item Add ${\bm x}_{\rm max}$ to the list of samples: ${\bm x} = [{\bm x}, {\bm x}_{\rm max}]$.
+
 \end{enumerate}
-These new waveforms can now be evaluated in parallel.
-
-[[The methods used here fall under the terms active learning and Bayesian optimization [[cite]]. The specific method being used is
-uncertainty sampling [[cite]]. The parallelization method I am using I have not seen in the literature. Is it actually new? Ask Niki.]]
-
-[[Another method you could try is k-means clustering of the mismatch for your validation set (using mismatch above some threshold), 
-where you have $k=N$, because there will
-be $\sim N$ local maxima of the mismatch. Within each cluster, choose the waveform with the largest mismatch and add it to the
-training set. This way you only use one new point per local maximum region. You could do the same thing with the Fisher matrix
-systematic error calculation instead of the mismatch if that's a better quantity to use.]]
+These new waveforms can now be evaluated in parallel. An updated surrogate can then be constructed with re-optimized hyperparameters $\btheta$ using the $N_{\rm old}+N_{\rm new}$ waveforms.
+
+Fig.~\ref{fig:LHD} shows the parameters of the 200 new waveforms chosen by the uncertainty sampling method.
+We note that the edges faces of the parameter space are more often chosen than the inner region.
+This is likely a consequence of the Mat\'{e}rn GPR kernel used.
+Ref.~[[]] found similar behavior when using a Mat\'{e}rn kernel instead of a squared-exponential kernel.
+Fig.~\ref{fig:uncsamp} shows the RMS error $\epsilon_{\rm rms}$ maximized over the parameters $\bx$ for each new sample added to the training set.
 
@@
 \includegraphics[width=0.49\textwidth]{uncertaintysampling.png}
-\caption{Estimated RMS phase error as function of number of sampled points}
+\caption{Estimated RMS phase error $\epsilon_{\rm rms}$ as function of the number of sampled waveforms (blue curve).
+The first 160 waveforms are from the initial training set chosen from the corners of parameter space and an LHD.
+The next 200 waveforms are chosen with uncertainty sampling.
+The final 1000 waveforms are drawn uniformly from parameter space.
+The markers are the comparisons of $\epsilon_{\rm rms}$ calculated by maximizing the error over a test set of 1000 waveforms instead of using Eq.~\eqref{eq:errorrms} for the error estimate.}
 \label{fig:uncsamp}
 \end{figure}
-
-%In constructing a surrogate model, we will need to interpolate several quantities as a function of the five
-%waveform parameters ${\bm x}$.
-%Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) 
-%require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality; 
-%the number of samples grows exponentially with the dimension $d$ ($N^d$ for $N$ samples per dimension).
-%In the 5-dimensional problem discussed here, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
-%At 10 minutes--1 hour per EOB waveform on a standard CPU, this is at the limit of what is reasonable.
-
-%For a smooth approximately monotonic function in $d$-dimensions, standard lore is that $10d$ samples are needed to interpolate 
-%a function with GPR [[cite]] [[To what accuracy]]. We therefore construct our initial training set with 50 waveform samples, and use 
-%these waveforms to construct the reduced basis and perform interpolation with GPR. Once an initial surrogate model is constructed,
-%we will use the GPR error estimate to choose new waveforms to add to the training set, and update the surrogate model.
-
-%A general overview of the process we have used is as follows:
-%\begin{enumerate}
-%\item Select an initial, sparse set of waveform parameters $\bx$ that cover the parameter space using a Latin Hypercube Design (LHD).
-%
-%\item Evaluate the initial training set of EOB waveforms $h(t;\bx)$ and condition them to produce smoothed, frequency-domain waveforms $\tilde h(f;\bx)$.
-%
-%%\item Construct separate orthonormal, reduced bases for the amplitude $A(f;\bx)$ and phase $\Phi(f;\bx)$ of the training set waveforms.
-Then, use the empirical interpolation method (EIM) to generate separate sets of frequency-dependent empirical interpolating functions for the amplitude $B^A_j(f)$ and phase $B^\Phi_j(f)$.
-%
-%\item Choose interpolation nodes.
-%
-%%\item For each empirical interpolating function, interpolate the amplitude $A(F^A_j;\bx)$ or phase $\Phi(F^\Phi_j;\bx)$ at the associated frequency nodes $F^A_j$ or $F^\Phi_j$, respectively, as a function of the waveform parameters $\bx$ using GPR.
-%
-%\item GPR
-%
-%
-%\item With the resulting surrogate model and GPR uncertainty estimates, iteratively choose the points in parameter space with the largest estimated waveform errors.
-Then, generate these waveforms and add them to the initial training set.
-%
-%\item Reconstruct the surrogate model with the updated training set. This can be repeated as necessary with additional training set waveforms.
-%
-%\end{enumerate}
 
@@
 
-Although optimizing the hyperparameters scales with number of samples $n$ as $\mathcal{O}(n^3)$
-due to the required matrix inversion, the evaluation time of a stored GPR scales as $\mathcal{O}(n)$.
-The evaluation time for the waveform functions will therefore be $\mathcal{O}(n(N_A+N_\Phi))$.
-
-The surrogate has a fixed time of $\sim 50$~ms to evaluate the GPR at each node then evaluate the amplitude
-and phase at the log-spaced frequencies. Then, there is an additional cost to resample the amplitude and phase
-at uniformly-spaced frequencies with interpolation, and finally evaluating $\tilde h_+$ and $\tilde h_\times$. 
-Below 50~Hz the surrogate is faster than all models except TaylorF2 [[you didn't try PhenomP etc.]], and below
-18~Hz it is faster than TaylorF2.
+The surrogate evaluation time can be roughly decomposed into two times.
+The first is the time needed to evaluate the residuals $\Delta\ln(A)(F_j; \bx)$ and $\Delta\Phi(F_j; \bx)$ at each of the $N_A+N_\Phi$ interpolating nodes.
+Although optimizing the hyperparameters scales with the number of samples $n$ as $\mathcal{O}(n^3)$ due to the required matrix inversion, the evaluation time of a stored GPR scales as $\mathcal{O}(n)$.
+The evaluation time for the residuals will therefore be fixed cost of $\mathcal{O}(n(N_A+N_\Phi))$.
+The second part is the time needed to resample the final surrogate (Eq.[[ref]]) at uniformly spaced frequency samples in physical units beginning at a starting frequency $f_{\rm low}$.
+
+In Fig.~\ref{fig:timing}, we show the surrogate evaluation time as a function of starting frequency $f_{\rm low}$.
+The surrogate has a fixed time of $\sim 50$~ms to evaluate the GPR at each node.
+The additional cost to resample the amplitude and phase at uniformly-spaced frequencies with spline interpolation dominates at starting frequencies below $\sim 30$~Hz.
+We also compare the evaluation time to several BNS waveform models that have been used for LIGO data analysis.
+We note that at $\sim 30$~Hz, the TEOBv4\_surrogate model is comparable to the SEOBNRv4\_ROM\_NRTidal and IMRPhenomD\_NRTidal models that were used in the analysis of the GW170817 event [[cite]].
+At lower frequencies, the surrogate is faster than all BNS waveform models except for TaylorF2.
 
@@
 
---Frequency domain allows for evaluating likelihood with non-uniform sampling.
-
---Can construct a ROQ of the model if needed.
-
-%\subsection{Parameter estimation}
-%
-%--Can only include this section if there's a LAL version. Should you bother?
-
 \section{Discussion and future work}
@@
 BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the Institute for Nuclear Theory where many of the methods used here were discussed.
-We thank Sylvain Marsat for fixing issues that caused the original TEOBv4 model to be numerically unstable.
-BL, MP, and AT were supported by...
+We also thank Sylvain Marsat for correcting numerical instabilities in the original TEOBv4 code.
+The authors were supported by...
 

### 2017-11-21 paper/paper.tex
Text edits.

@@
 
-%As discussed above in Sec.~II, the current implementation of the EOB model takes at least 10~minutes to evaluate.
-In this section we will describe in detail the method for training and evaluating our surrogate model.
-The first decision in building a waveform surrogate is choosing between building a surrogate of the time-domain waveform $h(t;\bx)$ or the frequency-domain waveform $\tilde h(f;\bx)$.
-A time-domain surrogate has the advantage that significantly less processing of the original waveform is needed, resulting in a surrogate that typically more accurately reproduces the original waveform model.
-A frequency-domain surrogate, on the other hand, requires windowing of the Fourier transformed waveform to avoid Gibbs oscillations.
-For the beginning of the waveform, one can simply begin at a lower frequency, window the waveform, then truncate the Fourier-transformed waveform above some desired starting frequency.
-At the end of the waveform, the BNS amplitude drops rapidly over $\sim 1$~GW cycle, resulting in Gibbs oscillations at the high-frequency end.
-One could window the end of the waveform over several cycles, however, this would remove available information about tidal interactions.
-As we will show in Sec.~[[]], these high-frequency Gibbs oscillations in the amplitude of the Fourier transformed waveform are a genuine feature of BNS waveforms unlike BBH waveforms which typically taper over a larger number of GW cycles during their ringdown.
-These oscillations make it difficult to construct a surrogate.
-However, we will show that smoothing these oscillations in the Fourier transformed waveform with a moving average filter leads to a surrogate that is sufficiently faithful to the original time-domain model.
-Another important advantage is that a surrogate of $\tilde h(f;\bx)$ bypasses the need to evaluate the Fourier transform at each waveform evaluation for parameter estimation.
-And, as we shall see, a frequency-domain surrogate can be trivially attached to analytic models that are accurate at low-frequencies, such as TaylorF2, making analysis for third-generation detectors which are sensitive at lower frequencies easier.
-Finally, methods for speeding up the likelihood evaluation in parameter estimation such as reduced order quadrature typically require a fast frequency-domain model for calculating the appropriate weights online [[cite]].
-For these reasons we will construct a surrogate of $\tilde h(f;\bx)$.
+In this section we describe a method for decomposing the waveform into smooth, slowly-varying functions and training a surrogate model for these functions.
+We will work with the frequency domain waveform $\tilde h(f; \bx)$.
+This is favorable for data analysis which is usually done in the frequency domain.
+It also allows us to extend the model down to arbitrarily low frequencies with analytic, Frequency-domain PN approximations.
+Unfortunately, Fourier transforming a finite length, discretely sampled waveform, leads to a surrogate with more noise.
+We will show below, however, that sufficient filtering can mostly solve this problem.
 
@@
 
-In this section we will describe the method for decomposing the waveform into smooth, slowly-varying functions and training a surrogate model for these functions.
-As discussed above, we work with the frequency domain waveform $\tilde h(f; \bx)$.
-This is favorable for data analysis which is usually done in the frequency domain.
-However, Fourier transforming a finite length, discretely sampled waveform, typically leads to a surrogate with more noise.
-We will show below, that sufficient filtering can solve most of these problems.
-Because the waveform $\tilde h(f; \bx)$ is an oscillatory function of $f$ and $\bx$, the waveform is usually decomposed into an amplitude and phase $\tilde h(f; \bx) = A(f; \bx) e^{i\Phi(f; \bx)}$, which are smoother, mostly monotonic functions of $f$ and $\bx$.
-The amplitude and phase of these waveforms are shown in Fig.~\ref{fig:h}.
+Because the waveform $\tilde h(f; \bx)$ is an oscillatory function of $f$ and $\bx$, the waveform is usually decomposed into an amplitude $A(f; \bx)$ and phase $\Phi(f; \bx)$ as $\tilde h(f; \bx) = A(f; \bx) e^{i\Phi(f; \bx)}$.
+These functions are smoother, mostly monotonic functions of $f$ and $\bx$.
+This can be seen in Fig.~\ref{fig:h} which shows the initial training set of waveforms.
+\red{[[Actually, you might want to show the 51 waveforms with only $S_1$ varied.
+Then show them as a function of $f$ and another plot as a function of $S_1$.]]} Unfortunately, the amplitude and phase still span a wide range of values.
+The phase, for example, spans about $10^4$~rad.
+To avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad.
+This is a difficult requirement to achieve in the 5-dimensional parameter space here.
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{h_trainingset.png}\\
-\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
-\caption{Training set}
+\includegraphics[width=0.49\textwidth]{h_trainingset.png}
+%\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
+\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
 \label{fig:h}
@@
 
-Although the amplitude and phase are relatively smooth functions, they still span a wide range of values.
-The phase, for example, spans about $10^4$~rad.
-Unfortunately, to avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad.
-This is a difficult requirement to achieve in the 5-dimensional parameter space here.
-However, for aligned-spin waveforms these functions are approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2 $\tilde h_{\rm F2}(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2} (f; \bx)}$ [[cite]].
-We can therefore evaluate the EOB waveform in terms of the residuals, $\Delta\ln(A)$ and $\Delta\Phi$, between EOB and TaylorF2
+For aligned-spin waveforms, we can solve this problem using that fact that the waveform is approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2 [[cite]] $\tilde h_{\rm F2}(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2} (f; \bx)}$.
+This allows us to write the EOB amplitude and phase in terms of the small residuals, $\Delta\ln(A)(f; \bx)$ and $\Delta\Phi(f; \bx)$, relative to TaylorF2:
 \begin{align}
-\tilde h(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2}(f; \bx)} e^{\Delta\ln(A)(f; \bx) + i  \Delta\Phi(f; \bx)},
+\Delta\ln(A)(f; \bx) &= \ln\left(\frac{ A(f; \bx) }{ A_{\rm F2}(f; \bx)}\right), \label{eq:ampres}\\
+\Delta\Phi(f; \bx) & = \Phi(f; \bx) - \Phi_{\rm F2}(f; \bx), \label{eq:phaseres}
+\end{align}
+such that
+\begin{align}
+\tilde h(f; \bx) = \tilde h_{\rm F2}(f; \bx) e^{\Delta\ln(A)(f; \bx) + i  \Delta\Phi(f; \bx)}.
 \label{eq:hdecomp}
 \end{align}
-where
-\begin{align}
-\Delta\ln(A)(f; \bx) &= \ln\left(\frac{ A(f; \bx) }{ A_{\rm F2}(f; \bx)}\right),\\
-\Delta\Phi(f; \bx) & = \Phi(f; \bx) - \Phi_{\rm F2}(f; \bx).
-\end{align}
-These residuals are shown in Fig.~\ref{fig:h}.
-For the amplitude, we use the log-ratio instead of the ratio because it guarantees that the amplitude of the reconstructed waveform in Eq.~\eqref{eq:hdecomp} is positive even if interpolation error leads to a negative $\Delta\ln(A)(f; \bx)$.
+%\begin{align}
+%\tilde h(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2}(f; \bx)} e^{\Delta\ln(A)(f; \bx) + i  \Delta\Phi(f; \bx)}.
+%\label{eq:hdecomp}
+%\end{align}
+These residuals are shown in Fig.~\ref{fig:dh}.
+For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that the amplitude of the reconstructed waveform in Eq.~\eqref{eq:hdecomp} will be positive even if interpolation error leads to a negative $\Delta\ln(A)(f; \bx)$.
 In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency of each waveform, the log-ratio captures this behavior.
-
-To demonstrate the usefulness of this approach, in Fig.~\ref{fig:phaserange}, we show the range of values of the phase $\Phi$ over the training set $\Phi_{\rm range} = \Phi_{\rm max} - \Phi_{\rm min}$ as well as the range of values of the phase residual $\Delta\Phi$ over the training set $\Delta\Phi_{\rm range} = \Delta\Phi_{\rm max}-\Delta\Phi_{\rm min}$.
+Comparing Fig.~\ref{fig:dh} to Fig.~\ref{fig:h}, we find that for most of the frequency range, the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$.
+
+\begin{figure}[htb]
+\centering
+%\includegraphics[width=0.49\textwidth]{h_trainingset.png}\\
+\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
+\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
+\label{fig:dh}
+\end{figure}
+
+%To demonstrate the usefulness of this approach, in Fig.~\ref{fig:phaserange}, we show the range of values of the phase $\Phi$ over the training set $\Phi_{\rm range} = \Phi_{\rm max} - \Phi_{\rm min}$ as well as the range of values of the phase residual $\Delta\Phi$ over the training set $\Delta\Phi_{\rm range} = \Delta\Phi_{\rm max}-\Delta\Phi_{\rm min}$.
 The range in $\Phi$ can be larger than 1000~radians.
 The range of $\Delta\Phi$ is 1--2 orders of magnitude smaller, and as a result, the interpolation accuracy requirements are 1--2 orders of magnitude smaller as well.
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.49\textwidth]{phaserange.png}
-\caption{Range of phase values for training set waveforms. The range is a factor of $\gtrsim 100$ larger for $\Phi$ than for $\Delta\Phi$, 
-meaning that the interpolation accuracy requirements are at least $\sim 2$ orders of magnitude less for $\Delta\Phi$.}
-\label{fig:phaserange}
-\end{figure}
+%
+%\begin{figure}[htb]
+%\centering
+%\includegraphics[width=0.49\textwidth]{phaserange.png}
+%\caption{Range of phase values for training set waveforms. The range is a factor of $\gtrsim 100$ larger for $\Phi$ than for $\Delta\Phi$, 
+%meaning that the interpolation accuracy requirements are at least $\sim 2$ orders of magnitude less for $\Delta\Phi$.}
+%\label{fig:phaserange}
+%\end{figure}
 
@@
 
-\subsection{Preconditioning the training set waveforms}
-
-We evaluate the EOB waveform with a starting frequency equivalent to 20Hz for a total mass of $M=2M_\odot$ ($Mfwi=0.000197$).
-In order to take a discrete Fourier transform, we window the start of the waveform with a Planck window [[cite]] in the interval $(Mfwi, Mfwf) = (0.000197, 0.00021)$ to avoid Gibbs oscillations.
-We then resample the waveform with a spacing $\Delta t/M = 5$, and pad the end of the waveform with zeros such that all waveforms in the training set have the exact same time samples.
-
-After evaluating the discrete Fourier transform, evaluate the residuals $\Delta\ln(A)$ and $\Delta\Phi$ between the EOB and TaylorF2 waveforms.
-Because waveforms have free time and phase parameters, subtract a linear fit to $\Delta\Phi$ at the beginning of the waveform in the window $(Mf_{fit,i}, Mf_{fit,f}) = 0.00021(1, 1.05)$.
-The resulting $\Delta\Phi$ is zero and has zero slope at $Mf_{fit,i}$, guaranteeing that the surrogate smoothly matches to TaylorF2 at lower frequencies.
-
-As a result of Gibbs oscillations in the Fourier transformed waveform that weren't completely removed by the Planck waveform as well as from the end of the waveform where the amplitude rapidly drops to zero amplitude after the peak amplitude over $\sim 1$ cycle, small amplitude, high frequency oscillations remain in the residuals $\Delta\ln(A)$ and $\Delta\Phi$.
-We remove these oscillations using a moving average filter centered on $Mf$ with width $[Mf(1-\Delta Mf), Mf(1+\Delta Mf)]$.
-We use a width of $\Delta Mf=0.1$ for the amplitude and $\Delta Mf=0.05$ for the phase.
-
-Finally, we truncate the data to the interval $(Mf_{trunc,i}, Mf_{trunc,f}) = (0.00021, 0.07)$.
-We note that the gravitational wave frequency at the Schwarzschild ISCO is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$.
-Although second generation detectors are only marginally sensitive to signals above this frequency, we use the high-frequency cutoff of $Mf_{trunc,f}=0.07$ because some of the EOB waveforms still have enough amplitude at these high frequencies to affect the structure of the surrogate waveform if inverse Fourier transformed back into the time domain.
+\subsection{Conditioning the training set waveforms}
+
+The accuracy of the final frequency-domain surrogate depends on how well the finite-length, numerical waveform is Fourier transformed and filtered to remove numerical artifacts.
+We now describe the procedure to condition the training-set waveforms used to construct the surrogate.
+
+We evaluate the EOB waveform with a starting frequency of $Mf_{{\rm win}, i}=0.000197$, equivalent to a physical frequency of 20~Hz for a binary with total mass $M=2M_\odot$.
+In order to take a discrete Fourier transform, we window the start of the waveform with a Planck window [[cite]] in the interval $[Mf_{{\rm win}, i}, Mf_{{\rm win}, f}] = [0.000197, 0.00021]$ to avoid Gibbs oscillations.
+The end of the waveform has zero amplitude, so the end does not need to be windowed.
+We then resample the waveform with a spacing $\Delta t/M = 5$ and pad the end of the waveform with zeros such that all waveforms in the training set have the exact same time samples.
+After evaluating the discrete Fourier transform, we calculate the residuals $\Delta\ln(A)(Mf)$ and $\Delta\Phi(Mf)$ between the EOB and TaylorF2 waveforms using Eqs.~\eqref{eq:ampres} and~\eqref{eq:phaseres}.
+
+Waveforms have free time and phase parameters $t_c$ and $\phi_c$, and in the frequency-domain this means that one can freely add a linear term $\phi_c + 2\pi (Mf) (t_c/M)$ to the phase $\Phi(Mf)$.
+We use this freedom to match the EOB waveform to the analytic TaylorF2 waveform near the starting frequency.
+We do this by subtracting a linear fit to $\Delta\Phi(Mf)$ at the beginning of the waveform in the window $[Mf_{{\rm fit}, i}, Mf_{{\rm fit}, f}] = 0.00021[1, 1.05]$.
+At $Mf=Mf_{{\rm fit}, i}$, the resulting phase residual $\Delta\Phi(Mf)$ is zero and has zero slope, guaranteeing that the surrogate smoothly matches to TaylorF2 below this frequency.
+
+The resulting waveforms still have some remaining Gibbs oscillations which can be seen as small-amplitude, high-frequency oscillations in the residuals $\Delta\ln(A)(Mf)$ and $\Delta\Phi(Mf)$.
+These come from two sources.
+The first is the fact that the Planck window at the beginning of the waveform was not sufficiently long.
+We could make this window longer, but that would require us to start the surrogate at a higher frequency.
+The second source comes from the end of the waveform where the amplitude rapidly drops to zero amplitude after the peak amplitude in $\sim 1$ cycle.
+The Gibbs oscillations at high frequencies are therefore a genuine feature of the EOB model.
+In an attempt to preserve all structure in the EOB waveform, we choose not to window the end of the time-domain waveform.
+We instead remove these oscillations in the frequency domain using a moving average filter centered on $Mf$ with width $[Mf(1-\Delta Mf), Mf(1+\Delta Mf)]$.
+We use a width of $\Delta Mf=0.1$ for the amplitude residual and $\Delta Mf=0.05$ for the phase residual.
+Smoothing these oscillations makes it significantly easier to interpolate the amplitude and phase residuals as functions of $Mf$ and $\bx$.
+
+Finally, we truncate the residuals outside the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}] = [0.00021, 0.07]$.
+We note that the gravitational-wave frequency at the Schwarzschild ISCO is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$.
+Although second generation detectors are only marginally sensitive to BNS signals above this frequency, we use the high-frequency cutoff of $Mf_{trunc,f}=0.07$ because the amplitude of some of the EOB waveforms is still large enough at these high frequencies to affect the structure of the surrogate waveform if inverse Fourier transformed back into the time domain.
 
@@
 
-We now seek to interpolate the conditioned and filtered residuals $\Delta\ln(A)(f; \bx)$ and $\Delta\Phi(f; \bx)$.
-We begin by interpolating as a function of $f$ for fixed $\bx$.
-The majority of previous papers on gravitational waveform surrogates have used an orthonormal basis of global functions $\hat e_i(f)$ for frequency (or time) interpolation [[cite]].
-This type of surrogate is referred to as a reduced order model.
-For a generic function $g(f; \bx)$, this can be written as $g(f; \bx) \approx \sum_{i=1}^N c_i(\bx) \hat e_i(f)$.
+We now seek to interpolate the conditioned residuals $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$.
+We begin by choosing a method for interpolating as a function of $Mf$ for fixed $\bx$.
+The majority of previous papers on gravitational-wave surrogates have used an orthonormal basis of global functions $\hat e_i(Mf)$ for interpolating as a function of frequency (or time) [[cite]].
+This type of surrogate, built from a reduced basis, is usually referred to as a reduced order model.
+For a generic function $g(Mf; \bx)$, this decomposition can be written as $g(Mf; \bx) \approx \sum_{i=1}^N c_i(\bx) \hat e_i(Mf)$.
 The coefficients $c_i(\bx)$ are then interpolated as functions of $\bx$ [[Puerrer]].
-Alternatively, using the empirical interpolation method [[cite]], one can re-express these basis functions $\hat e_i(f)$ in terms of empirical interpolating functions $B_j(f)$ and the value of $g(f; \bx)$ at empirical nodes $F_j$ as $g(f; \bx) \approx \sum_{j=1}^N B_j(f) g(F_j; \bx)$.
-The location of these nodes $F_j$ can be optimized to minimize interpolation errors.
+Alternatively, using the empirical interpolation method [[cite]], one can re-express these basis functions $\hat e_i(Mf)$ in terms of empirical interpolating functions $B_j(Mf)$ and the value of $g(Mf; \bx)$ at empirical nodes $MF_j$ as $g(Mf; \bx) \approx \sum_{j=1}^N B_j(Mf) g(MF_j; \bx)$.
+The location of these nodes $MF_j$ is then optimized to minimize interpolation errors.
 This method has been used in Refs.~[[cite]].
@@
 The residuals are small and smooth at low frequencies and large and noisy at high frequencies.
-Errors in evaluating $g(F_j;\bx)$ for high frequency can propagate to large errors at low frequencies.
+Errors in evaluating $g(MF_j;\bx)$ for high frequency can propagate to large errors at low frequencies.
 Instead, we find that spline interpolation works significantly better.
-We use 20 frequency nodes $F_j$ log-spaced the interval $[f_{trunc,i}, f_{trunc,f}]$, and evaluate the residuals $\Delta\ln(A)(F_j; \bx)$ and $\Delta\Phi(F_j; \bx)$ at these nodes using GPR discussed below.
-We then choose third-order spline interpolation to interpolate between these points.
-The local, third-order polynomials, that are only connected by the requirement of smoothness [[check this]] do not propagate high-frequency errors down to low-frequency errors as significantly.
+We use 20 frequency nodes $MF_j$ log-spaced in the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}]$.
+We evaluate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at these nodes as discussed below.
+We then interpolate between these frequencies using third-order splines.
+The local, third-order polynomials, that are only connected by the requirement of smoothness [[check this]] do not propagate high-frequency errors down to low-frequency errors as significantly as do global basis functions.
 
@@
 
-We now need a method to interpolate the residuals $\Delta\ln(A)(F_j; \bx)$ and $\Delta\Phi(F_j; \bx)$ at the frequency nodes $F_j$ as a function of the five waveform parameters $\bx$.
+Next we choose a method to interpolate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at each of the frequency nodes $MF_j$ as a function of the five waveform parameters $\bx$.
 Most multivariate interpolation techniques (e.g.
-tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality; the number of samples grows exponentially with the dimension $d$ ($N^d$ for $N$ samples per dimension).
+tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality; the number of samples grows exponentially with the dimension $d$ ($N^d$ samples for $N$ points per dimension).
 For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
-At 10 minutes--1 hour per EOB waveform on a standard CPU, this is at the limit of what is reasonable.
+At $\sim10$~minutes--1~hour per EOB waveform on a standard CPU, this is at the limit of what is reasonable.
 However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs.
@@
 One of the main aims of this paper is to build a surrogate with as few waveform evaluations as possible.
-EOB models are expensive, and we would like a method that extends to higher dimensions and more expensive simulations such as numerical relativity simulations.
+EOB models are expensive, and we would like a method that extends to higher dimensions and more expensive waveform models such as from numerical relativity simulations.
 Using the freedom provided by GPR to sample waveforms at arbitrary locations in parameter space, we try a set of more optimal designs.
@@
 
-With this initial surrogate model, we can choose new training set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion.
-The quantity that we use is the root-mean-square error in the phase at frequencies below $Mf=0.03$:
-\begin{equation}
-\epsilon_{\rm RMS} = \sqrt{\frac{1}{N} \sum_i (\sigma_{\Delta\Phi_i})^2},
+With this initial surrogate model, we can choose new training-set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion.
+The quantity that we use is the root-mean-squared error in the phase at the frequency nodes $Mf_j$ below $Mf=0.03$:
+\begin{equation}
+\epsilon_{\rm RMS} = \sqrt{\frac{1}{N} \sum_j (\sigma_{\Delta\Phi_j})^2},
 \label{eq:errorrms}
 \end{equation}
-where $\sigma_{\Delta\Phi_i}$ is the GPR estimate of the uncertainty in $\Delta\Phi(MF_i, \bx)$.
+where $\sigma_{\Delta\Phi_j}$ is the GPR estimate of the uncertainty in $\Delta\Phi(MF_j, \bx)$.
 [[justify the definition of error]].
 For an initial training set with $N$ samples, this function has $\sim N$ local maxima located in the voids between samples $\bx$.
-To find the global maximum, we use a basin jumping method [[cite]] to avoid getting stuck in local maxima.
+To find the global maximum, we use a basin jumping algorithm [[cite]] to avoid getting stuck in local maxima.
 
@@
 Reconstructing the surrogate with each new EOB waveform allows us to retune the hyperparameters $\btheta$ with each new waveform.
-However, if we hold the hyperparameters $\btheta$ fixed, we note that the GPR error estimate (Eq.~\eqref{eq:var}) does only depends on the samples $\bx$ and not on the data $\by$.
-Therefore, the approximate $\epsilon_{\rm rms}$ (Eq.~\eqref{eq:distinguish}) only depends on the parameters of the new waveform, not on the waveform data.
-The method uses the idea of uncertainty sampling [[cite]] but without updating the hyperparameters after each new point is chosen.
+However, if we hold the hyperparameters $\btheta$ fixed, we note that the GPR error estimate (Eq.~\eqref{eq:var}) only depends on the samples $\bx$ and not on the data $\by$.
+Therefore, the error $\epsilon_{\rm rms}$ (Eq.~\eqref{eq:errorrms}) only depends on the parameters of the new waveform, not on the waveform data.
+This method is sometimes referred to as uncertainty sampling [[cite]].
+However in our case, we do not update the hyperparameters after each new point is chosen, allowing us to choose the points then evaluate the waveforms in parallel.
 The algorithm for choosing the $N_{\rm new}$ new points is as follows.
@@
 Fig.~\ref{fig:LHD} shows the parameters of the 200 new waveforms chosen by the uncertainty sampling method.
-We note that the edges faces of the parameter space are more often chosen than the inner region.
+We note that the edges and faces of the parameter space are more often chosen than the inner region.
 This is likely a consequence of the Mat\'{e}rn GPR kernel used.
@@
 Fig.~\ref{fig:uncsamp} shows the RMS error $\epsilon_{\rm rms}$ maximized over the parameters $\bx$ for each new sample added to the training set.
+\red{[[Describe this figure...]]}
 
@@
 
-%\begin{figure}[htb]
-%\centering
-%\includegraphics[width=0.49\textwidth]{mismatchtrianglem10.png}
-%\caption{Mismatch for systems with $m_2=1M_\odot$.}
-%\label{fig:mismatchtriangle}
-%\end{figure}
-
-%--Fisher matrix systematic error. (See spin NSBH paper.)
-%
-%--Phase error. (needs to be significantly smaller than tidal effects) Use this for order of magnitude estimate of required
-%accuracy.
-%
-%\begin{figure}[htb]
-%\centering
-%\includegraphics[width=0.49\textwidth]{maxamphist.png}
-%\includegraphics[width=0.49\textwidth]{maxphasehist.png}
-%\caption{Amplitude and phase error.}
-%\label{fig:ampphaseerr}
-%\end{figure}
-%
-
-%--distinguishability
-%
-%--Fisher matrix systematic error.
-
 \subsection{Timing}
@@
 
+%% ________________________________________________________
 \section{Discussion and future work}
@@
 \end{document}
+
+%As discussed above in Sec.~II, the current implementation of the EOB model takes at least 10~minutes to evaluate.
+In this section we will describe in detail the method for training and evaluating our surrogate model.
+The first decision in building a waveform surrogate is choosing between building a surrogate of the time-domain waveform $h(t;\bx)$ or the frequency-domain waveform $\tilde h(f;\bx)$.
+A time-domain surrogate has the advantage that significantly less processing of the original waveform is needed, resulting in a surrogate that typically more accurately reproduces the original waveform model.
+A frequency-domain surrogate, on the other hand, requires windowing of the Fourier transformed waveform to avoid Gibbs oscillations.
+For the beginning of the waveform, one can simply begin at a lower frequency, window the waveform, then truncate the Fourier-transformed waveform above some desired starting frequency.
+At the end of the waveform, the BNS amplitude drops rapidly over $\sim 1$~GW cycle, resulting in Gibbs oscillations at the high-frequency end.
+One could window the end of the waveform over several cycles, however, this would remove available information about tidal interactions.
+As we will show in Sec.~[[]], these high-frequency Gibbs oscillations in the amplitude of the Fourier transformed waveform are a genuine feature of BNS waveforms unlike BBH waveforms which typically taper over a larger number of GW cycles during their ringdown.
+These oscillations make it difficult to construct a surrogate.
+However, we will show that smoothing these oscillations in the Fourier transformed waveform with a moving average filter leads to a surrogate that is sufficiently faithful to the original time-domain model.
+Another important advantage is that a surrogate of $\tilde h(f;\bx)$ bypasses the need to evaluate the Fourier transform at each waveform evaluation for parameter estimation.
+And, as we shall see, a frequency-domain surrogate can be trivially attached to analytic models that are accurate at low-frequencies, such as TaylorF2, making analysis for third-generation detectors which are sensitive at lower frequencies easier.
+Finally, methods for speeding up the likelihood evaluation in parameter estimation such as reduced order quadrature typically require a fast frequency-domain model for calculating the appropriate weights online [[cite]].
+For these reasons we will construct a surrogate of $\tilde h(f;\bx)$.
 

### 2017-11-22 paper/paper.tex
Changed Figs 1 and 2 and discussion. New surrogate with almost reviewed waveform model in progress.

@@
 
-In this section we describe a method for decomposing the waveform into smooth, slowly-varying functions and training a surrogate model for these functions.
-We will work with the frequency domain waveform $\tilde h(f; \bx)$.
+In this section we describe how we decompose the waveform into smooth, slowly-varying functions and train a surrogate model for these functions.
+We will work with the frequency-domain waveform $\tilde h(Mf; \bx)$.
 This is favorable for data analysis which is usually done in the frequency domain.
 It also allows us to extend the model down to arbitrarily low frequencies with analytic, Frequency-domain PN approximations.
-Unfortunately, Fourier transforming a finite length, discretely sampled waveform, leads to a surrogate with more noise.
-We will show below, however, that sufficient filtering can mostly solve this problem.
+Unfortunately, Fourier transforming a finite length, discretely sampled waveform leads to a surrogate with more noise.
+We will show below, however, that sufficient filtering can solve this problem.
 
@@
 
-Because the waveform $\tilde h(f; \bx)$ is an oscillatory function of $f$ and $\bx$, the waveform is usually decomposed into an amplitude $A(f; \bx)$ and phase $\Phi(f; \bx)$ as $\tilde h(f; \bx) = A(f; \bx) e^{i\Phi(f; \bx)}$.
-These functions are smoother, mostly monotonic functions of $f$ and $\bx$.
+%Because the waveform $\tilde h(Mf; \bx)$ is an oscillatory function of $Mf$ and $\bx$, the waveform is usually decomposed into an amplitude $A(Mf; \bx)$ and phase $\Phi(Mf; \bx)$ as $\tilde h(Mf; \bx) = A(Mf; \bx) e^{i\Phi(Mf; \bx)}$.
+These functions are smoother, mostly monotonic functions of $Mf$ and $\bx$.
 This can be seen in Fig.~\ref{fig:h} which shows the initial training set of waveforms.
@@
 
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.49\textwidth]{h_trainingset.png}
+%\begin{figure}[htb]
+%\centering
+%\includegraphics[width=0.49\textwidth]{h_trainingset.png}
+%%\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
+%\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
+%\label{fig:h}
+%\end{figure}
+
+%\begin{figure}[htb]
+%\centering
+%%\includegraphics[width=0.49\textwidth]{h_trainingset.png}\\
 %\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
-\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
-\label{fig:h}
-\end{figure}
-
-For aligned-spin waveforms, we can solve this problem using that fact that the waveform is approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2 [[cite]] $\tilde h_{\rm F2}(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2} (f; \bx)}$.
-This allows us to write the EOB amplitude and phase in terms of the small residuals, $\Delta\ln(A)(f; \bx)$ and $\Delta\Phi(f; \bx)$, relative to TaylorF2:
-\begin{align}
-\Delta\ln(A)(f; \bx) &= \ln\left(\frac{ A(f; \bx) }{ A_{\rm F2}(f; \bx)}\right), \label{eq:ampres}\\
-\Delta\Phi(f; \bx) & = \Phi(f; \bx) - \Phi_{\rm F2}(f; \bx), \label{eq:phaseres}
-\end{align}
-such that
-\begin{align}
-\tilde h(f; \bx) = \tilde h_{\rm F2}(f; \bx) e^{\Delta\ln(A)(f; \bx) + i  \Delta\Phi(f; \bx)}.
-\label{eq:hdecomp}
-\end{align}
-%\begin{align}
-%\tilde h(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2}(f; \bx)} e^{\Delta\ln(A)(f; \bx) + i  \Delta\Phi(f; \bx)}.
-%\label{eq:hdecomp}
-%\end{align}
-These residuals are shown in Fig.~\ref{fig:dh}.
-For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that the amplitude of the reconstructed waveform in Eq.~\eqref{eq:hdecomp} will be positive even if interpolation error leads to a negative $\Delta\ln(A)(f; \bx)$.
-In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency of each waveform, the log-ratio captures this behavior.
-Comparing Fig.~\ref{fig:dh} to Fig.~\ref{fig:h}, we find that for most of the frequency range, the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$.
-
-\begin{figure}[htb]
-\centering
-%\includegraphics[width=0.49\textwidth]{h_trainingset.png}\\
-\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
-\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
-\label{fig:dh}
-\end{figure}
+%\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
+%\label{fig:dh}
+%\end{figure}
 
@@
 
+Because the waveform $\tilde h(Mf; \bx)$ is an oscillatory function of $Mf$ and $\bx$, the waveform is usually decomposed into an amplitude $A(Mf; \bx)$ and phase $\Phi(Mf; \bx)$ as $\tilde h(Mf; \bx) = A(Mf; \bx) e^{i\Phi(Mf; \bx)}$.
+The amplitude and phase are smoother, mostly monotonic functions of frequency.
+This can be seen in Fig.~\ref{fig:hoff} where we show the waveforms for the 32 corners of the 5-dimensional parameter space.
+Unfortunately, the amplitude and phase still span a wide range of values.
+The phase, for example, spans about $10^4$~rad between waveforms with different parameters $\bx$ (see Fig.~\ref{fig:hoff}).
+To avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad.
+This is a difficult requirement to achieve in the 5-dimensional parameter space here.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.49\textwidth]{hoff.png}
+\caption{EOB waveforms $\tilde h(Mf; \bx)$ for the 32 corners of parameter space.
+The waveforms are filtered and Fourier transformed as described in Sec.~\ref{sec:condition}.
+Top two panels: the amplitude $A(Mf; \bx)$ and residual $\Delta\ln A(Mf; \bx)$ relative to TaylorF2 as defined in Eq.~\eqref{eq:ampres}.
+Bottom two panels: the phase $\Phi(Mf; \bx)$ and residual $\Delta\Phi(Mf; \bx)$ relative to TaylorF2 as defined in Eq.~\eqref{eq:phaseres}.
+Vertical dashed lines represent the frequency nodes $MF_j$ where the residuals are interpolated as functions of $\bx$ using Gaussian process regression.}
+\label{fig:hoff}
+\end{figure}
+
+For aligned-spin waveforms, we can solve this problem using the fact that the waveform is approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2 [[cite]] $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$.
+This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2:
+\begin{align}
+\Delta\ln(A)(Mf; \bx) &= \ln\left(\frac{ A(Mf; \bx) }{ A_{\rm F2}(Mf; \bx)}\right), \label{eq:ampres}\\
+\Delta\Phi(Mf; \bx) & = \Phi(Mf; \bx) - \Phi_{\rm F2}(Mf; \bx), \label{eq:phaseres}
+\end{align}
+such that
+\begin{align}
+\tilde h(Mf; \bx) = \tilde h_{\rm F2}(Mf; \bx) e^{\Delta\ln(A)(Mf; \bx) + i  \Delta\Phi(Mf; \bx)}.
+\label{eq:hdecomp}
+\end{align}
+These residuals are also shown in Fig.~\ref{fig:hoff}.
+For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that the amplitude of the reconstructed waveform (Eq.~\eqref{eq:hdecomp}) will be positive even if interpolation error leads to a negative value of $\Delta\ln(A)(Mf; \bx)$.
+In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency of each waveform, the log-ratio captures this behavior.
+Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies.
+In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $S_1$.
+These quantities are smooth functions of $\bx$, except for the amplitude residual at very high frequencies where the amplitude is small and numerical noise dominates, implying that interpolation as functions of $\bx$ should work.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.49\textwidth]{dhofs.png}
+\caption{Cross sections of the amplitude and phase residuals as functions of $S_1$ for the fixed frequencies $Mf=\{0.01, 0.03, 0.05\}$.
+The waveform parameters are $\bx = \{q, S_1, S_2, \Lambda_1, \Lambda_2\} = \{0.6, S_1, 0.2, 2000, 1000\}$.
+The waveforms are generated for 51 values of $S_1$ shown as dots.
+Also shown is the 1-dimensional cross section of the final surrogate using Gaussian process regression (solid black) as well as its estimated $1\sigma$ uncertainty (shaded gray region).}
+\label{fig:dhofs}
+\end{figure}
+
 The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
@@
 \subsection{Conditioning the training set waveforms}
+\label{sec:condition}
 
@@
 
-\red{[[Show plot (?) to validate that waveform is properly conditioned by plotting amplitude/phase as function of waveform parameters at specific frequencies and show that they're smooth functions?]]}
+To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of the waveform parameter $S_1$.
+The fact that the residuals are smooth functions of the waveform parameters indicates that most numerical noise has been removed.
+The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.05; \bx)$, where the amplitude is very small and dominated by numerical noise.
 

### 2017-11-24 paper/paper.tex
More paper edits.

@@
 \centering
-\includegraphics[width=0.49\textwidth]{hoff.png}
+\includegraphics[width=0.49\textwidth]{hoff.pdf}
 \caption{EOB waveforms $\tilde h(Mf; \bx)$ for the 32 corners of parameter space.
@@
 Bottom two panels: the phase $\Phi(Mf; \bx)$ and residual $\Delta\Phi(Mf; \bx)$ relative to TaylorF2 as defined in Eq.~\eqref{eq:phaseres}.
-Vertical dashed lines represent the frequency nodes $MF_j$ where the residuals are interpolated as functions of $\bx$ using Gaussian process regression.}
+Vertical dashed lines represent the frequency nodes $MF_j$ where the residuals are interpolated as functions of $\bx$ using Gaussian process regression.
+The vertical solid line at $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$ is the gravitational-wave frequency at the Schwarzschild ISCO.}
 \label{fig:hoff}
@@
 For aligned-spin waveforms, we can solve this problem using the fact that the waveform is approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2 [[cite]] $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$.
-This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2:
+This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2
 \begin{align}
@@
 These residuals are also shown in Fig.~\ref{fig:hoff}.
-For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that the amplitude of the reconstructed waveform (Eq.~\eqref{eq:hdecomp}) will be positive even if interpolation error leads to a negative value of $\Delta\ln(A)(Mf; \bx)$.
-In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency of each waveform, the log-ratio captures this behavior.
+For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that interpolation errors will not lead to a negative amplitude for the reconstructed waveform (Eq.~\eqref{eq:hdecomp}).
+In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency, the log-ratio captures this behavior.
 Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies.
 In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $S_1$.
-These quantities are smooth functions of $\bx$, except for the amplitude residual at very high frequencies where the amplitude is small and numerical noise dominates, implying that interpolation as functions of $\bx$ should work.
+These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work.
+The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise dominates.
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{dhofs.png}
-\caption{Cross sections of the amplitude and phase residuals as functions of $S_1$ for the fixed frequencies $Mf=\{0.01, 0.03, 0.05\}$.
-The waveform parameters are $\bx = \{q, S_1, S_2, \Lambda_1, \Lambda_2\} = \{0.6, S_1, 0.2, 2000, 1000\}$.
-The waveforms are generated for 51 values of $S_1$ shown as dots.
-Also shown is the 1-dimensional cross section of the final surrogate using Gaussian process regression (solid black) as well as its estimated $1\sigma$ uncertainty (shaded gray region).}
+\includegraphics[width=0.49\textwidth]{dhofs.pdf}
+\caption{Cross sections of the amplitude and phase residuals as functions of $S_1$ at three nodes $MF_j$.
+The waveforms are generated for 51 values of $S_1$ shown as dots with the other parameters held constant at $\{q, S_2, \Lambda_1, \Lambda_2\} = \{0.6, 0.2, 2000, 1000\}$.
+Also shown are 1-dimensional cross sections of the interpolated residuals using Gaussian process regression (solid black) as well as their estimated $1\sigma$ uncertainty (shaded gray region).}
 \label{fig:dhofs}
@@
 We note that the gravitational-wave frequency at the Schwarzschild ISCO is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$.
-Although second generation detectors are only marginally sensitive to BNS signals above this frequency, we use the high-frequency cutoff of $Mf_{trunc,f}=0.07$ because the amplitude of some of the EOB waveforms is still large enough at these high frequencies to affect the structure of the surrogate waveform if inverse Fourier transformed back into the time domain.
-
-To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of the waveform parameter $S_1$.
+Although second generation detectors are only marginally sensitive to BNS signals above this frequency, we use the high-frequency cutoff of $Mf_{{\rm trunc}, f}=0.07$ because the amplitude of some of the EOB waveforms is still large enough at these high frequencies to affect the structure of the surrogate waveform if inverse Fourier transformed back into the time domain.
+
+To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of one of the waveform parameters $S_1$.
 The fact that the residuals are smooth functions of the waveform parameters indicates that most numerical noise has been removed.
@@
 The residuals are small and smooth at low frequencies and large and noisy at high frequencies.
-Errors in evaluating $g(MF_j;\bx)$ for high frequency can propagate to large errors at low frequencies.
+With global basis functions, errors in evaluating $g(MF_j;\bx)$ at high frequencies can propagate to large errors between the nodes $MF_j$ at low frequencies.
 Instead, we find that spline interpolation works significantly better.
-We use 20 frequency nodes $MF_j$ log-spaced in the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}]$.
+We use 40 frequency nodes $MF_j$ log-spaced in the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}]$ (see Fig.~\ref{fig:hoff}).
 We evaluate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at these nodes as discussed below.
 We then interpolate between these frequencies using third-order splines.
-The local, third-order polynomials, that are only connected by the requirement of smoothness [[check this]] do not propagate high-frequency errors down to low-frequency errors as significantly as do global basis functions.
+The local, third-order polynomials, that are only connected by the requirement of smoothness do not propagate high-frequency errors down to low-frequency errors as significantly as do global basis functions.
 
 \subsection{Gaussian process regression for parameters $\bx$}
+
+\red{[[Make sure all vectors and matrices are bold.]]}
 
@@
 For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
-At $\sim10$~minutes--1~hour per EOB waveform on a standard CPU, this is at the limit of what is reasonable.
-However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs.
+At $\sim10$~minutes--1~day per EOB waveform on a standard CPU, this is at the limit of what is reasonable.
+However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below.
 The method we choose is Gaussian Process Regression (GPR).
 
-In GPR a function $f(\vec x)$ is described in terms of a mean $m(\vec x)$ and  covariance $k(\vec x, \vec x')$ between points [[cite RassmussenWilliamsChapter2, IntroArxivPaper]].
-\begin{equation}
-f(\vec x) \approx \mathcal{GP}(m(\vec x), k(\vec x, \vec x')).
-\end{equation}
-The mean $m(x)$ can be some parameterized function which is the standard version of regression, while $k(\vec x, \vec x')$ is a kernel with tunable hyperparameters that describes the covariance between the point $\vec x$ and the sampled point $\vec x'$.
-This covariance can be used to describe features of the function not captured by the parameterized mean as well as genuine noise in the data ${\bm y}$.
-For the problem here, we have already subtracted the TaylorF2 waveform from EOB waveform, so we set $m=0$ and model the remainder terms purely in terms of the covariance $k$.
-
-In a zero-mean Gaussian process, the data and function $({\bm y}, f_*)$ at the training set and new points $({\bm x}, x_*)$ are drawn from a multivariate normal distribution
+In GPR a function $g(\bx)$ is described in terms of a mean $m(\bx)$ and  covariance $k(\bx, \bx')$ between points [[cite RassmussenWilliamsChapter2, IntroArxivPaper]].
+\begin{equation}
+g(\bx) \approx \mathcal{GP}(m(\bx), k(\bx, \bx')).
+\end{equation}
+The mean $m(\bx)$ can be a parameterized fitting function while $k(\bx, \bx')$ is a kernel with tunable hyperparameters that describes the covariance between the point $\bx$ and the sampled point $\bx'$.
+This covariance can be used to describe features of the function not captured by the parameterized mean $m(\bx)$ as well as genuine noise in the data $\by$.
+For the problem here, we have already subtracted the TaylorF2 waveform from EOB waveform, so we set $m(\bx) = 0$ and model the residuals purely in terms of the kernel $k(\bx, \bx')$.
+
+In a zero-mean Gaussian process, the data $\by_i$ and function value $\by_*$ at the training set points $\bx_i$ and new point $\bx_*$, respectively, are drawn from a multivariate normal distribution
 \begin{equation}
@@
 {\bm y} \\
-f_* \\
+y_* \\
 \end{bmatrix}
@@
 \end{equation}
-where $K = K({\bm x}^i, {\bm x}^j)$ is a matrix, $K_* = K({\bm x}^i, {\bm x}_*)$ is a vector, and $K_{**} = K({\bm x}_*, {\bm x}_*)$ is a scalar.
-
-The conditional probability for $f_*$ given the training set examples ${\bm y}$ with hyperparameters $\theta$ is also a Gaussian
-\begin{equation}
-f_* | {\bm x}, x_*, {\bm y}, {\bm \theta} \sim \mathcal{N}(\bar f_*, {\rm var}(f_*))
+where $K_{ij} = k(\bx_i, \bx_j)$ is a matrix, $K_{*i} = k(\bx_i, \bx_*)$ is a vector, and $K_{**} = k(\bx_*, \bx_*)$ is a scalar.
+
+The conditional probability for $y_*$ given the training set examples $\by$ and kernel hyperparameters $\btheta$ is also a Gaussian
+\begin{equation}
+p(y_* | \bx_i, \bx_*, \by, \btheta) \sim \mathcal{N}(\bar y_*, {\rm Var}(y_*)),
 \end{equation}
@@
 \label{eq:mean}
-\bar f_* &= K_*^T K^{-1} y \\
+\bar y_* &= K_*^T K^{-1} y, \\
 \label{eq:var}
-{\rm var}(f_*) &= K_{**} - K_*^T K^{-1} K_*.
+{\rm Var}(y_*) &= K_{**} - K_*^T K^{-1} K_*.
 \end{align}
-Eq.~\eqref{eq:mean} is the estimate of the function, and Eq.~\eqref{eq:var} is the estimate of the uncertainty, where we note that the variance does not depend on the data $\bm{y}$.
+Eq.~\eqref{eq:mean} is the estimate of the function, and Eq.~\eqref{eq:var} is the estimate of the uncertainty. We note that the variance does not depend on the data $\by$.
 
@@
 \begin{equation}
-r^2 = (x-x')^T M (x-x'),
-\end{equation}
-where we choose the matrix M to be diagonal
+r^2 = (\bx - \bx')^T M (\bx - \bx'),
+\end{equation}
+where we choose the matrix $M$ to be diagonal
 \begin{equation}
@@
 \end{equation}
-The tunable hyperparameters $\ell_i$ represent the length scale over which the function $f(x)$ varies in 
-each coordinate $x_i$. Specifically, we use the Mat\'{e}rn class of kernels
+The tunable hyperparameters $\ell_i$ represent the length scale over which the function $g(\bx)$ varies in each coordinate $x_i$. Specifically, we use the Mat\'{e}rn class of kernels
 \begin{equation}
@@
 \end{equation}
-where $K_\nu(x)$ is a modified Bessel function. The value of $\nu$ parameterizes the smoothness of the 
-Gaussian Process $f(x)$, and is $k$ times mean-square differentiable if $\nu>k$ [[cite]]. For half-integer values of $\nu$, this 
-kernel has a computationally cheap form without special functions, and we have had good results with $\nu=5/2$, 
-resulting in a twice-differentiable function. The $\nu=5/2$ kernel is
+where $K_\nu(x)$ is a modified Bessel function.
+The value of $\nu$ parameterizes the smoothness of the Gaussian process, and is $k$ times mean-square differentiable if $\nu>k$ [[cite]].
+For half-integer values of $\nu$, this kernel has a computationally cheap form without special functions, and we have had good results with $\nu=5/2$, resulting in a twice-differentiable function\footnote{If desired, one can select the optimal class of kernels using cross validation, and we have also tried the more common, infinitely differentiable squared exponential kernel [[cite]].
+However, we found it difficult to reliably tune its hyperparameters for the problem here.}.
+The $\nu=5/2$ kernel is
 \begin{equation}
@@
 \begin{equation}
-k(x_p, x_q) = \sigma_f^2 k_{\rm Matern}^{\nu=5/2}(r) + \sigma_n^2 \delta_{pq}
-\end{equation}
-where $\sigma_f$ is a scale factor that describes the range of values that $f(x)$ takes over the domain, 
-and $\sigma_n$ is a noise parameter. The white noise kernel $\sigma_n^2 \delta_{pq}$ (also called a nugget), which is
-$\sigma_n^2$ when $x_p=x_q$ and zero otherwise, is generically used for numerical stability, but also parameterizes
-noise in the data ${\bm y}$. In our case, the training set waveforms which are evaluated, tapered, and 
-Fourier transformed numerically have numerical noise that we will estimate by optimizing the hyperparameters.
-The hyperparameters now are $\theta = \{\sigma_f, \ell_q, \ell_{S_1}, \ell_{S_2}, \ell_{\Lambda_1}, \ell_{\Lambda_2}, \sigma_n\}$.
-
-[[In principle, one can select the optimal class of kernels using k-fold or leave-one-out cross validation. We have also tried the 
-more common infinitely differentiable squared exponential kernel. However, we found it difficult to reliably tune the 
-hyperparameters for the problem here.]]
+k(\bx_i, \bx_j) = \sigma_f^2 k_{\rm Matern}^{\nu=5/2}(r) + \sigma_n^2 \delta_{ij}
+\end{equation}
+where $\sigma_f$ is a scale factor that describes the range of values that $g(\bx)$ takes over the domain, and $\sigma_n$ is a noise parameter.
+The white noise kernel $\sigma_n^2 \delta_{ij}$ (also called a nugget) is $\sigma_n^2$ when $\bx_i = \bx_j$ and zero otherwise, and parameterizes noise in the data $\by$.
+In our case, the training set waveforms have numerical noise that we will estimate by optimizing the hyperparameters.
+The full set of hyperparameters is now  $\btheta = \{\sigma_f, \ell_q, \ell_{S_1}, \ell_{S_2}, \ell_{\Lambda_1}, \ell_{\Lambda_2}, \sigma_n\}$.
 
@@
 \end{equation}
-The prior $p({\bm \theta})$ is typically uniform and used to set the bounds on ${\bm \theta}$. If interested in the distribution of the 
-hyperparamaters, one can sample this posterior with, for example, Markov chain Monte Carlo. However, for the problem here, we are
-simply interested in finding the maximum of the posterior. To do this, we use the \texttt{gaussian\_process} module in the
-\texttt{scikit-learn} package [[cite]]. Finally, with the optimized hyperparameters ${\bm \theta}$, Eq.~\eqref{eq:var} is the 
-interpolating function. We note that evaluating Eq.~\eqref{eq:var}, is an $\mathcal{O}(N)$ operation if the vector $K^{-1}{\bm y}$ is
-precomputed.
+The prior $p({\bm \theta})$ is typically uniform and used to set the bounds on ${\bm \theta}$.
+One can sample this posterior if interested in the distribution of hyperparameters $\btheta$.
+However, for the problem here, we simply want the maximum posterior.
+We do this using the \texttt{gaussian\_process} module in the \texttt{scikit-learn} package [[cite]].
+With these optimized hyperparameters, the final interpolating function is given by Eq.~\eqref{eq:mean} and its uncertainty by Eq.~\eqref{eq:var}.
+We note that once the vector $K^{-1}y$ is precomputed, evaluating Eq.~\eqref{eq:mean} is a fast $\mathcal{O}(N)$ operation.
 
 \subsection{Iterative construction of training set and surrogate}
-
-One of the main aims of this paper is to build a surrogate with as few waveform evaluations as possible.
+\label{sec:design}
+
+%One of the main aims of this paper is to build a surrogate with as few waveform evaluations as possible.
 EOB models are expensive, and we would like a method that extends to higher dimensions and more expensive waveform models such as from numerical relativity simulations.
@@
 
+One of the main aims of this paper is to build a surrogate with as few waveform evaluations as possible.
+Using the freedom provided by GPR to sample waveforms at arbitrary locations, we try a set of designs more efficient than uniform grids.
 One such design is a Latin Hypercube Design (LHD) [[cite]].
-An LHD with $N$ samples divides each of $d$ dimensions uniformly into $N$ grid points, for a total of $N^d$ grid points.
-However, for each dimension, the $N$ values are sampled exactly once.
-For an LHD there are $(N!)^d$ possible designs, and we choose one randomly\footnote{There are ways to better choose an LHD.
+An LHD with $N$ samples divides each of $d$ dimensions uniformly into $N$ grid points for a total of $N^d$ grid points.
+However, unlike a uniform grid, the $N$ values in each dimension are sampled exactly once instead of $N^{d-1}$ times.
+For an LHD there are $(N!)^d$ ways to choose these points, and we choose one randomly\footnote{There are ways to better choose an LHD.
 A standard requirement is that the LHD be space-filling, meaning that the points are as far apart as possible from each other.
 (See [[cite]] for a review of methods for optimizing the placement of samples.) One such definition of space filling is that the chosen locations maximize the minimum Euclidean distance between any two samples [[cite]].
-We have not studied this here.}.
-An LHD has the property that the samples are non-collapsing; a projection onto a subspace is still an LHD.
-This avoids wasting samples when one of the parameters has much less of an effect than the other parameters.
-As an example of the usefulness of non-collapsing designs, the tidal parameters have a minimal impact on the waveform at low frequencies, so it wouldn't make sense to densely sample a grid of tidal parameters at low frequencies.
+We have not experimented with these alternatives here.}.
+
+An LHD has the property that the samples are non-collapsing; a projection onto a subspace is still an LHD and no points are repeated in any dimension.
+This avoids wasting samples when one of the parameters has much less of an influence than the other parameters.
+As an example, the tidal parameters have a minimal impact on the waveform at low frequencies, so it would not make sense to densely sample a grid of tidal parameters at low frequencies.
 However, during the merger the tidal parameters can be more important than the other parameters.
@@
 We build an initial training set with 128 waveforms sampled with an LHD in our 5-dimensional parameter space given by $q\in[1/3, 1]$, $S_1, S_2 \in [-0.5, 0.5]$ and $\Lambda_1, \Lambda_2 \in [0, 5000]$.
-In addition, we find empirically that the GPR uncertainty estimates are largest at the corners of the parameter space, so we also sample the 32 corners.
+In addition, we find empirically that the GPR uncertainty estimates (Eq.~\eqref{eq:var} and Eq.~\eqref{eq:errorrms} below) are largest at the corners of the parameter space, so we also sample the 32 corners.
 We construct our initial surrogate with these 160 waveforms shown in Fig.~\ref{fig:LHD}.
@@
 \centering
-\includegraphics[width=0.49\textwidth]{trainingset3d.png}\\
-\includegraphics[width=0.49\textwidth]{trainingset2d.png}
+\includegraphics[width=0.49\textwidth]{trainingset3d.pdf}\\
+\includegraphics[width=0.49\textwidth]{trainingset2d.pdf}
 \caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, S_1, S_2\}$ (top) and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom).
-The 160 black circles were used for the initial training set (32 corner points + 128 LHD points) that generated the initial surrogate.
-The 200 red squares were generated using uncertainty sampling from the GPR error estimate of the RMS phase error at the interpolation nodes.
-The final 1000 waveforms are drawn uniformly from parameter space.
+The 160 blue circles were used for the initial training set (32 corners + 128 LHD points).
+The 200 red squares were generated using uncertainty sampling from the GPR error estimate (Eq.~\eqref{eq:errorrms}).
+The final 1000 green triangles were drawn uniformly from the parameter space.
 }
@@
 
-With this initial surrogate model, we can choose new training-set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion.
+%This method is sometimes referred to as uncertainty sampling [[cite]].
+However in our case, we do not update the hyperparameters after each new point is chosen, allowing us to choose the points then evaluate the waveforms in parallel.
+
+With this initial surrogate model, we can choose new training-set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion and update the surrogate with each new waveform.
+This method is sometimes referred to as uncertainty sampling [[cite]].
 The quantity that we use is the root-mean-squared error in the phase at the frequency nodes $Mf_j$ below $Mf=0.03$:
 \begin{equation}
-\epsilon_{\rm RMS} = \sqrt{\frac{1}{N} \sum_j (\sigma_{\Delta\Phi_j})^2},
+\epsilon_{\rm RMS}(\bx) = \sqrt{\frac{1}{N} \sum_{ \substack{j \\ MF_j \le 0.03} } [\sigma_{\Delta\Phi_j}(\bx)]^2},
+%\epsilon_{\rm RMS} = \sqrt{\frac{1}{N} \sum_{j=0}^{MF_j \le 0.03} (\sigma_{\Delta\Phi_j})^2},
 \label{eq:errorrms}
 \end{equation}
-where $\sigma_{\Delta\Phi_j}$ is the GPR estimate of the uncertainty in $\Delta\Phi(MF_j, \bx)$.
-[[justify the definition of error]].
-For an initial training set with $N$ samples, this function has $\sim N$ local maxima located in the voids between samples $\bx$.
+where $\sigma_{\Delta\Phi_j}(\bx)$ is the GPR estimate (Eq.~\eqref{eq:var}) of the uncertainty in $\Delta\Phi(MF_j, \bx)$.
+There are many possible ways to construct a scalar that estimates overall waveform error.
+An alternative would be an approximation to the mismatch discussed in Sec.~\ref{sec:accuracy}.
+However, the quantity $\epsilon_{\rm RMS}(\bx)$ describes the most important quantity, phase, and is fast to compute.
+For an initial training set with $N$ samples, $\epsilon_{\rm RMS}(\bx)$ has $\sim N$ local maxima located in the voids between samples $\bx$.
 To find the global maximum, we use a basin jumping algorithm [[cite]] to avoid getting stuck in local maxima.
 
-For low starting frequencies, reconstructing the surrogate for each new EOB waveform is smaller than the time needed to simulate a new EOB waveform.
-However, EOB waveforms are expensive enough that we would like a method to efficinetly choose $N_{\rm new}$ new waveforms in parallel.
-Reconstructing the surrogate with each new EOB waveform allows us to retune the hyperparameters $\btheta$ with each new waveform.
-However, if we hold the hyperparameters $\btheta$ fixed, we note that the GPR error estimate (Eq.~\eqref{eq:var}) only depends on the samples $\bx$ and not on the data $\by$.
-Therefore, the error $\epsilon_{\rm rms}$ (Eq.~\eqref{eq:errorrms}) only depends on the parameters of the new waveform, not on the waveform data.
-This method is sometimes referred to as uncertainty sampling [[cite]].
-However in our case, we do not update the hyperparameters after each new point is chosen, allowing us to choose the points then evaluate the waveforms in parallel.
+For low starting frequencies, EOB waveforms are expensive enough that we would like a method to efficiently choose $N_{\rm new}$ waveforms and evaluate them in parallel.
+We note that if we hold the hyperparameters $\btheta$ fixed, the GPR error estimate (Eq.~\eqref{eq:var}), and therefore $\epsilon_{\rm RMS}(\bx)$, only depends on the samples $\bx$ and not on the waveform data.
 The algorithm for choosing the $N_{\rm new}$ new points is as follows.
@@
 \end{enumerate}
-These new waveforms can now be evaluated in parallel. An updated surrogate can then be constructed with re-optimized hyperparameters $\btheta$ using the $N_{\rm old}+N_{\rm new}$ waveforms.
+These $N_{\rm new}$ waveforms can now be evaluated in parallel. An updated surrogate can then be constructed with re-optimized hyperparameters $\btheta$ using the $N + N_{\rm new}$ waveforms.
 
@@
 Fig.~\ref{fig:uncsamp} shows the RMS error $\epsilon_{\rm rms}$ maximized over the parameters $\bx$ for each new sample added to the training set.
-\red{[[Describe this figure...]]}
+Increasing the number of waveforms from 160 to 360 decreases the estimated RMS phase error by a factor of $\sim 4$.
 
@@
 \includegraphics[width=0.49\textwidth]{uncertaintysampling.png}
-\caption{Estimated RMS phase error $\epsilon_{\rm rms}$ as function of the number of sampled waveforms (blue curve).
-The first 160 waveforms are from the initial training set chosen from the corners of parameter space and an LHD.
-The next 200 waveforms are chosen with uncertainty sampling.
-The final 1000 waveforms are drawn uniformly from parameter space.
-The markers are the comparisons of $\epsilon_{\rm rms}$ calculated by maximizing the error over a test set of 1000 waveforms instead of using Eq.~\eqref{eq:errorrms} for the error estimate.}
+\caption{Estimated RMS phase error $\epsilon_{\rm RMS}$ (Eq.~\eqref{eq:errorrms}) maximized over parameter space $\bx$ as a function of the number of chosen parameters (blue curve).
+$\epsilon_{\rm RMS}$ depends on the chose samples $\bx_j$ and hyperparameters, not on the data, and can be calculated before generating the waveforms.
+Also shown for comparison are the maximum RMS phase errors at the nodes where $MF_j \le 0.03$ for the surrogate compared to a test set of 1000 waveforms.
+Blue circle: surrogate constructed from the 160 waveforms in the initial training set.
+Red square: surrogate after adding the 200 waveforms chosen with uncertainty sampling.
+Green triangle: final surrogate after also adding 1000 more training set waveforms drawn uniformly from parameter space.
+}
 \label{fig:uncsamp}
@@
 
-Given the interpolating functions for the amplitude and phase residuals $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ and $\{\mathcal{I}_{\rm GPR, j}[\Delta\Phi_j](\bx)\}$ evaluated at the frequency nodes $\{F_j\}$, we can now reconstruct the frequency-domain waveform.
-The surrogates for the residuals $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$ are constructed by interpolating between the nodes $\{F_j\}$ with cubic splines:
+Given the interpolating functions for the amplitude and phase residuals $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ and $\{\mathcal{I}_{\rm GPR, j}[\Delta\Phi_j](\bx)\}$ evaluated at the frequency nodes $MF_j$, we can now reconstruct the frequency-domain waveform.
+The surrogates for the residuals $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$ are constructed by interpolating between the nodes $MF_j$ with cubic splines:
 \begin{align}
@@
 \end{align}
-In physical units the $+$ and $\times$ polarizations of the waveform are then 
+In physical units, for an inclination angle $\iota$, the $+$ and $\times$ polarizations of the waveform are 
 \begin{align}
 \tilde h_+(f; \bx) &= \frac{1}{2}(1+\cos^2\iota) \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
-& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right)\right] \\
+& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right)\right], \\
 \tilde h_\times(f; \bx) &= \cos\iota \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
-& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right) + i \frac{\pi}{2}\right]
+& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right) + i \frac{\pi}{2}\right].
 \end{align}
-where $\iota$ is the inclination angle.
 
@@
 \subsection{Accuracy}
+\label{sec:accuracy}
 
@@
 \includegraphics[width=0.49\textwidth]{mismatch.png}
-\caption{Mismatch}
+\caption{Mismatches between the surrogate and 1000 test-set waveforms with randomly sampled parameters $\bx$.
+The total mass for each waveform is $M=2.8M_\odot$ and the PSD used was the aLIGO design sensitivity PSD.
+Histograms are shown for the initial surrogate with 160 waveforms (blue), the updated surrogate with 360 waveforms (red), and the final surrogate with 1360 waveforms (green).
+Also shown for comparison are the mismatches between TaylorF2 and the test set (black).}
 \label{fig:mismatch}
@@
 \includegraphics[width=0.49\textwidth]{htildemaxerror.png}
-\caption{Amplitude and phase errors as functions of frequency for the waveform parameters $\bx$ that gave the largest mismatch with the test-set waveforms.}
+\caption{Amplitude and phase errors between the surrogate and the test-set waveform with the largest mismatch.
+Blue curve: initial surrogate with 160 waveforms.
+Red curve: updated surrogate with 360 waveforms.
+Green curve: final surrogate with 1360 waveforms.
+The vertical solid line is the gravitational-wave frequency at the Schwarzschild ISCO.}
 \label{fig:maxmismatch}
@@
 In Fig.~\ref{fig:maxmismatchtd} we compare the surrogate to the test set waveform that had the largest mismatch.
-We align the two waveforms in time and phase by maximizing the overlap at early times in the interval $-200,000 < t/M < -100,000$.
+We align the two waveforms in time and phase by maximizing the overlap at early times in the interval $-2\times 10^5 < t/M < -1\times 10^5$.
 The two waveforms are nearly identical except during the last $\sim 2$ cycles where the surrogate peaks in amplitude before the EOB waveform.
@@
 \includegraphics[width=0.49\textwidth]{hmaxerror.png}
-\caption{Inverse Fourier transformed surrogate compared to the original EOB waveform.}
+\caption{Inverse Fourier transformed surrogate (green dashed) for the parameters $\bx$ that had the largest mismatch with the test-set EOB waveforms (black).
+The two waveforms are aligned in the time domain in the interval $-2\times 10^5 < t/M < -1\times 10^5$.}
 \label{fig:maxmismatchtd}

### 2017-11-26 paper/paper.tex
Fixing most figures.

@@
 The fact that the residuals are smooth functions of the waveform parameters indicates that most numerical noise has been removed.
-The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.05; \bx)$, where the amplitude is very small and dominated by numerical noise.
+The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.052; \bx)$, where the amplitude is very small and dominated by numerical noise.
 
@@
 We then interpolate between these frequencies using third-order splines.
-The local, third-order polynomials, that are only connected by the requirement of smoothness do not propagate high-frequency errors down to low-frequency errors as significantly as do global basis functions.
+The local, third-order polynomials, that are only connected by the requirement of smoothness, do not propagate high-frequency errors down to low-frequency errors as significantly as do global basis functions.
 
@@
 Most multivariate interpolation techniques (e.g.
-tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality; the number of samples grows exponentially with the dimension $d$ ($N^d$ samples for $N$ points per dimension).
+tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality: the number of samples grows exponentially with the dimension $d$ ($N^d$ samples for $N$ points per dimension).
 For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
@@
 
-In a zero-mean Gaussian process, the data $\by_i$ and function value $\by_*$ at the training set points $\bx_i$ and new point $\bx_*$, respectively, are drawn from a multivariate normal distribution
+In a zero-mean Gaussian process, the data $y_i$ and function value $y_*$ at the training set points $\bx_i$ and new point $\bx_*$, respectively, are drawn from a multivariate normal distribution
 \begin{equation}
@@
 \centering
-\includegraphics[width=0.49\textwidth]{uncertaintysampling.png}
+\includegraphics[width=0.49\textwidth]{uncertaintysampling.pdf}
 \caption{Estimated RMS phase error $\epsilon_{\rm RMS}$ (Eq.~\eqref{eq:errorrms}) maximized over parameter space $\bx$ as a function of the number of chosen parameters (blue curve).
@@
 \centering
-\includegraphics[width=0.49\textwidth]{mismatch.png}
+\includegraphics[width=0.49\textwidth]{mismatch.pdf}
 \caption{Mismatches between the surrogate and 1000 test-set waveforms with randomly sampled parameters $\bx$.
@@
 \centering
-\includegraphics[width=0.49\textwidth]{htildemaxerror.png}
+\includegraphics[width=0.49\textwidth]{htildemaxerror.pdf}
 \caption{Amplitude and phase errors between the surrogate and the test-set waveform with the largest mismatch.
@@
 \centering
-\includegraphics[width=0.49\textwidth]{hmaxerror.png}
+\includegraphics[width=0.49\textwidth]{hmaxerror.pdf}
 \caption{Inverse Fourier transformed surrogate (green dashed) for the parameters $\bx$ that had the largest mismatch with the test-set EOB waveforms (black).

### 2017-11-30 src/TEOB-LAL-implementation/LALSimIMR.h
Corrected TaylorF2 and verified with lalsimulation version.

@@ -1,298 +1,0 @@
-/*
- * Copyright (C) 2011 N. Fotopoulos <nickolas.fotopoulos@ligo.org>
- *
- *  This program is free software; you can redistribute it and/or modify
- *  it under the terms of the GNU General Public License as published by
- *  the Free Software Foundation; either version 2 of the License, or
- *  (at your option) any later version.
- *
- *  This program is distributed in the hope that it will be useful,
- *  but WITHOUT ANY WARRANTY; without even the implied warranty of
- *  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
- *  GNU General Public License for more details.
- *
- *  You should have received a copy of the GNU General Public License
- *  along with with program; see the file COPYING. If not, write to the
- *  Free Software Foundation, Inc., 59 Temple Place, Suite 330, Boston,
- *  MA  02111-1307  USA
- */
-
-#ifndef _LALSIMIMR_H
-#define _LALSIMIMR_H
-
-#include <lal/LALDatatypes.h>
-#include <lal/LALSimInspiral.h>
-
-#ifdef LAL_HDF5_ENABLED
-#include <lal/H5FileIO.h>
-#endif
-
-#if defined(__cplusplus)
-extern "C" {
-#elif 0
-} /* so that editors will match preceding brace */
-#endif
-
-/**
- * @defgroup LALSimIMR_h Header LALSimIMR.h
- * @ingroup lalsimulation_inspiral
- *
- * @brief Routines for generating inspiral-merger-ringdown waveforms.
- *
- * @{
- * @defgroup LALSimIMRPhenom_c                   LALSimIMRPhenom.c
- * @defgroup LALSimIMREOBNRv2_c                  LALSimIMREOBNRv2.c
- * @defgroup LALSimIMRSpinAlignedEOB_c           LALSimIMRSpinAlignedEOB.c
- * @defgroup LALSimIMRSpinPrecEOB_c              LALSimIMRSpinPrecEOB.c
- * @defgroup LALSimIMRSEOBNRROM_c                LALSimIMRSEOBNRvxROMXXX.c
- * @defgroup LALSimIMRSEOBNRv2ChirpTime_c        LALSimIMRSEOBNRv2ChirpTime.c
- * @defgroup LALSimIMRPSpinInspiralRD_c          LALSimIMRPSpinInspiralRD.c
[…]

### 2017-11-30 src/TEOB-LAL-implementation/LALSimIMREOBNRv2.h

@@ -1,520 +1,0 @@
-/*
-*  Copyright (C) 2010 Craig Robinson
-*
-*  This program is free software; you can redistribute it and/or modify
-*  it under the terms of the GNU General Public License as published by
-*  the Free Software Foundation; either version 2 of the License, or
-*  (at your option) any later version.
-*
-*  This program is distributed in the hope that it will be useful,
-*  but WITHOUT ANY WARRANTY; without even the implied warranty of
-*  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-*  GNU General Public License for more details.
-*
-*  You should have received a copy of the GNU General Public License
-*  along with with program; see the file COPYING. If not, write to the
-*  Free Software Foundation, Inc., 59 Temple Place, Suite 330, Boston,
-*  MA  02111-1307  USA
-*/
-
-
-/**
- * \author Craig Robinson, Andrea Taracchini
- *
- * \brief File containing most of the structures and prototypes which are
- * used in the generation of the EOBNRv2 waveform. These structures and
- * prototypes are used internally within the waveform generation code,
- * and shouldn't be needed to generate the waveforms from outside the
- * library. Therefore, people generating EOBNRv2 waveforms should only
- * need to include LALSimIMR.h.
- */
-
-#include <lal/LALConstants.h>
-#include <lal/LALSimIMR.h>
-#include <lal/LALSimInspiral.h>
-
-#ifndef _LALSIMIMREOBNRv2_H
-#define _LALSIMIMREOBNRv2_H
-
-#if defined(__cplusplus)
-extern "C"
-{
-#elif 0
-}				/* so that editors will match preceding brace */
-#endif
-
-/**
[…]

### 2017-11-30 src/TEOB-LAL-implementation/LALSimIMRSpinAlignedEOB.c

@@ -1,2216 +1,0 @@
-/*
-*  Copyright (C) 2011 Craig Robinson, Enrico Barausse, Yi Pan, Prayush Kumar
-*  (minor changes), Andrea Taracchini
-*
-*  This program is free software; you can redistribute it and/or modify
-*  it under the terms of the GNU General Public License as published by
-*  the Free Software Foundation; either version 2 of the License, or
-*  (at your option) any later version.
-*
-*  This program is distributed in the hope that it will be useful,
-*  but WITHOUT ANY WARRANTY; without even the implied warranty of
-*  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-*  GNU General Public License for more details.
-*
-*  You should have received a copy of the GNU General Public License
-*  along with with program; see the file COPYING. If not, write to the
-*  Free Software Foundation, Inc., 59 Temple Place, Suite 330, Boston,
-*  MA  02111-1307  USA
-*/
-
-
-#include <math.h>
-#include <complex.h>
-#include <lal/LALSimInspiral.h>
-#include <lal/LALSimIMR.h>
-#include <lal/Date.h>
-#include <lal/TimeSeries.h>
-#include <lal/Units.h>
-#include <lal/LALAdaptiveRungeKuttaIntegrator.h>
-#include <lal/SphericalHarmonics.h>
-#include <gsl/gsl_sf_gamma.h>
-
-#include "LALSimIMREOBNRv2.h"
-#include "LALSimIMRSpinEOB.h"
-#include "LALSimInspiralPrecess.h"
-
-/* Include all the static function files we need */
-#include "LALSimIMREOBHybridRingdown.c"
-#include "LALSimIMREOBFactorizedWaveform.c"
-#include "LALSimIMREOBNewtonianMultipole.c"
-#include "LALSimIMREOBNQCCorrection.c"
-#include "LALSimIMRSpinEOBInitialConditions.c"
-#include "LALSimIMRSpinEOBAuxFuncs.c"
-#include "LALSimIMRSpinAlignedEOBHcapDerivative.c"
-#include "LALSimIMRSpinEOBHamiltonian.c"
-#include "LALSimIMRSpinEOBFactorizedWaveform.c"
-#include "LALSimIMRSpinEOBFactorizedFlux.c"
-/* OPTIMIZED */
-#include "LALSimIMRSpinEOBHamiltonianOptimized.c"
[…]

### 2017-11-30 src/TEOB-LAL-implementation/LALSimIMRTEOBv4ROM.c

@@ -1,1226 +1,0 @@
-/*
- *  Copyright (C) 2017 Michael Puerrer, Ben Lackey
- *  Reduced Order Model for TEOBv4
- *
- *  This program is free software; you can redistribute it and/or modify
- *  it under the terms of the GNU General Public License as published by
- *  the Free Software Foundation; either version 2 of the License, or
- *  (at your option) any later version.
- *
- *  This program is distributed in the hope that it will be useful,
- *  but WITHOUT ANY WARRANTY; without even the implied warranty of
- *  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
- *  GNU General Public License for more details.
- *
- *  You should have received a copy of the GNU General Public License
- *  along with with program; see the file COPYING. If not, write to the
- *  Free Software Foundation, Inc., 59 Temple Place, Suite 330, Boston,
- *  MA  02111-1307  USA
- */
-
-#ifdef __GNUC__
-#define UNUSED __attribute__ ((unused))
-#else
-#define UNUSED
-#endif
-
-#include <stdio.h>
-#include <stdlib.h>
-#include <math.h>
-#include <complex.h>
-#include <time.h>
-#include <stdbool.h>
-#include <alloca.h>
-#include <string.h>
-#include <libgen.h>
-#include <assert.h>
-
-#include <gsl/gsl_errno.h>
-#include <gsl/gsl_bspline.h>
-#include <gsl/gsl_blas.h>
-#include <gsl/gsl_min.h>
-#include <gsl/gsl_spline.h>
-#include <gsl/gsl_poly.h>
-#include <lal/Units.h>
-#include <lal/SeqFactories.h>
-#include <lal/LALConstants.h>
-#include <lal/XLALError.h>
-#include <lal/FrequencySeries.h>
-#include <lal/Date.h>
-#include <lal/StringInput.h>
-#include <lal/Sequence.h>
-#include <lal/LALStdio.h>
-#include <lal/FileIO.h>
-
-
-#ifdef LAL_HDF5_ENABLED
-#include <lal/H5FileIO.h>
-// FIXME: change these to our datafile and version number 1.0.0
-//static const char ROMDataHDF5[] = "SEOBNRv4ROM_v2.0.hdf5";
-//static const char ROMDataHDF5[] = "/Users/mpuer/Documents/gpsurrogate/src/TEOB-LAL-implementation/TEOBv4_surrogate.hdf5";
[…]

### 2017-11-30 src/taylorf2.py

@@ -134,6 +134,25 @@ def taylorf2_amp_1pn(mf, eta):
     return 2.0*a00 * x**(-7./4.) * (1. + a10*x)
 
 
+def taylorf2_phase_35pn_spin(eta, chi1, chi2):
+    """3.5PN phase copied from LALSimInspiralPNCoefficients.c (line 694).
+    Spin-orbit terms derived from arXiv:1303.7412 (Eq. 3.15-16).
+    """
+    d = np.sqrt(1.0-4.0*eta)
+    X1 = 0.5*(1.0+d)
+    X2 = 0.5*(1.0-d)
+
+    # point particle term
+    #v7 = ( 77096675.0/254016.0 + 378515.0*eta/1512.0 - 74045.0*eta**2/756.0 )*np.pi
+
+    # aligned spin term
+    SL = X1**2 * chi1 + X2**2 * chi2
+    dSigmaL = d*(X2*chi2 - X1*chi1)
+    v7 = ( -8980424995.0/762048.0 + 6586595.0*eta/756.0 - 305.0*eta**2/36.0 )*SL
+    v7 = v7 - ( 170978035.0/48384.0 - 2876425.0*eta/672.0 - 4735.0*eta**2/144.0 )*dSigmaL
+    return v7
+
+
 def taylorf2_phase(mf, tbymc, phic, eta, chi1, chi2, lambda1, lambda2):
     """3.5PN point-particle phase.
     FFT sign convention is $\tilde h(f) = \int h(t) e^{-2 \pi i f t} dt$
@@ -161,12 +180,12 @@ def taylorf2_phase(mf, tbymc, phic, eta, chi1, chi2, lambda1, lambda2):
     + (75515.0/288.0 - 263245.0*eta/252.0 - 480.0*eta**2)*chia**2\
     + (75515.0/288.0 - 232415.0*eta/504.0 + 1255.0*eta**2/9.0)*chis**2
 
-    p7 = (-25150083775.0/3048192.0 + 26804935.0*eta/6040.0 - 1985.0*eta**2/48.0)*delta*chia\
-    + (-25150083775.0/3048192.0 + 10566655595.0*eta/762048.0 - 1042165.0*eta**2/3024.0 + 5345.0*eta**3/36.0)*chis\
-    + (14585.0/24.0 - 2380.0*eta)*delta*chia**3\
-    + (14585.0/24.0 - 475.0*eta/6.0 + 100.0*eta**2/3.0)*chis**3\
[…]

### 2017-11-30 src/TEOB-LAL-implementation/LALSimIMREOBNRv2.h
Restoring accidentally deleted files in TEOB-LAL-implementation

@@ -1,0 +1,520 @@
+/*
+*  Copyright (C) 2010 Craig Robinson
+*
+*  This program is free software; you can redistribute it and/or modify
+*  it under the terms of the GNU General Public License as published by
+*  the Free Software Foundation; either version 2 of the License, or
+*  (at your option) any later version.
+*
+*  This program is distributed in the hope that it will be useful,
+*  but WITHOUT ANY WARRANTY; without even the implied warranty of
+*  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+*  GNU General Public License for more details.
+*
+*  You should have received a copy of the GNU General Public License
+*  along with with program; see the file COPYING. If not, write to the
+*  Free Software Foundation, Inc., 59 Temple Place, Suite 330, Boston,
+*  MA  02111-1307  USA
+*/
+
+
+/**
+ * \author Craig Robinson, Andrea Taracchini
+ *
+ * \brief File containing most of the structures and prototypes which are
+ * used in the generation of the EOBNRv2 waveform. These structures and
+ * prototypes are used internally within the waveform generation code,
+ * and shouldn't be needed to generate the waveforms from outside the
+ * library. Therefore, people generating EOBNRv2 waveforms should only
+ * need to include LALSimIMR.h.
+ */
+
+#include <lal/LALConstants.h>
+#include <lal/LALSimIMR.h>
+#include <lal/LALSimInspiral.h>
+
+#ifndef _LALSIMIMREOBNRv2_H
+#define _LALSIMIMREOBNRv2_H
+
+#if defined(__cplusplus)
+extern "C"
+{
+#elif 0
+}				/* so that editors will match preceding brace */
+#endif
+
+/**
[…]

### 2017-11-30 src/TEOB-LAL-implementation/LALSimIMRSpinAlignedEOB.c

@@ -1,0 +1,2216 @@
+/*
+*  Copyright (C) 2011 Craig Robinson, Enrico Barausse, Yi Pan, Prayush Kumar
+*  (minor changes), Andrea Taracchini
+*
+*  This program is free software; you can redistribute it and/or modify
+*  it under the terms of the GNU General Public License as published by
+*  the Free Software Foundation; either version 2 of the License, or
+*  (at your option) any later version.
+*
+*  This program is distributed in the hope that it will be useful,
+*  but WITHOUT ANY WARRANTY; without even the implied warranty of
+*  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+*  GNU General Public License for more details.
+*
+*  You should have received a copy of the GNU General Public License
+*  along with with program; see the file COPYING. If not, write to the
+*  Free Software Foundation, Inc., 59 Temple Place, Suite 330, Boston,
+*  MA  02111-1307  USA
+*/
+
+
+#include <math.h>
+#include <complex.h>
+#include <lal/LALSimInspiral.h>
+#include <lal/LALSimIMR.h>
+#include <lal/Date.h>
+#include <lal/TimeSeries.h>
+#include <lal/Units.h>
+#include <lal/LALAdaptiveRungeKuttaIntegrator.h>
+#include <lal/SphericalHarmonics.h>
+#include <gsl/gsl_sf_gamma.h>
+
+#include "LALSimIMREOBNRv2.h"
+#include "LALSimIMRSpinEOB.h"
+#include "LALSimInspiralPrecess.h"
+
+/* Include all the static function files we need */
+#include "LALSimIMREOBHybridRingdown.c"
+#include "LALSimIMREOBFactorizedWaveform.c"
+#include "LALSimIMREOBNewtonianMultipole.c"
+#include "LALSimIMREOBNQCCorrection.c"
+#include "LALSimIMRSpinEOBInitialConditions.c"
+#include "LALSimIMRSpinEOBAuxFuncs.c"
+#include "LALSimIMRSpinAlignedEOBHcapDerivative.c"
+#include "LALSimIMRSpinEOBHamiltonian.c"
+#include "LALSimIMRSpinEOBFactorizedWaveform.c"
+#include "LALSimIMRSpinEOBFactorizedFlux.c"
+/* OPTIMIZED */
+#include "LALSimIMRSpinEOBHamiltonianOptimized.c"
[…]

### 2017-11-30 src/TEOB-LAL-implementation/LALSimIMRTEOBv4ROM.c

@@ -1,0 +1,1226 @@
+/*
+ *  Copyright (C) 2017 Michael Puerrer, Ben Lackey
+ *  Reduced Order Model for TEOBv4
+ *
+ *  This program is free software; you can redistribute it and/or modify
+ *  it under the terms of the GNU General Public License as published by
+ *  the Free Software Foundation; either version 2 of the License, or
+ *  (at your option) any later version.
+ *
+ *  This program is distributed in the hope that it will be useful,
+ *  but WITHOUT ANY WARRANTY; without even the implied warranty of
+ *  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+ *  GNU General Public License for more details.
+ *
+ *  You should have received a copy of the GNU General Public License
+ *  along with with program; see the file COPYING. If not, write to the
+ *  Free Software Foundation, Inc., 59 Temple Place, Suite 330, Boston,
+ *  MA  02111-1307  USA
+ */
+
+#ifdef __GNUC__
+#define UNUSED __attribute__ ((unused))
+#else
+#define UNUSED
+#endif
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <math.h>
+#include <complex.h>
+#include <time.h>
+#include <stdbool.h>
+#include <alloca.h>
+#include <string.h>
+#include <libgen.h>
+#include <assert.h>
+
+#include <gsl/gsl_errno.h>
+#include <gsl/gsl_bspline.h>
+#include <gsl/gsl_blas.h>
+#include <gsl/gsl_min.h>
+#include <gsl/gsl_spline.h>
+#include <gsl/gsl_poly.h>
+#include <lal/Units.h>
+#include <lal/SeqFactories.h>
+#include <lal/LALConstants.h>
+#include <lal/XLALError.h>
+#include <lal/FrequencySeries.h>
+#include <lal/Date.h>
+#include <lal/StringInput.h>
+#include <lal/Sequence.h>
+#include <lal/LALStdio.h>
+#include <lal/FileIO.h>
+
+
+#ifdef LAL_HDF5_ENABLED
+#include <lal/H5FileIO.h>
+// FIXME: change these to our datafile and version number 1.0.0
+//static const char ROMDataHDF5[] = "SEOBNRv4ROM_v2.0.hdf5";
[…]

### 2017-12-02 paper/paper.tex
Corrected TaylorF2 and verified with lalsimulation version.
Adding references.

@@
 
-For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.
+%For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.
 
@@
 
-The detection of the binary neutron star (BNS) system GW170817 [[cite]] by the Advanced LIGO [[cite]] and Advanced Virgo [[cite]] gravitational-wave detectors, has demonstrated that the sky position, distance, masses, spins, and tidal parameters can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant [[cite]], and understanding the structure of neutron stars (NSs) through tidal interactions [[cite]].
+The detection of the binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors, has demonstrated that the sky position, distance, masses, spins, and tidal parameters can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant~\cite{GW170817Hubble}, and understanding the structure of neutron stars (NSs) through tidal interactions~\cite{GW170817}.
 Measuring these parameters accurately will also be necessary to properly interpret the gamma ray burst [[cite]] and optical afterglow [[cite]] that was found in conjunction.
-Furthermore, an estimate of the BNS merger rate of x--y events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval) [[GW170817]] indicates that, when second generation detectors reach design sensitivity, including KAGRA [[cite]] and possibly LIGO-India [[cite]], we could eventually observe a population of tens or hundreds of events after a couple of years of observation [[2010Rates?]].
-This will allow us to estimate the NS mass distribution [[cite]], spin distribution [[cite]], and universal equation of state (EOS) that is common to all NSs [[DelPozzo, Lackey, ...]].
+Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} indicates that, when second generation detectors reach design sensitivity, including KAGRA~\cite{Somiya2012} and possibly LIGO-India~\cite{IyerSouradeepUnnikrishnan2011}, we could eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}.
+This will allow us to estimate the NS mass distribution [[cite]], spin distribution [[cite]], and universal equation of state (EOS) that is common to all NSs\cite{DelPozzoLiAgathos2013, LackeyWade2015}.
 
@@
 The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms.
-Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad.
+Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad~\cite{HindererLackeyLangRead2010}.
 Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]].
@@
 This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions.
-In previous work, this reduced basis has been constructed either with singular value decomposition (SVD) [[Michael's papers]] or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization [[All the other papers]].
+In previous work, this reduced basis has been constructed either with singular value decomposition (SVD)~\cite{Puerrer2014, Puerrer2015} or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
 
@@
 Previous work on surrogate models has focused mainly on parameter spaces $\bx$ with $\le 3$~dimensions.
-This allows one to approximate the coefficients $c_i(\bx)$ or waveform values $h(T_j; \bx)$ as functions of $\bx$ using simple fitting functions [[cite]] or standard interpolation techniques such as tensor spline [[Michael, others]] or Chebyshev interpolation [[Ben]].
+This allows one to approximate the coefficients $c_i(\bx)$ or waveform values $h(T_j; \bx)$ as functions of $\bx$ using simple fitting functions [[cite]] or standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}.
 These interpolation techniques typically require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
 For the 5-dimensional problem here, this is somewhat unfeasable.
-Recent work on optimally choosing waveforms for both EOB [[cite GPRpaper]] and NR [[cite CaltechPrecessing]] waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.
-
-In this paper we focus on breaking this curse of dimensionality by using an interpolation method known as Gaussian process regression (GPR) that does not require a regular grid.
+Recent work on optimally choosing waveforms for both EOB~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.
+
+In this paper we focus on breaking this curse of dimensionality by using an interpolation method known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid.
 Importantly, GPR also provides a convenient estimate of its own uncertainty.
 This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set.
-A similar approach has recently been used to optimally sample waveforms for a 2-dimensional parameter space using GPR [[DoctorEtAl]].
-GPR has also been used recently in GW data analysis to marginalize over waveform uncertainties in parameter estimation [[ChrisMoore2papers?]].
+Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of eccentric BBH mergers~\cite{HuertaMooreKumar2017}.
+GPR has also been used recently in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016}.
 
 In this paper, we will develop methods to construct a surrogate model for the inspiral through merger of an aligned spin BNS waveform with tidal interactions using the EOB tidal description described in Hinderer et al.~[[cite1, cite2]].
-This work is a continuation of the surrogate developed in [[cite]] for a nonspinning BNS waveform using a slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~[[cite, ...]].
-Whereas in [[Ben]] we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation.
-It also enables additional techniques such as reduced order quadrature [[cite]] to accelerate likelihood evaluation.
+This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS waveform using a slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~\cite{Bernuzzi:2014owa}.
+Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation.
+It also enables additional techniques such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} to accelerate likelihood evaluation.
 
@@
 
-%Because the waveform $\tilde h(Mf; \bx)$ is an oscillatory function of $Mf$ and $\bx$, the waveform is usually decomposed into an amplitude $A(Mf; \bx)$ and phase $\Phi(Mf; \bx)$ as $\tilde h(Mf; \bx) = A(Mf; \bx) e^{i\Phi(Mf; \bx)}$.
-These functions are smoother, mostly monotonic functions of $Mf$ and $\bx$.
-This can be seen in Fig.~\ref{fig:h} which shows the initial training set of waveforms.
-\red{[[Actually, you might want to show the 51 waveforms with only $S_1$ varied.
-Then show them as a function of $f$ and another plot as a function of $S_1$.]]} Unfortunately, the amplitude and phase still span a wide range of values.
-The phase, for example, spans about $10^4$~rad.
-To avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad.
-This is a difficult requirement to achieve in the 5-dimensional parameter space here.
-
-%\begin{figure}[htb]
-%\centering
-%\includegraphics[width=0.49\textwidth]{h_trainingset.png}
-%%\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
-%\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
-%\label{fig:h}
-%\end{figure}
-
-%\begin{figure}[htb]
-%\centering
-%%\includegraphics[width=0.49\textwidth]{h_trainingset.png}\\
-%\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
-%\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
-%\label{fig:dh}
-%\end{figure}
-
-%To demonstrate the usefulness of this approach, in Fig.~\ref{fig:phaserange}, we show the range of values of the phase $\Phi$ over the training set $\Phi_{\rm range} = \Phi_{\rm max} - \Phi_{\rm min}$ as well as the range of values of the phase residual $\Delta\Phi$ over the training set $\Delta\Phi_{\rm range} = \Delta\Phi_{\rm max}-\Delta\Phi_{\rm min}$.
-The range in $\Phi$ can be larger than 1000~radians.
-The range of $\Delta\Phi$ is 1--2 orders of magnitude smaller, and as a result, the interpolation accuracy requirements are 1--2 orders of magnitude smaller as well.
-%
-%\begin{figure}[htb]
-%\centering
-%\includegraphics[width=0.49\textwidth]{phaserange.png}
-%\caption{Range of phase values for training set waveforms. The range is a factor of $\gtrsim 100$ larger for $\Phi$ than for $\Delta\Phi$, 
-%meaning that the interpolation accuracy requirements are at least $\sim 2$ orders of magnitude less for $\Delta\Phi$.}
-%\label{fig:phaserange}
-%\end{figure}
-
 Because the waveform $\tilde h(Mf; \bx)$ is an oscillatory function of $Mf$ and $\bx$, the waveform is usually decomposed into an amplitude $A(Mf; \bx)$ and phase $\Phi(Mf; \bx)$ as $\tilde h(Mf; \bx) = A(Mf; \bx) e^{i\Phi(Mf; \bx)}$.
@@
 
-For aligned-spin waveforms, we can solve this problem using the fact that the waveform is approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2 [[cite]] $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$.
+For aligned-spin waveforms, we can solve this problem using the fact that the waveform is approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2~\cite{BuonannoIyerOchsner2009} $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$.
 This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2
@@
 \end{equation}
-where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{lalsuite}. The phase has the schematic form
+where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{lalsuite}~\cite{lal}. The phase has the schematic form
 \begin{align}
@@
 \end{align}
-We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~[[]] of Ref.~[[]], the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order in Eqs.~[[]] of Ref.~[[]] and Eqs.~[[]] of Ref.~[[]], and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order in Eqs.~[[]] of Ref.~[[]].
-These terms for the phase are exactly as used in the \texttt{lalsuite} waveform \texttt{TaylorF2} [[cite]].
+We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}.
+These terms for the phase are exactly as used in the \texttt{lalsuite} waveform \texttt{TaylorF2}.
 
@@
 We evaluate the EOB waveform with a starting frequency of $Mf_{{\rm win}, i}=0.000197$, equivalent to a physical frequency of 20~Hz for a binary with total mass $M=2M_\odot$.
-In order to take a discrete Fourier transform, we window the start of the waveform with a Planck window [[cite]] in the interval $[Mf_{{\rm win}, i}, Mf_{{\rm win}, f}] = [0.000197, 0.00021]$ to avoid Gibbs oscillations.
+In order to take a discrete Fourier transform, we window the start of the waveform with a Planck window~\cite{McKechanRobinsonSathyaprakash2010} in the interval $[Mf_{{\rm win}, i}, Mf_{{\rm win}, f}] = [0.000197, 0.00021]$ to avoid Gibbs oscillations.
 The end of the waveform has zero amplitude, so the end does not need to be windowed.
@@
 We begin by choosing a method for interpolating as a function of $Mf$ for fixed $\bx$.
-The majority of previous papers on gravitational-wave surrogates have used an orthonormal basis of global functions $\hat e_i(Mf)$ for interpolating as a function of frequency (or time) [[cite]].
+The majority of previous papers on gravitational-wave surrogates have used an orthonormal basis of global functions $\hat e_i(Mf)$ for interpolating as a function of frequency (or time)~\cite{Puerrer2014, Puerrer2015, FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
 This type of surrogate, built from a reduced basis, is usually referred to as a reduced order model.
 For a generic function $g(Mf; \bx)$, this decomposition can be written as $g(Mf; \bx) \approx \sum_{i=1}^N c_i(\bx) \hat e_i(Mf)$.
-The coefficients $c_i(\bx)$ are then interpolated as functions of $\bx$ [[Puerrer]].
-Alternatively, using the empirical interpolation method [[cite]], one can re-express these basis functions $\hat e_i(Mf)$ in terms of empirical interpolating functions $B_j(Mf)$ and the value of $g(Mf; \bx)$ at empirical nodes $MF_j$ as $g(Mf; \bx) \approx \sum_{j=1}^N B_j(Mf) g(MF_j; \bx)$.
+The coefficients $c_i(\bx)$ are then interpolated as functions of $\bx$~\cite{Puerrer2014, Puerrer2015}.
+Alternatively, using the empirical interpolation method~\cite{Barrault2004, Chaturantabut2010, FieldGalleyHesthaven2014}, one can re-express these basis functions $\hat e_i(Mf)$ in terms of empirical interpolating functions $B_j(Mf)$ and the value of $g(Mf; \bx)$ at empirical nodes $MF_j$ as $g(Mf; \bx) \approx \sum_{j=1}^N B_j(Mf) g(MF_j; \bx)$.
 The location of these nodes $MF_j$ is then optimized to minimize interpolation errors.
-This method has been used in Refs.~[[cite]].
 
@@
 
-\red{[[Make sure all vectors and matrices are bold.]]}
-
 Next we choose a method to interpolate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at each of the frequency nodes $MF_j$ as a function of the five waveform parameters $\bx$.
@@
 However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below.
-The method we choose is Gaussian Process Regression (GPR).
-
-In GPR a function $g(\bx)$ is described in terms of a mean $m(\bx)$ and  covariance $k(\bx, \bx')$ between points [[cite RassmussenWilliamsChapter2, IntroArxivPaper]].
+The method we choose is Gaussian Process Regression (GPR)~\cite{RasmussenWilliams2006}.
+
+In GPR a function $g(\bx)$ is described in terms of a mean $m(\bx)$ and  covariance $k(\bx, \bx')$ between points.
 \begin{equation}
@@
 where $K_\nu(x)$ is a modified Bessel function.
-The value of $\nu$ parameterizes the smoothness of the Gaussian process, and is $k$ times mean-square differentiable if $\nu>k$ [[cite]].
-For half-integer values of $\nu$, this kernel has a computationally cheap form without special functions, and we have had good results with $\nu=5/2$, resulting in a twice-differentiable function\footnote{If desired, one can select the optimal class of kernels using cross validation, and we have also tried the more common, infinitely differentiable squared exponential kernel [[cite]].
+The value of $\nu$ parameterizes the smoothness of the Gaussian process, and is $k$ times mean-square differentiable if $\nu>k$~\cite{RasmussenWilliams2006}.
+For half-integer values of $\nu$, this kernel has a computationally cheap form without special functions, and we have had good results with $\nu=5/2$, resulting in a twice-differentiable function\footnote{If desired, one can select the optimal class of kernels using cross validation~\cite{RasmussenWilliams2006}, and we have also tried the more common, infinitely differentiable squared exponential kernel~\cite{RasmussenWilliams2006}.
 However, we found it difficult to reliably tune its hyperparameters for the problem here.}.
@@
 However, for the problem here, we simply want the maximum posterior.
-We do this using the \texttt{gaussian\_process} module in the \texttt{scikit-learn} package [[cite]].
+We do this using the \texttt{gaussian\_process} module in the \texttt{scikit-learn} package~\cite{scikit-learn}.
 With these optimized hyperparameters, the final interpolating function is given by Eq.~\eqref{eq:mean} and its uncertainty by Eq.~\eqref{eq:var}.
 We note that once the vector $K^{-1}y$ is precomputed, evaluating Eq.~\eqref{eq:mean} is a fast $\mathcal{O}(N)$ operation.
-
-\subsection{Iterative construction of training set and surrogate}
-\label{sec:design}
-
-%One of the main aims of this paper is to build a surrogate with as few waveform evaluations as possible.
-EOB models are expensive, and we would like a method that extends to higher dimensions and more expensive waveform models such as from numerical relativity simulations.
-Using the freedom provided by GPR to sample waveforms at arbitrary locations in parameter space, we try a set of more optimal designs.
-
-One of the main aims of this paper is to build a surrogate with as few waveform evaluations as possible.
-Using the freedom provided by GPR to sample waveforms at arbitrary locations, we try a set of designs more efficient than uniform grids.
-One such design is a Latin Hypercube Design (LHD) [[cite]].
-An LHD with $N$ samples divides each of $d$ dimensions uniformly into $N$ grid points for a total of $N^d$ grid points.
-However, unlike a uniform grid, the $N$ values in each dimension are sampled exactly once instead of $N^{d-1}$ times.
-For an LHD there are $(N!)^d$ ways to choose these points, and we choose one randomly\footnote{There are ways to better choose an LHD.
-A standard requirement is that the LHD be space-filling, meaning that the points are as far apart as possible from each other.
-(See [[cite]] for a review of methods for optimizing the placement of samples.) One such definition of space filling is that the chosen locations maximize the minimum Euclidean distance between any two samples [[cite]].
-We have not experimented with these alternatives here.}.
-
-An LHD has the property that the samples are non-collapsing; a projection onto a subspace is still an LHD and no points are repeated in any dimension.
-This avoids wasting samples when one of the parameters has much less of an influence than the other parameters.
-As an example, the tidal parameters have a minimal impact on the waveform at low frequencies, so it would not make sense to densely sample a grid of tidal parameters at low frequencies.
-However, during the merger the tidal parameters can be more important than the other parameters.
-One therefore might want more samples for the tidal parameters and less samples for the other parameters.
-However, due to the expense of the waveforms we must use the same set of waveforms for low and high frequencies.
-An LHD with its noncollapsing property bypasses this problem.
-
-We build an initial training set with 128 waveforms sampled with an LHD in our 5-dimensional parameter space given by $q\in[1/3, 1]$, $S_1, S_2 \in [-0.5, 0.5]$ and $\Lambda_1, \Lambda_2 \in [0, 5000]$.
-In addition, we find empirically that the GPR uncertainty estimates (Eq.~\eqref{eq:var} and Eq.~\eqref{eq:errorrms} below) are largest at the corners of the parameter space, so we also sample the 32 corners.
-We construct our initial surrogate with these 160 waveforms shown in Fig.~\ref{fig:LHD}.
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.49\textwidth]{trainingset3d.pdf}\\
-\includegraphics[width=0.49\textwidth]{trainingset2d.pdf}
-\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, S_1, S_2\}$ (top) and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom).
-The 160 blue circles were used for the initial training set (32 corners + 128 LHD points).
-The 200 red squares were generated using uncertainty sampling from the GPR error estimate (Eq.~\eqref{eq:errorrms}).
-The final 1000 green triangles were drawn uniformly from the parameter space.
-}
-\label{fig:LHD}
-\end{figure}
-
-%This method is sometimes referred to as uncertainty sampling [[cite]].
-However in our case, we do not update the hyperparameters after each new point is chosen, allowing us to choose the points then evaluate the waveforms in parallel.
-
-With this initial surrogate model, we can choose new training-set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion and update the surrogate with each new waveform.
-This method is sometimes referred to as uncertainty sampling [[cite]].
-The quantity that we use is the root-mean-squared error in the phase at the frequency nodes $Mf_j$ below $Mf=0.03$:
-\begin{equation}
-\epsilon_{\rm RMS}(\bx) = \sqrt{\frac{1}{N} \sum_{ \substack{j \\ MF_j \le 0.03} } [\sigma_{\Delta\Phi_j}(\bx)]^2},
-%\epsilon_{\rm RMS} = \sqrt{\frac{1}{N} \sum_{j=0}^{MF_j \le 0.03} (\sigma_{\Delta\Phi_j})^2},
-\label{eq:errorrms}
-\end{equation}
-where $\sigma_{\Delta\Phi_j}(\bx)$ is the GPR estimate (Eq.~\eqref{eq:var}) of the uncertainty in $\Delta\Phi(MF_j, \bx)$.
-There are many possible ways to construct a scalar that estimates overall waveform error.
-An alternative would be an approximation to the mismatch discussed in Sec.~\ref{sec:accuracy}.
-However, the quantity $\epsilon_{\rm RMS}(\bx)$ describes the most important quantity, phase, and is fast to compute.
-For an initial training set with $N$ samples, $\epsilon_{\rm RMS}(\bx)$ has $\sim N$ local maxima located in the voids between samples $\bx$.
-To find the global maximum, we use a basin jumping algorithm [[cite]] to avoid getting stuck in local maxima.
-
-For low starting frequencies, EOB waveforms are expensive enough that we would like a method to efficiently choose $N_{\rm new}$ waveforms and evaluate them in parallel.
-We note that if we hold the hyperparameters $\btheta$ fixed, the GPR error estimate (Eq.~\eqref{eq:var}), and therefore $\epsilon_{\rm RMS}(\bx)$, only depends on the samples $\bx$ and not on the waveform data.
-The algorithm for choosing the $N_{\rm new}$ new points is as follows.
-For $i = 1, \dots, N_{\rm new}$:
-\begin{enumerate}
-
-\item Construct the GPR error estimator (Eq.~\eqref{eq:errorrms}).
-In practice this can be done by specifying the samples $\bx$, the hyperparameters $\btheta$, and dummy data for $\by$ since Eq.~\eqref{eq:errorrms} does not depend on $\by$.
-
-\item Find $({\bm x}_{\rm max}, \epsilon_{\rm rms})$ that maximizes Eq.~\eqref{eq:errorrms} over the parameter space.
-
-\item Add ${\bm x}_{\rm max}$ to the list of samples: ${\bm x} = [{\bm x}, {\bm x}_{\rm max}]$.
-
-\end{enumerate}
-These $N_{\rm new}$ waveforms can now be evaluated in parallel. An updated surrogate can then be constructed with re-optimized hyperparameters $\btheta$ using the $N + N_{\rm new}$ waveforms.
-
-Fig.~\ref{fig:LHD} shows the parameters of the 200 new waveforms chosen by the uncertainty sampling method.
-We note that the edges and faces of the parameter space are more often chosen than the inner region.
-This is likely a consequence of the Mat\'{e}rn GPR kernel used.
-Ref.~[[]] found similar behavior when using a Mat\'{e}rn kernel instead of a squared-exponential kernel.
-Fig.~\ref{fig:uncsamp} shows the RMS error $\epsilon_{\rm rms}$ maximized over the parameters $\bx$ for each new sample added to the training set.
-Increasing the number of waveforms from 160 to 360 decreases the estimated RMS phase error by a factor of $\sim 4$.
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.49\textwidth]{uncertaintysampling.pdf}
-\caption{Estimated RMS phase error $\epsilon_{\rm RMS}$ (Eq.~\eqref{eq:errorrms}) maximized over parameter space $\bx$ as a function of the number of chosen parameters (blue curve).
-$\epsilon_{\rm RMS}$ depends on the chose samples $\bx_j$ and hyperparameters, not on the data, and can be calculated before generating the waveforms.
-Also shown for comparison are the maximum RMS phase errors at the nodes where $MF_j \le 0.03$ for the surrogate compared to a test set of 1000 waveforms.
-Blue circle: surrogate constructed from the 160 waveforms in the initial training set.
-Red square: surrogate after adding the 200 waveforms chosen with uncertainty sampling.
-Green triangle: final surrogate after also adding 1000 more training set waveforms drawn uniformly from parameter space.
-}
-\label{fig:uncsamp}
-\end{figure}
 
@@
 
+\subsection{Iterative construction of training set and surrogate}
+\label{sec:design}
+
+One of the main aims of this paper is to build a surrogate with as few waveform evaluations as possible.
+Using the freedom provided by GPR to sample waveforms at arbitrary locations, we try a set of designs more efficient than uniform grids.
+One such design is a Latin Hypercube Design (LHD)~\cite{McKayBeckmanConover1979}.
+An LHD with $N$ samples divides each of $d$ dimensions uniformly into $N$ grid points for a total of $N^d$ grid points.
+However, unlike a uniform grid, the $N$ values in each dimension are sampled exactly once instead of $N^{d-1}$ times.
+For an LHD there are $(N!)^d$ ways to choose these points, and we choose one randomly\footnote{There are ways to better choose an LHD.
+A standard requirement is that the LHD be space-filling, meaning that the points are as far apart as possible from each other.
+(See Ref.~\cite{Husslage2011} for a review of methods for optimizing the placement of samples.) One such definition of space filling is that the chosen locations maximize the minimum Euclidean distance between any two samples.
+We have not experimented with these alternatives here.}.
+
+An LHD has the property that the samples are non-collapsing; a projection onto a subspace is still an LHD and no points are repeated in any dimension.
+This avoids wasting samples when one of the parameters has much less of an influence than the other parameters.
+As an example, the tidal parameters have a minimal impact on the waveform at low frequencies, so it would not make sense to densely sample a grid of tidal parameters at low frequencies.
+However, during the merger the tidal parameters can be more important than the other parameters.
+One therefore might want more samples for the tidal parameters and less samples for the other parameters.
+However, due to the expense of the waveforms we must use the same set of waveforms for low and high frequencies.
+An LHD with its noncollapsing property bypasses this problem.
+
+We build an initial training set with 128 waveforms sampled with an LHD in our 5-dimensional parameter space given by $q\in[1/3, 1]$, $S_1, S_2 \in [-0.5, 0.5]$ and $\Lambda_1, \Lambda_2 \in [0, 5000]$.
+In addition, we find empirically that the GPR uncertainty estimates (Eq.~\eqref{eq:var} and Eq.~\eqref{eq:errorrms} below) are largest at the corners of the parameter space, so we also sample the 32 corners.
+We construct our initial surrogate with these 160 waveforms shown in Fig.~\ref{fig:LHD}.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.49\textwidth]{trainingset3d.pdf}\\
+\includegraphics[width=0.49\textwidth]{trainingset2d.pdf}
+\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, S_1, S_2\}$ (top) and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom).
+The 160 blue circles were used for the initial training set (32 corners + 128 LHD points).
+The 200 red squares were generated using uncertainty sampling from the GPR error estimate (Eq.~\eqref{eq:errorrms}).
+The final 1000 green triangles were drawn uniformly from the parameter space.
+}
+\label{fig:LHD}
+\end{figure}
+
+With this initial surrogate model, we can choose new training-set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion and update the surrogate with each new waveform.
+This method is sometimes referred to as uncertainty sampling~\cite{BrochuCoraDeFreitas2010}.
+The quantity that we use is the root-mean-squared error in the phase at the frequency nodes $Mf_j$ below $Mf=0.03$:
+\begin{equation}
+\epsilon_{\rm RMS}(\bx) = \sqrt{\frac{1}{N} \sum_{ \substack{j \\ MF_j \le 0.03} } [\sigma_{\Delta\Phi_j}(\bx)]^2},
+\label{eq:errorrms}
+\end{equation}
+where $\sigma_{\Delta\Phi_j}(\bx)$ is the GPR estimate (Eq.~\eqref{eq:var}) of the uncertainty in $\Delta\Phi(MF_j, \bx)$.
+There are many possible ways to construct a scalar that estimates overall waveform error.
+An alternative would be an approximation to the mismatch discussed in Sec.~\ref{sec:accuracy}.
+However, the quantity $\epsilon_{\rm RMS}(\bx)$ describes the most important quantity, phase, and is fast to compute.
+For an initial training set with $N$ samples, $\epsilon_{\rm RMS}(\bx)$ has $\sim N$ local maxima located in the voids between samples $\bx$.
+To find the global maximum, we use a basin hopping algorithm~\cite{WalesDoye1998, scipy:basinhopping} to avoid getting stuck in local maxima.
+
+For low starting frequencies, EOB waveforms are expensive enough that we would like a method to efficiently choose $N_{\rm new}$ waveforms and evaluate them in parallel.
+We note that if we hold the hyperparameters $\btheta$ fixed, the GPR error estimate (Eq.~\eqref{eq:var}), and therefore $\epsilon_{\rm RMS}(\bx)$, only depends on the samples $\bx$ and not on the waveform data.
+The algorithm for choosing the $N_{\rm new}$ new points is as follows.
+
+\begin{algorithm}[H]
+\caption{Uncertainty Sampling}
+\label{alg:uc}
+\begin{algorithmic}[1]
+\State {\bf Input:} Initial surrogate 
+\For{$i = 1 \to N_{\rm New}$}
+
+\State Construct the GPR error estimator (Eq.~\eqref{eq:errorrms}).
+In practice this can be done by specifying the samples $\bx$, the hyperparameters $\btheta$, and dummy data for $\by$ since Eq.~\eqref{eq:errorrms} does not depend on $\by$.
+
+\State Find $({\bm x}_{\rm max}, \epsilon_{\rm rms})$ that maximizes Eq.~\eqref{eq:errorrms} over the parameter space.
+
+\State Add ${\bm x}_{\rm max}$ to the list of samples: ${\bm x} = [{\bm x}, {\bm x}_{\rm max}]$.
+
+\EndFor
+\State {\bf Output:} Updated surrogate
+\end{algorithmic}
+\end{algorithm}
+
+%{\scriptsize
+%\begin{algorithm}[H]
+%\caption{Empirical Interpolation (EI) Method}
+%\label{alg:eim}
+%\begin{algorithmic}[1]
+%%\State {\bf Input:} $\{ \vec{e}_i \}_{i=1}^n$, $\{t_i\}_{i=1}^L$
+%\State {\bf Input:} $\{ e_i \}_{i=1}^n$, $t := \{t_i\}_{i=1}^L$
+%\vskip 10pt
+%\State $i = \text{argmax} | e_1(t) |$ (\text{argmax} returns the largest entry of its argument). 
+%\State Set $T_1 = t_i$
+%\For{$j = 2 \to n$} 
+%\State Build ${\cal I}_{j-1} [e_j](t)$ from (\ref{eq:empinterpX1})--(\ref{eq:empinterpX2})
+%\State $\vec{r} = {\cal I}_{j-1} [e_j](t) -e_j(t)$
+%\State $i = \text{argmax} |\vec{r}|$
+%\State $T_j = t_i$
+%\EndFor
+%\vskip 10pt
+%\State {\bf Output:} EI nodes $\{ T_i \}_{i=1}^n$,  interpolant operator ${\cal I}_n$
+%\end{algorithmic}
+%\end{algorithm}
+%}
+
+For $i = 1, \dots, N_{\rm new}$:
+\begin{enumerate}
+
+\item Construct the GPR error estimator $\epsilon_{\rm rms}(\bx)$ (Eq.~\eqref{eq:errorrms}).
+In practice this can be done by specifying the samples $\bx$, the hyperparameters $\btheta$, and dummy data for $\by$ since Eq.~\eqref{eq:errorrms} does not depend on $\by$.
+
+\item Find $\bx_{\rm max}$ that maximizes $\epsilon_{\rm rms}(\bx)$ over the parameter space.
+
+\item Add $\bx_{\rm max}$ to the list of samples: ${\bm x} \to [{\bm x}, {\bm x}_{\rm max}]$.
+
+\end{enumerate}
+These $N_{\rm new}$ waveforms can now be evaluated in parallel. An updated surrogate can then be constructed with re-optimized hyperparameters $\btheta$ using the $N + N_{\rm new}$ waveforms.
+
+Fig.~\ref{fig:LHD} shows the parameters of the 200 new waveforms chosen by the uncertainty sampling method.
+We note that the edges and faces of the parameter space are more often chosen than the inner region.
+This is likely a consequence of the Mat\'{e}rn GPR kernel used.
+Ref.~\cite{DoctorFarrHolz2017} found similar behavior when using a Mat\'{e}rn kernel instead of a squared exponential kernel.
+Fig.~\ref{fig:uncsamp} shows the RMS error $\epsilon_{\rm rms}$ maximized over the parameters $\bx$ for each new sample added to the training set.
+Increasing the number of waveforms from 160 to 360 decreases the estimated RMS phase error by a factor of $\sim 4$.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.49\textwidth]{uncertaintysampling.pdf}
+\caption{Estimated 1--$\sigma$ RMS phase error $\epsilon_{\rm RMS}(\bx)$ (Eq.~\eqref{eq:errorrms}) maximized over parameter space $\bx$ as a function of the number of chosen parameters (blue curve).
+Also shown for comparison are the maximum RMS phase errors at the nodes where $MF_j \le 0.03$ for the surrogate compared to a test set of 1000 waveforms.
+Blue circle: surrogate constructed from the 160 waveforms in the initial training set.
+Red square: surrogate after adding the 200 waveforms chosen with uncertainty sampling.
+Green triangle: final surrogate after also adding 1000 more training set waveforms drawn uniformly from parameter space.
+}
+\label{fig:uncsamp}
+\end{figure}
+
 %% ________________________________________________________
@@
 We also compare the evaluation time to several BNS waveform models that have been used for LIGO data analysis.
-We note that at $\sim 30$~Hz, the TEOBv4\_surrogate model is comparable to the SEOBNRv4\_ROM\_NRTidal and IMRPhenomD\_NRTidal models that were used in the analysis of the GW170817 event [[cite]].
+We note that at $\sim 30$~Hz, the SEOBNRv4T\_surrogate model is comparable to the SEOBNRv4\_ROM\_NRTidal and IMRPhenomD\_NRTidal models that were used in the analysis of the GW170817 event~\cite{DietrichBernuzziTichy2017}.
 At lower frequencies, the surrogate is faster than all BNS waveform models except for TaylorF2.
@@
 We have constructed a fast frequency-domain surrogate of the most accurate BNS waveform currently available for data analysis.
-This aligned-spin model, TEOBv4, incorporates the tidally induced $\ell=2$ and 3 multipole moments as well as the effect of dynamical tides as the excitation approaches the $\ell=2$ and 3 f-mode frequencies.
+This aligned-spin model, SEOBNRv4T, incorporates the tidally induced $\ell=2$ and 3 multipole moments as well as the effect of dynamical tides as the excitation approaches the $\ell=2$ and 3 f-mode frequencies.
 We have achieved mismatches of no more than $7 \times 10^{-4}$ and phase errors of $\lesssim 1$~rad up to the merger frequency.
@@
 For current parameter estimation codes, this is sufficient.
-However, one could further improve run times using reduced order quadrature [[cite]] which requires frequency-domain waveforms, or adaptively sampled likelihood evaluation [[cite]].
+However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} which requires frequency-domain waveforms, or multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017}.
+Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^4)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.
 
 The hierarchical method presented here, where we begin with the analytic TaylorF2 reference model then make a surrogate of the residual, can be used to further improve the waveform model.
-For example, one could build a surrogate for numerical BNS simulations using TEOBv4\_surrogate as the base model, and constructing a surrogate of the residual.
-Such a model would have the accuracy of EOB below $\sim 400$~Hz and the accuracy of numerical simulations for the last several cycles before merger (once numerical simulations can be proven to be more accurate than EOB [[cite Tim?
-Kyohei?]]).
+For example, one could build a surrogate for numerical BNS simulations using SEOBNRv4T\_surrogate as the base model, and constructing a surrogate of the residual.
+Such a model would have the accuracy of EOB below $\sim 400$~Hz and the accuracy of numerical simulations for the last several cycles before merger.
+Current state of the art NR simulations have phase errors of several tenths of a radian over the last $\sim 20$ gravitational-wave cycles~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}.
 Using GPR and uncertainty sampling discussed above, one could optimally choose the waveform parameters for the numerical simulations, and run them in parallel to build the training set.
-Because the difference between TEOBv4 and numerical BNS simulations is exceptionally small, one would not need a high fractional accuracy for a surrogate of the difference, and 10--100 waveforms may be sufficient.
+Because the difference between SEOBNRv4T and numerical BNS simulations is exceptionally small, one would not need a high fractional accuracy for a surrogate of the difference, and 10--100 waveforms may be sufficient.
 
 This model notably does not include two effects that can be important for systems with large spin.
-The first is the spin-induced quadrupole moment $Q$ that effects the waveform phase at 2PN order [[cite Poisson]].
+The first is the spin-induced quadrupole moment $Q$ that effects the waveform phase at 2PN order~\cite{Poisson1998}.
 The easiest way to correct for this effect is to simply add this 2PN term to the TaylorF2 reference waveform.
@@
 Once this effect is incorporated into the EOB model, one could easily rebuild the surrogate presented in this paper to incorporate this effect.
+The quadrupole moment can be approximated in terms of the tidal parameter $\Lambda$ using a so-called universal relation~\cite{YagiYunes2013}.
 
 The second effect is precession for non-aligned spins.
-Although none of the EOB models currently available include both tidal effects and precession, there are ways to boost this waveform model to a precessing frame, and approximately incorporate precession.
+Although none of the EOB models currently available include both tidal effects and precession, there are ways to boost this waveform model to a precessing frame and approximately incorporate precession.
 For example, Chatziioanno et al.
-have analytically solved the 2PN-accurate precession equations for generic spins.
-This only works for systems that do not include transitional precession, as is unlikely for low mass ratio BNS systems [[cite]].
+have analytically solved the 2PN-accurate precession equations for generic spins~\cite{ChatziioannouKleinCornish2017a, ChatziioannouKleinCornish2017b}, with the exception of transitional precession which is unlikely for low mass-ratio and spin BNS systems.
 Using stationary uniform asymptotics, they can also analytically Fourier-transform the solution.
-Importantly, one is free to specify the frequency-domain amplitude and phase evolution, such as the TEOBv4\_surrogate here, for the waveform in the co-precessing frame.
+Importantly, one is free to specify the frequency-domain amplitude and phase evolution, such as the SEOBNRv4T\_surrogate here, for the waveform in the co-precessing frame.
 This approach would provide a fast, accurate model for BNS systems with tides and generic spins.
@@
 BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the Institute for Nuclear Theory where many of the methods used here were discussed.
-We also thank Sylvain Marsat for correcting numerical instabilities in the original TEOBv4 code.
+We also thank Sylvain Marsat for correcting numerical instabilities in the original SEOBNRv4T code in \texttt{lalsuite}.
 The authors were supported by...

### 2017-12-04 paper/paper.tex
Rewrote introduction. Added some more references.

@@
 
-%For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.
-
 \end{abstract}
@@
   04.30.Db,   % gravitational wave generation and sources
-  % 04.40.Dg,   % Relativistic stars: structure, stability, and oscillations
+  04.40.Dg,   % Relativistic stars: structure, stability, and oscillations
   % 04.70.Bw,   % classical black holes
@@
 
-\red{[[What is possible with BNS systems:]]}
-
-The detection of the binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors, has demonstrated that the sky position, distance, masses, spins, and tidal parameters can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant~\cite{GW170817Hubble}, and understanding the structure of neutron stars (NSs) through tidal interactions~\cite{GW170817}.
+The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors, has demonstrated that the parameters of the binary can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant~\cite{GW170817Hubble}, and understanding the structure of neutron stars (NSs) through tidal interactions~\cite{GW170817}.
 Measuring these parameters accurately will also be necessary to properly interpret the gamma ray burst [[cite]] and optical afterglow [[cite]] that was found in conjunction.
@@
 
-\red{[[Why we need more accurate waveform models to do these things:]]}
+These results are obtained using Bayesian analyses that compare the data to parameterized waveform models.
+To avoid systematic errors highly accurate models are needed, and they must include all relevant physical effects.
+This includes descriptions of the point-particle interactions, spins, and tides.
+Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models.
+These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5 and 6PN order.
+Unfortunately, the lack of point-particle terms above 3.5PN order leads to systematic errors in the tidal parameters that are as large as the statistical errors [[Favata, YagiYunes, Wade]].
+Alternative models exist with more accurate representations of the point particle terms.
+However, they do not always simultaneously include the spin and tidal effects as well.
+Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$ [[Favata]].
+In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].
 
-These results are obtained using a Bayesian analysis that compares the data to parameterized waveform templates.
-To avoid systematic errors highly accurate waveform templates are needed.
-In addition, the Bayesian parameter estimation alogrithms (Markov chain Monte Carlo and Nested Sampling [[parallel algorithms are becoming available]]) typically perform several million sequential waveform evaluations, requiring waveform evaluation times to be less than $\mathcal{O}(1s)$ in order to run in less than a month for each event.
+One such alternative to the PN waveforms is the effective one body (EOB) formalism (discussed in section~\ref{sec:eob}).
+It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations.
+Until recently, however, no version of the EOB formalism included the effects of spin and tides in the same model.
+Recently two versions have become available that include both tidal effects and aligned spin.
+These models are being implemented within the LIGO Algorithm Library (LAL) [[cite]] module \texttt{lalsimulation} under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}.
+The \texttt{SEOBNRv4T} version [[cite]] used in this paper, includes the tidally induced $\ell=2$ and 3  multipole moments as well as the induced $\ell=2$ and 3 $f$-mode resonances, but does not yet include the spin induced quadrupole moment that can be important for large spins~\cite{Poisson, HarryHinderer2017}.
+This model also agrees with NR simulations of BNS systems to a level consistent with the numerical accuracy of the simulations, which is less than 1~rad during the last $\sim 10$ orbits before merger~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}.
+The alternative version, \texttt{TEOBResumS} [[cite]], uses a slightly different point-particle and spin prescription.
+And, in addition, includes the $\ell = 2, 3$, and 4 tidal effects and the spin induced quadrupole moment, but does not include the effect of dynamical tides from the induced $f$-mode resonances.
+It does, however, use a method for re-summing the expressions for tidal interactions.
+It also shows similar consistency with NR simulations [[Is this true?
+cite]].
 
-%The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters.
-These parameters, measured with Bayesian parameter estimation techniques, require highly accurate waveform models to avoid systematic errors.
-For example, post-Newtonian (PN) waveform models, which are truncated at the currently known 3.5PN order, lead to uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz [[cite]].
-This is approximately the same size as the contribution to the phase from tidal effects, and as a result, the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]].
-This also biases the recovered NS equation of state and radius [[cite]], as well as the masses and spins [[cite]].
+In addition to accuracy, useful waveform models must be fast enough to be used in Bayesian parameter estimation algorithms such as Markov chain Monte Carlo (MCMC) and Nested Sampling that currently require $10^7$--$10^8$ sequential waveform evaluations.
+Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event.
+Most time-domain waveform implementations such as the PN approximants and the EOB models involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain.
+For PN waveform models, this can take several seconds to minutes, while for the \texttt{SEOBNRv4T} model discussed here, evaluation time can be more than a day for a starting frequency of 10~Hz.
+Significant work has been done to optimize existing implementations of EOB for BBH systems~\cite{DevineEtienneMcWilliams2016}.
+However, it is not clear that sufficient speed ups can be achieved for the BNS model discussed here with dynamical tides.
 
-\red{[[The problems with the current waveform models/templates.]]}
+For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates.
+The first was the stationary phase approximation PN waveform model known as TaylorF2 [[cite]].
+Because of the known systematic errors that result from PN waveforms, two additional models were produced for comparison: \texttt{IMRPhenomD\_NRTides} and \texttt{SEOBNRv4\_ROM\_NRTides}.
+Both of these models used an identical prescription for tides presented in Ref.~\cite{DietrichBernuzziTichy2017}; at low frequencies, the tidal effect is that of TaylorF2.
+It then transitions to a fit to the tidal effect measured by a small set of numerical simulations with varying mass ratio and aligned-spins.
+The point-particle and spin prescriptions were different.
+For \texttt{IMRPhenomD\_NRTides}, the waveform is based on the phenomenological \texttt{IMRPhenomD} BBH model [[cite]].
+It uses TaylorF2 at low frequencies then transitions to fits to the EOB waveform at mid frequencies and fits to NR simulations at high frequencies.
+The \texttt{SEOBNRv4\_ROM\_NRTides} model is based on the fast surrogate model \texttt{SEOBNRv4\_ROM} [[cite]] to the EOB waveform \texttt{SEOBNRv4} for BBH systems [[cite]].
+Both of these models are improvements to \texttt{TaylorF2}, but do not incorporate all known physical effects in a consistent way.
+We therefore build a fast surrogate that reproduces as accurately as possible \texttt{SEOBNRv4T} in the frequency domain.
 
-To date, there are a limited number BNS waveform models that include spin and tides, and that are fast enough for standard parameter estimation.  
-TaylorF2: (not so good point particle. okay tidal terms)
-For example, post-Newtonian (PN) waveform models, which are truncated at the currently known 3.5PN order, lead to uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz [[cite]].
-This is approximately the same size as the contribution to the phase from tidal effects, and as a result, the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]].
-This also biases the recovered NS equation of state and radius [[cite]], as well as the masses and spins [[cite]].
+Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of waveform models.
+The essential idea is to construct an interpolating function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$.
+Previous works have focused on efficiently representing the parameter space of waveforms using a reduced basis of orthonormal functions whose linear combination can reproduce any waveform in the parameter space to some desired accuracy.
+The two most common approaches have been singular value decomposition (SVD)~\cite{Puerrer2014, Puerrer2015} or a greedy method that selects a subset of the training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
+In this work, instead of using a reduced basis to efficiently represent the EOB waveform as a function of frequency, we will instead use the fact that the frequency-domain waveform is approximately known analytically via the TaylorF2 waveform.
+We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform.
+Because this residual is small, extremely high accuracy is not required, and we simply choose to represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.
 
-\*\_NRTides: (better/good point particle. better tidal terms)
-Initial parameter estimation results for the first BNS event, GW170817 using a variety of waveform templates (TaylorF2, SEOBNRv4\_ROM\_NRTides, and IMRPhenomD\_NRTides) showed that there are noticeable differences in the measured parameters between the different waveform templates [[cite]], and in particular the tidal parameter.
-[[NR\_Tides is a fit to NR waveforms at high frequency and has only been fit to a limited number of low spin, low tidal parameter NR simulations.]]
-
-TEOBResumS:
-
-\red{[[What is good about the new EOB model:]]}
-
-The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms.
-Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad~\cite{HindererLackeyLangRead2010}.
-Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]].
-
-Another model
-
-\red{[[Problem with EOB (speed). Attempts to speed it up with optimization. Alternative using surrogates.]]}
-
-Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]].
-Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable.
-Significant work has been done to optimize existing implementations of EOB for BBH systems [[EtienneMcWilliams]].
-However, it is not clear that these optimizations can be readily adapted for the BNS model discussed here with dynamical tides.
-We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.
-
-Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$.
-The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce any waveform in the parameter space to some desired accuracy.
-This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions.
-In previous work, this reduced basis has been constructed either with singular value decomposition (SVD)~\cite{Puerrer2014, Puerrer2015} or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
-
-\red{[[Overview of surrogates. Typical methods that have been used. Their problems. What you are doing that is new.]]}
-
-%The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms.
-Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad.
-Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]].
-Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]].
-Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable.
-Significant work has been done to optimize existing implementations of EOB for BBH systems [[EtienneMcWilliams]].
-However, it is not clear that these optimizations can be readily adapted for the BNS model discussed here with dynamical tides.
-We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.
-
-%Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$.
-The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce any waveform in the parameter space to some desired accuracy.
-This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions.
-In previous work, this reduced basis has been constructed either with singular value decomposition (SVD) [[Michael's papers]] or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization [[All the other papers]].
-
-Once this reduced basis is built, one can interpolate each of the coefficients $c_i(\bx)$ as a function of the waveform parameters $\bx$.
-This method has been used for frequency-domain surrogates of aligned spin BBH waveforms [[Michael's papers]].
-Alternatively, one can use the empirical interpolation method (EIM) [[cite]] which re-expresses the linear combination of orthonormal basis functions in terms of a sum of empirical interpolating functions $B_j(t)$ times the value of the waveform $h(T_j; \bx)$ at empirical nodes $T_j$.
-One then interpolates the waveform $h(T_j; \bx)$ as a function of $\bx$ at each node $T_j$.
-This method has been used for time domain-waveforms for nonspinning [[cite]], aligned-spin [[cite]], and precessing [[cite]] BBH waveforms as well as nonspinning BNS waveforms with tidal interactions [[cite]], and frequency-domain, precessing waveforms [[cite]].
-
-Previous work on surrogate models has focused mainly on parameter spaces $\bx$ with $\le 3$~dimensions.
-This allows one to approximate the coefficients $c_i(\bx)$ or waveform values $h(T_j; \bx)$ as functions of $\bx$ using simple fitting functions [[cite]] or standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}.
-These interpolation techniques typically require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
-For the 5-dimensional problem here, this is somewhat unfeasable.
+This solves the problem of efficiently representing waveforms as a function of frequency, but leaves the question of interpolating as a function of waveform parameters $\bx$ unanswered.
+Previous work on surrogate models have focused mainly on parameter spaces with three or fewer dimensions.
+This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}.
+These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
+For the problem here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation will be unfeasable.
 Recent work on optimally choosing waveforms for both EOB~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.
 
-In this paper we focus on breaking this curse of dimensionality by using an interpolation method known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid.
+In this paper we focus on an interpolation method known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid.
 Importantly, GPR also provides a convenient estimate of its own uncertainty.
@@
 Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of eccentric BBH mergers~\cite{HuertaMooreKumar2017}.
-GPR has also been used recently in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016}.
+GPR has also been used in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016} and to estimate noise power spectral densities [[cite]].
 
-In this paper, we will develop methods to construct a surrogate model for the inspiral through merger of an aligned spin BNS waveform with tidal interactions using the EOB tidal description described in Hinderer et al.~[[cite1, cite2]].
-This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS waveform using a slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~\cite{Bernuzzi:2014owa}.
+This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi et al.~\cite{Bernuzzi:2014owa}.
 Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation.
-It also enables additional techniques such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} to accelerate likelihood evaluation.
-
-\red{[[Obligitory summary:]]}
+It also enables additional techniques such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} and multi-banding~\cite{VinciguerraVeitchMandel2017} to accelerate likelihood evaluation.
+The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.
 
@@
 
-\red{[[In the introduction you should emphasize the parts of this paper that are new to the GW field: Applying GPR to a high dimensional parameter space, showing that you can choose several new training set samples then calculate them in parallel without reconstructing the reduced basis.
-Building a surrogate for the residual to reduce the required interpolation accuracy.
-The first BNS waveform that realistically tapers to zero amplitude.
-Active learning method that readily extends to higher dimensions.
-By attaching to TaylorF2, the surrogate is valid for all frequencies.]]}
-
-\red{[[We aim to construct a sufficiently accurate waveform surrogate using as few waveform evaluations as possible, to show how one could use NR simulations/go to higher dimensions.]]}
-
 %% ________________________________________________________
 \section{Aligned-spin, dynamical tides EOB model}
-
-\red{[[Say somewhere that this model is based off of what is known as SEOBNRv4 (and/or 2?) in lalsuite.]]}
-
-\red{[[The inspiral part is exactly as described in previous papers. The post-merger part is new to this paper. If this is true, we should write that somewhere.]]}
+\label{sec:eob}
 
@@
 
-We note that, to date, this is the only tidal EOB model that includes the effect of spin \red{[[Sebastiano also has a model that includes spin and tides.]]}\blue{[[Which is the Reference?]]}.
+%We note that, to date, this is the only tidal EOB model that includes the effect of spin \red{[[Sebastiano also has a model that includes spin and tides.]]}\blue{[[Which is the Reference?]]}.
+
 More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.\footnote{This model is implemented in the LIGO Algorithm Library as SEOBNRv4.} The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning.
@@
 Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.
-\red{[[Show how $\Delta_u$ enters the Hamiltonian.]]}\blue{[[There's no easy/illuminating way to explain how this potential enters the Hamiltonian]]}.
 In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
@@
 A set of NR simulations with a fixed mass ratio and tidal parameters but different total masses (and therefore different EOSs in order to keep the tidal parameters fixed) will be needed to determine the time when this approximation breaks down.
-
-%%% ________________________________________________________
-%\section{Aligned-spin, dynamical tides EOB model}
-%Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects.
-State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries.
-The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}.
-This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter --- typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs).
-The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case.
-In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides.
-Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star.
-Because of binary motion, the forcing tidal field varies at the orbital frequency.
-Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode.
-The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field.
-Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.
-%
-%In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
-%\begin{equation}
-%\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
-%\end{equation}
-%where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides.
-We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below.
-Let $A$ and $B$ be the body labels.
-Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.
-Explicitly, including quadrupolar and octupolar tidal effects, we have
-%\begin{widetext}
-%\begin{align}
-% \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
-% &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
-%\end{align}
-%\end{widetext}
-%where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
-Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies.
-In particular, the dimensionless dynamical tidal polarizability reads
-%\begin{equation}
-%\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\,,
-%\end{equation}
-%where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
-%\begin{equation}
-%\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
-%\end{equation}
-%with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star.
-Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole angular frequency, $\omega_{0\ell}^{A,B}$.
-%
-%Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
-%\begin{equation}
-%h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{insp-plunge, pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
-%\end{equation} 
-%where $h_{\ell m}^{\textrm{insp-plunge, pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs.
-(A14)-(A17) of Ref.~\cite{Damour:2012yf}.
-Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
-%
-%The inspiral-plunge orbital dynamics is obtained by numerically integrating Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s: see Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}.
-The numerical integration starts from quasicircular initial conditions that are built following the prescription of Ref.~\cite{Buonanno:2005xu}.
-We exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics.
-When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case.
-%
-%Let us now discuss the construction of the full waveform.
-In this work we only consider the dominant mode, $(\ell,m)=(2,2)$.
-The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$.
-Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when $|h_{22}^{\textrm{insp-plunge}}|$ is maximum.
-Let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum.
-Then we put
-%\begin{equation}
-%t_{\textrm{match}}^{22} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.
-%\end{equation}
-%We model the post-merger GW emission by a signal that tapers down in amplitude and has a constant GW frequency.
-In particular, we employ the amplitude of the merger-ringdown of the underlying BBH model~\cite{Bohe:2016gbl}, using as inputs the mass ratio and spins of the BNS/NSBH of interest.
-Explicitly,
-%\begin{equation}
-%|h_{22}^{\textrm{post-mrg}}(t)|=\nu \tilde{A}_{22}(t)\exp{\left[-(t-t^{22}_{\textrm{match}})/\tau_{220}\right]}\,,
-%\end{equation}
-%where $\nu=M_A M_B/M^2$ is the symmetric mass ratio, $\tilde{A}_{22}(t)$ is a monotonic and bounded function given in Eq.~(2.12) of Ref.~\cite{Bohe:2016gbl}, and $\tau_{220}$ is the decay time of the least-damped quasinormal mode of the remnant BH that forms after the BBH merger.
-This choice is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}.
-The constant value of the GW frequency is taken to be that of $h_{22}^{\textrm{insp-plunge}}$ at time $t=t_{\textrm{match}}^{22}$, so that the post-merger phase evolution reads
-%\begin{align}
-%\phi_{22}^{\textrm{post-mrg}}(t)&=\omega_{22}^{\textrm{insp-plunge}}(t_{\textrm{match}}^{22})(t-t_{\textrm{match}}^{22})\nonumber\\
-%&+\phi_{22}^{\textrm{inspiral-plunge}}(t_{\textrm{match}}^{22})\,
-%\end{align}
-%where $\phi_{22}^{\textrm{inspiral-plunge}}$ is the phase of $h_{22}^{\textrm{insp-plunge}}$.
-A more sophisticated, NR-informed, time-domain model of the post-merger emission will be part of future investigations.
-Finally, the full waveform reads
-%\begin{align}
-%h_{22}(t) &= h_{22}^{\textrm{insp-plunge}}(t)\, \theta\left(t_{\textrm{match}}^{22}-t\right)\nonumber\\
-%&+h_{22}^{\textrm{post-mrg}}(t)\, \theta\left(t-t_{\textrm{match}}^{22}\right)\,.
-%\end{align}
-%
-%Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state, finding that the model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state.
-Similar results hold for the tidal EOB model of Ref.~\cite{Bernuzzi:2014owa}, whose radial potential includes tidal effects in a form that is inspired by first-order gravitational-self-force (GSF) calculations of tidal invariants in the small-mass-ratio limit.
-Both the EOB model with dynamical tides and the GSF-inspired one are significantly more faithful to NR than the adiabatic tidal EOB model of Refs.~\cite{Damour:2009wj,Bini:2012gu}.
-%
-%We note that, to date, this is the only tidal EOB model that includes the effect of spin.
-More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.
-The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning.
-This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled.
-The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.
-%
-%As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless components of the spins $\boldsymbol{S}_{A,B}$ along the Newtonian angular momentum ${\boldsymbol{L}}_N$, $-1 \leq \chi_{A,B}=(\boldsymbol{S}_{A,B}\cdot \hat{\boldsymbol{L}}_N)/M_{A,B}^2 \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
-To reduce the dimensionality of the intrinsic parameter space, we exploit universal (i.e., equation-of-state-independent) relations that are available in the literature.
-In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} with the coefficients from the first row of Table I therein expresses $\Lambda_3^{A,B}$ as a function of $\Lambda_2^{A,B}$ with an accuracy to within about 10\%, depending on the equation of state and the value of $\Lambda_2^{A,B}$.
-While this relation has been fitted to NR data only for $1 \leq \Lambda_2^{A,B}\leq10^6$, we use it also outside such range.
-However, this formula is divergent at small values of $\Lambda_2^{A,B}$, thus in such regime we replace it with a polynomial function that vanishes at 0, equals the universal relation at $\Lambda_2^{A,B}=10^{-2}$, and best-fits it for $10^{-5}\leq\Lambda_2^{A,B}\leq10^{-2}$ Explicitly, putting $\zeta_{A,B}=\frac{\log\Lambda^{A,B}_2}{10}$,
-%\begin{widetext}
-%\begin{equation}
-%\Lambda_3^{A,B}=\left\{\begin{array}{ll}
-%\Lambda_2^{A,B}\left[0.441 - 34.6\,\Lambda_2^{A,B} + 1760(\Lambda_2^{A,B})^2\right]\,,&\textrm{if}\quad0\leq\Lambda_2^{A,B}\leq10^{-2}\\
-%&\\
-%\exp{\left[-1.15 +11.8\,\zeta_{A,B}+2.51\,\zeta_{A,B}^2-1.31\,\zeta_{A,B}^3+0.252\,\zeta_{A,B}^4\right]}\,,&\textrm{otherwise}\,.
-%\end{array}\right.
-%\end{equation}
-%\end{widetext}
-%Eq.
-(3.5) of Ref.~\cite{Chan:2014kua} with the coefficients from the first column of Table I therein gives $\omega_{02}^{A,B}$ as a function of $\Lambda_2^{A,B}$ to within a few percent, while the same equation but with the coefficients from the second column of Table I therein gives $\omega_{03}^{A,B}$ as a function of $\Lambda_3^{A,B}$ to within a few percent.
-Putting $\xi_{A,B}=\log\Lambda_2^{A,B}$, over the fitting range used by Ref.~\cite{Chan:2014kua} (i.e., $0\leq \xi_{A,B} \leq 9$), we set
-%\begin{align}
-%M_{A,B}\omega_{02}^{A,B}&=
-%0.182 -6.836\times10^{-3}\xi_{A,B}\nonumber\\
-%&-4.196\times10^{-3}\xi_{A,B}^2+5.215\times10^{-4}\xi_{A,B}^3\nonumber\\
-%&-1.857\times10^{-5}\xi_{A,B}^4\,;\label{URomega2}
-%\end{align}
-%outside this range, we set $M_{A,B}\omega_{02}^{A,B}=0.182$ if $\xi_{A,B} < 1$ and $M_{A,B}\omega_{02}^{A,B}=0.161$ if $\xi_{A,B} >9$, that is Eq.~(\ref{URomega2}) evaluated for $\xi_{A,B}=1$ and $\xi_{A,B}=9$, respectively.
-Moreover, putting $\Upsilon_{A,B}=\log\Lambda_3^{A,B}$, over the fitting range used by Ref.~\cite{Chan:2014kua} (i.e., $-1 \leq \Upsilon_{A,B} \leq 10$), we set
-%\begin{align}
-%M_{A,B}\omega_{03}^{A,B}&=
-%\,0.2245 -1.5\times10^{-2}\Upsilon_{A,B}\nonumber\\
-%&-1.412\times10^{-3}\Upsilon_{A,B}^2+1.832\times10^{-4}\Upsilon_{A,B}^3\nonumber\\
-%&-5.561\times10^{-6}\Upsilon_{A,B}^4\,;\label{URomega3}
-%\end{align}
-%outside this range, we set $M_{A,B}\omega_{03}^{A,B}=0.2379$ if $\Upsilon_{A,B} < -1$ and $M_{A,B}\omega_{03}^{A,B}=0.1165$ if $\Upsilon_{A,B} >10$, that is Eq.~(\ref{URomega3}) evaluated for $\Upsilon_{A,B}=-1$ and $\Upsilon_{A,B}=10$, respectively.
-%Thanks to these relations, the waveform only depends on 5 intrinsic parameters, which we denote as $\boldsymbol{x}=\{q, \chi_A,\chi_B, \Lambda_2^A,\Lambda_2^B\}$.
-We note that this list does not include the total mass $M$ of the system.
-Of course, the point-mass part of the model is scale-invariant.
-As to the inspiral-plunge phase, in this model tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$, and $\Lambda_2^{A,B}$, which is inherent to the post-Newtonian analysis of tidal effects~\cite{Flanagan:2007ix}.
-Finally, the simple model of the post-merger signal that we employ does not depend on $M$ either, although a more sophisticated description will have to include such dependence to account for the phenomenology that is observed in NR simulations of this phase.
-%
-%\red{A notable effect that has not been included is the spin induced quadrupole moment of each NS [[Poisson etc.]].
-The impact on detection and parameter estimation is discussed in [[HarryHinderer]].}
 

### 2017-12-05 src/TEOB-LAL-implementation/LALSimIMR.h
Update with latest code

@@ -212,8 +212,8 @@ int XLALSimIMRSEOBNRv4ROMFrequencyOfTime(REAL8 *frequency, REAL8 t, REAL8 m1SI, 
 
 /* in module LALSimIMRTEOBv4ROM.c */
 
-int XLALSimIMRTEOBv4ROM(struct tagCOMPLEX16FrequencySeries **hptilde, struct tagCOMPLEX16FrequencySeries **hctilde, REAL8 phiRef, REAL8 deltaF, REAL8 fLow, REAL8 fHigh, REAL8 fRef, REAL8 distance, REAL8 inclination, REAL8 m1SI, REAL8 m2SI, REAL8 chi1, REAL8 chi2, REAL8 lambda1, REAL8 lambda2);
-int XLALSimIMRTEOBv4ROMFrequencySequence(struct tagCOMPLEX16FrequencySeries **hptilde, struct tagCOMPLEX16FrequencySeries **hctilde, const REAL8Sequence *freqs, REAL8 phiRef, REAL8 fRef, REAL8 distance, REAL8 inclination, REAL8 m1SI, REAL8 m2SI, REAL8 chi1, REAL8 chi2, REAL8 lambda1, REAL8 lambda2);
+int XLALSimIMRSEOBNRv4TSurrogate(struct tagCOMPLEX16FrequencySeries **hptilde, struct tagCOMPLEX16FrequencySeries **hctilde, REAL8 phiRef, REAL8 deltaF, REAL8 fLow, REAL8 fHigh, REAL8 fRef, REAL8 distance, REAL8 inclination, REAL8 m1SI, REAL8 m2SI, REAL8 chi1, REAL8 chi2, REAL8 lambda1, REAL8 lambda2);
+int XLALSimIMRSEOBNRv4TSurrogateFrequencySequence(struct tagCOMPLEX16FrequencySeries **hptilde, struct tagCOMPLEX16FrequencySeries **hctilde, const REAL8Sequence *freqs, REAL8 phiRef, REAL8 fRef, REAL8 distance, REAL8 inclination, REAL8 m1SI, REAL8 m2SI, REAL8 chi1, REAL8 chi2, REAL8 lambda1, REAL8 lambda2);
 
 
 /* in module LALSimIMRPSpinInspiralRD.c */

### 2017-12-05 src/TEOB-LAL-implementation/LALSimIMRSEOBNRv4TSurrogate.c

@@ -1,0 +1,1178 @@
+/*
+ *  Copyright (C) 2017 Michael Puerrer, Ben Lackey
+ *  Reduced Order Model for SEOBNRv4T spin-aligned tidal effective-one-body model
+ *
+ *  This program is free software; you can redistribute it and/or modify
+ *  it under the terms of the GNU General Public License as published by
+ *  the Free Software Foundation; either version 2 of the License, or
+ *  (at your option) any later version.
+ *
+ *  This program is distributed in the hope that it will be useful,
+ *  but WITHOUT ANY WARRANTY; without even the implied warranty of
+ *  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+ *  GNU General Public License for more details.
+ *
+ *  You should have received a copy of the GNU General Public License
+ *  along with with program; see the file COPYING. If not, write to the
+ *  Free Software Foundation, Inc., 59 Temple Place, Suite 330, Boston,
+ *  MA  02111-1307  USA
+ */
+
+#ifdef __GNUC__
+#define UNUSED __attribute__ ((unused))
+#else
+#define UNUSED
+#endif
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <math.h>
+#include <complex.h>
+#include <time.h>
+#include <stdbool.h>
+#include <alloca.h>
+#include <string.h>
+#include <libgen.h>
+#include <assert.h>
+
+#include <gsl/gsl_errno.h>
+#include <gsl/gsl_bspline.h>
+#include <gsl/gsl_blas.h>
+#include <gsl/gsl_min.h>
+#include <gsl/gsl_spline.h>
+#include <gsl/gsl_poly.h>
+#include <lal/Units.h>
+#include <lal/SeqFactories.h>
+#include <lal/LALConstants.h>
+#include <lal/XLALError.h>
+#include <lal/FrequencySeries.h>
+#include <lal/Date.h>
+#include <lal/StringInput.h>
+#include <lal/Sequence.h>
+#include <lal/LALStdio.h>
+#include <lal/FileIO.h>
+
+
+#ifdef LAL_HDF5_ENABLED
+#include <lal/H5FileIO.h>
+static const char SurDataHDF5[] = "SEOBNRv4T_surrogate_v1.0.0.hdf5";
+static const INT4 SurDataHDF5_VERSION_MAJOR = 1;
+static const INT4 SurDataHDF5_VERSION_MINOR = 0;
+static const INT4 SurDataHDF5_VERSION_MICRO = 0;
[…]

### 2017-12-05 src/TEOB-LAL-implementation/LALSimInspiral.h

@@ -1,0 +1,823 @@
+/*
+ * Copyright (C) 2008 J. Creighton, S. Fairhurst, B. Krishnan, L. Santamaria, E. Ochsner, C. Pankow, 2104 A. Klein
+ *
+ *  This program is free software; you can redistribute it and/or modify
+ *  it under the terms of the GNU General Public License as published by
+ *  the Free Software Foundation; either version 2 of the License, or
+ *  (at your option) any later version.
+ *
+ *  This program is distributed in the hope that it will be useful,
+ *  but WITHOUT ANY WARRANTY; without even the implied warranty of
+ *  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+ *  GNU General Public License for more details.
+ *
+ *  You should have received a copy of the GNU General Public License
+ *  along with with program; see the file COPYING. If not, write to the
+ *  Free Software Foundation, Inc., 59 Temple Place, Suite 330, Boston,
+ *  MA  02111-1307  USA
+ */
+
+#ifndef _LALSIMINSPIRAL_H
+#define _LALSIMINSPIRAL_H
+
+#include <lal/LALDatatypes.h>
+#include <lal/LALSimSphHarmSeries.h>
+#include <lal/LALSimInspiralTestGRParams.h>
+#include <lal/LALSimInspiralWaveformFlags.h>
+#include <lal/LALSimInspiralWaveformParams.h>
+
+#include <gsl/gsl_matrix.h>
+
+#if defined(__cplusplus)
+extern "C" {
+#elif 0
+} /* so that editors will match preceding brace */
+#endif
+
+/**
+ * @addtogroup lalsimulation_inspiral
+ * @details
+ *
+ * Various types of inspiral approximants are supported for producing
+ * waveforms in the time- or frequency-domains.  The high-level routines
+ * for generating simulated inspiral waveforms are
+ * XLALSimInspiralChooseTDWaveform() (for time-domain waveforms) and
+ * XLALSimInspiralChooseFDWaveform() (for frequency-domain waveforms).
+ * The following examples show their basic usage.
+ *
+ * Generate a time-domain waveform:
+ * @code
+ * #include <lal/LALSimInspiral.h>
+ * #include <lal/LALConstants.h>
+ * ...
+ * double m1 = 1.4 * LAL_MSUN_SI; // mass of body 1
+ * double m2 = 1.4 * LAL_MSUN_SI; // mass of body 2
+ * double S1x = 0.0;              // x-component of dimensionless spin of body 1
[…]

### 2017-12-05 src/TEOB-LAL-implementation/fix_TEOBv4_sur.py

@@ -1,0 +1,20 @@
+import h5py
+import numpy as np
+
+fp = h5py.File('SEOBNRv4T_surrogate_v1.0.0.hdf5', 'r+')
+
+del fp['lambda1_bounds']
+del fp['lambda2_bounds']
+
+fp.create_dataset('lambda1_bounds', data=np.array([0.0, 5000.0]))
+fp.create_dataset('lambda2_bounds', data=np.array([0.0, 5000.0]))
+
+fp.attrs['Creator'] = "Ben Lackey, Michael Puerrer, Andrea Taracchini"
+fp.attrs['Email'] = "Ben.Lackey@ligo.org, Michael.Puerrer@ligo.org"
+fp.attrs['Description'] = fp.attrs['description']
+
+fp.attrs.create('version_major', 1, dtype='i4')
+fp.attrs.create('version_minor', 0, dtype='i4')
+fp.attrs.create('version_micro', 0, dtype='i4')
+
+fp.close()

### 2017-12-05 src/TEOB-LAL-implementation/LALSimIMRSEOBNRv4TSurrogate.c
Further cleanup

@@ -74,11 +74,6 @@ static const INT4 SurDataHDF5_VERSION_MICRO = 0;
 #endif
 
 
-// TODO:
-// * remove all debugging code: fprintf statements
-// * add more checks and tests
-// * take care of all FIXMEs
-
 
 #ifdef LAL_PTHREAD_LOCK
 static pthread_once_t Surrogate_is_initialized = PTHREAD_ONCE_INIT;
@@ -92,8 +87,6 @@ struct tagSurrogatedata_submodel
   gsl_matrix *hyp_phi;           // GP hyperparameters for dephasing
   gsl_matrix *kinv_dot_y_amp;    // kinv_dot_y for log amplitude
   gsl_matrix *kinv_dot_y_phi;    // kinv_dot_y for dephasing
-  gsl_matrix *B_amp;             // Reduced basis for log amplitude
-  gsl_matrix *B_phi;             // Reduced basis for dephasing
   gsl_matrix *x_train;           // Training points
   gsl_vector *mf_amp;            // location of spline nodes for log amplitude
   gsl_vector *mf_phi;            // location of spline nodes for dephasing
@@ -265,11 +258,6 @@ double kernel(
   double sigma_n = gsl_vector_get(hyperparams, hyperparams->size-1);
   gsl_vector ls = gsl_vector_subvector(hyperparams, 1, hyperparams->size-2).vector;
 
-  // fprintf(stderr, "** kernel **\n");
-  // fprintf(stderr, "x1 = [%g, %g, %g, %g, %g]\n", gsl_vector_get(x1, 0), gsl_vector_get(x1, 1), gsl_vector_get(x1, 2), gsl_vector_get(x1, 3), gsl_vector_get(x1, 4));
-  // fprintf(stderr, "x2 = [%g, %g, %g, %g, %g]\n", gsl_vector_get(x2, 0), gsl_vector_get(x2, 1), gsl_vector_get(x2, 2), gsl_vector_get(x2, 3), gsl_vector_get(x2, 4));
-  // fprintf(stderr, "ls = [%g, %g, %g, %g, %g]\n", gsl_vector_get(&ls, 0), gsl_vector_get(&ls, 1), gsl_vector_get(&ls, 2), gsl_vector_get(&ls, 3), gsl_vector_get(&ls, 4));
-
   XLAL_CHECK((x1->size == x2->size) && (x1->size == ls.size), XLAL_EDIMS,
[…]

### 2017-12-15 paper/paper.tex
Rewrote Section II. Other minor edits.

@@
 We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with $\sim 1000$ waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods.
+\red{[[Mismatch is ...]]}
 
@@
 %% ________________________________________________________
+\red{Choose point-mass or point-particle.}
 \section{Introduction}
@@
 These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5 and 6PN order.
-Unfortunately, the lack of point-particle terms above 3.5PN order leads to systematic errors in the tidal parameters that are as large as the statistical errors [[Favata, YagiYunes, Wade]].
+Unfortunately, the lack of point-particle terms above 3.5PN order leads to systematic errors in the tidal parameters that are as large as the statistical errors~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
 Alternative models exist with more accurate representations of the point particle terms.
 However, they do not always simultaneously include the spin and tidal effects as well.
-Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$ [[Favata]].
+Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$~\cite{Favata2014}.
 In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].
 
-One such alternative to the PN waveforms is the effective one body (EOB) formalism (discussed in section~\ref{sec:eob}).
+One such alternative to the PN waveforms is the effective one body (EOB) formalism discussed in section~\ref{sec:eob}.
 It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations.
@@
 Recently two versions have become available that include both tidal effects and aligned spin.
-These models are being implemented within the LIGO Algorithm Library (LAL) [[cite]] module \texttt{lalsimulation} under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}.
-The \texttt{SEOBNRv4T} version [[cite]] used in this paper, includes the tidally induced $\ell=2$ and 3  multipole moments as well as the induced $\ell=2$ and 3 $f$-mode resonances, but does not yet include the spin induced quadrupole moment that can be important for large spins~\cite{Poisson, HarryHinderer2017}.
-This model also agrees with NR simulations of BNS systems to a level consistent with the numerical accuracy of the simulations, which is less than 1~rad during the last $\sim 10$ orbits before merger~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}.
-The alternative version, \texttt{TEOBResumS} [[cite]], uses a slightly different point-particle and spin prescription.
+These models are being implemented within the LIGO Algorithm Library (\texttt{LAL})~\cite{lal} under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}.
+The \texttt{SEOBNRv4T} model~\cite{Hinderer:2016eia,Steinhoff:2016rfi} used in this paper, includes the tidally induced $\ell=2$ and 3  multipole moments as well as the induced $\ell=2$ and 3 $f$-mode resonances, but does not yet include the spin induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2017}.
+This model also agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}.
+The alternative model, \texttt{TEOBResumS}~\cite{Nagar2017}, uses a slightly different point-particle and spin prescription.
+For tidal interactions, the model includes the $\ell = 2, 3$, and 4 terms and uses a method for re-summing these expressions.
+It does not include the effect of dynamical tides from the induced $f$-mode resonances, but does include the spin induced quadrupole moment.
+
+%The alternative version, \texttt{TEOBResumS}~\cite{Nagar2017}, uses a slightly different point-particle and spin prescription.
 And, in addition, includes the $\ell = 2, 3$, and 4 tidal effects and the spin induced quadrupole moment, but does not include the effect of dynamical tides from the induced $f$-mode resonances.
 It does, however, use a method for re-summing the expressions for tidal interactions.
-It also shows similar consistency with NR simulations [[Is this true?
-cite]].
 
@@
 Most time-domain waveform implementations such as the PN approximants and the EOB models involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain.
-For PN waveform models, this can take several seconds to minutes, while for the \texttt{SEOBNRv4T} model discussed here, evaluation time can be more than a day for a starting frequency of 10~Hz.
-Significant work has been done to optimize existing implementations of EOB for BBH systems~\cite{DevineEtienneMcWilliams2016}.
+For PN waveform models, this can take several seconds for a starting frequency of 10~Hz, while for the \texttt{SEOBNRv4T} model discussed here, the equivalent evaluation time can be a day.
+Significant work has been done to optimize existing implementations of EOB for binary black hole (BBH) systems by a factor of several hundred~\cite{DevineEtienneMcWilliams2016}.
 However, it is not clear that sufficient speed ups can be achieved for the BNS model discussed here with dynamical tides.
@@
 For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates.
-The first was the stationary phase approximation PN waveform model known as TaylorF2 [[cite]].
+The first was the stationary phase approximation PN waveform model known as \texttt{TaylorF2}~\cite{BuonannoIyerOchsner2009}.
 Because of the known systematic errors that result from PN waveforms, two additional models were produced for comparison: \texttt{IMRPhenomD\_NRTides} and \texttt{SEOBNRv4\_ROM\_NRTides}.
-Both of these models used an identical prescription for tides presented in Ref.~\cite{DietrichBernuzziTichy2017}; at low frequencies, the tidal effect is that of TaylorF2.
+Both of these models used an identical prescription for tides presented in Ref.~\cite{DietrichBernuzziTichy2017}.
+At low frequencies, the tidal effect is that of \texttt{TaylorF2}.
 It then transitions to a fit to the tidal effect measured by a small set of numerical simulations with varying mass ratio and aligned-spins.
 The point-particle and spin prescriptions were different.
-For \texttt{IMRPhenomD\_NRTides}, the waveform is based on the phenomenological \texttt{IMRPhenomD} BBH model [[cite]].
-It uses TaylorF2 at low frequencies then transitions to fits to the EOB waveform at mid frequencies and fits to NR simulations at high frequencies.
-The \texttt{SEOBNRv4\_ROM\_NRTides} model is based on the fast surrogate model \texttt{SEOBNRv4\_ROM} [[cite]] to the EOB waveform \texttt{SEOBNRv4} for BBH systems [[cite]].
+For \texttt{IMRPhenomD\_NRTides}, the waveform is based on the phenomenological \texttt{IMRPhenomD} BBH model~\cite{HusaKhanHannam2016, KhanHusaHannam2016}.
+It uses \texttt{TaylorF2} at low frequencies then transitions to fits to the EOB waveform at mid frequencies and fits to NR simulations at high frequencies.
+The \texttt{SEOBNRv4\_ROM\_NRTides} model is based on the fast surrogate model \texttt{SEOBNRv4\_ROM}~\cite{Puerrer2015} for the EOB waveform \texttt{SEOBNRv4} for BBH systems~\cite{Bohe:2016gbl}.
 Both of these models are improvements to \texttt{TaylorF2}, but do not incorporate all known physical effects in a consistent way.
@@
 Previous works have focused on efficiently representing the parameter space of waveforms using a reduced basis of orthonormal functions whose linear combination can reproduce any waveform in the parameter space to some desired accuracy.
-The two most common approaches have been singular value decomposition (SVD)~\cite{Puerrer2014, Puerrer2015} or a greedy method that selects a subset of the training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
+The two most common approaches have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method that selects a subset of training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
 In this work, instead of using a reduced basis to efficiently represent the EOB waveform as a function of frequency, we will instead use the fact that the frequency-domain waveform is approximately known analytically via the TaylorF2 waveform.
@@
 
-This solves the problem of efficiently representing waveforms as a function of frequency, but leaves the question of interpolating as a function of waveform parameters $\bx$ unanswered.
-Previous work on surrogate models have focused mainly on parameter spaces with three or fewer dimensions.
+The final step is to interpolate between waveform parameters $\bx$.
+Previous work on surrogate models focused on parameter spaces with three or fewer dimensions.
 This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}.
-These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
-For the problem here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation will be unfeasable.
+These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
+For the problem here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible.
 Recent work on optimally choosing waveforms for both EOB~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.
 
-In this paper we focus on an interpolation method known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid.
+In this paper we focus on a technique known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid.
 Importantly, GPR also provides a convenient estimate of its own uncertainty.
@@
 Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of eccentric BBH mergers~\cite{HuertaMooreKumar2017}.
-GPR has also been used in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016} and to estimate noise power spectral densities [[cite]].
+GPR has also been used in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016}.
 
-This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi et al.~\cite{Bernuzzi:2014owa}.
+This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}.
 Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation.
-It also enables additional techniques such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} and multi-banding~\cite{VinciguerraVeitchMandel2017} to accelerate likelihood evaluation.
+It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
 The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.
@@
 We organize the paper as follows.
-In Section II we provide an overview of the EOB waveform model used for constructing the surrogate and the approximations we use to reduce the dimensionality of the parameter space.
+In Section II we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space.
 In section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model.
 We then compare the accuracy and speed of the surrogate to the original model in section IV.
-Then, finally discuss future improvements and applications in Section V.
+Finally, we discuss applications and future improvements in Section V.
 
 \textit{Conventions:} Unless explicitly stated, we use units where $G=c=1$.
-To agree with the GPR literature, we will refer to the waveform parameters as $\bx$ and the hyperparameters in the GPR model as $\btheta$.
 
@@
 
-The effective-one-body (EOB) approach to the general-relativistic 2-body problem, first described in Ref.~\cite{Buonanno:1998gg}, has proven successful in modeling the dynamics and GW emission of binaries of compact objects.
-State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of aligned-spin binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to 0.98 for equal-mass) binaries.
+The EOB approach to the general-relativistic 2-body problem, first described in Ref.~\cite{Buonanno:1998gg}, has proven successful in modeling the dynamics and GW emission of compact binaries.
+State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of NR simulations of aligned-spin BBH systems for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to 0.98 for equal-mass) binaries.
 The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}.
-This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter---BNS or neutron-star-black-hole (NSBH) binaries.
+
+This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu} for binaries that contain NSs.
 The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case.
 In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides.
-Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion---be it a black hole or another neutron star.
-Because of binary motion, the forcing tidal field varies at the orbital frequency.
-Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of the f-mode.
+Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion.
+The forcing tidal field varies at a multiple of the orbital frequency.
+Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of the $f$-mode.
 The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field.
 
-%We note that, to date, this is the only tidal EOB model that includes the effect of spin \red{[[Sebastiano also has a model that includes spin and tides.]]}\blue{[[Which is the Reference?]]}.
-
-More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.\footnote{This model is implemented in the LIGO Algorithm Library as SEOBNRv4.} The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning.
-This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled.
-The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.
+More precisely, spin effects are those of the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins.
+The model, however, does not account for the spin-induced quadrupole moment~\cite{Poisson1998}.
+This effect appears at 2PN order as compared to the tidal effects that first appear at 5PN order, but is only significant for systems with large spin.
+The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{HarryHinderer2017}.
 
@@
 Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.
-In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
+We adopt the tidally-augmented expression for $\Delta_u$ discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}:
 \begin{equation}
@@
 \end{equation}
-where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides.
-We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below.
-Let $A$ and $B$ be the body labels.
-Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.
-Explicitly, including quadrupolar and octupolar tidal effects, we have
+where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass EOB term (Eq.~(2.2) of Ref.~\cite{Bohe:2016gbl}), and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides.
+If either component of the binary is a black hole, then we set the tidal polarizabilities to zero.
+Let $M_{A,B}$ be the masses bodies $A$ and $B$, and $M = M_A + M_B$ be the total mass of the binary.
+The tidal contribution, including quadrupolar and octupolar dynamical tides, is
 \begin{widetext}
@@
  \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
- &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
+ &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,.
 \end{align}
 \end{widetext}
-where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$ and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
-Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$.
+Here, $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
+Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$\footnote{Here, $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.}.
 In particular, the dimensionless dynamical tidal polarizability reads
@@
 \end{equation}
-with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star.
-Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.\footnote{Here $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.} This enhancement factor is 1 for low orbital frequencies, and increases to about 2 as $\ell$ times the orbital frequency approaches the $f$-mode frequency.
+with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$.
+Here, $\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor (Eq.~(11) of Ref.~\cite{Dietrich:2017feu}), which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.
+This enhancement factor is 1 for low orbital frequencies, and increases to $\sim 2$ as $\ell$ times the orbital frequency approaches the $f$-mode frequency.
 
@@
 Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
-The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the GW flux (Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived.
+The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the gravitational-wave flux (Eqs.~(5) and~(6) of Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived.
+The Hamiltonian equations of motion with this radiation-reaction force are numerically integrated beginning with the quasicircular initial conditions described in Ref.~\cite{Buonanno:2005xu}.
+The inspiral-plunge waveform is obtained by evaluating Eq.~(\ref{hlminspplunge}) using the solution to the orbital dynamics.
+We only include the dominant $(\ell,m)=(2,2)$ mode in constructing the final waveform, although we use all modes when evaluating the radiation reaction force.
 
-Let us now discuss the construction of the full waveform.
-The Hamiltonian equations of motion with the above radiation-reaction force are numerically integrated beginning with the quasicircular initial conditions described in Ref.~\cite{Buonanno:2005xu}.
-The inspiral-plunge waveform is obtained by evaluating Eq.~(\ref{hlminspplunge}) using the solution to the orbital dynamics.
-Let $\Omega$ be the instantaneous orbital frequency, $r$ be orbital separation, $p_r$ be the radial momentum, $f_s$ be the waveform sampling frequency in Hz, $\omega_{\textrm{peak}}^{\textrm{NR}}$ be the NR fit for the (2,2)-mode amplitude peak of a BNS merger defined in Eq.~(2) of Ref.~\cite{Bernuzzi:2015rla}~\footnote{Note that this fit was tuned to NR simulations only for $0 \leq \kappa_2^T \leq 500$, where $\kappa_2^T$ is defined by Eq.~(1) of Ref.~\cite{Bernuzzi:2015rla}.
+%Let $\Omega$ be the instantaneous orbital frequency, $r$ be orbital separation, $p_r$ be the radial momentum, $f_s$ be the waveform sampling frequency in Hz, $\omega_{\textrm{peak}}^{\textrm{NR}}$ be the NR fit for the (2,2)-mode amplitude peak of a BNS merger defined in Eq.~(2) of Ref.~\cite{Bernuzzi:2015rla}~\footnote{Note that this fit was tuned to NR simulations only for $0 \leq \kappa_2^T \leq 500$, where $\kappa_2^T$ is defined by Eq.~(1) of Ref.~\cite{Bernuzzi:2015rla}.
 In our model, for $\kappa_2^T >500$ we use the value of this fit at $\kappa_2^T=500$.}.
@@
 
-Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the earliest time when the amplitude of the $(\ell,m)=(2,2)$ mode $|h_{22}^{\textrm{insp-plunge}}|$ peaks, and let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum.
-Then
+We now look for a suitable time $t_{\rm match}$ to stop the numerical integration and match with a post-merger waveform.
+In the point-mass model \texttt{SEOBNRv4}, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks, and $t_{\rm match}$ is chosen to be the time of the peak amplitude $t_{\rm peak}^{\rm amp}$.
+Here, we do not use nonquasicircular corrections tuned to BNS simulations \red{[[Ambiguous.
+Do we use them or not.
+Find out.]]}, so we have less control over the behavior of Eq.~(\ref{hlminspplunge}) in the late inspiral.
+We therefore choose the following definition for $t_{\rm match}$.
+Let $t_{\rm peak}^{\rm amp}$ be the earliest time when the amplitude $|h_{22}^{\textrm{insp-plunge}}|$ of the (2, 2) mode peaks, and let $t_{\rm peak}^{\rm freq}$ be the time when the frequency $\omega_{22}^{\rm insp-plunge}$ of the (2, 2) mode peaks.
+Then,
 \begin{equation}
-t_{\textrm{match}}^{22} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.\label{tmatch}
+t_{\textrm{match}} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.\label{tmatch}
 \end{equation}
-If $|h_{22}^{\textrm{insp-plunge}}|$  does not peak, then we choose $t_{\textrm{peak}}^{\textrm{amp}}$ to be the earliest time when $\partial_t |h_{22}^{\textrm{insp-plunge}}|$ reaches a minimum after having reached a peak.
+For some waveforms, the amplitude does not peak.
+In this case, we choose $t_{\rm peak}^{\rm amp}$ to be the earliest time when the slope of the amplitude $\partial_t |h_{22}^{\rm insp-plunge}|$ reaches a minimum after having reached a peak.
 
-In the BBH model, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks at a prescribed time and has prescribed amplitude and frequency, hence $t_{\textrm{match}}^{22}=t_{\textrm{peak}}^{\textrm{amp}}$.
+%In the BBH model, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks at a prescribed time and has prescribed amplitude and frequency, hence $t_{\textrm{match}}^{22}=t_{\textrm{peak}}^{\textrm{amp}}$.
 Here, however, we do not impose such corrections because of limited input from NR BNS simulations, therefore we have little control over the behavior of Eq.~(\ref{hlminspplunge}) in the late inspiral.
@@
 To help bracket the stopping time, we exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics.
-%When such frequency is being approached (to within some tolerance), we stop the numerical evolution if a peak in the orbital frequency is detected or if the binary stops to monotonically plunge, similarly to what is done in the BBH case.
-In our model we only include the dominant $(\ell,m)=(2,2)$ mode in constructing the final waveform, although we use all modes when evaluating the radiation reaction force.
 
-Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state, finding that the model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state.
-Similar results hold for the tidal EOB model of Ref.~\cite{Bernuzzi:2014owa}, whose radial potential includes tidal effects in a form that is inspired by first-order gravitational-self-force (GSF) calculations of tidal invariants in the small-mass-ratio limit.
-Both the EOB model with dynamical tides and the GSF-inspired one are significantly more faithful to NR than the adiabatic tidal EOB model of Refs.~\cite{Damour:2009wj,Bini:2012gu}.
+%Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state.
+The model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state.
+\red{[[To what accuracy?
+Is this paragraph already in the intro and can be deleted?]]}
 
@@
 
-We model the post-merger GW emission that tapers to zero amplitude in terms of analytic functions for the amplitude $A_{22}^{\textrm{post-mrg}}(t)$ and phase $\phi^{\textrm{post-mrg}}(t)$ such that $h_{22}^{\textrm{post-mrg}}(t) = A_{22}^{\textrm{post-mrg}}(t) \exp{[i\phi_{22}^{\textrm{post-mrg}}(t)]}$.
-For the amplitude, we define the following linear extension of the inspiral-plunge amplitude
+In BNS simulations, the waveform rapidly tapers to zero amplitude in 1--2 gravitational-wave cycles after reaching peak amplitude.
+The BNS system then either undergoes prompt collapse or post-merger oscillations of the remnant.
+We model this tapering in terms of analytic functions for the amplitude $A^{\textrm{post-mrg}}(t)$ and phase $\phi^{\textrm{post-mrg}}(t)$ such that $h_{22}^{\textrm{post-mrg}}(t) = A^{\textrm{post-mrg}}(t) \exp{[i\phi^{\textrm{post-mrg}}(t)]}$.
+
+For the amplitude, we smoothly extend the waveform after $t_{\rm match}$ with a linear fit, and then taper the resulting amplitude. The linear extension is defined by
 \begin{align}
 \hat{A}(t)=\left\{\begin{array}{ll}
-|h_{22}^{\textrm{insp-plunge}}(t)|, & t \leq t^{22}_{\textrm{match}}\,, \\
-a + b (t-t^{22}_{\textrm{match}}) , & t > t^{22}_{\textrm{match}}\,,
+|h_{22}^{\rm insp-plunge}(t)|, & t \leq t_{\rm match}\,, \\
+a + b (t - t_{\rm match}), & t > t_{\rm match}\,,
 \end{array}\right.
 \end{align}
-where $a=|h_{22}^{\textrm{insp-plunge}}(t^{22}_{\textrm{match}})|$ and $b=\partial_t |h_{22}^{\textrm{insp-plunge}}(t^{22}_{\textrm{match}})|$.
-Then we introduce the tapering function \begin{equation}
-W(t) = \frac{1}{1+\exp{[-(t- t^{22}_{\textrm{match}}-15M)/\tau]}}\,,
+where $a = |h_{22}^{\rm insp-plunge}(t_{\rm match})|$ and $b=\partial_t |h_{22}^{\rm insp-plunge}(t_{\rm match})|$.
+The tapering function is centered $15M$ after $t_{\rm match}$ and has a decay time $\tau = \pi/(2\omega_{\rm match})$ (where $\omega_{\rm match} = \omega_{22}^{\rm insp-plunge}(t_{\rm match})$) of one gravitational wave period.
+\red{[[As written, this is a decay time of 1/4 of a period.
+Figure out what is correct.]]} It is given by
+\begin{equation}
+W(t) = \frac{1}{1+\exp{[(t- t^{22}_{\textrm{match}}-15M)/\tau]}}\,.
 \end{equation}
-with $\tau=\pi/(2\omega_{\textrm{match}})$ for $\omega_{\textrm{match}} =\omega_{22}^{\textrm{insp-plunge}}(t^{22}_{\textrm{insp-plunge}})$.
-The amplitude at all times is then given by $|h_{22}(t)|=\hat{A}(t)W(t)$.
+The final amplitude after windowing is then given by $|h_{22}(t)| = \hat A(t) W(t)$.
 
-Let $\phi_{22}^{\textrm{inspiral-plunge}}$ be the phase of $h_{22}^{\textrm{insp-plunge}}$. We define a time $t_{\textrm{freq}}=t^{22}_{\textrm{insp-plunge}}-20M$. The phase at all times is given by
+For the phase, we smoothly extend the waveform frequency such that it agrees with the inspiral frequency at a time $t_{\rm freq} = t_{\rm match} - 20M$ before the matching time $t_{\rm match}$, but then stretches out the frequency evolution such that it only approaches $\omega_{\rm match}$ asymptotically.
+We define this frequency evolution as
 \begin{align}
-\phi_{22}(t)=\left\{\begin{array}{ll}
-\phi_{22}^{\textrm{inspiral-plunge}}(t), & t \leq t_{\textrm{freq}}\,, \\
-\phi_{\textrm{freq}} + \omega_{\textrm{match}}(t-t_{\textrm{freq}})  & \\
-+\frac{\Delta\omega^2}{\dot{\omega}_{\textrm{freq}}} [\exp{(-\dot{\omega}_{\textrm{freq}}(t-t_{\textrm{freq}})/\Delta\omega)}-1]\,,& t > t_{\textrm{freq}}\,,
+\omega(t) = \omega_{\rm match} - \Delta\omega \exp\left[ -(t - t_{\rm freq}) / (20M) \right],
+\end{align}
+where $\Delta\omega = \omega_{\rm match} - \omega_{\rm freq}$. Integrating and requiring continuity at $t_{\rm freq}$ results in the final expression for the phase
+\begin{align}
+\phi_{22}(t) = \left\{\begin{array}{ll}
+\phi_{22}^{\textrm{insp-plunge}}(t), & t \leq t_{\rm freq} \,, \\
+\phi_{\rm freq} + \omega_{\rm match}(t - t_{\rm freq}) + 20M \Delta\omega \left\{ \right. & \\
+  \left. \exp\left[ -(t - t_{\rm freq}) / (20M) \right] - 1\right\} \,,& t > t_{\rm freq} \, .
 \end{array}\right.
 \end{align}
-where $\phi_{\textrm{freq}}=_{22}^{\textrm{inspiral-plunge}}(t_{\textrm{freq}})$, $\Delta\omega=\omega_{\textrm{match}} - \omega_{\textrm{freq}}$, $\omega_{\textrm{freq}}=\omega_{22}^{\textrm{insp-plunge}}(t_{\textrm{freq}})$, and $\dot{\omega}_{\textrm{freq}}=\partial_t\omega_{22}^{\textrm{insp-plunge}}(t_{\textrm{freq}})$.
 
@@
 
-As described above, an aligned-spin EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless components of the spins along the Newtonian orbital angular momentum $-1 \leq S_{A,B} \leq 1$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
+After rescaling with the total mass $M$, an aligned-spin EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless components of the spins along the Newtonian orbital angular momentum $-1 \leq S_{A,B} \leq 1$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
 To reduce the dimensionality of the intrinsic parameter space, we exploit nearly EOS-independent fits for the higher-order effects in terms of $\Lambda_2$ usually referred to as universal relations.
-In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} provides a fit for $\Lambda_3$ as a function of $\Lambda_2$ with an accuracy of about 10\%, depending on the EOS.
+In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} provides a fit for $\Lambda_3$ as a function of $\Lambda_2$ with an accuracy of about 10\% depending on the EOS.
 This relation has been fit for a sample of EOSs and for tidal parameters in the range $1 \leq \Lambda_2\leq10^6$.
@@
 \end{equation}
-where $\xi = \ln\Lambda_2$, $\{a, b, c, d, e\} = \{-1.15, 1.18, 2.51\times 10^{-2}, -1.31\times 10^{-3}, 2.52\times 10^{-5}\}$ from Ref.~\cite{Yagi:2013sva}, and $\{f, g, h\}=\{0.4406491912035266, -34.63232296075433,1762.112913125107\}$.
+where $\xi = \ln\Lambda_2$, $\{a, b, c, d, e\} = \{-1.15, 1.18, 2.51\times 10^{-2}, -1.31\times 10^{-3}, 2.52\times 10^{-5}\}$ from Ref.~\cite{Yagi:2013sva}, and $\{f, g, h\}=\{0.440649, -34.632322, 1762.112913\}$.
 
@@
 The point-mass part of the model is scale-invariant, and the tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$ and $\Lambda_2^{A,B}$.
-Finally, the simple model of the post-merger signal that we employ does not depend on $M$ either.
+Finally, the simple model of the post-merger signal that we employ rescales with $M$ as well.
 It is not clear at what point in the transition from the inspiral to the post-merger this approximation breaks down.
-For example, the frequency of post-merger oscillations depends sensitively on the total mass of the hypermassive remnant.
+For example, whether the merging binary undergoes prompt collapse or forms a hypermassive remnant depends sensitively on the total mass.
 A set of NR simulations with a fixed mass ratio and tidal parameters but different total masses (and therefore different EOSs in order to keep the tidal parameters fixed) will be needed to determine the time when this approximation breaks down.
+
+\red{State range of validity for this model.}
+
+\red{State that you will use the notation $\Lambda_1$, $\Lambda_2$ for the remainder of the paper.}
 
@@
 
-In this section we describe how we decompose the waveform into smooth, slowly-varying functions and train a surrogate model for these functions.
+In this section we describe how we decompose the EOB waveform into smooth, slowly-varying functions and train a surrogate model for these functions.
 We will work with the frequency-domain waveform $\tilde h(Mf; \bx)$.
 This is favorable for data analysis which is usually done in the frequency domain.
-It also allows us to extend the model down to arbitrarily low frequencies with analytic, Frequency-domain PN approximations.
+It also allows us to extend the model down to arbitrarily low frequencies with analytic, frequency-domain PN approximations.
 Unfortunately, Fourier transforming a finite length, discretely sampled waveform leads to a surrogate with more noise.
@@
 To avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad.
-This is a difficult requirement to achieve in the 5-dimensional parameter space here.
+This is a difficult requirement to achieve for 5-dimensional interpolation.
 
@@
 
-For aligned-spin waveforms, we can solve this problem using the fact that the waveform is approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2~\cite{BuonannoIyerOchsner2009} $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$.
+For aligned-spin waveforms, we can solve this problem by using the fact that the waveform can be approximated with the analytic \texttt{TaylorF2} waveform $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$.
 This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2
@@
 These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work.
-The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise dominates.
+The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise is significant.
 
@@
 \end{equation}
-where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{lalsuite}~\cite{lal}. The phase has the schematic form
+where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{LAL}~\cite{lal}. The phase has the schematic form
 \begin{align}
@@
 We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}.
-These terms for the phase are exactly as used in the \texttt{lalsuite} waveform \texttt{TaylorF2}.
+These terms for the phase are exactly as used in the \texttt{LAL} waveform \texttt{TaylorF2}.
 
@@
 we show results for the smaller mass $M_B$ fixed at $1M_\odot$ or fixed at $2M_\odot$.
-[[This is all wrong. Fix it.]]
+\red{[[This is all wrong. Fix it.]]}
 We compare this mismatch to that of th
@@
 In Fig.~\ref{fig:maxmismatchtd} we compare the surrogate to the test set waveform that had the largest mismatch.
-We align the two waveforms in time and phase by maximizing the overlap at early times in the interval $-2\times 10^5 < t/M < -1\times 10^5$.
-The two waveforms are nearly identical except during the last $\sim 2$ cycles where the surrogate peaks in amplitude before the EOB waveform.
+We align the two waveforms in time and phase by maximizing the overlap at early times in the interval $t/M \in [-2\times 10^5, -1\times 10^5]$ [[cite Read et al.
+for details.]].
+The two waveforms are nearly identical except during the last $\sim 2$ cycles.
 
@@
 \caption{Inverse Fourier transformed surrogate (green dashed) for the parameters $\bx$ that had the largest mismatch with the test-set EOB waveforms (black).
-The two waveforms are aligned in the time domain in the interval $-2\times 10^5 < t/M < -1\times 10^5$.}
+The two waveforms are aligned in the time domain in the interval $t/M \in [-2\times 10^5, -1\times 10^5]$.}
 \label{fig:maxmismatchtd}
@@
 We also compare the evaluation time to several BNS waveform models that have been used for LIGO data analysis.
-We note that at $\sim 30$~Hz, the SEOBNRv4T\_surrogate model is comparable to the SEOBNRv4\_ROM\_NRTidal and IMRPhenomD\_NRTidal models that were used in the analysis of the GW170817 event~\cite{DietrichBernuzziTichy2017}.
+We note that at $\sim 30$~Hz, the \texttt{SEOBNRv4T\_surrogate} model is comparable to the \texttt{SEOBNRv4\_ROM\_NRTidal} and \texttt{IMRPhenomD\_NRTidal} models that were used in the analysis of the GW170817 event~\cite{DietrichBernuzziTichy2017}.
 At lower frequencies, the surrogate is faster than all BNS waveform models except for TaylorF2.
@@
 We have constructed a fast frequency-domain surrogate of the most accurate BNS waveform currently available for data analysis.
-This aligned-spin model, SEOBNRv4T, incorporates the tidally induced $\ell=2$ and 3 multipole moments as well as the effect of dynamical tides as the excitation approaches the $\ell=2$ and 3 f-mode frequencies.
+This aligned-spin model, \texttt{SEOBNRv4T}, incorporates the tidally induced $\ell=2$ and 3 multipole moments as well as the effect of dynamical tides as the excitation approaches the $\ell=2$ and 3 $f$-mode frequencies.
 We have achieved mismatches of no more than $7 \times 10^{-4}$ and phase errors of $\lesssim 1$~rad up to the merger frequency.
@@
 The hierarchical method presented here, where we begin with the analytic TaylorF2 reference model then make a surrogate of the residual, can be used to further improve the waveform model.
-For example, one could build a surrogate for numerical BNS simulations using SEOBNRv4T\_surrogate as the base model, and constructing a surrogate of the residual.
+For example, one could build a surrogate for numerical BNS simulations using \texttt{SEOBNRv4T\_surrogate} as the base model, and constructing a surrogate of the residual.
 Such a model would have the accuracy of EOB below $\sim 400$~Hz and the accuracy of numerical simulations for the last several cycles before merger.
@@
 Using GPR and uncertainty sampling discussed above, one could optimally choose the waveform parameters for the numerical simulations, and run them in parallel to build the training set.
-Because the difference between SEOBNRv4T and numerical BNS simulations is exceptionally small, one would not need a high fractional accuracy for a surrogate of the difference, and 10--100 waveforms may be sufficient.
+Because the difference between \texttt{SEOBNRv4T} and numerical BNS simulations is exceptionally small, one would not need a high fractional accuracy for a surrogate of the difference, and 10--100 waveforms may be sufficient.
 
@@
 Using stationary uniform asymptotics, they can also analytically Fourier-transform the solution.
-Importantly, one is free to specify the frequency-domain amplitude and phase evolution, such as the SEOBNRv4T\_surrogate here, for the waveform in the co-precessing frame.
+Importantly, one is free to specify the frequency-domain amplitude and phase evolution, such as the \texttt{SEOBNRv4T\_surrogate} here, for the waveform in the co-precessing frame.
 This approach would provide a fast, accurate model for BNS systems with tides and generic spins.
@@
 BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the Institute for Nuclear Theory where many of the methods used here were discussed.
-We also thank Sylvain Marsat for correcting numerical instabilities in the original SEOBNRv4T code in \texttt{lalsuite}.
+We also thank Sylvain Marsat for correcting numerical instabilities in the original \texttt{SEOBNRv4T} code in \texttt{LAL}.
 The authors were supported by...
@@
 %% ________________________________________________________
-%\bibliographystyle{revtex}     
 \bibliography{paper,refs}
@@
 
-%As discussed above in Sec.~II, the current implementation of the EOB model takes at least 10~minutes to evaluate.
-In this section we will describe in detail the method for training and evaluating our surrogate model.
-The first decision in building a waveform surrogate is choosing between building a surrogate of the time-domain waveform $h(t;\bx)$ or the frequency-domain waveform $\tilde h(f;\bx)$.
-A time-domain surrogate has the advantage that significantly less processing of the original waveform is needed, resulting in a surrogate that typically more accurately reproduces the original waveform model.
-A frequency-domain surrogate, on the other hand, requires windowing of the Fourier transformed waveform to avoid Gibbs oscillations.
-For the beginning of the waveform, one can simply begin at a lower frequency, window the waveform, then truncate the Fourier-transformed waveform above some desired starting frequency.
-At the end of the waveform, the BNS amplitude drops rapidly over $\sim 1$~GW cycle, resulting in Gibbs oscillations at the high-frequency end.
-One could window the end of the waveform over several cycles, however, this would remove available information about tidal interactions.
-As we will show in Sec.~[[]], these high-frequency Gibbs oscillations in the amplitude of the Fourier transformed waveform are a genuine feature of BNS waveforms unlike BBH waveforms which typically taper over a larger number of GW cycles during their ringdown.
-These oscillations make it difficult to construct a surrogate.
-However, we will show that smoothing these oscillations in the Fourier transformed waveform with a moving average filter leads to a surrogate that is sufficiently faithful to the original time-domain model.
-Another important advantage is that a surrogate of $\tilde h(f;\bx)$ bypasses the need to evaluate the Fourier transform at each waveform evaluation for parameter estimation.
-And, as we shall see, a frequency-domain surrogate can be trivially attached to analytic models that are accurate at low-frequencies, such as TaylorF2, making analysis for third-generation detectors which are sensitive at lower frequencies easier.
-Finally, methods for speeding up the likelihood evaluation in parameter estimation such as reduced order quadrature typically require a fast frequency-domain model for calculating the appropriate weights online [[cite]].
-For these reasons we will construct a surrogate of $\tilde h(f;\bx)$.
-
-%\section{Old Junk text}
+%\begin{widetext}
+%\begin{align}
+%\Lambda_{2,\textrm{dyn}}^{\textrm{(A,B)}}(u) = &\frac{\lambda_{2}^{\textrm{(1,2)}}}{M^5} \left\{\frac{1}{4} + \frac{3}{4} \left(\frac{\omega^{(1,2)}_{f,2}}{2\Omega}\right)^2\left[\frac{4\Omega^2}{(\omega^{(1,2)}_{f,2})^2 - 4\Omega^2} + \frac{10}{3Q}\right.\right.
+\nonumber\\
+%&\left.\left.+\sqrt{\frac{\pi}{3}}\frac{1}{\epsilon_{1,2}}\left[(1 + 2\mathcal{S}{(y_{1,2})})\cos{\left(\frac{\pi}{2}y^2_{1,2}\right)} - (1 + 2\mathcal{C}{(y_{1,2})})\sin{\left(\frac{\pi}{2}y_{1,2}^2\right)}\right]\right]\right\}\,,\\
+%\Lambda_{3,\textrm{eff}}^{\textrm{(1,2)}} (u)= &\frac{\lambda_{3}^{\textrm{(1,2)}}}{M^7} \left\{\frac{3}{8} + \frac{(\omega^{(1,2)}_{f,3})^2}{\Omega^2}\left(\frac{25}{48 O_{1,2}} + \frac{5}{72}\frac{9\Omega^2}{(\omega^{(1,2)}_{f,3})^2-9\Omega^2}\right)\right.
+\nonumber\\
+%&\left.+ O_{1,2}\left[\cos{(w^2)}\left(\frac{1}{2} +\mathcal{S}(sqrt2pi*w)\right) - \sin{(w^2)}\left(\frac{1}{2} + \mathcal{C}(sqrt2pi*w)\right)\right]\right\}\,,
+%\end{align}
+%\end{widetext}
+%where $\lambda_{2}^{\textrm{(1,2)}}$ are the (constant) adiabatic values of the quadrupolar tidal polarizabilities, $\lambda_{3}^{\textrm{(1,2)}}$ are the (constant) adiabatic values of the octupolar tidal polarizabilities, $\omega^{(1,2)}_{f,2}$ are the quadrupolar $f$-mode frequencies, $\omega^{(1,2)}_{f,3}$ are the octupolar $f$-mode frequencies, $\mathcal{S}$ is the Fresnel sine integral, $\mathcal{C}$ is the Fresnel cosine integral, $\epsilon_{1,2} = 2^{19/3}\nu (M\omega^{(1,2)}_{f,2})^{5/3}/5$,   $y_{1,2}=\sqrt{3/(\pi \epsilon)} Q_{1,2}/5$, and $Q_{1,2} = 4- 2^{1/3}(M\omega^{(1,2)}_{f,2})^{5/3}u^{-5/2}$.
+%, $w_{1,2}=O_{1,2}/(4\,3^{2/3}\sqrt{10}(M\omega^{(1,2)}_{f,3})^{5/6}\sqrt{\nu})$, $O_{1,2}=5\sqrt{5\pi}/u^3(M\omega^{(1,2)}_{f,3})^{7/6}/(192\,3^{2/3}\sqrt{\nu})$.
 %
-%
-%
-%\sout{A time-domain surrogate has the advantage that significantly less processing of the original waveform is needed, resulting in a surrogate that typically more accurately reproduces the original waveform model.}
-%%A frequency-domain surrogate, on the other hand, requires windowing of the Fourier transformed waveform to avoid Gibbs oscillations.
-For the beginning of the waveform, one can simply begin at a lower frequency, window the waveform, then truncate the Fourier-transformed waveform above some desired starting frequency.
-%
-%
-%\sout{As we will show in Sec.~[[]], these high-frequency Gibbs oscillations in the amplitude of the Fourier transformed waveform are a genuine feature of BNS waveforms unlike BBH waveforms which typically taper over a larger number of GW cycles during their ringdown.
-These oscillations make it difficult to construct a surrogate.
-However, we will show that smoothing these oscillations in the Fourier transformed waveform with a moving average filter leads to a surrogate that is sufficiently faithful to the original time-domain model.
-Another important advantage is that a surrogate of $\tilde h(f;\bx)$ bypasses the need to evaluate the Fourier transform at each waveform evaluation for parameter estimation.
-And, as we shall see, a frequency-domain surrogate can be trivially attached to analytic models that are accurate at low-frequencies, such as TaylorF2, making analysis for third-generation detectors which are sensitive at lower frequencies easier.
-Finally, methods for speeding up the likelihood evaluation in parameter estimation such as reduced order quadrature typically require a fast frequency-domain model for calculating the appropriate weights online [[cite]].
-For these reasons we will construct a surrogate of $\tilde h(f;\bx)$.}
-
-\begin{widetext}
-\begin{align}
-\Lambda_{2,\textrm{dyn}}^{\textrm{(A,B)}}(u) = &\frac{\lambda_{2}^{\textrm{(1,2)}}}{M^5} \left\{\frac{1}{4} + \frac{3}{4} \left(\frac{\omega^{(1,2)}_{f,2}}{2\Omega}\right)^2\left[\frac{4\Omega^2}{(\omega^{(1,2)}_{f,2})^2 - 4\Omega^2} + \frac{10}{3Q}\right.\right.
-\nonumber\\
-&\left.\left.+\sqrt{\frac{\pi}{3}}\frac{1}{\epsilon_{1,2}}\left[(1 + 2\mathcal{S}{(y_{1,2})})\cos{\left(\frac{\pi}{2}y^2_{1,2}\right)} - (1 + 2\mathcal{C}{(y_{1,2})})\sin{\left(\frac{\pi}{2}y_{1,2}^2\right)}\right]\right]\right\}\,,\\
-\Lambda_{3,\textrm{eff}}^{\textrm{(1,2)}} (u)= &\frac{\lambda_{3}^{\textrm{(1,2)}}}{M^7} \left\{\frac{3}{8} + \frac{(\omega^{(1,2)}_{f,3})^2}{\Omega^2}\left(\frac{25}{48 O_{1,2}} + \frac{5}{72}\frac{9\Omega^2}{(\omega^{(1,2)}_{f,3})^2-9\Omega^2}\right)\right.
-\nonumber\\
-&\left.+ O_{1,2}\left[\cos{(w^2)}\left(\frac{1}{2} +\mathcal{S}(sqrt2pi*w)\right) - \sin{(w^2)}\left(\frac{1}{2} + \mathcal{C}(sqrt2pi*w)\right)\right]\right\}\,,
-\end{align}
-\end{widetext}
-where $\lambda_{2}^{\textrm{(1,2)}}$ are the (constant) adiabatic values of the quadrupolar tidal polarizabilities, $\lambda_{3}^{\textrm{(1,2)}}$ are the (constant) adiabatic values of the octupolar tidal polarizabilities, $\omega^{(1,2)}_{f,2}$ are the quadrupolar $f$-mode frequencies, $\omega^{(1,2)}_{f,3}$ are the octupolar $f$-mode frequencies, $\mathcal{S}$ is the Fresnel sine integral, $\mathcal{C}$ is the Fresnel cosine integral, $\epsilon_{1,2} = 2^{19/3}\nu (M\omega^{(1,2)}_{f,2})^{5/3}/5$,   $y_{1,2}=\sqrt{3/(\pi \epsilon)} Q_{1,2}/5$, and $Q_{1,2} = 4- 2^{1/3}(M\omega^{(1,2)}_{f,2})^{5/3}u^{-5/2}$.
-, $w_{1,2}=O_{1,2}/(4\,3^{2/3}\sqrt{10}(M\omega^{(1,2)}_{f,3})^{5/6}\sqrt{\nu})$, $O_{1,2}=5\sqrt{5\pi}/u^3(M\omega^{(1,2)}_{f,3})^{7/6}/(192\,3^{2/3}\sqrt{\nu})$.

### 2017-12-16 src/TEOB-LAL-implementation/timings/timings.py
Timing data for various waveforms
Includes SEOBNRv4T timing data down to f_min=10Hz.

@@ -1,0 +1,200 @@
+#!/usr/bin/env python
+
+import numpy as np
+import lal
+import lalsimulation as LS
+import sys
+
+
+def FD_waveform_test(Mtot, x, approximant, fLow=20.0, fHigh=16384.0, deltaF=0):
+    q = 1.0/x[0]
+    chi1 = x[1]
+    chi2 = x[2]
+    lambda1 = x[3]
+    lambda2 = x[4]
+    
+    phiRef, fRef = 0.0, fLow
+    distance, inclination = 1.0, 0.0
+    
+    m1 = q/(1.0+q)*Mtot
+    m2 = 1.0/(1.0+q)*Mtot
+    m1SI, m2SI, chi1, chi2, lambda1, lambda2, nk_max = m1*lal.MSUN_SI, m2*lal.MSUN_SI, chi1, chi2, lambda1, lambda2, -1
+
+    longAscNodes, eccentricity, meanPerAno = 0,0,0
+
+    LALpars = lal.CreateDict()
+    LS.SimInspiralWaveformParamsInsertTidalLambda1(LALpars, lambda1)
+    LS.SimInspiralWaveformParamsInsertTidalLambda2(LALpars, lambda2)
+
+    # Nyquist frequency is set by fHigh
+    # Can set deltaF = 0 to figure out required frequency spacing; the chosen deltaF depends on fLow
+    
+    # Documentation from LALSimInspiral.c
+    #
+    #  * This routine can generate TD approximants and transform them into the frequency domain.
+    #  * Waveforms are generated beginning at a slightly lower starting frequency and tapers
+    #  * are put in this early region so that the waveform smoothly turns on.
+    #  *
+    #  * If an FD approximant is used, this routine applies tapers in the frequency domain
+    #  * between the slightly-lower frequency and the requested f_min.  Also, the phase of the
+    #  * waveform is adjusted to introduce a time shift.  This time shift should allow the
+    #  * resulting waveform to be Fourier transformed into the time domain without wrapping
+    #  * the end of the waveform to the beginning.
+    #  *
+    #  * This routine assumes that f_max is the Nyquist frequency of a corresponding time-domain
+    #  * waveform, so that deltaT = 0.5 / f_max.  If deltaF is set to 0 then this routine computes
+    #  * a deltaF that is small enough to represent the Fourier transform of a time-domain waveform.
[…]

### 2017-12-16 src/TEOB-LAL-implementation/timings.py
sync

@@ -1,200 +1,0 @@
-#!/usr/bin/env python
-
-import numpy as np
-import lal
-import lalsimulation as LS
-import sys
-
-
-def FD_waveform_test(Mtot, x, approximant, fLow=20.0, fHigh=16384.0, deltaF=0):
-    q = 1.0/x[0]
-    chi1 = x[1]
-    chi2 = x[2]
-    lambda1 = x[3]
-    lambda2 = x[4]
-    
-    phiRef, fRef = 0.0, fLow
-    distance, inclination = 1.0, 0.0
-    
-    m1 = q/(1.0+q)*Mtot
-    m2 = 1.0/(1.0+q)*Mtot
-    m1SI, m2SI, chi1, chi2, lambda1, lambda2, nk_max = m1*lal.MSUN_SI, m2*lal.MSUN_SI, chi1, chi2, lambda1, lambda2, -1
-
-    longAscNodes, eccentricity, meanPerAno = 0,0,0
-
-    LALpars = lal.CreateDict()
-    LS.SimInspiralWaveformParamsInsertTidalLambda1(LALpars, lambda1)
-    LS.SimInspiralWaveformParamsInsertTidalLambda2(LALpars, lambda2)
-
-    # Nyquist frequency is set by fHigh
-    # Can set deltaF = 0 to figure out required frequency spacing; the chosen deltaF depends on fLow
-    
-    # Documentation from LALSimInspiral.c
-    #
-    #  * This routine can generate TD approximants and transform them into the frequency domain.
-    #  * Waveforms are generated beginning at a slightly lower starting frequency and tapers
-    #  * are put in this early region so that the waveform smoothly turns on.
-    #  *
-    #  * If an FD approximant is used, this routine applies tapers in the frequency domain
-    #  * between the slightly-lower frequency and the requested f_min.  Also, the phase of the
-    #  * waveform is adjusted to introduce a time shift.  This time shift should allow the
-    #  * resulting waveform to be Fourier transformed into the time domain without wrapping
-    #  * the end of the waveform to the beginning.
-    #  *
-    #  * This routine assumes that f_max is the Nyquist frequency of a corresponding time-domain
-    #  * waveform, so that deltaT = 0.5 / f_max.  If deltaF is set to 0 then this routine computes
-    #  * a deltaF that is small enough to represent the Fourier transform of a time-domain waveform.
[…]

### 2017-12-16 src/TEOB-LAL-implementation/timings/timings.sh
Add new environment for timings

@@ -1,0 +1,10 @@
+#!/bin/bash
+
+#source /home/mpuer/lsc/TEOBBNS/etc/lalsuiterc
+source ~/pe/master.sh
+export LAL_DATA_PATH=/home/mpuer/SEOBNRROMData:$LAL_DATA_PATH
+
+export PYTHONPATH=$PYTHONPATH:/home/mpuer/projects/
+
+LOG=timings.log
+python timings.py > ${LOG} 2>&1 &

### 2017-12-16 src/TEOB-LAL-implementation/timings/timings.py
rename datafile

@@ -172,9 +172,9 @@ x = np.array([1.0/1.5, 0.35, -0.12, 2000.0, 2750.0])
 xnT = np.array([1.0/1.5, 0.35, -0.12, 0.0, 0.0])
 xNSBH = np.array([1.0/1.5, 0.35, -0.12, 0.0, 2500.0])
 
-#tim_sur = np.array([timeit_ipython("FD_waveform_test(Mtot, x, LS.SEOBNRv4T_surrogate, fLow=%f, fHigh=2048.0, deltaF=0)" %(f_min)) 
-# for f_min in f_mins])
-#np.save('tim_sur', tim_sur)
+tim_sur = np.array([timeit_ipython("FD_waveform_test(Mtot, x, LS.SEOBNRv4T_surrogate, fLow=%f, fHigh=2048.0, deltaF=0)" %(f_min)) 
+ for f_min in f_mins])
+np.save('tim_sur', tim_sur)
 
 tim_TF2 = np.array([timeit_ipython("FD_waveform_test(Mtot, x, LS.TaylorF2, fLow=%f, fHigh=2048.0, deltaF=0)" %(f_min)) 
  for f_min in f_mins])

### 2018-01-15 paper/paper.tex
Add comments on text (except EOB section)

@@
 %%
+\newcommand{\MP}[1]{{\textcolor{blue}{\texttt{MP: #1}} }}
 \newcommand{\bs}[1]{{\textcolor{green}{\texttt{SB: #1}} }}
@@
 Alternative models exist with more accurate representations of the point particle terms.
-However, they do not always simultaneously include the spin and tidal effects as well.
+\MP{What are these alternative models besides EOB?} However, they do not always simultaneously include the spin and tidal effects as well.
 Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$~\cite{Favata2014}.
@@
 These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
-For the problem here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible.
-Recent work on optimally choosing waveforms for both EOB~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.
+For the problem here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible \MP{I think that this is too strong; Blackman17a also perform 5D TP interpolation which contradicts this.}.
+Recent work on optimally choosing waveforms for both EOB\MP{We used IMRPhenomD}~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.
 
@@
 Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation.
-It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
+\MP{Is it really true that this is a significant speedup?
+FFT may still be cheaper than surrogate waveform generation.} It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} \MP{TODO: add citation to PhenomP ROQ paper here and later} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
 The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.
@@
 Because the waveform $\tilde h(Mf; \bx)$ is an oscillatory function of $Mf$ and $\bx$, the waveform is usually decomposed into an amplitude $A(Mf; \bx)$ and phase $\Phi(Mf; \bx)$ as $\tilde h(Mf; \bx) = A(Mf; \bx) e^{i\Phi(Mf; \bx)}$.
-The amplitude and phase are smoother, mostly monotonic functions of frequency.
+The amplitude and phase are smoother, mostly monotonic functions of frequency \MP{the phase is not guaranteed to be monotonic}.
 This can be seen in Fig.~\ref{fig:hoff} where we show the waveforms for the 32 corners of the 5-dimensional parameter space.
@@
 In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $S_1$.
-These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work.
+\MP{TODO: replace $S_i$ by $\chi_i = \hat L_N \cdot \vec S_i / m_i^2$.} These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work.
 The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise is significant.
@@
 These terms for the phase are exactly as used in the \texttt{LAL} waveform \texttt{TaylorF2}.
+\MP{add git hash?
+LAL's TF2 will change in the future when new terms become available.}
 
@@
 Although second generation detectors are only marginally sensitive to BNS signals above this frequency, we use the high-frequency cutoff of $Mf_{{\rm trunc}, f}=0.07$ because the amplitude of some of the EOB waveforms is still large enough at these high frequencies to affect the structure of the surrogate waveform if inverse Fourier transformed back into the time domain.
+\MP{Say that Mf=0.07 is a (generous?) upper bound on the NR BNS merger frequency and thus a safe choice.}
 
@@
 The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.052; \bx)$, where the amplitude is very small and dominated by numerical noise.
+\MP{The discussion of Fig.~\ref{fig:dhofs} appears twice in the text; here and one section earlier.}
 
@@
 For the problem here, we have found that global basis functions do not work well.
-The residuals are small and smooth at low frequencies and large and noisy at high frequencies.
+\MP{EIM did not work well; is this also true for the SVD basis and projection coefficient method?} The residuals are small and smooth at low frequencies and large and noisy at high frequencies.
 With global basis functions, errors in evaluating $g(MF_j;\bx)$ at high frequencies can propagate to large errors between the nodes $MF_j$ at low frequencies.
@@
 For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
-At $\sim10$~minutes--1~day per EOB waveform on a standard CPU, this is at the limit of what is reasonable.
+At $\sim10$~minutes--1~day per EOB waveform on a standard CPU \MP{For which $f_\min$ are these times given?}, this is at the limit of what is reasonable.
 However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below.
@@
 
+\MP{Say that we use sparse frequency grids based on power-law $\Delta f$ in the evaluation of the surrogate; perhaps describe in more detail how the model is evaluated in the LAL code.}
+\MP{Discuss linear vs cubic splines if we decide to use them. Tradeoff in accuracy and speed.}
+
 Given the interpolating functions for the amplitude and phase residuals $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ and $\{\mathcal{I}_{\rm GPR, j}[\Delta\Phi_j](\bx)\}$ evaluated at the frequency nodes $MF_j$, we can now reconstruct the frequency-domain waveform.
@@
 An LHD with its noncollapsing property bypasses this problem.
+\MP{Note (perhaps in a footnotes) that it is possible to have separate surrogates for low and high frequency which are then joined smoothly as done for \texttt{SEOBNRv4\_ROM}.}
 
@@
 This method is sometimes referred to as uncertainty sampling~\cite{BrochuCoraDeFreitas2010}.
-The quantity that we use is the root-mean-squared error in the phase at the frequency nodes $Mf_j$ below $Mf=0.03$:
+The quantity that we use is the root-mean-squared error in the phase at the frequency nodes $Mf_j$ below $Mf=0.03$: \MP{Add reason why higher frequencies are not included in this error measure.}
 \begin{equation}
@@
 
+\MP{Add cost of complex exponential sincos function: look at profiling results}
+\MP{Potentially, explore approximation to sincos for further speedup; need to guarantee that errors don't add up; range reduction is important for accuracy}
+
 The surrogate evaluation time can be roughly decomposed into two times.
@@
 \end{figure}
+
+\subsection{Parameter estimation}
+
+\MP{TODO: add 2D density plot of which quantities? masses, spins, lambdas?
+or mass-ratio, eff spin, eff lambda?}
 

### 2018-01-15 paper/paper.tex
make paper compile

@@
 For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
-At $\sim10$~minutes--1~day per EOB waveform on a standard CPU \MP{For which $f_\min$ are these times given?}, this is at the limit of what is reasonable.
+At $\sim10$~minutes--1~day per EOB waveform on a standard CPU \MP{For which $f_\text{min}$ are these times given?}, this is at the limit of what is reasonable.
 However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below.

### 2018-01-16 paper/paper.tex
Change notation for spins: S_i -> \chi_i

@@
 
-With these relations the waveform only depends on 5 intrinsic parameters $\bx=\{q, S_A, S_B, \Lambda_2^A,\Lambda_2^B\}$.
+With these relations the waveform only depends on 5 intrinsic parameters $\bx=\{q, \chi_A, \chi_B, \Lambda_2^A,\Lambda_2^B\}$.
+The aligned component spins are defined as $\chi_i = \hat L_N \cdot \vec S_i / m_i^2$ where $\vec S_i$ are the dimensionfull spin vectors, $\hat L_N$ the Newtonian orbital angular momentum and $m_i$ the component masses.
 We note that this list does not include the total mass $M$ of the system.
@@
 Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies.
-In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $S_1$.
-\MP{TODO: replace $S_i$ by $\chi_i = \hat L_N \cdot \vec S_i / m_i^2$.} These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work.
+In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $\chi_1$.
+These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work.
 The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise is significant.
@@
 \includegraphics[width=0.49\textwidth]{dhofs.pdf}
-\caption{Cross sections of the amplitude and phase residuals as functions of $S_1$ at three nodes $MF_j$.
-The waveforms are generated for 51 values of $S_1$ shown as dots with the other parameters held constant at $\{q, S_2, \Lambda_1, \Lambda_2\} = \{0.6, 0.2, 2000, 1000\}$.
+\caption{Cross sections of the amplitude and phase residuals as functions of $\chi_1$ at three nodes $MF_j$.
+The waveforms are generated for 51 values of $\chi_1$ shown as dots with the other parameters held constant at $\{q, S_2, \Lambda_1, \Lambda_2\} = \{0.6, 0.2, 2000, 1000\}$.
 Also shown are 1-dimensional cross sections of the interpolated residuals using Gaussian process regression (solid black) as well as their estimated $1\sigma$ uncertainty (shaded gray region).}
@@
 \Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
-                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, S_1, S_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
+                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, \chi_1, \chi_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
 \end{align}
-We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}.
+We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, \chi_1, \chi_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}.
 These terms for the phase are exactly as used in the \texttt{LAL} waveform \texttt{TaylorF2}.
@@
 
-To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of one of the waveform parameters $S_1$.
+To validate that the waveforms are sufficiently conditioned, we plot cross-sections of the residuals in Fig.~\ref{fig:dhofs} for fixed frequencies as functions of one of the waveform parameters $\chi_1$.
 The fact that the residuals are smooth functions of the waveform parameters indicates that most numerical noise has been removed.
@@
 In our case, the training set waveforms have numerical noise that we will estimate by optimizing the hyperparameters.
-The full set of hyperparameters is now  $\btheta = \{\sigma_f, \ell_q, \ell_{S_1}, \ell_{S_2}, \ell_{\Lambda_1}, \ell_{\Lambda_2}, \sigma_n\}$.
+The full set of hyperparameters is now  $\btheta = \{\sigma_f, \ell_q, \ell_{\chi_1}, \ell_{\chi_2}, \ell_{\Lambda_1}, \ell_{\Lambda_2}, \sigma_n\}$.
 
@@
 
-We build an initial training set with 128 waveforms sampled with an LHD in our 5-dimensional parameter space given by $q\in[1/3, 1]$, $S_1, S_2 \in [-0.5, 0.5]$ and $\Lambda_1, \Lambda_2 \in [0, 5000]$.
+We build an initial training set with 128 waveforms sampled with an LHD in our 5-dimensional parameter space given by $q\in[1/3, 1]$, $\chi_1, \chi_2 \in [-0.5, 0.5]$ and $\Lambda_1, \Lambda_2 \in [0, 5000]$.
 In addition, we find empirically that the GPR uncertainty estimates (Eq.~\eqref{eq:var} and Eq.~\eqref{eq:errorrms} below) are largest at the corners of the parameter space, so we also sample the 32 corners.
@@
 \includegraphics[width=0.49\textwidth]{trainingset2d.pdf}
-\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, S_1, S_2\}$ (top) and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom).
+\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, \chi_1, \chi_2\}$ (top) and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom).
 The 160 blue circles were used for the initial training set (32 corners + 128 LHD points).

### 2018-01-16 paper/paper.tex
Add PhenomP ROQ paper reference

@@
 \MP{Is it really true that this is a significant speedup?
-FFT may still be cheaper than surrogate waveform generation.} It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} \MP{TODO: add citation to PhenomP ROQ paper here and later} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
+FFT may still be cheaper than surrogate waveform generation.} It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
 The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.
@@
 For current parameter estimation codes, this is sufficient.
-However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} which requires frequency-domain waveforms, or multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017}.
+However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, or multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017}.
 Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^4)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.

### 2018-01-16 paper/paper.tex
Initial surrogate of EOB with quadrupole-monopole effect. There's still some discrepency in the phase that needs to be checked.

@@
 \author{
-Benjamin D. Lackey$^1$, 
-Michael P\"{u}rrer$^1$
-Andrea Taracchini$^1$
+Benjamin D. Lackey, 
+Michael P\"{u}rrer, 
+Andrea Taracchini,
+Sylvain Marsat
 }
 \affiliation{
-$^1$Max Planck Institute for Gravitational Physics, Albert Einstein Institute, D-14476 Golm, Germany
+Max Planck Institute for Gravitational Physics, Albert Einstein Institute, D-14476 Golm, Germany
 }
@@
 
-Fast and accurate waveform models are necessary for measuring properties of inspiraling binary neutron star systems such as the masses, spins, and tidal parameters.
+Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as the recently discovered event GW170817.
 The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as the effect of dynamical tides, and the formalism is accurate enough to avoid significant biases in the measured parameters.
@@
 We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with $\sim 1000$ waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods.
-\red{[[Mismatch is ...]]}
+\red{[[Mismatch is ...]] [[It can be used in arbitrary frequency range by matching to analytic TaylorF2.]] [[Improvement to previously published EOB model [cite] by including spin-induced quadrupole moment and prescription for waveform tapering at merger.]] }
 
@@
 Smoothing these oscillations makes it significantly easier to interpolate the amplitude and phase residuals as functions of $Mf$ and $\bx$.
+\red{[[$\Delta MF$ is a difference not a fraction.
+maybe try $p$ or something similar?
+Mention that 0.1 for the amplitude is a bit aggressive, but only matters for high frequency where the amplitude is crazy anyway.]]}
 
@@
 This covariance can be used to describe features of the function not captured by the parameterized mean $m(\bx)$ as well as genuine noise in the data $\by$.
-For the problem here, we have already subtracted the TaylorF2 waveform from EOB waveform, so we set $m(\bx) = 0$ and model the residuals purely in terms of the kernel $k(\bx, \bx')$.
+For the problem here, we have already subtracted the TaylorF2 waveform from the EOB waveform, so we set $m(\bx) = 0$ and model the residuals purely in terms of the kernel $k(\bx, \bx')$.
 

### 2018-01-16 src/lalwaveform.py

@@ -10,7 +10,7 @@ import lalsimulation
 
 from constants import *
 import waveform as wave
-
+import taylorf2
 
 ################################################################################
 #                         Names of available approximants                      #
@@ -46,9 +46,10 @@ def print_fd_approximants():
 #            Wrappers for lalsimulation TD and FD waveform generators          #
 ################################################################################
 
-def lalsim_td_waveform(long_asc_nodes=0.0, eccentricity=0.0, mean_per_ano=0.0,
-                       phi_ref=0.0, f_ref=None,
-                       phase_order=-1, amplitude_order=-1, spin_order=-1, tidal_order=-1, **p):
+def lalsim_td_waveform(
+    long_asc_nodes=0.0, eccentricity=0.0, mean_per_ano=0.0,
+    phi_ref=0.0, f_ref=None,
+    phase_order=-1, amplitude_order=-1, spin_order=-1, tidal_order=-1, **p):
     """Wrapper for lalsimulation.SimInspiralChooseTDWaveform.
     Simplified version of pycbc.waveform.get_td_waveform wrapper.
 
@@ -100,9 +101,11 @@ def lalsim_td_waveform(long_asc_nodes=0.0, eccentricity=0.0, mean_per_ano=0.0,
     return wave.Waveform.from_hp_hc(xs, hp.data.data, hc.data.data)
 
 
-def lalsim_fd_waveform(long_asc_nodes=0.0, eccentricity=0.0, mean_per_ano=0.0,
-                       phi_ref=0.0, f_ref=None,
-                       phase_order=-1, amplitude_order=-1, spin_order=-1, tidal_order=-1, **p):
+def lalsim_fd_waveform(
+    long_asc_nodes=0.0, eccentricity=0.0, mean_per_ano=0.0,
+    phi_ref=0.0, f_ref=None,
+    phase_order=-1, amplitude_order=-1, spin_order=-1, tidal_order=-1,
+    quad1=None, quad2=None, **p):
     """Wrapper for lalsimulation.SimInspiralChooseTDWaveform.
     Simplified version of pycbc.waveform.get_td_waveform wrapper.
 
@@ -132,6 +135,18 @@ def lalsim_fd_waveform(long_asc_nodes=0.0, eccentricity=0.0, mean_per_ano=0.0,
         lalsimulation.SimInspiralWaveformParamsInsertTidalLambda1(lal_pars, p['lambda1'])
     if p['lambda2']:
[…]

### 2018-01-16 src/taylorf2.py

@@ -2,12 +2,36 @@ import numpy as np
 from constants import euler_gamma
 import waveform as wave
 
+def quad_of_lambda_fit(lam):
+    """Universal relation for dimensionless quadrupole in terms of dimensionless tidal parameter.
+    For point particle (Kerr), quad = 1.
+    For lambda >= 1, this is Eq. 15 of arXiv: 1608.02582.
+    For lambda < 1, this is Sylvain Marsat's continuation.
+    Copy of XLALSimUniversalRelationQuadMonVSlambda2Tidal() from LALSimUniversalRelations.c
+    """
+    if lam<1.0:
+        b = 0.427688866723244
+        c = -0.324336526985068
+        d = 0.1107439432180572
+        return 1.0 + b*lam + c*lam**2 + d*lam**3
+    else:
+        xi = np.log(lam)
+        a = 0.1940
+        b = 0.09163
+        c = 0.04812
+        d = -4.283e-3
+        e = 1.245e-4
+        ps = a + b*xi + c*xi**2 + d*xi**3 + e*xi**4
+        return np.exp(ps)
+
+
 def lamtilde_of_eta_lam1_lam2(eta, lam1, lam2):
     """$\tilde\Lambda(\eta, \Lambda_1, \Lambda_2)$.
     Lambda_1 is assumed to correspond to the more massive (primary) star m_1.
     Lambda_2 is for the secondary star m_2.
     """
     return (8.0/13.0)*((1.0+7.0*eta-31.0*eta**2)*(lam1+lam2) + np.sqrt(1.0-4.0*eta)*(1.0+9.0*eta-11.0*eta**2)*(lam1-lam2))
+
 
 def deltalamtilde_of_eta_lam1_lam2(eta, lam1, lam2):
     """This is the definition found in Les Wade's paper.
@@ -21,79 +45,6 @@ def deltalamtilde_of_eta_lam1_lam2(eta, lam1, lam2):
                       + (1.0 - 15910.0*eta/1319.0 + 32850.0*eta**2/1319.0 + 3380.0*eta**3/1319.0)*(lam1-lam2)
                       )
 
-
[…]

### 2018-01-16 src/trainingseteob.py

@@ -138,20 +138,20 @@ class ConditionedWaveform(object):
     def log_spaced_in_frequency_domain(self, trunc_i, trunc_f, npoints=10000):
         wave.resample_uniform(self.h, xi=trunc_i, xf=trunc_f, npoints=npoints, spacing='log', order=2)
 
-    def difference_with_taylorf2(self, params):
-        """Compare with TaylorF2 waveform.
-        """
-        q, spin1z, spin2z, lambda1, lambda2 = params
-        mf = self.h.x
-        h_ref = taylorf2.dimensionless_taylorf2_waveform(
-            mf=mf, q=q,
-            spin1z=spin1z, spin2z=spin2z,
-            lambda1=lambda1, lambda2=lambda2)
-
-        h_ref.add_phase(remove_start_phase=True)
-        npoints = len(mf)
-        dh = train.lnamp_phase_difference(self.h, h_ref, npoints=npoints, spacing='log', order=2)
-        return h_ref, dh
+    # def difference_with_taylorf2(self, params):
+    #     """Compare with TaylorF2 waveform.
+    #     """
+    #     q, spin1z, spin2z, lambda1, lambda2 = params
+    #     mf = self.h.x
+    #     h_ref = taylorf2.dimensionless_taylorf2_waveform(
+    #         mf=mf, q=q,
+    #         spin1z=spin1z, spin2z=spin2z,
+    #         lambda1=lambda1, lambda2=lambda2)
+    #
+    #     h_ref.add_phase(remove_start_phase=True)
+    #     npoints = len(mf)
+    #     dh = train.lnamp_phase_difference(self.h, h_ref, npoints=npoints, spacing='log', order=2)
+    #     return h_ref, dh
 
 
 def moving_average_geometric_range(fs, ys, dfbyf, fbound_low=None, fbound_high=None):
@@ -204,7 +204,7 @@ def moving_average_geometric_range(fs, ys, dfbyf, fbound_low=None, fbound_high=N
     return ymovavg
 
 
-def difference_with_taylorf2(h, params):
+def difference_with_taylorf2(h, params, quad1=None, quad2=None):
     """Compare with TaylorF2 waveform.
     """
     q, spin1z, spin2z, lambda1, lambda2 = params
@@ -212,7 +212,8 @@ def difference_with_taylorf2(h, params):
     h_ref = taylorf2.dimensionless_taylorf2_waveform(
         mf=mf, q=q,
[…]

### 2018-01-16 paper/PE-analysis/bounded_KDE.py
Start analyzing PE results

@@ -1,0 +1,269 @@
+import numpy as np
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+from matplotlib import patheffects as PathEffects
+
+import pandas as pd
+from six import string_types
+from scipy.stats import gaussian_kde as kde
+
+
+
+class Bounded_2d_kde(kde):
+    r"""Represents a two-dimensional Gaussian kernel density estimator
+    for a probability distribution function that exists on a bounded
+    domain."""
+
+    def __init__(self, pts, xlow=None, xhigh=None, ylow=None, yhigh=None, *args, **kwargs):
+        """Initialize with the given bounds.  Either ``low`` or
+        ``high`` may be ``None`` if the bounds are one-sided.  Extra
+        parameters are passed to :class:`gaussian_kde`.
+
+        :param xlow: The lower x domain boundary.
+
+        :param xhigh: The upper x domain boundary.
+
+        :param ylow: The lower y domain boundary.
+
+        :param yhigh: The upper y domain boundary.
+        """
+        pts = np.atleast_2d(pts)
+
+        assert pts.ndim == 2, 'Bounded_kde can only be two-dimensional'
+
+        super(Bounded_2d_kde, self).__init__(pts.T, *args, **kwargs)
+
+        self._xlow = xlow
+        self._xhigh = xhigh
+        self._ylow = ylow
+        self._yhigh = yhigh
+
+    @property
+    def xlow(self):
+        """The lower bound of the x domain."""
+        return self._xlow
+
+    @property
+    def xhigh(self):
+        """The upper bound of the x domain."""
+        return self._xhigh
+
+    @property
+    def ylow(self):
+        """The lower bound of the y domain."""
+        return self._ylow
+
+    @property
+    def yhigh(self):
+        """The upper bound of the y domain."""
+        return self._yhigh
+
+    def evaluate(self, pts):
+        """Return an estimate of the density evaluated at the given
+        points."""
+        pts = np.atleast_2d(pts)
+        assert pts.ndim == 2, 'points must be two-dimensional'
+
+        pdf = super(Bounded_2d_kde, self).evaluate(pts.T)
+        
+        if np.any([self.xlow, self.xhigh, self.ylow, self.yhigh]):
+            pts_p = pts.copy()
+            
[…]

### 2018-06-28 paper/paper.tex
Updated surrogate with change of variables lambda->xi.

@@
 %% ________________________________________________________
+
 \red{Choose point-mass or point-particle.}
+
+\red{Use $\chi$ for dimensionless spin instead of $S$.}
+
+\red{Choose notation $\Lambda_2^A$ or $\Lambda_1$.}
+
 \section{Introduction}
@@
 
-\subsection{Approximation of higher-order tidal parameters}
+\subsection{Reducing the number of matter parameters}
 
-After rescaling with the total mass $M$, an aligned-spin EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless components of the spins along the Newtonian orbital angular momentum $-1 \leq S_{A,B} \leq 1$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
-To reduce the dimensionality of the intrinsic parameter space, we exploit nearly EOS-independent fits for the higher-order effects in terms of $\Lambda_2$ usually referred to as universal relations.
-In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} provides a fit for $\Lambda_3$ as a function of $\Lambda_2$ with an accuracy of about 10\% depending on the EOS.
+After rescaling with the total mass $M$, the aligned-spin EOB waveform with dynamical tides depends on 13 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless spin components along the orbital angular momentum $\chi_{A,B}$, the two spin-induced quadrupole moments $\bar Q_{A,B}$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
+To reduce the dimensionality of the intrinsic parameter space, we use nearly EOS-independent fits (universal relations) for these parameters in terms of $\Lambda_2$ as discussed below.
+This reduces the number of matter parameters from ten to two.
+
+Eq.~(15) of Ref.~\cite{YagiYunes2017} provides a fit for $\bar Q$ as a function of $\Lambda_2$.
+This relation has been fit for a sample of EOSs and for tidal parameters in the range $1 \leq \Lambda_2\leq10^4$.
+This formula, however, diverges at small values of $\Lambda_2$, so we replace it with a polynomial function that approaches the Kerr value of $\bar Q = 1$ at $\Lambda_2 = 0$.
+We require this extension to be continuous at $\Lambda_2 = 1$.
+The exact relation we use is
+\begin{equation}
+\bar Q = \left\{\begin{array}{ll}
+1 + f \Lambda_2 + g \Lambda_2^2 + h \Lambda_2^3\, , & 0 \leq \Lambda_2 \le 1 \\
+e^{a + b \xi + c \xi^2 + d \xi^3 + e \xi^4}\, , & \Lambda_2 > 1
+\end{array}\right.,
+\end{equation}
+where $\xi = \ln\Lambda_2$.
+The coefficients $\{a, b, c, d, e\} = \{ 0.194, 0.09163, 0.04812, -4.283\times 10^{-3}, 1.245\times 10^{-4} \}$ are from Table I of Ref.~\cite{YagiYunes2017}, and $\{f, g, h\} = \{0.42769, -0.32434, 0.11074\}$.
+
+Eq.~(60) of Ref.~\cite{Yagi:2013sva} provides a fit for $\Lambda_3$ as a function of $\Lambda_2$ with an accuracy of about 10\% depending on the EOS.
 This relation has been fit for a sample of EOSs and for tidal parameters in the range $1 \leq \Lambda_2\leq10^6$.
@@
 
-With these relations the waveform only depends on 5 intrinsic parameters $\bx=\{q, \chi_A, \chi_B, \Lambda_2^A,\Lambda_2^B\}$.
-The aligned component spins are defined as $\chi_i = \hat L_N \cdot \vec S_i / m_i^2$ where $\vec S_i$ are the dimensionfull spin vectors, $\hat L_N$ the Newtonian orbital angular momentum and $m_i$ the component masses.
+With these relations the waveform only depends on the five intrinsic parameters $\bx=\{q, \chi_A, \chi_B, \Lambda_2^A,\Lambda_2^B\}$. 
+%The aligned component spins are defined as $\chi_i = \hat L_N \cdot \vec S_i / m_i^2$ where $\vec S_i$ are the dimensionfull spin vectors, $\hat L_N$ the Newtonian orbital angular momentum and $m_i$ the component masses.
 We note that this list does not include the total mass $M$ of the system.
@@
 For example, whether the merging binary undergoes prompt collapse or forms a hypermassive remnant depends sensitively on the total mass.
-A set of NR simulations with a fixed mass ratio and tidal parameters but different total masses (and therefore different EOSs in order to keep the tidal parameters fixed) will be needed to determine the time when this approximation breaks down.
-
-\red{State range of validity for this model.}
-
-\red{State that you will use the notation $\Lambda_1$, $\Lambda_2$ for the remainder of the paper.}
+NR simulations will be needed to determine when after the merger this approximation breaks down.
 
@@
 We will show below, however, that sufficient filtering can solve this problem.
+The \texttt{SEOBNRv4T} waveform model, as implemented, works for mass ratios in the range $q \in [1/3, 1]$, spins $\chi_{1,2} \in [-0.5, 0.5]$ and tidal parameters $\Lambda_{1,2} \in [0, 5000]$.
+Our surrogate of this model, \texttt{SEOBNRv4T\_surrogate}, will be valid for the same range of parameters.
 
@@
 \end{align}
-These residuals are also shown in Fig.~\ref{fig:hoff}.
+The exact functional form of the TaylorF2 waveform is given in Appendix~\ref{sec:taylorf2}, and the residuals are shown in Fig.~\ref{fig:hoff}.
+
 For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that interpolation errors will not lead to a negative amplitude for the reconstructed waveform (Eq.~\eqref{eq:hdecomp}).
-In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency, the log-ratio captures this behavior.
+In addition, because the waveform amplitude spans several orders of magnitude at high frequencies, the log-ratio better captures this behavior.
 Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies.
@@
 
-The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
+Finally, we have found that the functions $\Delta\ln A(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$ sometimes vary rapidly as a function of the tidal parameters $\Lambda_1$ and $\Lambda_2$ for tidal parameters in the range [0, 1000].
+We therefore perform a change of variables that stretches out the parameter space for small values of the tidal parameters:
 \begin{equation}
-A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} x^{-7/4} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)x \right],
+\xi = \log_{10}(\Lambda/100 + 1).
 \end{equation}
-where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{LAL}~\cite{lal}. The phase has the schematic form
-\begin{align}
-\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
-                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, \chi_1, \chi_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
-\end{align}
-We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, \chi_1, \chi_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}.
-These terms for the phase are exactly as used in the \texttt{LAL} waveform \texttt{TaylorF2}.
-\MP{add git hash?
-LAL's TF2 will change in the future when new terms become available.}
+With $\bx=\{q, \chi_A, \chi_B, \xi_A,\xi_B\}$, the functions $\Delta\ln A(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$ are smoother functions of the parameters, making them easier to fit.
 
@@
 
-\begin{figure}[htb]
+\begin{figure*}[htb]
 \centering
-\includegraphics[width=0.49\textwidth]{hmaxerror.pdf}
+\includegraphics[width=0.99\textwidth]{hmaxerror.pdf}
 \caption{Inverse Fourier transformed surrogate (green dashed) for the parameters $\bx$ that had the largest mismatch with the test-set EOB waveforms (black).
@@
 \label{fig:maxmismatchtd}
-\end{figure}
+\end{figure*}
 
@@
 
+\appendix
+\section{TaylorF2 reference waveform}
+\label{sec:taylorf2}
+
+The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
+\begin{equation}
+A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} x^{-7/4} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)x \right],
+\end{equation}
+where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{LAL}~\cite{lal}. The phase has the schematic form
+\begin{align}
+\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
+                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, \chi_1, \chi_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
+\end{align}
+We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, \chi_1, \chi_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}.
+These terms for the phase are exactly as used in the \texttt{LAL} waveform \texttt{TaylorF2}.
+
+The spin terms are
+
+\red{Because the \texttt{SEOBNRv4T} waveform does not include 3PN spin-spin effects, we remove the 3PN spin-spin term in the \texttt{TaylorF2} waveform as well.
+The main reason for doing this is because including only common physical effects in both \texttt{SEOBNRv4T} and \texttt{TaylorF2} makes the residual $\Delta\Phi(Mf; \bx)$ smaller and have less variation, making it easier to fit at a given error requirement.}
+
+$SL = X_1^2\chi_1 + X_2^2\chi_2$
+
+$d\Sigma L = (X_1-X_2)(X_2\chi_2-X_1\chi_1)$
+
+$\gamma = \left(\frac{554345}{1134}+\frac{110\eta}{9}\right)SL + \left(\frac{13915}{84} - \frac{10\eta}{3}\right)d\Sigma L$
+
+$SS15 = \frac{188}{3}SL + 25 d\Sigma L$
+
+$SS2 = -10 \Sigma (not same sigma)$
+
+$SS25 = -\gamma -3\gamma\ln(v)$
+
+$SS3 = \frac{\pi}{3}\left[3760(X_1^2\chi_1 + X_2^2\chi_2) + 1490(X_1-X_2)(X_2\chi_2 - X_1\chi_1)\right]$
+
+\begin{equation}
+\tilde\Lambda = \frac{16}{13} \frac{(X_1+12X_2)X_1^4\Lambda_1 + (12X_1+X_2)X_2^4\Lambda_2}{(X_1+X_2)^5}
+\end{equation}
+
+\begin{align}
+\delta\tilde\Lambda =& \frac{(1319X_1^2 - 7996X_1X_2 - 11005X_2^2)X_1^4\Lambda_1}{1319(X_1+X_2)^6} \\
+& + \frac{(11005X_1^2 + 7996X_1X_2 - 1319X_2^2)X_2^4\Lambda_2}{1319(X_1+X_2)^6}
+\end{align}
+
+$a_5 = -\frac{39}{2}\tilde\Lambda$
+
+$a_6 = -\frac{3115}{64}\tilde\Lambda + \frac{6595}{364}(X_1-X_2)\delta\tilde\Lambda$
+
 %% ________________________________________________________

### 2018-06-28 src/surrogate.py

@@ -253,172 +253,173 @@ def physical_to_pycbc_frequency_series(freq, h_plus, h_cross):
     return hp_fs, hc_fs
 
 
-class GPSurrogate(object):
-    def __init__(self, mf_amp, mf_phase, Bamp, Bphase, damp_gp_list, dphase_gp_list):
-        # Lists of interpolating functions
-        self.Bamp = Bamp
-        self.Bphase = Bphase
-        # Lists of GPR functions
-        self.damp_gp_list = damp_gp_list
-        self.dphase_gp_list = dphase_gp_list
-        # Waveform samples
-        self.mf_amp = mf_amp
-        self.mf_phase = mf_phase
-        self.mf = self.Bamp[0].x
-        self.mf_a = self.Bamp[0].x[0]
-        self.mf_b = self.Bamp[0].x[-1]
+# class GPSurrogate(object):
+#     def __init__(self, mf_amp, mf_phase, Bamp, Bphase, damp_gp_list, dphase_gp_list):
+#         # Lists of interpolating functions
+#         self.Bamp = Bamp
+#         self.Bphase = Bphase
+#         # Lists of GPR functions
+#         self.damp_gp_list = damp_gp_list
+#         self.dphase_gp_list = dphase_gp_list
+#         # Waveform samples
+#         self.mf_amp = mf_amp
+#         self.mf_phase = mf_phase
+#         self.mf = self.Bamp[0].x
+#         self.mf_a = self.Bamp[0].x[0]
+#         self.mf_b = self.Bamp[0].x[-1]
+#
+#     @classmethod
+#     def load(cls, Bamp_filename, Bphase_filename, damp_gp_filename, dphase_gp_filename):
+#         """Load surrogate model from 4 hdf5 data files.
+#         """
+#         Bamp = ws.HDF5WaveformSet(Bamp_filename)
+#         Bphase = ws.HDF5WaveformSet(Bphase_filename)
+#         damp_gp_list = gpr.load_gaussian_process_regression_list(damp_gp_filename)
+#         dphase_gp_list = gpr.load_gaussian_process_regression_list(dphase_gp_filename)
+#         mf_amp = Bamp.parameters()[:, 0]
+#         mf_phase = Bphase.parameters()[:, 0]
+#         return GPSurrogate(mf_amp, mf_phase, Bamp, Bphase, damp_gp_list, dphase_gp_list)
+#
+#     ############# Evaluate waveform quantities in geometric units #############
+#
[…]

### 2018-06-28 src/taylorf2.py

@@ -187,17 +187,17 @@ def taylorf2_phase_30pn_spin(eta, chi1, chi2, quad1, quad2, spin_spin=True):
         ss3 = (326.75/1.12 + 557.5/1.8*eta)*eta*chi1L*chi2L
         ss3 += ((4703.5/8.4 + 2935.0/6.0*X1 - 120.0*X1**2)*quad1 + (-4108.25/6.72 - 108.5/1.2*X1 + 125.5/3.6*X1**2)) * X1**2 * chi1sq
         ss3 += ((4703.5/8.4 + 2935.0/6.0*X2 - 120.0*X2**2)*quad2 + (-4108.25/6.72 - 108.5/1.2*X2 + 125.5/3.6*X2**2)) * X2**2 * chi2sq
-    elif spin_spin=='point':
-        # Use the value for point particles
-        quad1 = 1.0
-        quad2 = 1.0
-        ss3 = (326.75/1.12 + 557.5/1.8*eta)*eta*chi1L*chi2L
-        ss3 += ((4703.5/8.4 + 2935.0/6.0*X1 - 120.0*X1**2)*quad1 + (-4108.25/6.72 - 108.5/1.2*X1 + 125.5/3.6*X1**2)) * X1**2 * chi1sq
-        ss3 += ((4703.5/8.4 + 2935.0/6.0*X2 - 120.0*X2**2)*quad2 + (-4108.25/6.72 - 108.5/1.2*X2 + 125.5/3.6*X2**2)) * X2**2 * chi2sq
+    # elif spin_spin=='point':
+    #     # Use the value for point particles
+    #     quad1 = 1.0
+    #     quad2 = 1.0
+    #     ss3 = (326.75/1.12 + 557.5/1.8*eta)*eta*chi1L*chi2L
+    #     ss3 += ((4703.5/8.4 + 2935.0/6.0*X1 - 120.0*X1**2)*quad1 + (-4108.25/6.72 - 108.5/1.2*X1 + 125.5/3.6*X1**2)) * X1**2 * chi1sq
+    #     ss3 += ((4703.5/8.4 + 2935.0/6.0*X2 - 120.0*X2**2)*quad2 + (-4108.25/6.72 - 108.5/1.2*X2 + 125.5/3.6*X2**2)) * X2**2 * chi2sq
     else:
         ss3 = 0.0
 
-    #print ss3
+    #print 'ss3={}'.format(ss3)
 
[…]

### 2018-07-04 paper/paper.tex
Fairly major edits of paper. Still in progress.

@@
 
-\title{Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression}
+\title{Surrogate waveform model using Gaussian process regression for an aligned-spin effective one body model of binary neutron star inspirals}
 
@@
 
-Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as the recently discovered event GW170817.
+%Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as the recently discovered event GW170817.
 The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as the effect of dynamical tides, and the formalism is accurate enough to avoid significant biases in the measured parameters.
@@
 \red{[[Mismatch is ...]] [[It can be used in arbitrary frequency range by matching to analytic TaylorF2.]] [[Improvement to previously published EOB model [cite] by including spin-induced quadrupole moment and prescription for waveform tapering at merger.]] }
+
+Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as the recently discovered event GW170817.
+The effective one body formalism was previously shown to produce waveforms that agree with numerical simulations of aligned-spin BNS systems during the inspiral.
+This model includes aligned-spin effects, the quadrupolar and octopolar adiabatic tides, and the effect of dynamical tides.
+We improve this model by adding the spin-induced quadrupole moment and a prescription for tapering the end of the waveform to qualitatively agree with numerical simulations of the merger.
+The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters.
+This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters.
+Unfortunately, the current implementation takes several minutes to hours to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations.
+We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression to efficiently sample the parameter space.
+The resulting surrogate has a maximum mismatch of \red{$\sim 5\times 10^{-4}$} for the Advanced LIGO detector, and requires less than \red{0.3~s} to evaluate for a waveform with a starting frequency of 10~Hz (\red{0.04~s} for a starting frequency of 30~Hz).
+With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).
 
@@
 
-The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors, has demonstrated that the parameters of the binary can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant~\cite{GW170817Hubble}, and understanding the structure of neutron stars (NSs) through tidal interactions~\cite{GW170817}.
+%The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors, has demonstrated that the parameters of the binary can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant~\cite{GW170817Hubble}, and understanding the structure of neutron stars (NSs) through tidal interactions~\cite{GW170817}.
 Measuring these parameters accurately will also be necessary to properly interpret the gamma ray burst [[cite]] and optical afterglow [[cite]] that was found in conjunction.
@@
 
-These results are obtained using Bayesian analyses that compare the data to parameterized waveform models.
+%\red{Purpose: Intro. Justify this field of research.}
+
+The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors has demonstrated that tidal interactions during the inspiral can be measured with enough precision to constrain the radius and equation of state~\cite{DeEtA, PEPaper, EOSPaperl} of NSs.
+Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further.
+These multiple events could then be stacked to significantly improve measurements of the radius and EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}.
+\red{[[Mention A+ being funded?
+How will this improve each parameter?]]}
+
+%These results are obtained using a Bayesian analysis that compares the data to parameterized waveform models.
 To avoid systematic errors highly accurate models are needed, and they must include all relevant physical effects.
@@
 Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models.
-These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5 and 6PN order.
+These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5PN and 6PN order.
 Unfortunately, the lack of point-particle terms above 3.5PN order leads to systematic errors in the tidal parameters that are as large as the statistical errors~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
@@
 
-One such alternative to the PN waveforms is the effective one body (EOB) formalism discussed in section~\ref{sec:eob}.
-It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations.
-Until recently, however, no version of the EOB formalism included the effects of spin and tides in the same model.
-Recently two versions have become available that include both tidal effects and aligned spin.
-These models are being implemented within the LIGO Algorithm Library (\texttt{LAL})~\cite{lal} under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}.
-The \texttt{SEOBNRv4T} model~\cite{Hinderer:2016eia,Steinhoff:2016rfi} used in this paper, includes the tidally induced $\ell=2$ and 3  multipole moments as well as the induced $\ell=2$ and 3 $f$-mode resonances, but does not yet include the spin induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2017}.
-This model also agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}.
-The alternative model, \texttt{TEOBResumS}~\cite{Nagar2017}, uses a slightly different point-particle and spin prescription.
-For tidal interactions, the model includes the $\ell = 2, 3$, and 4 terms and uses a method for re-summing these expressions.
-It does not include the effect of dynamical tides from the induced $f$-mode resonances, but does include the spin induced quadrupole moment.
+%\red{Purpose: Explain the problem. Fast and accurate waveforms are needed}
+
+These results are obtained using a Bayesian analysis that compares the data to parameterized waveform models. 
+Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations.
+Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event.
+Most time-domain waveform implementations involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain.
+Depending on the model, this can take seconds to hours for long BNS waveforms, so analytic frequency-domain approximations are often used instead.
+
+%In addition, to avoid systematic errors, highly accurate models are needed that include all relevant physical effects.
+This includes descriptions of the point-particle interactions, spins, and tides.
+Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models [[cite some examples]].
+These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5PN and 6PN order.
+Unfortunately, the lack of point-particle terms above 3.5PN order can lead to systematic errors in the tidal parameters that are as large as the statistical errors~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
+\red{[[Describe what was found for GW170817 in PE paper?
+This seemed to be the case for large tidal parameters but not for the actual event.
+Also you didn't compare different PN waveforms there.]]}
+
+%\red{Purpose: Explain what waveforms have been used for GW170817 and other works and why yet another one is needed.}
+
+In addition, to avoid systematic errors that can be as large as statistical errors, highly accurate models are needed that include all relevant physical effects, including point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
+For GW170817, a set of four frequency-domain analytic or semi-analytic waveform models were used as templates (see Table I of~\cite{PEPaper}) for the main results.
+These models are based on spinning binary black hole (BBH) waveform models.
+They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2~\cite{}), the aligned-spin effective one body (EOB) formalism (SEOBNRT~\cite{}) or the phenomenological formalism for aligned-spin (PhenomDNRT~\cite{}) and precessing spin (PhenomPNRT~\cite{}).
+Tidal interactions, on the other hand, were included simply by adding a correction to the phase of each waveform.
+For the PN TaylorF2 waveform the analytic 5PN and 6PN order terms were added.
+For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in~\cite{PEPaper}) was used for the phase correction~\cite{DietrichBernuzziTichy2017, NewNRTidalPaper}.
+
+%For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates~\cite{PEPaper}.
+The first was the analytic PN waveform model known as \texttt{TaylorF2}~\cite{BuonannoIyerOchsner2009}.
+Because of the known systematic errors that result from PN waveforms, three additional models were produced for comparison: \texttt{IMRPhenomD\_NRTides}, \texttt{SEOBNRv4\_ROM\_NRTides}, and \texttt{IMRPhenomP\_NRTides}.
+These three models used an identical prescription for tides presented in Ref.~\cite{DietrichBernuzziTichy2017, NewNRTidalPaper}.
+At low frequencies, the tidal effect is that of \texttt{TaylorF2}.
+It then transitions to a fit to the tidal effect measured by a small set of numerical simulations with varying mass ratio and aligned-spins.
+The point-particle and spin prescriptions were different.
+For \texttt{IMRPhenomD\_NRTides}, the waveform is based on the phenomenological \texttt{IMRPhenomD} BBH model~\cite{HusaKhanHannam2016, KhanHusaHannam2016}.
+It uses \texttt{TaylorF2} at low frequencies then transitions to fits to the EOB waveform at mid frequencies and fits to NR simulations at high frequencies.
+The \texttt{SEOBNRv4\_ROM\_NRTides} model is based on the fast surrogate model \texttt{SEOBNRv4\_ROM}~\cite{Puerrer2015} for the EOB waveform \texttt{SEOBNRv4} for BBH systems~\cite{Bohe:2016gbl}.
+Finally, the \texttt{IMRPhenomP\_NRTides} is an extension of \texttt{IMRPhenomD\_NRTides}, but also includes a spin-induced quadrupole moment and an approximates precession with a single effective spin parameter~\cite{}.
+These models are improvements to \texttt{TaylorF2}, but only treat matter effects as an additive phase correction to BBH waveform models.
+\red{[[Are all these details needed?]]} [[These waveform models were in broad agreement with each other for $\tilde\Lambda$ (and other parameters?) to 10\% error.]]
+
+%We therefore build a fast surrogate that reproduces as accurately as possible \texttt{SEOBNRv4T} in the frequency domain.
+
+%********************
+
+%Alternative models exist with more accurate representations of the point particle terms.
+\MP{What are these alternative models besides EOB?} However, they do not always simultaneously include the spin and tidal effects as well.
+Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$~\cite{Favata2014}.
+In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].
 
@@
 
-In addition to accuracy, useful waveform models must be fast enough to be used in Bayesian parameter estimation algorithms such as Markov chain Monte Carlo (MCMC) and Nested Sampling that currently require $10^7$--$10^8$ sequential waveform evaluations.
+%In addition to accuracy, useful waveform models must be fast enough to be used in Bayesian parameter estimation algorithms such as Markov chain Monte Carlo (MCMC) and Nested Sampling that currently require $10^7$--$10^8$ sequential waveform evaluations.
 Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event.
@@
 For PN waveform models, this can take several seconds for a starting frequency of 10~Hz, while for the \texttt{SEOBNRv4T} model discussed here, the equivalent evaluation time can be a day.
-Significant work has been done to optimize existing implementations of EOB for binary black hole (BBH) systems by a factor of several hundred~\cite{DevineEtienneMcWilliams2016}.
-However, it is not clear that sufficient speed ups can be achieved for the BNS model discussed here with dynamical tides.
-
-For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates.
+
+%For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates.
 The first was the stationary phase approximation PN waveform model known as \texttt{TaylorF2}~\cite{BuonannoIyerOchsner2009}.
@@
 
-Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of waveform models.
-The essential idea is to construct an interpolating function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$.
-Previous works have focused on efficiently representing the parameter space of waveforms using a reduced basis of orthonormal functions whose linear combination can reproduce any waveform in the parameter space to some desired accuracy.
-The two most common approaches have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method that selects a subset of training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
-In this work, instead of using a reduced basis to efficiently represent the EOB waveform as a function of frequency, we will instead use the fact that the frequency-domain waveform is approximately known analytically via the TaylorF2 waveform.
+%\red{[[Purpose: The EOB waveform that's expensive but solves all of humanity's problems.]]}
+
+In this paper we will use a new implementation of the effective one body (EOB) formalism, that treats matter effects consistently with the other parameters.
+It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations.
+%Until recently, however, no version of the EOB formalism included the effects of spin and tides in the same model. 
+%Recently two versions have become available that include both tidal effects and aligned spin and are implemented under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}. 
+The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2017}.
+This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}.
+It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations.
+We also note an alternative model, \texttt{TEOBResumS}~\cite{Nagar2017, NewPaper2018}, that uses a slightly different point-particle and spin prescription.
+For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions.
+It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances or a tapering to zero amplitude after merger.
+\red{[[Must mention evaluation time to segue into next 2 paragraphs.]]}
+
+%One such alternative to these waveforms that do treat matter effects consistently with the other parameters is the effective one body (EOB) formalism discussed in section~\ref{sec:eob}.
+It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations.
+Until recently, however, no version of the EOB formalism included the effects of spin and tides in the same model.
+Recently two versions have become available that include both tidal effects and aligned spin and are implemented under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}.
+The \texttt{SEOBNRv4T} model~\cite{Hinderer:2016eia,Steinhoff:2016rfi} used in this paper, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2017}.
+This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}.
+It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations.
+The other model, \texttt{TEOBResumS}~\cite{Nagar2017, NewPaper2018}, uses a slightly different point-particle and spin prescription.
+For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions.
+It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances or a tapering to zero amplitude after merger.
+
+%\red{[[Purpose: Why other attempts for using EOB with parameter estimation have been insufficient.]]}
+
+Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible.
+For GW170817~\cite{PEPaper}, an alternative highly parallelized code, \texttt{RapidPE}, was used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{RapidPE, RIFT}.
+Unfortunately, while the results were broadly consistent with the MCMC results when using the same waveform models, there were noticeable differences.
+In particular the shape of the PDF was different, and the credible intervals for the tidal parameters were shifted downward by $\sim 10$\%~\cite{PEPaper}.
+Another limitation is that RapidPE does not currently marginalize over calibration errors.
+As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved.
+In addition, different methods of formulating the problem can allow for faster integration~\cite{RushTheInspiral}.
+However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the BNS model discussed here with dynamical tides.
+As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T}, discussed below, that bypasses the need for specialized parameter estimation tools and optimization techniques.
+
+%*********** Part 2 of intro: Surrogates ************
+
+Surrogate modeling techniques have had significant success in gravitational-wave data analysis for rapidly evaluating waveform models.
+The essential idea is to construct a fitting function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$.
+Previous works have focused on efficiently representing the space of all waveforms using a reduced basis of orthonormal functions.
+% whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. 
+The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}
+%method that selects a subset of training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
+In this work, instead of using a reduced basis, we will instead use the fact that the Fourier transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model.
 We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform.
-Because this residual is small, extremely high accuracy is not required, and we simply choose to represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.
-
-The final step is to interpolate between waveform parameters $\bx$.
-Previous work on surrogate models focused on parameter spaces with three or fewer dimensions.
+Because this residual is small, extremely high accuracy is not required, and we can efficiently represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.
+
+The final step in building a surrogate is to interpolate between waveform parameters $\bx$.
+Previous work focused on parameter spaces with three or fewer dimensions.
 This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}.
 These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
-For the problem here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible \MP{I think that this is too strong; Blackman17a also perform 5D TP interpolation which contradicts this.}.
-Recent work on optimally choosing waveforms for both EOB\MP{We used IMRPhenomD}~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.
+For the EOB waveform here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible
+%\MP{I think that this is too strong; Blackman17a also perform 5D TP interpolation which contradicts this.}. 
+Recent work on optimally choosing waveforms for analytic~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.
 
@@
 
-This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}.
-Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation.
-\MP{Is it really true that this is a significant speedup?
-FFT may still be cheaper than surrogate waveform generation.} It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
+This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}.
+Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that does not require an online Fourier transform for each waveform evaluation.
+%\MP{Is it really true that this is a significant speedup? FFT may still be cheaper than surrogate waveform generation.} 
+It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
 The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.
@@
 In section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model.
-We then compare the accuracy and speed of the surrogate to the original model in section IV.
+We then compare the accuracy and speed of the surrogate to the original model in Section IV.
+In section Section V we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817.
 Finally, we discuss applications and future improvements in Section V.
@@
 This is favorable for data analysis which is usually done in the frequency domain.
-It also allows us to extend the model down to arbitrarily low frequencies with analytic, frequency-domain PN approximations.
-Unfortunately, Fourier transforming a finite length, discretely sampled waveform leads to a surrogate with more noise.
+It also allows us to extend the model down to arbitrarily low frequencies with the analytic, frequency-domain \texttt{TaylorF2} model.
+Unfortunately, Fourier transforming a finite length waveform leads to a surrogate with more noise.
 We will show below, however, that sufficient filtering can solve this problem.
@@
 Because the waveform $\tilde h(Mf; \bx)$ is an oscillatory function of $Mf$ and $\bx$, the waveform is usually decomposed into an amplitude $A(Mf; \bx)$ and phase $\Phi(Mf; \bx)$ as $\tilde h(Mf; \bx) = A(Mf; \bx) e^{i\Phi(Mf; \bx)}$.
-The amplitude and phase are smoother, mostly monotonic functions of frequency \MP{the phase is not guaranteed to be monotonic}.
+The amplitude and phase are smoother, mostly monotonic functions of frequency.
+%\MP{the phase is not guaranteed to be monotonic}. 
 This can be seen in Fig.~\ref{fig:hoff} where we show the waveforms for the 32 corners of the 5-dimensional parameter space.
@@
 \end{align}
-The exact functional form of the TaylorF2 waveform is given in Appendix~\ref{sec:taylorf2}, and the residuals are shown in Fig.~\ref{fig:hoff}.
+The exact functional form of the TaylorF2 waveform we use is given in Appendix~\ref{sec:taylorf2}, and the residuals are shown in Fig.~\ref{fig:hoff}.
 
@@
 Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies.
-In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $\chi_1$.
+%In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $\chi_1$.
 These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work.
@@
 
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.49\textwidth]{dhofs.pdf}
-\caption{Cross sections of the amplitude and phase residuals as functions of $\chi_1$ at three nodes $MF_j$.
-The waveforms are generated for 51 values of $\chi_1$ shown as dots with the other parameters held constant at $\{q, S_2, \Lambda_1, \Lambda_2\} = \{0.6, 0.2, 2000, 1000\}$.
-Also shown are 1-dimensional cross sections of the interpolated residuals using Gaussian process regression (solid black) as well as their estimated $1\sigma$ uncertainty (shaded gray region).}
-\label{fig:dhofs}
-\end{figure}
-
 Finally, we have found that the functions $\Delta\ln A(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$ sometimes vary rapidly as a function of the tidal parameters $\Lambda_1$ and $\Lambda_2$ for tidal parameters in the range [0, 1000].
@@
 \begin{equation}
-\xi = \log_{10}(\Lambda/100 + 1).
+\xi_{A, B} = \log_{10}(\Lambda_{A, B}/100 + 1).
 \end{equation}
@@
 We evaluate the EOB waveform with a starting frequency of $Mf_{{\rm win}, i}=0.000197$, equivalent to a physical frequency of 20~Hz for a binary with total mass $M=2M_\odot$.
-In order to take a discrete Fourier transform, we window the start of the waveform with a Planck window~\cite{McKechanRobinsonSathyaprakash2010} in the interval $[Mf_{{\rm win}, i}, Mf_{{\rm win}, f}] = [0.000197, 0.00021]$ to avoid Gibbs oscillations.
+In order to take a discrete Fourier transform, we window the start of the waveform with a Planck window~\cite{McKechanRobinsonSathyaprakash2010} in the interval $[Mf_{{\rm win}, i}, Mf_{{\rm win}, f}] = [0.000197, 0.00021]$ to reduce Gibbs oscillations.
 The end of the waveform has zero amplitude, so the end does not need to be windowed.
@@
 In an attempt to preserve all structure in the EOB waveform, we choose not to window the end of the time-domain waveform.
-We instead remove these oscillations in the frequency domain using a moving average filter centered on $Mf$ with width $[Mf(1-\Delta Mf), Mf(1+\Delta Mf)]$.
-We use a width of $\Delta Mf=0.1$ for the amplitude residual and $\Delta Mf=0.05$ for the phase residual.
+We instead remove these oscillations in the frequency domain using a moving average filter centered on $Mf$ with interval $[Mf(1-\alpha), Mf(1+\alpha)]$.
+We use a width of $\alpha=0.1$ for the amplitude residual and $\alpha=0.05$ for the phase residual.
 Smoothing these oscillations makes it significantly easier to interpolate the amplitude and phase residuals as functions of $Mf$ and $\bx$.
-\red{[[$\Delta MF$ is a difference not a fraction.
-maybe try $p$ or something similar?
-Mention that 0.1 for the amplitude is a bit aggressive, but only matters for high frequency where the amplitude is crazy anyway.]]}
 
 Finally, we truncate the residuals outside the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}] = [0.00021, 0.07]$. 
+We note that the gravitational-wave frequency at the Schwarzschild ISCO is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$ and higher frequencies represent only the last $\sim 1$ cycle of the waveform during the merger.
+Thus, the high-frequency cutoff of $Mf_{{\rm trunc}, f}=0.07$ is a very conservative upper bound for data-analysis purposes.
+However, truncating the waveform at lower frequencies can visibly effect the shape of some waveforms if inverse Fourier transformed back into the time domain.
+
+%Finally, we truncate the residuals outside the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}] = [0.00021, 0.07]$.
 We note that the gravitational-wave frequency at the Schwarzschild ISCO is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$.
@@
 The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.052; \bx)$, where the amplitude is very small and dominated by numerical noise.
-\MP{The discussion of Fig.~\ref{fig:dhofs} appears twice in the text; here and one section earlier.}
+%\MP{The discussion of Fig.~\ref{fig:dhofs} appears twice in the text; here and one section earlier.}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.49\textwidth]{dhofs.pdf}
+\caption{Cross sections of the amplitude and phase residuals as functions of $\chi_1$ at three nodes $MF_j$.
+The waveforms are generated for 51 values of $\chi_1$, shown as dots, with the other parameters held constant at $\{q, S_2, \Lambda_1, \Lambda_2\} = \{0.6, 0.2, 2000, 1000\}$.
+Also shown are 1-dimensional cross sections of the interpolated residuals using Gaussian process regression (solid black) as well as their estimated $1\sigma$ uncertainty (shaded gray region).}
+\label{fig:dhofs}
+\end{figure}
 
@@
 
-For the problem here, we have found that global basis functions do not work well.
-\MP{EIM did not work well; is this also true for the SVD basis and projection coefficient method?} The residuals are small and smooth at low frequencies and large and noisy at high frequencies.
-With global basis functions, errors in evaluating $g(MF_j;\bx)$ at high frequencies can propagate to large errors between the nodes $MF_j$ at low frequencies.
+For the problem here, we have found that global basis functions are not necessarily the optimal solution.
+The residuals are small and smooth at low frequencies and large and noisy at high frequencies.
+If using the empirical interpolation method with global basis functions, for example, errors in evaluating $g(MF_j;\bx)$ at high frequencies can propagate to large errors between the nodes $MF_j$ at low frequencies.
 Instead, we find that spline interpolation works significantly better.
@@
 For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
-At $\sim10$~minutes--1~day per EOB waveform on a standard CPU \MP{For which $f_\text{min}$ are these times given?}, this is at the limit of what is reasonable.
+At a starting frequency of 20~Hz, \red{$\sim x$--y}~hours are needed to evaluate a single \texttt{SEOBNRv4T} waveform on a standard CPU, so this is at the limit of what is reasonable.
 However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below.
-The method we choose is Gaussian Process Regression (GPR)~\cite{RasmussenWilliams2006}.
-
-In GPR a function $g(\bx)$ is described in terms of a mean $m(\bx)$ and  covariance $k(\bx, \bx')$ between points.
-\begin{equation}
-g(\bx) \approx \mathcal{GP}(m(\bx), k(\bx, \bx')).
-\end{equation}
-The mean $m(\bx)$ can be a parameterized fitting function while $k(\bx, \bx')$ is a kernel with tunable hyperparameters that describes the covariance between the point $\bx$ and the sampled point $\bx'$.
-This covariance can be used to describe features of the function not captured by the parameterized mean $m(\bx)$ as well as genuine noise in the data $\by$.
-For the problem here, we have already subtracted the TaylorF2 waveform from the EOB waveform, so we set $m(\bx) = 0$ and model the residuals purely in terms of the kernel $k(\bx, \bx')$.
-
-In a zero-mean Gaussian process, the data $y_i$ and function value $y_*$ at the training set points $\bx_i$ and new point $\bx_*$, respectively, are drawn from a multivariate normal distribution
+The method we choose is Gaussian process regression (GPR)~\cite{RasmussenWilliams2006}.
+
+In GPR, the values of a function $g(\bx)$ at the points $\bx$ are assumed to be a realization of a Gaussian process
+\begin{equation}
+g(\bx) \sim \mathcal{GP}(m(\bx), k(\bx, \bx')),
+\end{equation}
+where $m(\bx)$ is a mean function at each point $\bx$ and $k(\bx, \bx')$ is a covariance function between the points $\bx$ and $\bx'$.
+When modeling data with GPR, one often attempts to subtract out the mean by first fitting the data with a parameterized function.
+Then, the residual is represented by a zero-mean Gaussian process.
+The covariance $k(\bx, \bx')$ is described by a kernel function with tunable hyperparameters.
+For the problem here, we have already subtracted the TaylorF2 waveform from the EOB waveform, so we model the residual in terms of a zero-mean Gaussian process.
+
+Explicitly, we are interested in predicting the function value $y_*$ at the point $\bx_*$ given the sampled function values $y_i$ at the points $\bx_i$.
+In a zero-mean Gaussian process, this is represented by the following multivariate Gaussian distribution
 \begin{equation}
@@
 \begin{bmatrix}
-{\bm y} \\
+y_i \\
 y_* \\
@@
 
-The conditional probability for $y_*$ given the training set examples $\by$ and kernel hyperparameters $\btheta$ is also a Gaussian
-\begin{equation}
-p(y_* | \bx_i, \bx_*, \by, \btheta) \sim \mathcal{N}(\bar y_*, {\rm Var}(y_*)),
+From Eq.~\eqref{eq:gaussian}, the conditional probability for $y_*$ given the training set examples $y_i$ and kernel hyperparameters $\btheta$ is also a Gaussian
+\begin{equation}
+p(y_* | \bx_i, \bx_*, y_i, \btheta) = \mathcal{N}(\bar y_*, {\rm var}(y_*)),
 \end{equation}
@@
 \label{eq:mean}
-\bar y_* &= K_*^T K^{-1} y, \\
+\bar y_* &= K_{*i} (K^{-1})_{ij} y_j, \\
 \label{eq:var}
-{\rm Var}(y_*) &= K_{**} - K_*^T K^{-1} K_*.
-\end{align}
-Eq.~\eqref{eq:mean} is the estimate of the function, and Eq.~\eqref{eq:var} is the estimate of the uncertainty. We note that the variance does not depend on the data $\by$.
+{\rm var}(y_*) &= K_{**} - K_{*i} (K^{-1})_{ij} K_{*j}.
+\end{align}
+Eq.~\eqref{eq:mean} is the estimate of the function, and Eq.~\eqref{eq:var} is the estimate of the uncertainty.
 
@@
 \end{equation}
-The tunable hyperparameters $\ell_i$ represent the length scale over which the function $g(\bx)$ varies in each coordinate $x_i$. Specifically, we use the Mat\'{e}rn class of kernels
+The tunable hyperparameters ${\bf \ell}$ represent the length scale over which the function $g(\bx)$ varies in each coordinate. Specifically, we use the Mat\'{e}rn class of kernels
 \begin{equation}
@@
 The value of $\nu$ parameterizes the smoothness of the Gaussian process, and is $k$ times mean-square differentiable if $\nu>k$~\cite{RasmussenWilliams2006}.
-For half-integer values of $\nu$, this kernel has a computationally cheap form without special functions, and we have had good results with $\nu=5/2$, resulting in a twice-differentiable function\footnote{If desired, one can select the optimal class of kernels using cross validation~\cite{RasmussenWilliams2006}, and we have also tried the more common, infinitely differentiable squared exponential kernel~\cite{RasmussenWilliams2006}.
+For half-integer values of $\nu$, this kernel has a computationally cheap form without special functions, and we have had good results with $\nu=5/2$, resulting in a twice-differentiable function\footnote{We have also tried the more common, infinitely differentiable squared exponential kernel~\cite{RasmussenWilliams2006}.
 However, we found it difficult to reliably tune its hyperparameters for the problem here.}.
@@
 where $\sigma_f$ is a scale factor that describes the range of values that $g(\bx)$ takes over the domain, and $\sigma_n$ is a noise parameter.
-The white noise kernel $\sigma_n^2 \delta_{ij}$ (also called a nugget) is $\sigma_n^2$ when $\bx_i = \bx_j$ and zero otherwise, and parameterizes noise in the data $\by$.
+The white noise kernel $\sigma_n^2 \delta_{ij}$ (also called a nugget) parameterizes the noise in the data $y_i$.
 In our case, the training set waveforms have numerical noise that we will estimate by optimizing the hyperparameters.
-The full set of hyperparameters is now  $\btheta = \{\sigma_f, \ell_q, \ell_{\chi_1}, \ell_{\chi_2}, \ell_{\Lambda_1}, \ell_{\Lambda_2}, \sigma_n\}$.
+The full set of hyperparameters is now  $\btheta = \{\sigma_f, \ell_q, \ell_{\chi_1}, \ell_{\chi_2}, \ell_{\xi_1}, \ell_{\xi_2}, \sigma_n\}$ for our parameter space $\bx = \{q, \chi_1, \chi_2, \xi_1, \xi_2\}$.
 
 In order to estimate the hyperparameters, we use the above assumption (Eq.~\eqref{eq:gaussian}) that the joint distribution of 
-the data ${\bm y}$ is a multivariate Gaussian which has the following distribution
-\begin{equation}
-\ln p({\bm y} | {\bm x}, {\bm \theta}) = -\frac{1}{2}{\bm y}^T K^{-1} {\bm y} - \frac{1}{2} \ln |K| - \frac{d}{2} \ln 2\pi.
-\end{equation}
-This is the log-likelihood for ${\bm y}$ given the hyperparameters ${\bm \theta}$, and we can find the posterior for ${\bm \theta}$
-given ${\bm y}$ using Bayes' theorem
-\begin{equation}
-p({\bm \theta} | {\bm x}, {\bm y}) \propto p({\bm \theta}) p({\bm y} | {\bm x}, {\bm \theta}).
+the data $y_i$ is a multivariate Gaussian with the following distribution:
+\begin{equation}
+\ln p(y_i | \bx_i, {\bm \theta}) = -\frac{1}{2} y_i (K^{-1})_{ij} y_j - \frac{1}{2} \ln |K| - \frac{d}{2} \ln 2\pi.
+\end{equation}
+This is the log-likelihood for $y_i$ given the hyperparameters ${\bm \theta}$, and we can find the posterior for ${\bm \theta}$
+given $y_i$ using Bayes' theorem
+\begin{equation}
+p({\bm \theta} | \bx_i, y_i) \propto p({\bm \theta}) p(y_i | \bx_i, {\bm \theta}).
 \end{equation}
@@
 With these optimized hyperparameters, the final interpolating function is given by Eq.~\eqref{eq:mean} and its uncertainty by Eq.~\eqref{eq:var}.
-We note that once the vector $K^{-1}y$ is precomputed, evaluating Eq.~\eqref{eq:mean} is a fast $\mathcal{O}(N)$ operation.
+We note that once the vector $(K^{-1})_{ij}y_j$ is precomputed, evaluating Eq.~\eqref{eq:mean} is a fast $\mathcal{O}(N)$ operation.
 
@@
 
-\MP{Say that we use sparse frequency grids based on power-law $\Delta f$ in the evaluation of the surrogate; perhaps describe in more detail how the model is evaluated in the LAL code.}
-\MP{Discuss linear vs cubic splines if we decide to use them. Tradeoff in accuracy and speed.}
-
-Given the interpolating functions for the amplitude and phase residuals $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ and $\{\mathcal{I}_{\rm GPR, j}[\Delta\Phi_j](\bx)\}$ evaluated at the frequency nodes $MF_j$, we can now reconstruct the frequency-domain waveform.
+Given the GPR fits for the amplitude and phase residuals, we can now reconstruct the frequency-domain waveform.
+We label the set of GPR fits at the frequency nodes $MF_j$ by $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ for the amplitude residuals and by $\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}$ for the phase residuals.
 The surrogates for the residuals $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$ are constructed by interpolating between the nodes $MF_j$ with cubic splines:
@@
 \Delta\ln A_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}](Mf),\\
-\Delta\Phi_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\mathcal{I}_{\rm GPR}[\{\Delta\Phi_j](\bx)\}](Mf).
-\end{align}
-We set these functions to 0 below the first frequency node $MF_0 =  Mf_{trunc,i}$ so that the waveform transitions to TaylorF2 at lower frequencies.
+\Delta\Phi_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}](Mf).
+\end{align}
+We set these functions to 0 below the first frequency node $MF_0 =  Mf_{trunc,i}$ so that the waveform transitions to \texttt{TaylorF2} at lower frequencies.
 With the analytic expressions for $A_{\rm F2}(Mf;\bx)$ and $\Phi_{\rm F2}(Mf;\bx)$ and the interpolated expressions $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$, the final surrogates for the amplitude and phase are
 \begin{align}
+\label{eq:amp_sur}
 A_S(Mf;\bx) &= A_{\rm F2}(Mf;\bx) \exp\left[ \Delta\ln A_S(Mf;\bx) \right] \\
+\label{eq:phase_sur}
 \Phi_S(Mf;\bx) &= \Phi_{\rm F2}(Mf;\bx) + \Delta\Phi_S(Mf;\bx).
@@
 \begin{align}
+\label{eq:hplus_sur}
 \tilde h_+(f; \bx) &= \frac{1}{2}(1+\cos^2\iota) \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
 & \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right)\right], \\
+\label{eq:hcross_sur}
 \tilde h_\times(f; \bx) &= \cos\iota \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
@@
 
+\red{[[Michael: Add your discussion of resampling the waveform here.]]}
+%\MP{Say that we use sparse frequency grids based on power-law $\Delta f$ in the evaluation of the surrogate; perhaps describe in more detail how the model is evaluated in the LAL code.}
+%\MP{Discuss linear vs cubic splines if we decide to use them. Tradeoff in accuracy and speed.}
+
 \subsection{Iterative construction of training set and surrogate}
 \label{sec:design}
+\red{[[Can't finish this section until the final sampling method is settled on.]]}
 
@@
 
-%{\scriptsize
-%\begin{algorithm}[H]
-%\caption{Empirical Interpolation (EI) Method}
-%\label{alg:eim}
-%\begin{algorithmic}[1]
-%%\State {\bf Input:} $\{ \vec{e}_i \}_{i=1}^n$, $\{t_i\}_{i=1}^L$
-%\State {\bf Input:} $\{ e_i \}_{i=1}^n$, $t := \{t_i\}_{i=1}^L$
-%\vskip 10pt
-%\State $i = \text{argmax} | e_1(t) |$ (\text{argmax} returns the largest entry of its argument). 
-%\State Set $T_1 = t_i$
-%\For{$j = 2 \to n$} 
-%\State Build ${\cal I}_{j-1} [e_j](t)$ from (\ref{eq:empinterpX1})--(\ref{eq:empinterpX2})
-%\State $\vec{r} = {\cal I}_{j-1} [e_j](t) -e_j(t)$
-%\State $i = \text{argmax} |\vec{r}|$
-%\State $T_j = t_i$
-%\EndFor
-%\vskip 10pt
-%\State {\bf Output:} EI nodes $\{ T_i \}_{i=1}^n$,  interpolant operator ${\cal I}_n$
-%\end{algorithmic}
-%\end{algorithm}
-%}
-
 For $i = 1, \dots, N_{\rm new}$:
@@
 %% ________________________________________________________
-\section{Results}
+\section{Performance of the surrogate}
 
@@
 
-A common measure of the surrogate model accuracy is the mismatch between 
-the surrogate model and the original EOB waveform.
-The mismatch represents the loss in signal-to-noise ratio that would result 
-from using the surrogate model instead of the original EOB waveform. 
-It is defined by the deviation from a perfect overlap after aligning the two waveforms
-using the time and phase free parameters $t_0$ and $\phi_0$:
+A common measure of the surrogate model accuracy is the mismatch between the surrogate model and the original EOB waveform.
+The mismatch represents the loss in signal-to-noise ratio that would result from using the surrogate model instead of the original EOB waveform.
+It is defined as the deviation from a perfect overlap after aligning the two waveforms using the time and phase free parameters $t_0$ and $\phi_0$:
 \begin{equation}
@@
 \end{equation}
-The inner product here  is the integral of the Fourier transformed waveforms $\tilde h(f)$ weighted by the noise power spectral 
-density (PSD) $S_n(f)$ of the detector:
-\begin{equation}
-(h_1, h_2) = 4 \Re \int_{f_{\rm low}}^{f_{\rm high}} \frac{\tilde h_1(f) \tilde h^*_2(f)} {S_n(f)} df.
-\end{equation}
-
-In Fig.~\ref{fig:mismatch}, we show the distribution of mismatch $\mathcal{M}$ between our surrogate
-and the $10^3$ randomly sampled EOB waveforms. We use the design sensitivity aLIGO PSD~\cite{Aasi:2013wya} and
-a sampling rate of 4096~Hz. Our integration bounds are $f_{\rm low} = x$~Hz and the Nyquist frequency
-$f_{\rm high} = 2048$~Hz. Because the surrogate can be rescaled with mass,
-we show results for the smaller mass $M_B$ fixed at $1M_\odot$ or fixed at $2M_\odot$.
-\red{[[This is all wrong. Fix it.]]}
-We compare this mismatch to that of th
+The inner product between waveforms $h_1$ and $h_2$ is
+\begin{equation}
+(h_1, h_2) = 4 \Re \int_{f_{\rm low}}^{f_{\rm high}} \frac{\tilde h_1(f) \tilde h^*_2(f)} {S_n(f)} df,
+\end{equation}
+where the Fourier transformed waveforms are weighted by the noise power spectral density (PSD) $S_n(f)$ of the detector.
+
+In Fig.~\ref{fig:mismatch}, we show the distribution of mismatches $\mathcal{M}$ between our surrogate and the 1000 randomly sampled EOB waveforms.
+We use the design sensitivity aLIGO PSD~\cite{Aasi:2013wya} and a total mass of $M=2.8M_\odot$.
+Our integration bounds are $Mf_{\rm low} = 0.00021$ and $Mf_{\rm high} = 0.07$, corresponding to physical units of \red{15.2}~Hz and \red{5076}~Hz respectively.
+As we increase the number of samples in our training set, the mismatches decrease and, for our final surrogate, we find a maximum mismatch of \red{$5.5\times 10^{-4}$}.
 
@@
 The vertical solid line is the gravitational-wave frequency at the Schwarzschild ISCO.}
-\label{fig:maxmismatch}
+\label{fig:maxmismatchfd}
 \end{figure}
 
-To demonstrate that we can correctly recover the behavior of the original unfiltered time-domain EOB waveform, we inverse Fourier transform the surrogate and compare it to the EOB waveform.
-In Fig.~\ref{fig:maxmismatchtd} we compare the surrogate to the test set waveform that had the largest mismatch.
-We align the two waveforms in time and phase by maximizing the overlap at early times in the interval $t/M \in [-2\times 10^5, -1\times 10^5]$ [[cite Read et al.
-for details.]].
-The two waveforms are nearly identical except during the last $\sim 2$ cycles.
+In Fig.~\ref{fig:maxmismatchfd} we show the amplitude and phase errors as a function of frequency for the waveforms with the largest mismatch.
+As we add waveforms with the various sampling methods, the errors generally decrease.
+However, above a frequency of $Mf \sim 0.03$, the amplitude of the training set waveforms are noisy, and it is thus difficult to produce a good fit.
+This is not a problem, however, because aLIGO is insensitive to the waveform morphology at such high frequencies.
+
+Finally, to demonstrate that we can correctly recover the behavior of the original, time-domain EOB waveform, we inverse Fourier transform the surrogate and compare it to the time-domain waveform.
+In Fig.~\ref{fig:maxmismatchtd} we compare the final surrogate to the test set waveform that has the largest mismatch.
+We align the two waveforms in time and phase by maximizing the overlap at early times in the interval $t/M \in [-2\times 10^5, -1\times 10^5]$ using the method of~\cite{Read2009}.
+The two waveforms are nearly identical.
 
@@
 
-\MP{Add cost of complex exponential sincos function: look at profiling results}
-\MP{Potentially, explore approximation to sincos for further speedup; need to guarantee that errors don't add up; range reduction is important for accuracy}
-
-The surrogate evaluation time can be roughly decomposed into two times.
-The first is the time needed to evaluate the residuals $\Delta\ln(A)(F_j; \bx)$ and $\Delta\Phi(F_j; \bx)$ at each of the $N_A+N_\Phi$ interpolating nodes.
-Although optimizing the hyperparameters scales with the number of samples $n$ as $\mathcal{O}(n^3)$ due to the required matrix inversion, the evaluation time of a stored GPR scales as $\mathcal{O}(n)$.
-The evaluation time for the residuals will therefore be fixed cost of $\mathcal{O}(n(N_A+N_\Phi))$.
-The second part is the time needed to resample the final surrogate (Eq.[[ref]]) at uniformly spaced frequency samples in physical units beginning at a starting frequency $f_{\rm low}$.
+The surrogate evaluation time is the sum of two parts.
+The first part is the time needed to evaluate the residuals $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ and $\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}$ at each of the $N_A+N_\Phi$ interpolating nodes using GPR.
+Although optimizing the hyperparameters for GPR scales with the number of samples $N$ as $\mathcal{O}(N^3)$, the evaluation time of a stored GPR scales as $\mathcal{O}(N)$.
+The evaluation time for all the residuals therefore has a cost of $\mathcal{O}(N(N_A+N_\Phi))$.
+The second part is the time needed to resample the final surrogate (Eq.~\eqref{eq:amp_sur}--Eq.~\eqref{eq:hcross_sur}) at uniformly spaced frequency samples in physical units beginning at a starting frequency $f_{\rm low}$.
+\red{[[Michael: If you want, add details of how the specific implementation of resampling the waveform affects timing here.]]}
+%\MP{Add cost of complex exponential sincos function: look at profiling results}
+%\MP{Potentially, explore approximation to sincos for further speedup; need to guarantee that errors don't add up; range reduction is important for accuracy}
 
 In Fig.~\ref{fig:timing}, we show the surrogate evaluation time as a function of starting frequency $f_{\rm low}$.
-The surrogate has a fixed time of $\sim 50$~ms to evaluate the GPR at each node.
-The additional cost to resample the amplitude and phase at uniformly-spaced frequencies with spline interpolation dominates at starting frequencies below $\sim 30$~Hz.
+The surrogate has a fixed time of \red{$\sim 50$~ms} to evaluate the GPR all of the nodes.
+\red{[[Details of how starting frequency depends on $f_{\rm low}$.]]} The additional cost to resample the amplitude and phase at uniformly-spaced frequencies with spline interpolation dominates at starting frequencies below $\sim 30$~Hz.
 We also compare the evaluation time to several BNS waveform models that have been used for LIGO data analysis.
-We note that at $\sim 30$~Hz, the \texttt{SEOBNRv4T\_surrogate} model is comparable to the \texttt{SEOBNRv4\_ROM\_NRTidal} and \texttt{IMRPhenomD\_NRTidal} models that were used in the analysis of the GW170817 event~\cite{DietrichBernuzziTichy2017}.
-At lower frequencies, the surrogate is faster than all BNS waveform models except for TaylorF2.
+We note that at $\sim 30$~Hz, the \texttt{SEOBNRv4T\_surrogate} model is comparable to the \texttt{SEOBNRv4\_ROM\_NRTidal} and \texttt{IMRPhenomD\_NRTidal} models that were used in the analysis of the GW170817 event~\cite{PEPaper}.
+At lower frequencies, the surrogate is faster than all BNS waveform models except for \texttt{TaylorF2}.
 
@@
 
-\subsection{Parameter estimation}
-
-\MP{TODO: add 2D density plot of which quantities? masses, spins, lambdas?
-or mass-ratio, eff spin, eff lambda?}
+\section{Parameter estimation}
+\label{sec:pe}
+
+\subsection{Injection runs}
+
+\subsection{GW170817}
 
@@
 
-This model notably does not include two effects that can be important for systems with large spin.
-The first is the spin-induced quadrupole moment $Q$ that effects the waveform phase at 2PN order~\cite{Poisson1998}.
-The easiest way to correct for this effect is to simply add this 2PN term to the TaylorF2 reference waveform.
-One can use the universal relation between the quadrupole moment $Q$ and the tidal parameter $\Lambda$ as was done with the other NS matter parameters.
-Once this effect is incorporated into the EOB model, one could easily rebuild the surrogate presented in this paper to incorporate this effect.
-The quadrupole moment can be approximated in terms of the tidal parameter $\Lambda$ using a so-called universal relation~\cite{YagiYunes2013}.
-
-The second effect is precession for non-aligned spins.
+This model notably does not include precession for non-aligned spins.
 Although none of the EOB models currently available include both tidal effects and precession, there are ways to boost this waveform model to a precessing frame and approximately incorporate precession.
@@
 This approach would provide a fast, accurate model for BNS systems with tides and generic spins.
+We note, however, that including precession using the \texttt{IMRPhenomP\_NRTidal} model did not appear to affect the results in~\cite{PEPaper}.
 

### 2018-09-26 paper/paper.tex
Surrogate model used for PE runs. Many paper edits. References now complete.

@@
 
+\red{only use texttt for actual programs and functions. TaylorF2 should usually be standard text}
+
 \section{Introduction}
@@
 
-The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors has demonstrated that tidal interactions during the inspiral can be measured with enough precision to constrain the radius and equation of state~\cite{DeEtA, PEPaper, EOSPaperl} of NSs.
+The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors has demonstrated that tidal interactions during the inspiral can be measured with enough precision to constrain the radius and equation of state~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, BNSPE, EOSPaper} of NSs.
 Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further.
 These multiple events could then be stacked to significantly improve measurements of the radius and EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}.
-\red{[[Mention A+ being funded?
-How will this improve each parameter?]]}
+%\red{[[Mention A+ being funded? How will this improve each parameter?]]}
 
@@
 
-These results are obtained using a Bayesian analysis that compares the data to parameterized waveform models. 
+These results are obtained using Bayesian analyses that compare the data to parameterized waveform models.
 Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations.
@@
 
-In addition, to avoid systematic errors that can be as large as statistical errors, highly accurate models are needed that include all relevant physical effects, including point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
-For GW170817, a set of four frequency-domain analytic or semi-analytic waveform models were used as templates (see Table I of~\cite{PEPaper}) for the main results.
-These models are based on spinning binary black hole (BBH) waveform models.
-They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2~\cite{}), the aligned-spin effective one body (EOB) formalism (SEOBNRT~\cite{}) or the phenomenological formalism for aligned-spin (PhenomDNRT~\cite{}) and precessing spin (PhenomPNRT~\cite{}).
+In addition, to avoid systematic errors that can be as large as statistical errors, highly accurate waveform models are needed that include all relevant physical effects, including point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
+For GW170817, a set of four frequency-domain waveform models were used as templates for the main results.
+These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein).
+They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT).
 Tidal interactions, on the other hand, were included simply by adding a correction to the phase of each waveform.
-For the PN TaylorF2 waveform the analytic 5PN and 6PN order terms were added.
-For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in~\cite{PEPaper}) was used for the phase correction~\cite{DietrichBernuzziTichy2017, NewNRTidalPaper}.
+For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added.
+For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was used for the phase correction~\cite{DietrichBernuzziTichy2017}.
 
@@
 In this paper we will use a new implementation of the effective one body (EOB) formalism, that treats matter effects consistently with the other parameters.
-It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations.
-%Until recently, however, no version of the EOB formalism included the effects of spin and tides in the same model. 
-%Recently two versions have become available that include both tidal effects and aligned spin and are implemented under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}. 
-The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2017}.
+It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations.
+The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}.
 This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}.
 It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations.
-We also note an alternative model, \texttt{TEOBResumS}~\cite{Nagar2017, NewPaper2018}, that uses a slightly different point-particle and spin prescription.
+We also note an alternative model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a slightly different point-particle and spin prescription.
 For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions.
 It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances or a tapering to zero amplitude after merger.
-\red{[[Must mention evaluation time to segue into next 2 paragraphs.]]}
+%\red{[[Must mention evaluation time to segue into next 2 paragraphs.]]}
 
@@
 Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible.
-For GW170817~\cite{PEPaper}, an alternative highly parallelized code, \texttt{RapidPE}, was used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{RapidPE, RIFT}.
+For GW170817~\cite{BNSPE}, an alternative highly parallelized code, \texttt{RapidPE}, was used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}.
 Unfortunately, while the results were broadly consistent with the MCMC results when using the same waveform models, there were noticeable differences.
-In particular the shape of the PDF was different, and the credible intervals for the tidal parameters were shifted downward by $\sim 10$\%~\cite{PEPaper}.
+In particular the shape of the PDF was different, and the credible intervals for the tidal parameters were shifted downward by $\sim 10$\%~\cite{BNSPE}.
 Another limitation is that RapidPE does not currently marginalize over calibration errors.
 As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved.
-In addition, different methods of formulating the problem can allow for faster integration~\cite{RushTheInspiral}.
-However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the BNS model discussed here with dynamical tides.
-As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T}, discussed below, that bypasses the need for specialized parameter estimation tools and optimization techniques.
+In addition, different methods of formulating the problem can allow for faster integration~\cite{NagarNettegno2018}.
+However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides.
+As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.
 
@@
 % whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. 
-The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}
+The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
 %method that selects a subset of training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
@@
 This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}.
-These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
-For the EOB waveform here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible
-%\MP{I think that this is too strong; Blackman17a also perform 5D TP interpolation which contradicts this.}. 
+These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality: for $N$ points per dimension $d$,  $N^d$ waveforms are needed.
+For the EOB waveform here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible.
 Recent work on optimally choosing waveforms for analytic~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.
@@
 This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set.
-Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of eccentric BBH mergers~\cite{HuertaMooreKumar2017}.
+Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of nonspinning, eccentric BBH mergers~\cite{HuertaMooreKumar2017}.
 GPR has also been used in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016}.
@@
 This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}.
-Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that does not require an online Fourier transform for each waveform evaluation.
-%\MP{Is it really true that this is a significant speedup? FFT may still be cheaper than surrogate waveform generation.} 
+Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation.
 It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
-The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.
+%The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.
 
@@
 In Section II we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space.
-In section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model.
+In Section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model.
 We then compare the accuracy and speed of the surrogate to the original model in Section IV.
@@
 \label{sec:eob}
+
+\subsection{Inspiral-plunge waveform}
 
@@
 
-More precisely, spin effects are those of the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins.
-The model, however, does not account for the spin-induced quadrupole moment~\cite{Poisson1998}.
+For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are those of the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins.
+\red{It includes spin-orbit interactions up to 3PN order and spin-spin interactions up to 2PN order.
+In addition, for the \texttt{SEOBNRv4T} model, the spin-induced quadrupole moment~\cite{Poisson1998} is also included.
+This effect appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin.} \red{\sout{The model, however, does not account for the spin-induced quadrupole moment~\cite{Poisson1998}.
 This effect appears at 2PN order as compared to the tidal effects that first appear at 5PN order, but is only significant for systems with large spin.
-The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{HarryHinderer2017}.
-
-\subsection{Inspiral-plunge waveform}
+The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{HarryHinderer2018}}}.
 
@@
 If either component of the binary is a black hole, then we set the tidal polarizabilities to zero.
-Let $M_{A,B}$ be the masses bodies $A$ and $B$, and $M = M_A + M_B$ be the total mass of the binary.
 The tidal contribution, including quadrupolar and octupolar dynamical tides, is
@@
 \end{widetext}
-Here, $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
+Here, $M_{A,B}$ are the masses of bodies $A$ and $B$, $M = M_A + M_B$ is the total mass, $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
 Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$\footnote{Here, $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.}.
@@
 \begin{equation}
-\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
+\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
 \end{equation}
@@
 with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$.
-Here, $\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor (Eq.~(11) of Ref.~\cite{Dietrich:2017feu}), which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.
+Here, $\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor (Eq.~(11) of Ref.~\cite{Dietrich:2017feu}), which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.
 This enhancement factor is 1 for low orbital frequencies, and increases to $\sim 2$ as $\ell$ times the orbital frequency approaches the $f$-mode frequency.
@@
 
+\subsection{Post-merger waveform}
+
 We now look for a suitable time $t_{\rm match}$ to stop the numerical integration and match with a post-merger waveform.
-In the point-mass model \texttt{SEOBNRv4}, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks, and $t_{\rm match}$ is chosen to be the time of the peak amplitude $t_{\rm peak}^{\rm amp}$.
-Here, we do not use nonquasicircular corrections tuned to BNS simulations \red{[[Ambiguous.
-Do we use them or not.
-Find out.]]}, so we have less control over the behavior of Eq.~(\ref{hlminspplunge}) in the late inspiral.
+In the point-mass model, \texttt{SEOBNRv4}, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks, and $t_{\rm match}$ is chosen to be the time of the peak amplitude $t_{\rm peak}^{\rm amp}$.
+Here, we do not use nonquasicircular corrections tuned to BNS simulations, so we have less control over the behavior of Eq.~(\ref{hlminspplunge}) in the late inspiral.
 We therefore choose the following definition for $t_{\rm match}$.
@@
 Is this paragraph already in the intro and can be deleted?]]}
-
-\subsection{Post-merger waveform}
 
@@
 where $a = |h_{22}^{\rm insp-plunge}(t_{\rm match})|$ and $b=\partial_t |h_{22}^{\rm insp-plunge}(t_{\rm match})|$.
-The tapering function is centered $15M$ after $t_{\rm match}$ and has a decay time $\tau = \pi/(2\omega_{\rm match})$ (where $\omega_{\rm match} = \omega_{22}^{\rm insp-plunge}(t_{\rm match})$) of one gravitational wave period.
-\red{[[As written, this is a decay time of 1/4 of a period.
-Figure out what is correct.]]} It is given by
+The tapering function is centered $15M$ after $t_{\rm match}$ and has a decay time $\tau = 2\pi / \omega_{\rm match}$ (where $\omega_{\rm match} = \omega_{22}^{\rm insp-plunge}(t_{\rm match})$) of one gravitational wave period.
+It is given by
 \begin{equation}
@@
 \end{align}
-where $\Delta\omega = \omega_{\rm match} - \omega_{\rm freq}$. Integrating and requiring continuity at $t_{\rm freq}$ results in the final expression for the phase
+where $\Delta\omega = \omega_{\rm match} - \omega_{\rm freq}$. Integrating the frequency, and requiring continuity at $t_{\rm freq}$, results in the final expression for the phase
 \begin{align}
@@
 For example, whether the merging binary undergoes prompt collapse or forms a hypermassive remnant depends sensitively on the total mass.
-NR simulations will be needed to determine when after the merger this approximation breaks down.
+NR simulations will be needed to determine when after the merger this approximation is no longer valid.
 
@@
 
-In this section we describe how we decompose the EOB waveform into smooth, slowly-varying functions and train a surrogate model for these functions.
+In this Section we describe how we decompose the EOB waveform into smooth, slowly-varying functions and train a surrogate model for these functions.
 We will work with the frequency-domain waveform $\tilde h(Mf; \bx)$.
 This is favorable for data analysis which is usually done in the frequency domain.
-It also allows us to extend the model down to arbitrarily low frequencies with the analytic, frequency-domain \texttt{TaylorF2} model.
+It also allows us to extend the model down to arbitrarily low frequencies with an analytic, frequency-domain model.
 Unfortunately, Fourier transforming a finite length waveform leads to a surrogate with more noise.
@@
 The amplitude and phase are smoother, mostly monotonic functions of frequency.
-%\MP{the phase is not guaranteed to be monotonic}. 
 This can be seen in Fig.~\ref{fig:hoff} where we show the waveforms for the 32 corners of the 5-dimensional parameter space.
@@
 The phase, for example, spans about $10^4$~rad between waveforms with different parameters $\bx$ (see Fig.~\ref{fig:hoff}).
-To avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad.
+To avoid systematic errors in the tidal parameters, for example, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad.
 This is a difficult requirement to achieve for 5-dimensional interpolation.
@@
 Vertical dashed lines represent the frequency nodes $MF_j$ where the residuals are interpolated as functions of $\bx$ using Gaussian process regression.
-The vertical solid line at $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$ is the gravitational-wave frequency at the Schwarzschild ISCO.}
+The vertical solid line at $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$ is the gravitational-wave frequency at the Schwarzschild ISCO, and the top axis is the frequency for a binary with a total mass of $2.8M_\odot$.}
 \label{fig:hoff}
@@
 
-For aligned-spin waveforms, we can solve this problem by using the fact that the waveform can be approximated with the analytic \texttt{TaylorF2} waveform $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$.
+For aligned-spin waveforms, we can solve this problem by using the fact that the waveform can be approximated with the analytic TaylorF2 waveform $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$.
 This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2
@@
 Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies.
-%In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $\chi_1$.
-These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work.
-The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise is significant.
 
@@
 \begin{equation}
-\xi_{A, B} = \log_{10}(\Lambda_{A, B}/100 + 1).
+\xi_{A, B} = \log_{10}\left(\frac{\Lambda_{A, B}}{100} + 1\right).
 \end{equation}
@@
 We could make this window longer, but that would require us to start the surrogate at a higher frequency.
-The second source comes from the end of the waveform where the amplitude rapidly drops to zero amplitude after the peak amplitude in $\sim 1$ cycle.
-The Gibbs oscillations at high frequencies are therefore a genuine feature of the EOB model.
-In an attempt to preserve all structure in the EOB waveform, we choose not to window the end of the time-domain waveform.
-We instead remove these oscillations in the frequency domain using a moving average filter centered on $Mf$ with interval $[Mf(1-\alpha), Mf(1+\alpha)]$.
+The second source comes from the end of the waveform where the amplitude rapidly drops to zero amplitude during the $\sim 1$ cycle after the peak amplitude.
+Some contribution to the Gibbs oscillations at high frequencies is therefore a genuine feature of the EOB model.
+We reduce these oscillations in the frequency domain using a moving average filter centered on $Mf$ with interval $[Mf(1-\alpha), Mf(1+\alpha)]$.
 We use a width of $\alpha=0.1$ for the amplitude residual and $\alpha=0.05$ for the phase residual.
@@
 Finally, we truncate the residuals outside the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}] = [0.00021, 0.07]$. 
-We note that the gravitational-wave frequency at the Schwarzschild ISCO is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$ and higher frequencies represent only the last $\sim 1$ cycle of the waveform during the merger.
+We note that the gravitational-wave frequency at the Schwarzschild innermost stable circular orbit (ISCO) is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$, and higher frequencies represent only the last $\sim 1$ cycle of the waveform during the merger.
 Thus, the high-frequency cutoff of $Mf_{{\rm trunc}, f}=0.07$ is a very conservative upper bound for data-analysis purposes.
 However, truncating the waveform at lower frequencies can visibly effect the shape of some waveforms if inverse Fourier transformed back into the time domain.
-
-%Finally, we truncate the residuals outside the interval $[Mf_{{\rm trunc}, i}, Mf_{{\rm trunc}, f}] = [0.00021, 0.07]$.
-We note that the gravitational-wave frequency at the Schwarzschild ISCO is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$.
-Although second generation detectors are only marginally sensitive to BNS signals above this frequency, we use the high-frequency cutoff of $Mf_{{\rm trunc}, f}=0.07$ because the amplitude of some of the EOB waveforms is still large enough at these high frequencies to affect the structure of the surrogate waveform if inverse Fourier transformed back into the time domain.
-\MP{Say that Mf=0.07 is a (generous?) upper bound on the NR BNS merger frequency and thus a safe choice.}
 
@@
 The exception is the noisy amplitude residual at high frequencies, $\Delta\ln(A)(Mf=0.052; \bx)$, where the amplitude is very small and dominated by numerical noise.
-%\MP{The discussion of Fig.~\ref{fig:dhofs} appears twice in the text; here and one section earlier.}
 
@@
 \caption{Cross sections of the amplitude and phase residuals as functions of $\chi_1$ at three nodes $MF_j$.
-The waveforms are generated for 51 values of $\chi_1$, shown as dots, with the other parameters held constant at $\{q, S_2, \Lambda_1, \Lambda_2\} = \{0.6, 0.2, 2000, 1000\}$.
-Also shown are 1-dimensional cross sections of the interpolated residuals using Gaussian process regression (solid black) as well as their estimated $1\sigma$ uncertainty (shaded gray region).}
+Test-set waveforms are generated for 51 values of $\chi_1$, shown as dots, with the other parameters held constant at $\{q, \chi_2, \Lambda_1, \Lambda_2\} = \{0.6, 0.2, 2000, 1000\}$.
+Also shown are the interpolated values for the final surrogate, discussed in Sec.~\ref{sec:accuracy}, using Gaussian process regression (solid black) and the corresponding $1\sigma$ uncertainty estimates (shaded gray region).}
 \label{fig:dhofs}
@@
 \end{bmatrix}
-\right)
+\right),
 \end{equation}
@@
 
-We use a radial kernel which expresses the covariance in terms of a distance $r$ between points
+We use radial kernels, which express the covariance in terms of a distance $r$ between points
 \begin{equation}
@@
 \end{equation}
-The tunable hyperparameters ${\bf \ell}$ represent the length scale over which the function $g(\bx)$ varies in each coordinate. Specifically, we use the Mat\'{e}rn class of kernels
-\begin{equation}
-k_{\rm Matern}(r) = \frac{2^{1-\nu}}{\Gamma(\nu)} \left(\sqrt{2\nu} r\right)^\nu K_\nu \left(\sqrt{2\nu} r \right).
+The tunable hyperparameters ${\bf \ell}$ represent the length scale over which the function $g(\bx)$ varies in each coordinate.
+
+We try two classes of radial kernels. The first is the Mat\'{e}rn kernel
+\begin{equation}
+k_{\rm radial}(r) = \frac{2^{1-\nu}}{\Gamma(\nu)} \left(\sqrt{2\nu} r\right)^\nu K_\nu \left(\sqrt{2\nu} r \right),
 \end{equation}
@@
 The value of $\nu$ parameterizes the smoothness of the Gaussian process, and is $k$ times mean-square differentiable if $\nu>k$~\cite{RasmussenWilliams2006}.
-For half-integer values of $\nu$, this kernel has a computationally cheap form without special functions, and we have had good results with $\nu=5/2$, resulting in a twice-differentiable function\footnote{We have also tried the more common, infinitely differentiable squared exponential kernel~\cite{RasmussenWilliams2006}.
-However, we found it difficult to reliably tune its hyperparameters for the problem here.}.
+For half-integer values of $\nu$, this kernel has a computationally cheap form without special functions, and we have had good results with $\nu=5/2$, resulting in a twice-differentiable function.
 The $\nu=5/2$ kernel is
 \begin{equation}
-k_{\rm Matern}^{\nu=5/2}(r) = \left(1+\sqrt{5}r + \frac{5r^2}{3}\right) \exp\left(-\sqrt{5}r\right).
+k_{\rm radial}(r) = \left(1+\sqrt{5}r + \frac{5r^2}{3}\right) \exp\left(-\sqrt{5}r\right).
+\end{equation}
+The second class we try is the more common squared exponential kernel that results in an infinitely differentiable function~\cite{RasmussenWilliams2006}
+\begin{equation}
+k_{\rm radial}(r) = e^{-r^2/2}.
 \end{equation}
@@
 \begin{equation}
-k(\bx_i, \bx_j) = \sigma_f^2 k_{\rm Matern}^{\nu=5/2}(r) + \sigma_n^2 \delta_{ij}
+k(\bx_i, \bx_j) = \sigma_f^2 k_{\rm radial}(r) + \sigma_n^2 \delta_{ij},
 \end{equation}
@@
 \end{align}
-We set these functions to 0 below the first frequency node $MF_0 =  Mf_{trunc,i}$ so that the waveform transitions to \texttt{TaylorF2} at lower frequencies.
+We set these functions to 0 below the first frequency node $MF_0 =  Mf_{{\rm trunc},i}$ so that the waveform transitions to \texttt{TaylorF2} at lower frequencies.
 With the analytic expressions for $A_{\rm F2}(Mf;\bx)$ and $\Phi_{\rm F2}(Mf;\bx)$ and the interpolated expressions $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$, the final surrogates for the amplitude and phase are
@@
 \label{eq:amp_sur}
-A_S(Mf;\bx) &= A_{\rm F2}(Mf;\bx) \exp\left[ \Delta\ln A_S(Mf;\bx) \right] \\
+A_S(Mf;\bx) &= A_{\rm F2}(Mf;\bx) \exp\left[ \Delta\ln A_S(Mf;\bx) \right], \\
 \label{eq:phase_sur}
@@
 
-\red{[[Michael: Add your discussion of resampling the waveform here.]]}
+%\red{[[Michael: Add your discussion of resampling the waveform here.]]}
+
 %\MP{Say that we use sparse frequency grids based on power-law $\Delta f$ in the evaluation of the surrogate; perhaps describe in more detail how the model is evaluated in the LAL code.}
@@
 \label{sec:design}
-\red{[[Can't finish this section until the final sampling method is settled on.]]}
 
@@
 However, unlike a uniform grid, the $N$ values in each dimension are sampled exactly once instead of $N^{d-1}$ times.
-For an LHD there are $(N!)^d$ ways to choose these points, and we choose one randomly\footnote{There are ways to better choose an LHD.
+For an LHD there are $(N!)^d$ ways to choose these points, and we choose one randomly\footnote{There are additional ways to choose an LHD.
 A standard requirement is that the LHD be space-filling, meaning that the points are as far apart as possible from each other.
@@
 We have not experimented with these alternatives here.}.
-
 An LHD has the property that the samples are non-collapsing; a projection onto a subspace is still an LHD and no points are repeated in any dimension.
 This avoids wasting samples when one of the parameters has much less of an influence than the other parameters.
-As an example, the tidal parameters have a minimal impact on the waveform at low frequencies, so it would not make sense to densely sample a grid of tidal parameters at low frequencies.
+%As an example, the tidal parameters have a minimal impact on the waveform at low frequencies, so it would not make sense to densely sample a grid of tidal parameters at low frequencies.
 However, during the merger the tidal parameters can be more important than the other parameters.
@@
 An LHD with its noncollapsing property bypasses this problem.
-\MP{Note (perhaps in a footnotes) that it is possible to have separate surrogates for low and high frequency which are then joined smoothly as done for \texttt{SEOBNRv4\_ROM}.}
-
-We build an initial training set with 128 waveforms sampled with an LHD in our 5-dimensional parameter space given by $q\in[1/3, 1]$, $\chi_1, \chi_2 \in [-0.5, 0.5]$ and $\Lambda_1, \Lambda_2 \in [0, 5000]$.
-In addition, we find empirically that the GPR uncertainty estimates (Eq.~\eqref{eq:var} and Eq.~\eqref{eq:errorrms} below) are largest at the corners of the parameter space, so we also sample the 32 corners.
-We construct our initial surrogate with these 160 waveforms shown in Fig.~\ref{fig:LHD}.
+%\MP{Note (perhaps in a footnotes) that it is possible to have separate surrogates for low and high frequency which are then joined smoothly as done for \texttt{SEOBNRv4\_ROM}.}
+
+We build an initial training set with 128 waveforms sampled with an LHD for the five parameters $\bx = \{q, \chi_1, \chi_2, \xi_1, \xi_2\}$. 
+In addition, we find empirically that the GPR uncertainty estimates (Eq.~\eqref{eq:var})
+are largest at the corners of the parameter space, so we also sample the 32 corners. We construct our initial surrogate with these 160 waveforms shown in Fig.~\ref{fig:LHD}.
 
@@
 \includegraphics[width=0.49\textwidth]{trainingset2d.pdf}
-\caption{Projection of the sampled waveforms onto the three-dimensional subspace $\{q, \chi_1, \chi_2\}$ (top) and the two-dimensional subspace $\{\Lambda_1, \Lambda_2\}$ (bottom).
-The 160 blue circles were used for the initial training set (32 corners + 128 LHD points).
-The 200 red squares were generated using uncertainty sampling from the GPR error estimate (Eq.~\eqref{eq:errorrms}).
-The final 1000 green triangles were drawn uniformly from the parameter space.
+\caption{Projection of the training-set parameters onto the three-dimensional subspace $\{q, \chi_1, \chi_2\}$ (top) and the two-dimensional subspace $\{\xi_1, \xi_2\}$ (bottom).
+The 160 blue triangles were used for the initial training set (32 corners and 128 LHD points).
+The 400 red squares were generated using uncertainty sampling from the GPR error estimate (Eq.~\eqref{eq:errorrms}) with the Mat\'{e}rn kernel.
+The 400 green circles were sampled uniformly from the parameter space.
 }
@@
 
-With this initial surrogate model, we can choose new training-set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion and update the surrogate with each new waveform.
-This method is sometimes referred to as uncertainty sampling~\cite{BrochuCoraDeFreitas2010}.
-The quantity that we use is the root-mean-squared error in the phase at the frequency nodes $Mf_j$ below $Mf=0.03$: \MP{Add reason why higher frequencies are not included in this error measure.}
+With this initial surrogate, we compare two methods for adding samples and improving the accuracy of the surrogate.
+The first is to simply sample the parameter space $\bx$ with a uniform distribution.
+We choose 400 samples shown in Fig.~\ref{fig:LHD}.
+The second method is sometimes referred to as uncertainty sampling~\cite{BrochuCoraDeFreitas2010}.
+In this method, we can choose new training-set samples by iteratively searching the parameter space for new points ${\bm x}$ that maximize some error criterion, then adding a new sample at that point.
+The quantity that we use is the root-mean-squared (RMS) phase error at the frequency nodes $MF_j$:
 \begin{equation}
@@
 \end{equation}
-where $\sigma_{\Delta\Phi_j}(\bx)$ is the GPR estimate (Eq.~\eqref{eq:var}) of the uncertainty in $\Delta\Phi(MF_j, \bx)$.
-There are many possible ways to construct a scalar that estimates overall waveform error.
-An alternative would be an approximation to the mismatch discussed in Sec.~\ref{sec:accuracy}.
+where $\sigma_{\Delta\Phi_j}(\bx)$ is the GPR estimate (Eq.~\eqref{eq:var}) of the uncertainty in $\Delta\Phi(MF_j, \bx)$ at node $MF_j$.
+We only include nodes below $Mf=0.03$ because there is very little signal at higher frequencies, and the RMS error would otherwise be dominated by higher frequencies.
+There are many alternative ways to construct a scalar that estimates overall waveform error.
+One alternative would be an approximation to the mismatch discussed in Sec.~\ref{sec:accuracy}.
 However, the quantity $\epsilon_{\rm RMS}(\bx)$ describes the most important quantity, phase, and is fast to compute.
-For an initial training set with $N$ samples, $\epsilon_{\rm RMS}(\bx)$ has $\sim N$ local maxima located in the voids between samples $\bx$.
+For a training set with $N$ samples, $\epsilon_{\rm RMS}(\bx)$ has $\sim N$ local maxima located in the voids between samples $\bx$.
 To find the global maximum, we use a basin hopping algorithm~\cite{WalesDoye1998, scipy:basinhopping} to avoid getting stuck in local maxima.
@@
 
-\begin{algorithm}[H]
-\caption{Uncertainty Sampling}
-\label{alg:uc}
-\begin{algorithmic}[1]
-\State {\bf Input:} Initial surrogate 
-\For{$i = 1 \to N_{\rm New}$}
-
-\State Construct the GPR error estimator (Eq.~\eqref{eq:errorrms}).
+%\begin{algorithm}[H]
+%\caption{Uncertainty Sampling}
+%\label{alg:uc}
+%\begin{algorithmic}[1]
+%\State {\bf Input:} Initial surrogate 
+%\For{$i = 1 \to N_{\rm New}$}
+%
+%\State Construct the GPR error estimator (Eq.~\eqref{eq:errorrms}).
 In practice this can be done by specifying the samples $\bx$, the hyperparameters $\btheta$, and dummy data for $\by$ since Eq.~\eqref{eq:errorrms} does not depend on $\by$.
-
-\State Find $({\bm x}_{\rm max}, \epsilon_{\rm rms})$ that maximizes Eq.~\eqref{eq:errorrms} over the parameter space.
-
-\State Add ${\bm x}_{\rm max}$ to the list of samples: ${\bm x} = [{\bm x}, {\bm x}_{\rm max}]$.
-
-\EndFor
-\State {\bf Output:} Updated surrogate
-\end{algorithmic}
-\end{algorithm}
+%
+%\State Find $({\bm x}_{\rm max}, \epsilon_{\rm rms})$ that maximizes Eq.~\eqref{eq:errorrms} over the parameter space.
+%
+%\State Add ${\bm x}_{\rm max}$ to the list of samples: ${\bm x} = [{\bm x}, {\bm x}_{\rm max}]$.
+%
+%\EndFor
+%\State {\bf Output:} Updated surrogate
+%\end{algorithmic}
+%\end{algorithm}
 
@@
 
-\item Construct the GPR error estimator $\epsilon_{\rm rms}(\bx)$ (Eq.~\eqref{eq:errorrms}).
-In practice this can be done by specifying the samples $\bx$, the hyperparameters $\btheta$, and dummy data for $\by$ since Eq.~\eqref{eq:errorrms} does not depend on $\by$.
-
-\item Find $\bx_{\rm max}$ that maximizes $\epsilon_{\rm rms}(\bx)$ over the parameter space.
+\item Construct the GPR error estimate $\epsilon_{\rm RMS}(\bx)$ (Eq.~\eqref{eq:errorrms}).
+In practice this can be done by specifying the samples $\bx$, the hyperparameters $\btheta$ from the initial surrogate, and dummy data for $\by$ since Eq.~\eqref{eq:errorrms} does not depend on $\by$.
+
+\item Find the point $\bx_{\rm max}$ that maximizes $\epsilon_{\rm RMS}(\bx)$ over the parameter space.
 
@@
 
-Fig.~\ref{fig:LHD} shows the parameters of the 200 new waveforms chosen by the uncertainty sampling method.
-We note that the edges and faces of the parameter space are more often chosen than the inner region.
-This is likely a consequence of the Mat\'{e}rn GPR kernel used.
-Ref.~\cite{DoctorFarrHolz2017} found similar behavior when using a Mat\'{e}rn kernel instead of a squared exponential kernel.
-Fig.~\ref{fig:uncsamp} shows the RMS error $\epsilon_{\rm rms}$ maximized over the parameters $\bx$ for each new sample added to the training set.
-Increasing the number of waveforms from 160 to 360 decreases the estimated RMS phase error by a factor of $\sim 4$.
+Fig.~\ref{fig:LHD} shows the parameters of the $N_{\rm new}=400$ new waveforms chosen by the uncertainty sampling method when using a Mat\'{e}rn kernel.
+We note that the edges of the parameter space are more often chosen than the inner region.
+We find this is true for both the Mat\'{e}rn and squared exponential kernels.
+Ref.~\cite{DoctorFarrHolz2017} found similar behavior for 1- and 2-dimensional problems.
+As discussed there, this results because the edges have fewer nearby samples than the interior, resulting in a larger error estimate.
+The uncertainty sampling algorithm compensates by adding samples near the edges where the error estimate is largest.
+
+Fig.~\ref{fig:uncsamp} shows the estimated RMS error $\epsilon_{\rm RMS}(\bx)$ maximized over the parameters $\bx$ for each new sample added to the training set.
+This is done for both the Mat\'{e}rn (black) and squared exponential (gray) kernels.
+For each kernel, we constructed an initial surrogate with optimized hyperparameters, then fixed the hyperparameters before choosing the 400 new samples using uncertainty sampling.
+The non-smoothness of the curves results because the basin-hopping algorithm does not always find the global maximum of $\epsilon_{\rm RMS}$ at each iteration.
+We find that increasing the number of waveforms from 160 to 560 decreases the global maximum of $\epsilon_{\rm RMS}$ by a factor of \red{$\sim 4$} for the Mat\'{e}rn kernel and a factor of \red{$\sim 10$} for the squared exponential kernel.
+In Section~\ref{sec:accuracy} we will examine how well this estimated error agrees with the true error.
 
@@
 \includegraphics[width=0.49\textwidth]{uncertaintysampling.pdf}
-\caption{Estimated 1--$\sigma$ RMS phase error $\epsilon_{\rm RMS}(\bx)$ (Eq.~\eqref{eq:errorrms}) maximized over parameter space $\bx$ as a function of the number of chosen parameters (blue curve).
-Also shown for comparison are the maximum RMS phase errors at the nodes where $MF_j \le 0.03$ for the surrogate compared to a test set of 1000 waveforms.
-Blue circle: surrogate constructed from the 160 waveforms in the initial training set.
-Red square: surrogate after adding the 200 waveforms chosen with uncertainty sampling.
-Green triangle: final surrogate after also adding 1000 more training set waveforms drawn uniformly from parameter space.
-}
+\caption{Estimated $1\sigma$ RMS phase error $\epsilon_{\rm RMS}(\bx)$ (Eq.~\eqref{eq:errorrms}) at the frequency nodes $MF_j$ maximized over parameter space $\bx$ as a function of the number of samples.
+The black curve represents the estimated error using the Mat\'{e}rn kernel, and the gray curve represents the estimated error using the squared exponential kernel.
+Also shown are box plots of the maximum RMS phase errors at the nodes $MF_j \le 0.03$ for the surrogate compared to the test set of 1000 waveforms (Sec.~\ref{sec:accuracy}).
+The box plot represents the 25\%, 50\%, and 75\% quartiles.
+The whiskers contain all samples within 1.5 times the interquartile range.
+The black dots are outliers.
+Blue box plots: phase error from surrogates constructed using the initial design (160 corner and LHD waveforms) using either the Mat\'{e}rn or squared exponential kernel.
+Green box plots: surrogates constructed with the 160 initial waveforms and 400 uniformly sampled waveforms.
+Red box plots: surrogates constructed with the 160 initial waveforms and 400 waveforms chosen with uncertainty sampling.
+Magenta box plots: surrogates constructed with the 960 waveforms combining all sampling methods.}
 \label{fig:uncsamp}
@@
 
+Finally, after generating waveforms for the initial surrogate, the uncertainty sampling, and the uniform distribution, we combine all 960 waveforms.
+We then build the final surrogate with re-optimized hyperparameters.
+
 %% ________________________________________________________
@@
 
-The accuracy of the surrogate can be assessed by comparing it to a test set of waveforms with randomly sampled parameters.
-We evaluate an additional 1000 waveforms randomly sampled in parameter space using the same distribution discussed above to generate the last 1000 waveforms for the training set.
-
-A common measure of the surrogate model accuracy is the mismatch between the surrogate model and the original EOB waveform.
-The mismatch represents the loss in signal-to-noise ratio that would result from using the surrogate model instead of the original EOB waveform.
-It is defined as the deviation from a perfect overlap after aligning the two waveforms using the time and phase free parameters $t_0$ and $\phi_0$:
-\begin{equation}
-\mathcal{M} = 1 - \max_{t_0, \phi_0} \frac{(h_{\rm EOB}, h_{\rm Sur})} {\sqrt{(h_{\rm EOB}, h_{\rm EOB}) (h_{\rm Sur}, h_{\rm Sur})}}.
+The accuracy of the surrogate can be assessed by comparing it to a test set of waveforms.
+We generate 1000 waveforms with the original parameters sampled uniformly in the ranges $q \in [1/3, 1]$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$.
+In Fig.~\ref{fig:uncsamp}, we show box plots of the RMS phase error of the test set waveforms for each of 8 surrogate models generated.
+The first two surrogates are constructed using the initial design with either the Mat\'{e}rn or squared exponential kernel.
+The second two surrogates are constructed with the initial design and the 400 uniformly distributed samples and either the Mat\'{e}rn or squared exponential kernel.
+The next two surrogates use the initial design and the 400 samples chosen by the uncertainty sampling algorithm for each kernel.
+The final two surrogates use all 960 waveforms for each kernel.
+We find that for the Mat\'{e}rn kernel, the global maximum of $\epsilon_{\rm RMS}(\bx)$ is a reasonable estimate of the maximum test-set RMS phase errors.
+However, for the squared exponential kernel, the global maximum of $\epsilon_{\rm RMS}(\bx)$ significantly underestimates the true maximum test-set error.
+Additionally, we find that the surrogates constructed with the uniform distribution have smaller median test-set errors.
+However, the surrogates constructed with uncertainty sampling have smaller maximum test-set errors.
+
+Another common measure of the surrogate model accuracy is the mismatch between the surrogate model and the test-set waveforms.
+The mismatch represents the loss in signal-to-noise ratio that results from using the surrogate model $h_{\rm Sur}$ instead of the original waveform $h$.
+It is defined as the deviation from a perfect overlap after aligning the two waveforms using the time and phase free parameters $t_c$ and $\phi_c$:
+\begin{equation}
+\mathcal{M} = 1 - \max_{t_c, \phi_c} \frac{(h, h_{\rm Sur})} {\sqrt{(h, h) (h_{\rm Sur}, h_{\rm Sur})}}.
 \end{equation}
@@
 
-In Fig.~\ref{fig:mismatch}, we show the distribution of mismatches $\mathcal{M}$ between our surrogate and the 1000 randomly sampled EOB waveforms.
+In Fig.~\ref{fig:mismatch}, we show box plots of the mismatches $\mathcal{M}$ between the surrogate and the 1000 test-set waveforms for each of our 8 surrogates.
 We use the design sensitivity aLIGO PSD~\cite{Aasi:2013wya} and a total mass of $M=2.8M_\odot$.
 Our integration bounds are $Mf_{\rm low} = 0.00021$ and $Mf_{\rm high} = 0.07$, corresponding to physical units of \red{15.2}~Hz and \red{5076}~Hz respectively.
-As we increase the number of samples in our training set, the mismatches decrease and, for our final surrogate, we find a maximum mismatch of \red{$5.5\times 10^{-4}$}.
+As we increase the number of samples in our training set, the mismatches decrease.
+However, it is less clear which kernel and sampling strategy is optimal.
+For our final surrogate, we simply choose the surrogate that has the smallest maximum and median mismatches, and this is the surrogate with the Mat\'{e}rn kernel and all 960 waveforms.
+This surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$}.
 
@@
 \includegraphics[width=0.49\textwidth]{mismatch.pdf}
-\caption{Mismatches between the surrogate and 1000 test-set waveforms with randomly sampled parameters $\bx$.
-The total mass for each waveform is $M=2.8M_\odot$ and the PSD used was the aLIGO design sensitivity PSD.
-Histograms are shown for the initial surrogate with 160 waveforms (blue), the updated surrogate with 360 waveforms (red), and the final surrogate with 1360 waveforms (green).
-Also shown for comparison are the mismatches between TaylorF2 and the test set (black).}
+\caption{
+Mismatch between the surrogates and the 1000 test-set waveforms.
+The total mass for each waveform is $M=2.8M_\odot$, and the PSD is for the aLIGO design sensitivity configuration.
+Conventions for the box plots are the same as in Fig.~\ref{fig:uncsamp}.
+}
 \label{fig:mismatch}
@@
 \includegraphics[width=0.49\textwidth]{htildemaxerror.pdf}
-\caption{Amplitude and phase errors between the surrogate and the test-set waveform with the largest mismatch.
-Blue curve: initial surrogate with 160 waveforms.
-Red curve: updated surrogate with 360 waveforms.
-Green curve: final surrogate with 1360 waveforms.
-The vertical solid line is the gravitational-wave frequency at the Schwarzschild ISCO.}
+\caption{Fractional amplitude and phase errors between the 4 surrogates with a Mat\'{e}rn kernel and the test-set waveform with the largest mismatch for each surrogate.
+The vertical solid line is the gravitational-wave frequency at the Schwarzschild ISCO, and the top axis is the physical frequency for a binary with a total mass of $2.8M_\odot$.
+Blue curve: surrogate with initial waveforms (160 corner and LHD waveforms).
+Green curve: surrogate with initial waveforms and 400 uniformly sampled waveforms.
+Red curve: surrogate with initial waveforms and 400 waveforms chosen with uncertainty sampling.
+Magenta curve: final surrogate with the 960 waveforms combining all sampling methods.}
 \label{fig:maxmismatchfd}
@@
 
-In Fig.~\ref{fig:maxmismatchfd} we show the amplitude and phase errors as a function of frequency for the waveforms with the largest mismatch.
-As we add waveforms with the various sampling methods, the errors generally decrease.
-However, above a frequency of $Mf \sim 0.03$, the amplitude of the training set waveforms are noisy, and it is thus difficult to produce a good fit.
+We also show the surrogate amplitude and phase errors as a function of frequency for the parameters with the largest mismatch relative to the test-set in Fig.~\ref{fig:maxmismatchfd}.
+As we add waveforms to the training set with the various sampling methods, the errors generally decrease.
+However, above a frequency of $Mf \sim 0.03$, the amplitude of the training set waveforms are noisy (see Figs.~\ref{fig:hoff} and~\ref{fig:dhofs}), and it is thus difficult to produce a good fit.
 This is not a problem, however, because aLIGO is insensitive to the waveform morphology at such high frequencies.
+With our final surrogate, we are able to achieve a maximum phase error of $\lesssim 1$~radian over the entire frequency range relevant to aLIGO.
 
@@
 In Fig.~\ref{fig:maxmismatchtd} we compare the final surrogate to the test set waveform that has the largest mismatch.
-We align the two waveforms in time and phase by maximizing the overlap at early times in the interval $t/M \in [-2\times 10^5, -1\times 10^5]$ using the method of~\cite{Read2009}.
-The two waveforms are nearly identical.
+We align the two waveforms in time and phase by maximizing the overlap at early times in the interval $(t-t_{\rm merger})/M \in [-10^7, -10^6]$ using the method of~\cite{ReadMarkakisShibata2009}.
+The two waveforms remain nearly identical up to merger, far from the interval where they were aligned.
 
@@
 \includegraphics[width=0.99\textwidth]{hmaxerror.pdf}
-\caption{Inverse Fourier transformed surrogate (green dashed) for the parameters $\bx$ that had the largest mismatch with the test-set EOB waveforms (black).
-The two waveforms are aligned in the time domain in the interval $t/M \in [-2\times 10^5, -1\times 10^5]$.}
+\caption{Inverse Fourier transformed surrogate (red dashed) for the parameters $\{q, \chi_1, \chi_2, \Lambda_1, \Lambda_2\} = \{0.61, -0.23, -0.33, 180, 4390\}$ that have the largest mismatch with the test-set EOB waveforms (black).
+The two waveforms are aligned in the time domain, before the start of the plot, in the interval $(t-t_{\rm merger})/M \in [-10^7, -10^6]$.}
 \label{fig:maxmismatchtd}
@@
 We also compare the evaluation time to several BNS waveform models that have been used for LIGO data analysis.
-We note that at $\sim 30$~Hz, the \texttt{SEOBNRv4T\_surrogate} model is comparable to the \texttt{SEOBNRv4\_ROM\_NRTidal} and \texttt{IMRPhenomD\_NRTidal} models that were used in the analysis of the GW170817 event~\cite{PEPaper}.
+We note that at $\sim 30$~Hz, the \texttt{SEOBNRv4T\_surrogate} model is comparable to the \texttt{SEOBNRv4\_ROM\_NRTidal} and \texttt{IMRPhenomD\_NRTidal} models that were used in the analysis of the GW170817 event~\cite{BNSPE}.
 At lower frequencies, the surrogate is faster than all BNS waveform models except for \texttt{TaylorF2}.
@@
 
+As a final test of the surrogate, we perform parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template.
+For the two aLIGO detectors we use the design sensitivity PSD~\cite{}, and for the advanced Virgo detectors we also use its design sensitivity PSD.
+We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$, corresponding to the MPA1 equation of state.
+The sky location and inclination angle are fixed at an arbitrary value, and the distance is chosen such that the optimal network signal-to-noise ratio (SNR) is either 30 or 60.
+We inject the waveforms into zero noise data so that the recovered parameters do not depend on the specific noise realization~\cite{}.
+
+As discussed in more detail in~\cite{BNSPE}, the posterior $p(\theta | d)$ for the parameters $\theta$ is given by Bayes' theorem:
+\begin{equation}
+p(\theta | d) \propto p(\theta) \mathcal{L}(d | \theta),
+\end{equation}
+where $p(\theta)$ is the prior and $\mathcal{L}(d | \theta)$ is the likelihood.
+We choose for our priors the sky position and orientation uniform on the unit sphere.
+The distance is uniform in co-moving volume.
+The masses, spins, and tidal parameters are uniform in the ranges $m_{1, 2} \in [0.7, 2]M_\odot$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$.
+We sample this posterior with the parallel-tempered MCMC code in LALInference.
+We sample the data and waveform at \red{8192}~Hz such that the Nyquist frequency is 4096~Hz.
+\red{[[We note that published analyses for BNS systems typically use a sampling frequency of 4096~Hz.
+However, the \texttt{SEOBNRv4T} waveform model still has some signal above the corresponding Nyquist frequency of 2048~Hz (See Fig.~\ref{fig:hoff}).]]} We integrate the likelihood function in the interval $[20, 4096]$~Hz.
+Approximately \red{$10^x$} iterations are needed to produce \red{$\sim 5000$} statistically independent samples.
+
+For data with zero noise, the fit between the injected waveform and the template is best when the likelihood is maximum.
+However, as a result of the priors that are bounded and not flat for some parameters, the posterior, and especially the marginalized posteriors, will not be maximized at the injected parameters even with a perfect template.
+We therefore use the difference between the parameters that maximize the likelihood $\theta_{\rm max}$ and the injected parameters $\theta_{\rm inj}$ as our estimate for the systematic error due to the surrogate.
+
+Marginalized 1-dimensional posteriors for four key parameters are shown in Fig.~\ref{fig:pe}.
+These are the chirp mass $\mathcal{M} = (m_1 m_1)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$.
+The maximum value of the likelihood found for each run is dashed, and the injected value is solid.
+Because the sampled parameters do not necessarily contain the parameters for the maximum likelihood, the difference between these parameters $\Delta\theta_{\rm sys} = \theta_{\rm max} - \theta_{\rm inj}$ is an upper bound on the systematic error that results from using the surrogate as the template instead of the original \texttt{SEOBNRv4T} waveform.
+We can compare this systematic error to the statistical error from the finite sensitivity of the detectors.
+We will define this as half the distance between the 5\% and 95\% percentiles of each parameter: $\Delta\theta_{\rm stat} = (\theta_{95} - \theta_{5})/2$.
+Table~\ref{tab:syserrors} shows that the systematic error is small compared to the statistical errors even for the injection with SNR=60.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.39\textwidth]{chirpmass.pdf}\\
+\includegraphics[width=0.39\textwidth]{q.pdf}\\
+\includegraphics[width=0.39\textwidth]{chieff.pdf}\\
+\includegraphics[width=0.39\textwidth]{tildelambda.pdf}
+\caption{Marginalized PDFs for two waveform injections with zero noise.
+The parameters are identical except for the distances which are set such that the network SNRs are 30 (blue) or 60 (red).
+The shaded regions are bounded by the 5 and 95 percentiles.
+The vertical solid black line represents the injected parameter value.
+The vertical dashed lines represent the parameter value with maximum likelihood.}
+\label{fig:pe}
+\end{figure}
+
+\begin{table}[htb]
+\centering    
+\caption{Estimated systematic errors due to surrogate model and statistical errors from finite detector sensitivity.
+The ratio $\Delta\theta_{\rm sys} / \Delta\theta_{\rm stat}$ is given for the two injections with network SNRs of 30 and 60.}
+\begin{tabular}{ccc}        
+\hline
+$\theta$ & SNR=30 & SNR=60 \\
+\hline
+$\mathcal{M}$ &   &  \\
+$q$ &   &  \\
+$\chi_{\rm eff}$ &   &  \\
+$\tilde\Lambda$ &   &  \\
+\hline
+\end{tabular}
+\label{tab:syserrors}
+\end{table}
+
 \subsection{GW170817}
@@
 
-We have constructed a fast frequency-domain surrogate of the most accurate BNS waveform currently available for data analysis.
+We have constructed a fast, frequency-domain surrogate of one of the most accurate BNS waveform models to date.
 This aligned-spin model, \texttt{SEOBNRv4T}, incorporates the tidally induced $\ell=2$ and 3 multipole moments as well as the effect of dynamical tides as the excitation approaches the $\ell=2$ and 3 $f$-mode frequencies.
-We have achieved mismatches of no more than $7 \times 10^{-4}$ and phase errors of $\lesssim 1$~rad up to the merger frequency.
+We have achieved mismatches of no more than \red{$4.5 \times 10^{-4}$} and phase errors of \red{$\lesssim 1$~rad} up to the merger frequency.
 These are sufficient to not bias results in any of the parameters.
 
-The evaluation time has a flat cost of $\sim 0.02$~s to perform the GPR interpolation at each node.
+The evaluation time has a flat cost of \red{$\sim 0.02$~s} to perform the GPR interpolation at each node.
 The rest of the time is spent resampling the waveform with spline interpolation.
-For a starting frequency of 10~Hz, this takes 0.3~s when the waveform is matched with data uniformly sampled at 4096Hz.
-For a starting frequency of 30~Hz, the total evaluation time is 0.04~s.
+For a starting frequency of 10~Hz, this takes \red{0.3~s} when the waveform is matched with data uniformly sampled at 4096Hz.
+For a starting frequency of 30~Hz, the total evaluation time is \red{0.04~s}.
 For current parameter estimation codes, this is sufficient.
@@
 Although none of the EOB models currently available include both tidal effects and precession, there are ways to boost this waveform model to a precessing frame and approximately incorporate precession.
-For example, Chatziioanno et al.
-have analytically solved the 2PN-accurate precession equations for generic spins~\cite{ChatziioannouKleinCornish2017a, ChatziioannouKleinCornish2017b}, with the exception of transitional precession which is unlikely for low mass-ratio and spin BNS systems.
+For example, Chatziioanno {\it et al}.\ have analytically solved the 2PN-accurate precession equations for generic spins~\cite{ChatziioannouKleinCornish2017a, ChatziioannouKleinCornish2017b}, with the exception of transitional precession which is unlikely for low mass-ratio and spin BNS systems.
 Using stationary uniform asymptotics, they can also analytically Fourier-transform the solution.
@@
 This approach would provide a fast, accurate model for BNS systems with tides and generic spins.
-We note, however, that including precession using the \texttt{IMRPhenomP\_NRTidal} model did not appear to affect the results in~\cite{PEPaper}.
+We note, however, that including precession using the \texttt{IMRPhenomP\_NRTidal} model did not significantly affect the results for GW170817~\cite{BNSPE}.
 
@@
 
-BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the Institute for Nuclear Theory where many of the methods used here were discussed.
-We also thank Sylvain Marsat for correcting numerical instabilities in the original \texttt{SEOBNRv4T} code in \texttt{LAL}.
-The authors were supported by...
+BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the Institute for Nuclear Theory where many of the methods used here were discussed. The authors were supported by...
 

### 2018-09-26 src/trainingseteob.py

@@ -284,12 +284,12 @@ def condition_eob_waveform(
     """
     condition = ConditionedWaveform(h)
 
-    # Resample the waveform
-    condition.resample_uniform(delta_t)
-
     # Window the waveform
     condition.window_freq_on(winon_i, winon_f, win=win)
 
+    # Resample the waveform
+    condition.resample_uniform(delta_t)
+    
     if plots:
         # Windowed time-domain waveform
         fig, ax = wave.plot_waveforms([condition.h], hc=True, xi=h.x[-1]-1000, npoints=10000)
@@ -477,7 +477,7 @@ def condition_eob_training_set_from_list(
             win=win,
             filter_dfbyf_amp=filter_dfbyf_amp,
             filter_dfbyf_phase=filter_dfbyf_phase,
-            quad1=quad1, quad2=quad2, spin_spin=spin_spin, 
+            quad1=quad1, quad2=quad2, spin_spin=spin_spin,
             plots=plots)
 
         h_ts.set_waveform(i, h_cond, p)

### 2018-09-26 src/waveform.py

@@ -322,7 +322,7 @@ def fourier_transform_uniform_sampled_waveform(h):
 
 ##################### Functions for plotting waveforms ####################
 
-def plot_waveforms(waveforms, xi=-np.inf, xf=np.inf, npoints=10000, amp=True, hp=True, hc=False):
+def plot_waveforms(waveforms, xi=-np.inf, xf=np.inf, npoints=10000, amp=True, hp=True, hc=False, order=2):
     """
     """
     fig, axes = plt.subplots(1, figsize=(16, 3))
@@ -330,7 +330,7 @@ def plot_waveforms(waveforms, xi=-np.inf, xf=np.inf, npoints=10000, amp=True, hp
     for h in waveforms:
         xiplot = max(xi, h.x[0])
         xfplot = min(xf, h.x[-1])
-        hcomp = h.interpolate_complex()
+        hcomp = h.interpolate_complex(order=order)
         times = np.linspace(xiplot, xfplot, npoints)
         hs = hcomp(times)
         if amp:

### 2018-09-26 src/waveformset.py

@@ -138,23 +138,33 @@ def waveform_set_from_list(filename, h_list, params_list):
     h_set.close()
 
 
-def join_waveform_sets(filename1, filename2, filename_join):
+def join_waveform_sets(filename1, filename2, filename_join, list1=None, list2=None):
     """Join two HDF5WaveformSets together into a new HDF5WaveformSet.
+
+    parameters
+    ----------
+    list1 : list of waveform indices to use in WaveformSet 1.
+    list2 : list of waveform indices to use in WaveformSet 2.
     """
     h_set1 = HDF5WaveformSet(filename1)
     h_set2 = HDF5WaveformSet(filename2)
     h_join = HDF5WaveformSet(filename_join)
 
+    if list1 is None:
+        list1 = list(range(len(h_set1)))
+    if list2 is None:
+        list2 = list(range(len(h_set2)))
+
     j = 0
     # Write the waveforms in h_set1 to h_join
-    for i in range(len(h_set1)):
+    for i in list1:
         h = h_set1.get_waveform(i, data='waveform')
         p = h_set1.get_waveform(i, data='parameters')
         h_join.set_waveform(j, h, p)
         j += 1
 
     # Append the waveforms in h_set2 to h_join
-    for i in range(len(h_set2)):
+    for i in list2:
         h = h_set2.get_waveform(i, data='waveform')
         p = h_set2.get_waveform(i, data='parameters')
         h_join.set_waveform(j, h, p)

### 2018-09-26 src/window.py

@@ -131,18 +131,27 @@ def window_freq_on(h, freqi, freqf, win='planck', order=2, eps=1.0e-4):
     toff = interpolate_time_of_frequency(h, order=order)
     fmin = toff.get_knots()[0]
     fmax = toff.get_knots()[-1]
-    if freqi<(1.0-eps)*fmin:
-        raise Exception, 'freqi='+str(freqi)+' is lower than lowest frequency '+str(fmin)+'.'
-    if freqf>(1.0+eps)*fmax:
-        raise Exception, 'freqi='+str(freqf)+' is higher than highest frequency '+str(fmax)+'.'
 
-    xi = float(toff(freqi))
-    xf = float(toff(freqf))
+    # Get the starting time for windowing. Make surue you don't go below the starting time of the waveform.
+    if freqi >= fmin:
+        xi = float(toff(freqi))
+    else:
+        xi = h.x[0]
+        print 'Warning: freqi={:} < fmin={:}. Using starting time instead.'.format(freqi, fmin)
+
+    # Get the ending time for windowing. Make surue you don't go above the ending time of the waveform.
+    if freqf <= fmax:
+        xf = float(toff(freqf))
+    else:
+        xf = h.x[-1]
+        print 'Warning: freqf={:} > fmax={:}. Using ending time instead.'.format(freqf, fmax)
+
+    # Do the windowing with the chosen start and end times
     return window_on(h, xi, xf, win=win)
 
 
-def window_freq_off(h, freqi, freqf, win='planck', order=2, eps=1.0e-4):
-    """Smoothly turn off the waveform amplitude from A(x) to 0 over the frequency window [freqi, freqf].
+def window_freq_off(h, freqi, freqf, win='planck', order=2):
+    """Smoothly turn on the waveform amplitude from 0 to A(x) over the frequency window [freqi, freqf].
     eps :
         Fractional tolerance when determining if [freqi, freqf] are within
         the range of frequencies of h.
@@ -150,11 +159,59 @@ def window_freq_off(h, freqi, freqf, win='planck', order=2, eps=1.0e-4):
     toff = interpolate_time_of_frequency(h, order=order)
     fmin = toff.get_knots()[0]
     fmax = toff.get_knots()[-1]
-    if freqi<(1.0-eps)*fmin:
-        raise Exception, 'freqi='+str(freqi)+' is lower than lowest frequency '+str(fmin)+'.'
-    if freqf>(1.0+eps)*fmax:
[…]

### 2018-10-02 paper/paper.tex
Updates to PE section.

@@
 
-As a final test of the surrogate, we perform parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template.
-For the two aLIGO detectors we use the design sensitivity PSD~\cite{}, and for the advanced Virgo detectors we also use its design sensitivity PSD.
-We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$, corresponding to the MPA1 equation of state.
+As a final test of the surrogate accuracy, we perform parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template.
+We use the design sensitivity PSDs for the two advanced LIGO detectors~\cite{Aasi:2013wya} and the advanced Virgo detector~\cite{virgoPSD}.
+We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$ corresponding to the equation of state named MS1b in~\cite{ReadLackey2009}.
 The sky location and inclination angle are fixed at an arbitrary value, and the distance is chosen such that the optimal network signal-to-noise ratio (SNR) is either 30 or 60.
-We inject the waveforms into zero noise data so that the recovered parameters do not depend on the specific noise realization~\cite{}.
-
-As discussed in more detail in~\cite{BNSPE}, the posterior $p(\theta | d)$ for the parameters $\theta$ is given by Bayes' theorem:
-\begin{equation}
-p(\theta | d) \propto p(\theta) \mathcal{L}(d | \theta),
-\end{equation}
-where $p(\theta)$ is the prior and $\mathcal{L}(d | \theta)$ is the likelihood.
+We inject the waveforms into zero noise data so that the recovered parameters do not depend on the specific noise realization~\cite{NissankeHolzHughes2010}.
+
+As discussed in more detail in~\cite{BNSPE}, the posterior $p(\vec\vartheta | d)$ for the parameters $\vec\vartheta$ given the detector data $\vec d$ is given by Bayes' theorem:
+\begin{equation}
+p(\vec\vartheta | \vec d) \propto p(\vec\vartheta) \mathcal{L}(\vec d | \vec\vartheta),
+\end{equation}
+where $p(\vec\vartheta)$ is the prior and $\mathcal{L}(\vec d | \vec\vartheta)$ is the likelihood.
 We choose for our priors the sky position and orientation uniform on the unit sphere.
@@
 The masses, spins, and tidal parameters are uniform in the ranges $m_{1, 2} \in [0.7, 2]M_\odot$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$.
-We sample this posterior with the parallel-tempered MCMC code in LALInference.
-We sample the data and waveform at \red{8192}~Hz such that the Nyquist frequency is 4096~Hz.
-\red{[[We note that published analyses for BNS systems typically use a sampling frequency of 4096~Hz.
-However, the \texttt{SEOBNRv4T} waveform model still has some signal above the corresponding Nyquist frequency of 2048~Hz (See Fig.~\ref{fig:hoff}).]]} We integrate the likelihood function in the interval $[20, 4096]$~Hz.
+The mass ratio is restricted to the range $q\in[0.5, 1]$.
+We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal}.
+As was done in~\cite{BNSPE}, we sample the data and waveform at \red{4096}~Hz, such that the Nyquist frequency is \red{2048}~Hz, and integrate the likelihood function in the interval \red{$[20, 2048]$~Hz}.
 Approximately \red{$10^x$} iterations are needed to produce \red{$\sim 5000$} statistically independent samples.
-
-For data with zero noise, the fit between the injected waveform and the template is best when the likelihood is maximum.
-However, as a result of the priors that are bounded and not flat for some parameters, the posterior, and especially the marginalized posteriors, will not be maximized at the injected parameters even with a perfect template.
-We therefore use the difference between the parameters that maximize the likelihood $\theta_{\rm max}$ and the injected parameters $\theta_{\rm inj}$ as our estimate for the systematic error due to the surrogate.
+We note that the \texttt{SEOBNRv4T} waveform model still has some signal above \red{2048~Hz} (see Fig.~\ref{fig:hoff}).
+To test if this had any effect on the recovered posteriors, we also performed runs with a sampling frequency of \red{8192}~Hz and integrated the likelihood function in the interval \red{$[20, 4096]$}~Hz, but did not see a significant difference in marginalized posteriors.
 
@@
 These are the chirp mass $\mathcal{M} = (m_1 m_1)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$.
+We note that the peaks of the marginalized posteriors do not line up with the injected values.
+One reason is that the priors are not flat in these parameters.
+More importantly, because the posterior is bounded by the prior for many parameters, strong correlations cause a shift in other parameters.
+We therefore need another measure to examine the systematic error due to the surrogate.
+
+For data with zero noise, the fit between the injected waveform and the template is best when the likelihood is maximum.
+We therefore use the difference $\Delta\vec\vartheta_{\rm sys} = \vec\vartheta_{\rm max} - \vec\vartheta_{\rm inj}$ between the parameters that maximize the likelihood $\vec\vartheta_{\rm max}$ and the injected parameters $\vec\vartheta_{\rm inj}$ as our estimate for the systematic error due to the surrogate.
+In Fig.~\ref{fig:pe}, the maximum sampled likelihood is dashed, and the injected value is solid.
+This difference is a conservative upper limit because the maximum sampled likelihood does not necessarily correspond to the actual maximum.
+%This is especially true for strongly correlated parameters such as the pair $q$ and $\chi_{\rm eff}$. 
+Finally, we can compare this systematic error to the statistical error which we define as the 90\% credible interval shown by the shaded region.
+We find that the systematic error is significantly less that the statistical error, even for the SNR=60 signal.
+
+%For data with zero noise, the fit between the injected waveform and the template is best when the likelihood is maximum.
+However, as a result of the priors that are bounded and not flat for some parameters, the posterior, and especially the marginalized posteriors, will not be maximized at the injected parameters even with a perfect template.
+We therefore use the difference between the parameters that maximize the likelihood $\vec\vartheta_{\rm max}$ and the injected parameters $\vec\vartheta_{\rm inj}$ as our estimate for the systematic error due to the surrogate.
+%
+%Marginalized 1-dimensional posteriors for four key parameters are shown in Fig.~\ref{fig:pe}.
+These are the chirp mass $\mathcal{M} = (m_1 m_1)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$.
 The maximum value of the likelihood found for each run is dashed, and the injected value is solid.
-Because the sampled parameters do not necessarily contain the parameters for the maximum likelihood, the difference between these parameters $\Delta\theta_{\rm sys} = \theta_{\rm max} - \theta_{\rm inj}$ is an upper bound on the systematic error that results from using the surrogate as the template instead of the original \texttt{SEOBNRv4T} waveform.
-We can compare this systematic error to the statistical error from the finite sensitivity of the detectors.
-We will define this as half the distance between the 5\% and 95\% percentiles of each parameter: $\Delta\theta_{\rm stat} = (\theta_{95} - \theta_{5})/2$.
+Because the sampled parameters do not necessarily contain the parameters for the maximum likelihood, the difference between these parameters $\Delta\vec\vartheta_{\rm sys} = \vec\vartheta_{\rm max} - \vec\vartheta_{\rm inj}$ is an upper bound on the systematic error that results from using the surrogate as the template instead of the original \texttt{SEOBNRv4T} waveform.
+We can compare this systematic error to the statistical error $\Delta\vec\vartheta_{\rm stat}$ from the finite sensitivity of the detectors.
+Here, we will $\Delta\vec\vartheta_{\rm stat}$ as half the width of the 90\% credible interval.
 Table~\ref{tab:syserrors} shows that the systematic error is small compared to the statistical errors even for the injection with SNR=60.
@@
 The parameters are identical except for the distances which are set such that the network SNRs are 30 (blue) or 60 (red).
-The shaded regions are bounded by the 5 and 95 percentiles.
+The shaded regions give the 90\% credible interval.
 The vertical solid black line represents the injected parameter value.
@@
 
-\begin{table}[htb]
-\centering    
-\caption{Estimated systematic errors due to surrogate model and statistical errors from finite detector sensitivity.
-The ratio $\Delta\theta_{\rm sys} / \Delta\theta_{\rm stat}$ is given for the two injections with network SNRs of 30 and 60.}
-\begin{tabular}{ccc}        
-\hline
-$\theta$ & SNR=30 & SNR=60 \\
-\hline
-$\mathcal{M}$ &   &  \\
-$q$ &   &  \\
-$\chi_{\rm eff}$ &   &  \\
-$\tilde\Lambda$ &   &  \\
-\hline
-\end{tabular}
-\label{tab:syserrors}
-\end{table}
+%\begin{table}[htb]
+%\centering    
+%\caption{Estimated systematic errors due to surrogate model and statistical errors from finite detector sensitivity.
+The ratio $\Delta\vec\vartheta_{\rm sys} /\Delta\vec\vartheta_{\rm stat}$ is given for the two injections with network SNRs of 30 and 60.}
+%\begin{tabular}{ccc}        
+%\hline
+%$\vartheta$ & SNR=30 & SNR=60 \\
+%\hline
+%$\mathcal{M}$ &  $<$ & $<$ \\
+%$q$ & $<$  & $<$ \\
+%$\chi_{\rm eff}$ & $<$  & $<$ \\
+%$\tilde\Lambda$ &  $<$ & $<$ \\
+%\hline
+%\end{tabular}
+%\label{tab:syserrors}
+%\end{table}
 
@@
 
-BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the Institute for Nuclear Theory where many of the methods used here were discussed. The authors were supported by...
+BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the University of Washington for useful discussions at the start of this work.
+The authors were supported by...
+\red{[[Get AEI funding info.]]}
 

### 2018-10-16 src/TEOB-LAL-implementation/timings/timings.py
New timing data for final surrogate dataset

@@ -167,10 +167,14 @@ def inner(_it, _timer%(init)s):
 
 
 f_mins = np.arange(10.0, 100.0, 5)
-Mtot = 2*1.35
-x = np.array([1.0/1.5, 0.35, -0.12, 2000.0, 2750.0])
-xnT = np.array([1.0/1.5, 0.35, -0.12, 0.0, 0.0])
-xNSBH = np.array([1.0/1.5, 0.35, -0.12, 0.0, 2500.0])
+Mtot = 2*1.4
+q = 1.0/1.01
+chi1 = 0.1
+chi2 = 0.1
+lambda1 = 1286.0
+lambda2 = 1286.0
+x = np.array([q, chi1, chi2, lambda1, lambda2])
+xnT = np.array([q, chi1, chi2, 0, 0])
 
 tim_sur = np.array([timeit_ipython("FD_waveform_test(Mtot, x, LS.SEOBNRv4T_surrogate, fLow=%f, fHigh=2048.0, deltaF=0)" %(f_min)) 
  for f_min in f_mins])
@@ -184,15 +188,15 @@ tim_SEOBNRv4_ROM = np.array([timeit_ipython("FD_waveform_test(Mtot, xnT, LS.SEOB
  for f_min in f_mins])
 np.save('tim_SEOBNRv4_ROM', tim_SEOBNRv4_ROM)
 
-tim_SEOBNRv4_ROM_NRTidal = np.array([timeit_ipython("FD_waveform_test(Mtot, xNSBH, LS.SEOBNRv4_ROM_NRTidal, fLow=%f, fHigh=2048.0, deltaF=0)" %(f_min)) 
+tim_SEOBNRv4_ROM_NRTidal = np.array([timeit_ipython("FD_waveform_test(Mtot, x, LS.SEOBNRv4_ROM_NRTidal, fLow=%f, fHigh=2048.0, deltaF=0)" %(f_min)) 
  for f_min in f_mins])
 np.save('tim_SEOBNRv4_ROM_NRTidal', tim_SEOBNRv4_ROM_NRTidal)
 
-tim_IMRPhenomD_NRTidal = np.array([timeit_ipython("FD_waveform_test(Mtot, xNSBH, LS.IMRPhenomD_NRTidal, fLow=%f, fHigh=2048.0, deltaF=0)" %(f_min)) 
[…]

### 2018-10-16 src/TEOB-LAL-implementation/timings/timings_surrogate.py
Additional timing data for surrogate using same df as SEOBNRv4T

@@ -1,0 +1,191 @@
+#!/usr/bin/env python
+
+import numpy as np
+import lal
+import lalsimulation as LS
+import sys
+
+
+def FD_waveform_test(Mtot, x, approximant, fLow=20.0, fHigh=16384.0, deltaF=0):
+    q = 1.0/x[0]
+    chi1 = x[1]
+    chi2 = x[2]
+    lambda1 = x[3]
+    lambda2 = x[4]
+    
+    phiRef, fRef = 0.0, fLow
+    distance, inclination = 1.0, 0.0
+    
+    m1 = q/(1.0+q)*Mtot
+    m2 = 1.0/(1.0+q)*Mtot
+    m1SI, m2SI, chi1, chi2, lambda1, lambda2, nk_max = m1*lal.MSUN_SI, m2*lal.MSUN_SI, chi1, chi2, lambda1, lambda2, -1
+
+    longAscNodes, eccentricity, meanPerAno = 0,0,0
+
+    LALpars = lal.CreateDict()
+    LS.SimInspiralWaveformParamsInsertTidalLambda1(LALpars, lambda1)
+    LS.SimInspiralWaveformParamsInsertTidalLambda2(LALpars, lambda2)
+
+    # Nyquist frequency is set by fHigh
+    # Can set deltaF = 0 to figure out required frequency spacing; the chosen deltaF depends on fLow
+    
+    # Documentation from LALSimInspiral.c
+    #
+    #  * This routine can generate TD approximants and transform them into the frequency domain.
+    #  * Waveforms are generated beginning at a slightly lower starting frequency and tapers
+    #  * are put in this early region so that the waveform smoothly turns on.
+    #  *
+    #  * If an FD approximant is used, this routine applies tapers in the frequency domain
+    #  * between the slightly-lower frequency and the requested f_min.  Also, the phase of the
+    #  * waveform is adjusted to introduce a time shift.  This time shift should allow the
+    #  * resulting waveform to be Fourier transformed into the time domain without wrapping
+    #  * the end of the waveform to the beginning.
+    #  *
+    #  * This routine assumes that f_max is the Nyquist frequency of a corresponding time-domain
+    #  * waveform, so that deltaT = 0.5 / f_max.  If deltaF is set to 0 then this routine computes
+    #  * a deltaF that is small enough to represent the Fourier transform of a time-domain waveform.
+    #  * If deltaF is specified but f_max / deltaF is not a power of 2, and the waveform approximant
[…]

### 2018-10-16 src/TEOB-LAL-implementation/timings/timings_surrogate.sh

@@ -1,0 +1,9 @@
+#!/bin/bash
+
+source /home/mpuer/lsc/SEOBNRv4T_surrogate_master/etc/lalsuite-user-env.sh
+export LAL_DATA_PATH=/home/mpuer/SEOBNRROMData:$LAL_DATA_PATH
+
+export PYTHONPATH=$PYTHONPATH:/home/mpuer/projects/
+
+LOG=timings_surrogate.log
+python timings_surrogate.py > ${LOG} 2>&1 &

### 2018-10-17 paper/paper.tex
Adding PE discussion to paper. Fixing coalescence time problem with LAL version.

@@
 \caption{Inverse Fourier transformed surrogate (red dashed) for the parameters $\{q, \chi_1, \chi_2, \Lambda_1, \Lambda_2\} = \{0.61, -0.23, -0.33, 180, 4390\}$ that have the largest mismatch with the test-set EOB waveforms (black).
-The two waveforms are aligned in the time domain, before the start of the plot, in the interval $(t-t_{\rm merger})/M \in [-10^7, -10^6]$.}
+The two waveforms are aligned in the time domain, before the start of the plot, in the interval $(t-t_{\rm Merger})/M \in [-10^7, -10^6]$.}
 \label{fig:maxmismatchtd}
@@
 
+Having confirmed that the surrogate accurately recovers the injected waveform parameters, we now perform a parameter estimation run for GW170817.
+In the PE paper and catalog papers~\cite{}, the SEOBNRv4T surrogate was used with the parallelized PE code RapidPE.
+However, there were uncertainties about its ability to accurately recover the tidal parameter.
+
+We use the priors discussed in~cite{O2CatalogPaper}.
+
 %% ________________________________________________________

### 2018-10-22 paper/paper.tex
Preliminary discussion of quadrupole monopole

@@
 \newcommand{\esube}{\end{subequations}}
+%
+\newcommand{\nin}{\noindent}
+\newcommand{\nnm}{\nonumber}
+\newcommand{\doe}{\partial}
+%\newcommand{\be}{\begin{equation}}
+%\newcommand{\ee}{\end{equation}}
+\newcommand{\bea}{\begin{eqnarray}}
+\newcommand{\eea}{\end{eqnarray}}
+\newcommand{\bdm}{\begin{displaymath}}
+\newcommand{\edm}{\end{displaymath}}
+\newcommand{\bse}{\begin{subequations}}
+\newcommand{\ese}{\end{subequations}}
+\newcommand{\tb}{\textbf}
+\newcommand{\mr}{\mathrm}
+\newcommand{\tr}{\textrm}
+\newcommand{\mc}{\mathcal}
+\newcommand{\mb}{\mathbf}
+\newcommand{\py}{\phantom{yo}}
+\newcommand{\ve}{\varepsilon}
+\newcommand{\ind}{\indices}
+\newcommand{\bs}{\boldsymbol}
+\newcommand{\st}{{}^{*\!}}
 
@@
 \newcommand{\MP}[1]{{\textcolor{blue}{\texttt{MP: #1}} }}
-\newcommand{\bs}[1]{{\textcolor{green}{\texttt{SB: #1}} }}
+%\newcommand{\bs}[1]{{\textcolor{green}{\texttt{SB: #1}} }}
 \newcommand{\bl}[1]{{\textcolor{blue}{\texttt{BL: #1}} }}
@@
 In order to avoid triggering these stopping conditions at low frequency, where residual orbital eccentricity due to imperfect quasicircular initial conditions could be present, we exploit the $\omega_{\textrm{peak}}^{\textrm{NR}}$ fit to define a radius below which the stopping conditions can be triggered; such threshold value is taken to be $1.5 (\omega_{\textrm{peak}}^{\textrm{NR}}/2)^{-2/3}$.
+
+\subsection{Quadrupole-monopole terms}
+
+When we have a binary with two material bodies, like two neutron stars, 1 and 2, which have spin-induced mass quadrupole moments $Q_{1,\mr{SS}}^{ij}=-\kappa_1 S_1^{\langle i}S_1^{j\rangle}/m_1$, similarly with $1\to 2$, as opposed to the case with two black holes where $\kappa_1=\kappa_2=1$, the SEOB model should be modified as follows in order to take this into account at leading order.
+
+The SEOB effective Hamiltonian is given by (5.70) of \cite{Barausse:2009xi} as
+\be
+H_\mr{eff}=H_S+\beta^ip_i+\alpha\sqrt{\mu^2+\gamma^{ij}p_ip_j+\mc Q_4(p)}+H^\mr{BBH}_\mr{extra},
+\ee
+where
+\be
+H^\mr{BBH}_\mr{extra}=\frac{1}{2r^3}(3n_in_j-\delta_{ij})\frac{\mu}{M}S_*^iS_*^j,\qquad\qquad S_*^i=\frac{m_2}{m_1} S_1^i+\frac{m_1}{m_2}S_2^i.
+\ee
+This should be replaced with
+\begin{alignat}{3}
+H^\mr{BNS}_\mr{extra}&=\frac{1}{2r^3}(3n_in_j-\delta_{ij})\bigg[\frac{\mu}{M}S_*^iS_*^j+(\kappa_1-1)\frac{m_2}{m_1}S_1^i S_1^j+(\kappa_2-1)\frac{m_1}{m_2}S_2^i S_2^j\bigg] \label{eq:HBNSextra}
+\\\nnm
+&=\frac{1}{2r^3}(3n_in_j-\delta_{ij})\bigg[2\frac{\mu}{M}S_1^iS_2^j+\left(\kappa_1-1+\frac{m_2^2}{M^2}\right)\frac{m_2}{m_1}S_1^i S_1^j+\left(\kappa_2-1+\frac{m_1^2}{M^2}\right)\frac{m_1}{m_2}S_2^i S_2^j\bigg].
+\end{alignat}
+This could also be expressed solely in terms of $\bs S_*$ and $\bs S=\bs S_1+\bs S_2$, instead of $\bs S_1$ and $\bs S_2$, but then the mass difference $m_1-m_2$ would appear in denominators, which might cause problems with dividing by zero in the equal mass case.
+
+In the waveform, since we are focusing only on the leading 2PN order, the only contribution to consider will be in the mode $h_{22}$.
+The \emph{total} leading-order spin-squared contribution, for aligned-spin circular orbits, is given by
+\be
+h_{22}^\tr{LO-SS}=-\frac{8\nu M^3\omega^2}{R}\sqrt{\frac{\pi}{5}}e^{-2i\Phi}\mc A,
+\ee
+and the corresponding contribution to the flux is
+\be
+\mc F^\tr{LO-SS}=\frac{32\nu^2}{5}x^7\left[2\mc A+\frac{1}{16M^2}\left(\frac{S_1}{m_1}-\frac{S_2}{m_2}\right)^2\,\right],\qquad\qquad x=(M\omega)^{2/3},
+\ee
+where, for a BBH,
+\be\label{eq:ABBH}
+\mc A^\mr{BBH}=\frac{1}{M^2}\left(\frac{S_1^2}{m_1^2}+2\frac{S_1S_2}{m_1m_2}+\frac{S_2^2}{m_2^2}\right).
+\ee
+For a BNS, this should be replaced with
+\be\label{eq:ABNS}
+\mc A^\mr{BNS}=\frac{1}{M^2}\left(\kappa_1\frac{S_1^2}{m_1^2}+2\frac{S_1S_2}{m_1m_2}+\kappa_2\frac{S_2^2}{m_2^2}\right).
+\ee
+The $h_{22}$ then matches (4.26), (4.27b), (3.4) of \cite{Buonanno:2012rv}, and the flux matches e.g.\ (6.18)--(6.19) of \cite{Marsat:2014xea}.
+
+In the EOB factorized waveform (see e.g.
+(17) in~\cite{Taracchini:2012ig}), spin contributions in the effective source, tail factors and phase contribution would enter only at higher order, so that the only modification is in the $\rho_{22}$ factor.
+We have simply
+\be\label{eq:rho22}
+	\rho_{22}^{\textrm{LO-SS}} = \frac{1}{2}\mathcal{A} v^{4} \,,
+\ee
+with $\mathcal{A}$ the quantity given by~\eqref{eq:ABBH} or~\eqref{eq:ABNS} and $v=x^{1/2}$.
+
+%Though this should be ultimately irrelevant for our purposes, just for completeness, recall the following:  The LO-SS terms in the EOB effective Hamiltonian, specifically the $S_\mr{Kerr}^2$ terms, are not in the usual PN ``gauge.''  This is seen e.g.\ in (5.55) and (5.60) of \cite{Barausse:2009xi}; the last three lines of (5.60) are equivalent  to zero via a  canonical transformation [the one in (5.15), (5.20)].
+If all we care about is getting the SS terms in the Hamiltonian correct at LO, regardless of gauge, then the above prescriptions are sufficient, even though some SS terms are in a different gauge than others.
+(The BBH SEOB Hamiltonian already has this feature anyway.)
+
+In the Hamiltonian, we separate in Eq.~\eqref{eq:HBNSextra} the additional contribution for BNS, as
+\be
+	\Delta H_{\rm SS}^{\rm BNS} = H^{\rm BNS}_{\rm extra} - H^{\rm BBH}_{\rm extra} = \frac{1}{2r^3}(3n_in_j-\delta_{ij})\bigg[(\kappa_1-1)\frac{m_2}{m_1}S_1^i S_1^j+(\kappa_2-1)\frac{m_1}{m_2}S_2^i S_2^j\bigg] \,.
+\ee
+This contribution is computed by the function \texttt{XLALSimIMRTEOBdeltaHssQuadMonLO}, called inside \texttt{XLALSimIMRSpinEOBHamiltonian} in \texttt{LALSimIMRSpinEOBHamiltonian.c}.
+The expression implemented is kept generic although the $S\cdot n$ terms are zero for aligned spins.
+
+In the factorized waveform, after subtracting the quadratic-in-spin contribution for BBH, the additional contribution for BNS in~\eqref{eq:rho22} is given by:
+\be
+  \Delta \rho_{22}^{\rm BNS} = \rho_{22}^{\rm SS} - (\rho_{22}^{\rm SS})_{\kappa_{1,2}=1} = v^{4}\left[ \left(\chi_s^2+\chi_a^2\right) \left(\frac{\delta  \kappa_a}{2} -\kappa_s \nu+\frac{\kappa_s}{2} +\nu-\frac{1}{2}\right) + \chi_a \chi_s \left(-2 \nu  \kappa_a+\kappa_a-\delta +\delta  \kappa_s\right) \right]\,,
+\ee
+where we used the variables $\kappa_{s,a} \equiv (\kappa_{1} \pm \kappa_{2})/2$, $\chi_{s,a} \equiv (\chi_{1} \pm \chi_{2})/2$.
+One can readily check that this is zero in the binary black hole case, for $\kappa_{s}=1$ and $\kappa_{a}=0$.
+This contribution is added, if at least one of the $\kappa_{1,2}$ is different from one, in the function \texttt{XLALSimIMREOBCalcSpinFacWaveformCoefficients} of \texttt{LALSimIMRSpinEOBFactorizedWaveform.c}
+
+To compute the quadrupole-monopole parameters $\kappa_{1,2}$ entering these contributions (represented by their deviation from 1 as \texttt{dQuadMon1}, \texttt{dQuadMon2}), we use a fit taken from~\cite{Yagi:2016bkt} (see Eq.~(15), with coefficients given in the third row of Table~I), with an extension to ensure the black-hole limit with a $C^{2}$-continuity at the matching point at a dimensionless quadrupolar tidal deformability $\overline{\lambda}_{2}=1$ (chosen arbitrarily).
+This extended universal relation reads
+\be
+	\kappa(\overline{\lambda}_{2}) = 
+	  \begin{cases}
+	    \exp \left[\sum_{i=0}^{4} a_{i}\xi^{i} \right] & \text{ if } \overline{\lambda}_{2} \geq 1 \,,\\
+	    1 + 0.428\overline{\lambda}_{2} - 0.324\overline{\lambda}_{2}^{2} + 0.111\overline{\lambda}_{2}^{3} & \text{ if } 0 \leq \overline{\lambda}_{2} < 1 \,,
+	  \end{cases}
+\ee
+with $\{a_{i}\}_{i=0,\dots, 4} = \{0.1940, 0.09163, 0.04812, -4.283.10^{-3}, 1.245.10^{-4}\}$, $\xi = \ln \overline{\lambda}_{2}$, and where we rounded off the numerical coefficients of the quadratic extension.
+This universal relation is implemented in \texttt{XLALSimUniversalRelationQuadMonVSlambda2Tidal} in \texttt{LALSimUniversalRelations.c}.
 

### 2018-10-23 paper/paper.tex
Paper : added a paragraph on quadrupole-monopole terms.

--- 2018-10-17 paper/paper.tex
@@
 
-For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are those of the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins.
-\red{It includes spin-orbit interactions up to 3PN order and spin-spin interactions up to 2PN order.
-In addition, for the \texttt{SEOBNRv4T} model, the spin-induced quadrupole moment~\cite{Poisson1998} is also included.
-This effect appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin.} \red{\sout{The model, however, does not account for the spin-induced quadrupole moment~\cite{Poisson1998}.
-This effect appears at 2PN order as compared to the tidal effects that first appear at 5PN order, but is only significant for systems with large spin.
-The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{HarryHinderer2018}}}.
-
 Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.
@@
 In order to avoid triggering these stopping conditions at low frequency, where residual orbital eccentricity due to imperfect quasicircular initial conditions could be present, we exploit the $\omega_{\textrm{peak}}^{\textrm{NR}}$ fit to define a radius below which the stopping conditions can be triggered; such threshold value is taken to be $1.5 (\omega_{\textrm{peak}}^{\textrm{NR}}/2)^{-2/3}$.
+
+\subsection{Spin-induced quadrupole-monopole terms}
+
+\red{[check notations have been introduced]} For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are based on the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins.
+It includes consistently spin-orbit interactions up to 3.5PN order and spin-spin interactions up to 2PN order.
+In addition, the \texttt{SEOBNRv4T} model accounts for the spin-induced quadrupole moment of neutron stars~\cite{Poisson1998}, an extended-body effect distinguishing neutron stars from black holes.
+This effect is quadratic in the spins, appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin~\cite{HarryHinderer2018}.
+Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only, see~\cite{}.}, at the moment the effect is only consistently included at the leading 2PN order.
+We leave for future work an extension to higher orders\footnote{\red{complete this} We note that TaylorF2 models do include all spin-square terms at 3PN, see App.~\ref{sec:taylorf2}.
+They are also present in the IMRPhenomPv2+NRTides model, but not in the SEOBNRv4ROM+NRTides model.
+TEOBResumS is also limited to the leading 2PN order.}.
+
+In the compact binary system we consider, the two material bodies, e.g.
+two neutron stars, with masses and spins $m_{A}$ and $\bs{S}_{A}$ for $A=1,2$, have spin-induced mass quadrupole moments $Q_{A,\mr{SS}}^{ij}=-\kappa_A S_A^{\langle i}S_A^{j\rangle}/m_A$ \cite{Poisson1998}, with $\kappa_{A}$ the quadrupole-monopole parameter.
+The case of two black holes is recovered for $\kappa_1=\kappa_2=1$, while for a neutron star $\kappa$ can be larger, of order 10 for hard equations of state.
+
+The SEOB model should be modified as follows in order to take this into account at leading order.
+The SEOB effective Hamiltonian is related to the real Hamiltonian by (7.2) of~\cite{Barausse:2009xi}.
+Its structure is given by (5.70) of~\cite{Barausse:2009xi} as
+\be
+H_\mr{eff}=H_S+\beta^ip_i+\alpha\sqrt{\mu^2+\gamma^{ij}p_ip_j+\mc Q_4(p)}+H^\mr{BBH}_\mr{extra} \,,
+\ee
+where
+\be
+H^\mr{BBH}_\mr{extra}=\frac{1}{2r^3}(3n_in_j-\delta_{ij})\frac{\mu}{M}S_*^iS_*^j \,,
+\ee
+in terms of the tota mass $M$, the reduced mass $\mu = M \eta$, and the spin combination
+\be	
+	S_*^i=\frac{m_2}{m_1} S_1^i+\frac{m_1}{m_2}S_2^i \,.
+\ee
+We refer to~\cite{Barausse:2009xi} for the notations and meaning of the other terms. We find that, for systems including neutron stars, the latter contribution should be replaced with
+\begin{align}
+	H^\mr{BNS}_\mr{extra} &= \frac{1}{2r^3}(3n_in_j-\delta_{ij}) \nnm\\
+	& \; \cdot \bigg[\frac{\mu}{M}S_*^iS_*^j+(\kappa_1-1)\frac{m_2}{m_1}S_1^i S_1^j+(\kappa_2-1)\frac{m_1}{m_2}S_2^i S_2^j\bigg] \label{eq:HBNSextra}
+%\\\nnm
+%&=\frac{1}{2r^3}(3n_in_j-\delta_{ij})\bigg[2\frac{\mu}{M}S_1^iS_2^j+\left(\kappa_1-1+\frac{m_2^2}{M^2}\right)\frac{m_2}{m_1}S_1^i S_1^j+\left(\kappa_2-1+\frac{m_1^2}{M^2}\right)\frac{m_1}{m_2}S_2^i S_2^j\bigg],
+\end{align}
+where we chose to use the spin variables $\bs S_1$ and $\bs S_2$ instead of $\bs S_*$ and $\bs S=\bs S_1+\bs S_2$ to avoid the occurence of the mass difference $m_{1} - m_{2}$ in denominators.
+This expression generalizes the black hole case and reduces to it for $\kappa_{1} = \kappa_{2} = 1$.
+
+In the waveform, at the leading 2PN order, the only contribution to consider will be in the mode $h_{22}$.
+The total leading-order spin-squared contribution, for aligned-spin circular orbits, is given by~\cite{Kidder:1995zr, Will:1996zj, Buonanno:2012rv}
+\be
+h_{22}^\tr{LO-SS}=-\frac{8\nu M^3\omega^2}{R}\sqrt{\frac{\pi}{5}}e^{-2i\Phi}\mc A \,,
+\ee
+with $\Phi$ the orbital phase, $\omega = \dot{\Phi}$ the orbital frequency and $R$ the distance to the observer. Here, for a BBH,
+\be\label{eq:ABBH}
+\mc A^\mr{BBH}=\frac{1}{M^2}\left(\frac{S_1^2}{m_1^2}+2\frac{S_1S_2}{m_1m_2}+\frac{S_2^2}{m_2^2}\right) \,,
+\ee
+while for a BNS, this should be replaced with
+\be\label{eq:ABNS}
+\mc A^\mr{BNS}=\frac{1}{M^2}\left(\kappa_1\frac{S_1^2}{m_1^2}+2\frac{S_1S_2}{m_1m_2}+\kappa_2\frac{S_2^2}{m_2^2}\right) \,.
+\ee
+The total spin-squared contribution to the flux also includes squares of spin-orbit terms, and reads~\cite{Kidder:1995zr, Will:1996zj, Poisson1998}
+\be
+\mc F^\tr{LO-SS}=\frac{32\eta^2}{5} v^{14} \left[2\mc A+\frac{1}{16M^2}\left(\frac{S_1}{m_1}-\frac{S_2}{m_2}\right)^2\,\right] \,,
+\ee
+with $v=(M\omega)^{1/3}$.
+
+When translating the above for the EOB factorized waveform (see e.g.
+(17) in~\cite{Taracchini:2012ig}), spin contributions in the effective source, tail factors and phase contribution would enter only at higher order, so that the only modification to consider is in the $\rho_{\ell m}$ factor for $\ell = 2$, $m = 2$.
+We have simply
+\be\label{eq:rho22}
+	\rho_{22}^{\textrm{LO-SS}} = \frac{1}{2}\mathcal{A} v^{4} \,,
+\ee
+with $\mathcal{A}$ the quantity given by~\eqref{eq:ABBH} for a BBH or~\eqref{eq:ABNS} for a BNS.
 
@@
 
-After rescaling with the total mass $M$, the aligned-spin EOB waveform with dynamical tides depends on 13 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless spin components along the orbital angular momentum $\chi_{A,B}$, the two spin-induced quadrupole moments $\bar Q_{A,B}$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
+After rescaling with the total mass $M$, the aligned-spin EOB waveform with dynamical tides depends on 13 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless spin components along the orbital angular momentum $\chi_{A,B}$, the two spin-induced quadrupole-monopole parameters $\kappa_{A,B}$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
 To reduce the dimensionality of the intrinsic parameter space, we use nearly EOS-independent fits (universal relations) for these parameters in terms of $\Lambda_2$ as discussed below.
@@
 
-Eq.~(15) of Ref.~\cite{YagiYunes2017} provides a fit for $\bar Q$ as a function of $\Lambda_2$.
+Eq.~(15) of Ref.~\cite{YagiYunes2017} provides a fit for $\kappa$ as a function of $\Lambda_2$.
 This relation has been fit for a sample of EOSs and for tidal parameters in the range $1 \leq \Lambda_2\leq10^4$.
-This formula, however, diverges at small values of $\Lambda_2$, so we replace it with a polynomial function that approaches the Kerr value of $\bar Q = 1$ at $\Lambda_2 = 0$.
+This formula, however, diverges at small values of $\Lambda_2$, so we replace it with a polynomial function that approaches the Kerr value of $\kappa = 1$ at $\Lambda_2 = 0$.
 We require this extension to be continuous at $\Lambda_2 = 1$.
@@
 \begin{equation}
-\bar Q = \left\{\begin{array}{ll}
+\kappa = \left\{\begin{array}{ll}
 1 + f \Lambda_2 + g \Lambda_2^2 + h \Lambda_2^3\, , & 0 \leq \Lambda_2 \le 1 \\
@@
 The authors were supported by...
-\red{[[Get AEI funding info.]]}
+\red{[[Get AEI funding info.]]} \red{[Tanja Hinderer and Justin Vines derived the quadrupole-monopole terms for SEOB]}
 

### 2018-10-24 paper/paper.tex
Minor changes to PE section.

@@
 This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters.
-Unfortunately, the current implementation takes several minutes to hours to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations.
+Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations.
 We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression to efficiently sample the parameter space.
-The resulting surrogate has a maximum mismatch of \red{$\sim 5\times 10^{-4}$} for the Advanced LIGO detector, and requires less than \red{0.3~s} to evaluate for a waveform with a starting frequency of 10~Hz (\red{0.04~s} for a starting frequency of 30~Hz).
+The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz.
 With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).
@@
 With these optimized hyperparameters, the final interpolating function is given by Eq.~\eqref{eq:mean} and its uncertainty by Eq.~\eqref{eq:var}.
-We note that once the vector $(K^{-1})_{ij}y_j$ is precomputed, evaluating Eq.~\eqref{eq:mean} is a fast $\mathcal{O}(N)$ operation.
+%We note that once the vector $(K^{-1})_{ij}y_j$ is precomputed, evaluating Eq.~\eqref{eq:mean} is a fast $\mathcal{O}(N)$ operation.
 
@@
 The 400 red squares were generated using uncertainty sampling from the GPR error estimate (Eq.~\eqref{eq:errorrms}) with the Mat\'{e}rn kernel.
-The 400 green circles were sampled uniformly from the parameter space.
-}
+The 400 green circles were sampled uniformly from the parameter space.}
 \label{fig:LHD}
@@
 
-The surrogate evaluation time is the sum of two parts.
+The surrogate evaluation time is dominated by two parts.
 The first part is the time needed to evaluate the residuals $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ and $\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}$ at each of the $N_A+N_\Phi$ interpolating nodes using GPR.
-Although optimizing the hyperparameters for GPR scales with the number of samples $N$ as $\mathcal{O}(N^3)$, the evaluation time of a stored GPR scales as $\mathcal{O}(N)$.
+Although optimizing the hyperparameters for GPR scales with the number of training set samples $N$ as $\mathcal{O}(N^3)$, the evaluation time of a stored GPR scales as $\mathcal{O}(N)$ (see Eq.~\eqref{eq:mean}).
 The evaluation time for all the residuals therefore has a cost of $\mathcal{O}(N(N_A+N_\Phi))$.
-The second part is the time needed to resample the final surrogate (Eq.~\eqref{eq:amp_sur}--Eq.~\eqref{eq:hcross_sur}) at uniformly spaced frequency samples in physical units beginning at a starting frequency $f_{\rm low}$.
-\red{[[Michael: If you want, add details of how the specific implementation of resampling the waveform affects timing here.]]}
-%\MP{Add cost of complex exponential sincos function: look at profiling results}
-%\MP{Potentially, explore approximation to sincos for further speedup; need to guarantee that errors don't add up; range reduction is important for accuracy}
-
-In Fig.~\ref{fig:timing}, we show the surrogate evaluation time as a function of starting frequency $f_{\rm low}$.
-The surrogate has a fixed time of \red{$\sim 50$~ms} to evaluate the GPR all of the nodes.
-\red{[[Details of how starting frequency depends on $f_{\rm low}$.]]} The additional cost to resample the amplitude and phase at uniformly-spaced frequencies with spline interpolation dominates at starting frequencies below $\sim 30$~Hz.
-We also compare the evaluation time to several BNS waveform models that have been used for LIGO data analysis.
-We note that at $\sim 30$~Hz, the \texttt{SEOBNRv4T\_surrogate} model is comparable to the \texttt{SEOBNRv4\_ROM\_NRTidal} and \texttt{IMRPhenomD\_NRTidal} models that were used in the analysis of the GW170817 event~\cite{BNSPE}.
-At lower frequencies, the surrogate is faster than all BNS waveform models except for \texttt{TaylorF2}.
+The second part is the time needed to resample the final surrogate (Eqs.~\eqref{eq:amp_sur}--~\eqref{eq:hcross_sur}) at uniformly spaced frequency samples in physical units beginning at a starting frequency $f_{\rm min}$.
+
+In Fig.~\ref{fig:timing}, we show the surrogate evaluation time as a function of the starting frequency $f_{\rm min}$ for a equal mass $1.4M_\odot$--$1.4M_\odot$ binary evaluated on a 3.5~GHz Intel Xeon processor.
+The waveform was sampled with a uniform frequency spacing corresponding to a sampling rate of 4096~Hz.
+At large starting frequencies the evaluation time is limited by the \red{$\sim 10$~ms} needed to calculate the GPR fits at each frequency node.
+As $f_{\rm min}$ decreases, the waveform becomes longer and a smaller frequency spacing is needed.
+The evaluation time is then dominated by the spline interpolation needed to resample the waveform.
+We also compare the evaluation time to three BNS waveform models used recently in the analysis of GW170817, (\texttt{TaylorF2}, \texttt{SEOBNRv4\_ROM\_NRTidal}, \texttt{IMRPhenomD\_NRTidal})~\cite{BNSPE} as well as the BBH model \texttt{SEOBNRv4\_ROM}.
+Only the analytic \texttt{TaylorF2} model is noticeably faster.
+In the bottom panel, we compare the evaluation time to the original time-domain \texttt{SEOBNRv4T} waveform model and find it is $\sim 1500$--6000 times faster.
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{timing.png}
-\caption{Waveform evaluation time as a function of the waveform starting frequency. The time-domain waveforms
-were sampled at 4096~Hz then Fourier transformed. The frequency-domain waveforms used the
-same frequency samples as the Fourier transformed time-domain waveforms.}
+\includegraphics[width=0.49\textwidth]{../src/TEOB-LAL-implementation/timings/Models_timings_FD_monotonic.pdf}
+\includegraphics[width=0.49\textwidth]{../src/TEOB-LAL-implementation/timings/SEOBNRv4T_surrogate_speedup_deltaF_monontonic.pdf}
+\caption{Waveform evaluation time as a function of the waveform starting frequency $f_{\rm min}$.
+The time-domain \texttt{SEOBNRv4T} waveform was sampled at 4096~Hz then Fourier transformed.
+The frequency-domain waveforms used the same frequency samples as the Fourier transformed time-domain waveforms.
+Top panel: Evaluation times of the frequency-domain waveforms.
+Bottom panel: Speedup of \texttt{SEOBNRv4T\_surrogate} relative to the original \texttt{SEOBNRv4T}.}
 \label{fig:timing}
@@
 
-Having confirmed that the surrogate accurately recovers the injected waveform parameters, we now perform a parameter estimation run for GW170817.
-In the PE paper and catalog papers~\cite{}, the SEOBNRv4T surrogate was used with the parallelized PE code RapidPE.
+Having confirmed that the surrogate accurately recovers the injected waveform parameters, we now do parameter estimation for GW170817.
+In the PE paper and catalog papers~\cite{}, the original \texttt{SEOBNRv4T} model was used with the parallelized PE code RapidPE.
 However, there were uncertainties about its ability to accurately recover the tidal parameter.

### 2018-11-15 paper/paper.tex
Paper : worked on appendix on TF2 phasing.

@@
 \begin{equation}
-A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} x^{-7/4} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)x \right],
-\end{equation}
-where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{LAL}~\cite{lal}. The phase has the schematic form
-\begin{align}
-\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
-                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, \chi_1, \chi_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
-\end{align}
-We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, \chi_1, \chi_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}.
-These terms for the phase are exactly as used in the \texttt{LAL} waveform \texttt{TaylorF2}.
-
-The spin terms are
-
-\red{Because the \texttt{SEOBNRv4T} waveform does not include 3PN spin-spin effects, we remove the 3PN spin-spin term in the \texttt{TaylorF2} waveform as well.
-The main reason for doing this is because including only common physical effects in both \texttt{SEOBNRv4T} and \texttt{TaylorF2} makes the residual $\Delta\Phi(Mf; \bx)$ smaller and have less variation, making it easier to fit at a given error requirement.}
-
-$SL = X_1^2\chi_1 + X_2^2\chi_2$
-
-$d\Sigma L = (X_1-X_2)(X_2\chi_2-X_1\chi_1)$
-
-$\gamma = \left(\frac{554345}{1134}+\frac{110\eta}{9}\right)SL + \left(\frac{13915}{84} - \frac{10\eta}{3}\right)d\Sigma L$
-
-$SS15 = \frac{188}{3}SL + 25 d\Sigma L$
-
-$SS2 = -10 \Sigma (not same sigma)$
-
-$SS25 = -\gamma -3\gamma\ln(v)$
-
-$SS3 = \frac{\pi}{3}\left[3760(X_1^2\chi_1 + X_2^2\chi_2) + 1490(X_1-X_2)(X_2\chi_2 - X_1\chi_1)\right]$
-
-\begin{equation}
-\tilde\Lambda = \frac{16}{13} \frac{(X_1+12X_2)X_1^4\Lambda_1 + (12X_1+X_2)X_2^4\Lambda_2}{(X_1+X_2)^5}
-\end{equation}
-
-\begin{align}
+A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} v^{-7/2} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)v^{2} \right],
+\end{equation}
+where $v=(\pi M f)^{1/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{LAL}~\cite{lal} \red{[explain]}. The phase has the schematic form
+\begin{align}
+\Phi_{\rm F2} =& -2\pi f t_{0} + \phi_{0} + \frac{\pi}{4} - \frac{3}{128\eta}v^{-5} \left[\Phi_{\rm F2}^{\rm PP}(\eta) \right. \nonumber\\
+	& \left. \qquad + \Phi_{\rm F2}^{\rm Spin} (\eta, \chi_1, \chi_2) + \Phi_{\rm F2}^{\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right] \,,
+\end{align}
+where $t_{0}$ and $\phi_{0}$ are constants related to the freedom of choosing the time and phase of coalescence.
+We separated point particle terms $\Phi_{\rm F2}^{\rm PP}(\eta)$, spin terms $\Phi_{\rm F2}^{\rm Spin}(\eta, \chi_{1}, \chi_{2})$ and tidal terms $\Phi_{\rm F2}^{\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$.
+We include point particle terms to the highest know order of 3.5PN~\cite{}.
+
+The spin terms can be decomposed into spin-orbit and spin-spin terms (ignoring cubic-in-spin contributions at 3.5PN~\cite{}) as $\Phi_{\rm F2}^{\rm Spin} = \Phi_{\rm F2}^{\rm SO}  + \Phi_{\rm F2}^{\rm SS} $.
+We include spin-orbit terms up to 3.5PN~\cite{}.
+For spin-square terms, as the \texttt{SEOBNRv4T} waveform does not include yet 3PN spin-spin effects (see Sec.~\ref{}), we stop at  the leading 2PN in the \texttt{TaylorF2} waveform and ignore the known 3PN terms.
+The main reason for doing this is because including only common physical effects in both \texttt{SEOBNRv4T} and \texttt{TaylorF2} makes the residual $\Delta\Phi(Mf; \bx)$ smaller and have less variation, making it easier to fit at a given error requirement.
+
+The tidal terms $\Phi_{\rm F2}^{\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ are known to 6PN order~\cite{VinesFlanaganHinderer2011}.
+These tidal contributions to the phase are exactly as used in the \texttt{LAL} waveform \texttt{TaylorF2}.
+
+For completeness, we give below explicit expressions of the PN coefficients entering the phasing, using the notations $\delta \equiv (m_{1} - m_{2})/(m_{1} + m_{2})$ for the mass difference, and $\chi_{s,a} = (\chi_{1} \pm \chi_{2})/2$, $\kappa_{s,a} = (\kappa_{1} \pm \kappa_{2})/2$ for the symmetrized and antisymmetrized combinations of $\chi_{A}$, $\kappa_{A}$.
+
+The point particle terms coefficients up to 3.5PN can be found e.g. in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}. The non-zero coefficients read
+\begin{align}
+	a_{2} &= \frac{55 \nu }{9}+\frac{3715}{756} \nonumber\\
+	a_{3} &= -16 \pi \nonumber\\
+	a_{4} &= \frac{3085 \nu ^2}{72}+\frac{27145 \nu }{504}+\frac{15293365}{508032} \nonumber\\
+	a_{5} &= -\frac{65 \pi  \nu }{9}-\frac{65}{3} \pi  \nu   \ln v +\frac{38645}{252} \pi   \ln v +\frac{38645 \pi }{756} \nonumber\\
+	a_{6} &= -\frac{127825 \nu ^3}{1296}+\frac{76055 \nu ^2}{1728}+\frac{2255 \pi ^2 \nu }{12} \nonumber\\ 
+	& -\frac{15737765635 \nu }{3048192}-\frac{6848  \ln v }{21}-\frac{640 \pi ^2}{3}-\frac{6848 \gamma_{E} }{21} \nonumber\\ 
+	& +\frac{11583231236531}{4694215680}-\frac{13696 \log (2)}{21} \nonumber\\
+	a_{7} &= -\frac{74045 \pi  \nu ^2}{756}+\frac{378515 \pi  \nu }{1512}+\frac{77096675 \pi }{254016}
+\end{align}
+
+The spin-orbit and spin-spin corrections have the structure
+\begin{align}
+	\Phi_{\rm F2}^{\rm SO}(\eta, \chi_{1}, \chi_{2}) &= \sum_{k=3}^{7} b_{k} v^{k} + \mathcal{O}(v^{8})\nonumber\\
+	\Phi_{\rm F2}^{\rm SS}(\eta, \chi_{1}, \chi_{2}) &= c_{4} v^{4} + c_{6} v^{6} + \mathcal{O}(v^{7}) \,.
+\end{align}
+\begin{widetext}
+The non-zero spin-orbit coefficients are
+\begin{align}
+	b_{3} &= \frac{113 \delta   \chi_a }{3}+\left(\frac{113}{3}-\frac{76 \nu }{3}\right)  \chi_s \nonumber\\
+	b_{5} &=  \chi_s  \left(\frac{340 \nu ^2}{9}+\frac{24260 \nu }{81}+\frac{340}{3} \nu ^2  \ln v +\frac{24260}{27} \nu   \ln v -\frac{732985  \ln v }{756}-\frac{732985}{2268}\right) \nonumber\\
+	& \quad + \chi_a  \left(-\frac{140 \delta  \nu }{9}-\frac{732985 \delta }{2268}-\frac{140}{3} \delta  \nu   \ln v -\frac{732985}{756} \delta   \ln v \right)  \nonumber\\
+	b_{6} &= \frac{2270 \pi  \delta   \chi_a }{3}+\left(\frac{2270 \pi }{3}-520 \pi  \nu \right)  \chi_s \nonumber\\
+	b_{7} &=  \chi_a  \left(-\frac{1985 \delta  \nu ^2}{48}+\frac{26804935 \delta  \nu }{6048}-\frac{25150083775 \delta }{3048192}\right)+\left(\frac{5345 \nu ^3}{36}-\frac{1042165 \nu ^2}{3024}+\frac{10566655595 \nu }{762048}-\frac{25150083775}{3048192}\right)  \chi_s \,,
+\end{align}
+while the leading-order quadratic-in-spin term reads
+\begin{align}
+	c_{4} &= \chi_s ^2 \left(-50 \delta   \kappa_a +100  \kappa_s  \nu -50  \kappa_s -\frac{195 \nu }{2}-\frac{5}{8}\right) + \chi_a   \chi_s  \left(-100 \delta   \kappa_s -\frac{5 \delta }{4}+200  \kappa_a  \nu -100  \kappa_a \right) \nonumber\\
+	& + \chi_a ^2 \left(-50 \delta   \kappa_a +100  \kappa_s  \nu -50  \kappa_s +100 \nu -\frac{5}{8}\right) \,.
+\end{align}
+As explained above, to be consistent with the fact that this PN information has not yet been incorporated in \texttt{SEOBNRv4T}, we do not include the next-to-leading 3PN term, which for reference is given by
+\begin{align}
+	c_{6} &=  \chi_s ^2 \left(-\frac{1495 \delta   \kappa_a  \nu }{6}+\frac{26015 \delta   \kappa_a }{28}-240  \kappa_s  \nu ^2-\frac{44255  \kappa_s  \nu }{21}+\frac{26015  \kappa_s }{28}+\frac{3415 \nu ^2}{9}+\frac{829705 \nu }{504}-\frac{1344475}{2016}\right) \nonumber\\
+	& + \chi_a   \chi_s  \left(-\frac{1495 \delta   \kappa_s  \nu }{3}+\frac{26015 \delta   \kappa_s }{14}+\frac{745 \delta  \nu }{18}-\frac{1344475 \delta }{1008}-480  \kappa_a  \nu ^2-\frac{88510  \kappa_a  \nu }{21}+\frac{26015  \kappa_a }{14}\right) \nonumber\\
+	& + \chi_a ^2 \left(-\frac{1495 \delta   \kappa_a  \nu }{6}+\frac{26015 \delta   \kappa_a }{28}-240  \kappa_s  \nu ^2-\frac{44255  \kappa_s  \nu }{21}+\frac{26015  \kappa_s }{28}-240 \nu ^2+\frac{267815 \nu }{252}-\frac{1344475}{2016}\right)\,.
+\end{align}
+\end{widetext}
+
+Finally, the tidal contributions to the phasing are given by~\cite{VinesFlanaganHinderer2011}
+\begin{equation}
+	\Phi_{\rm Tidal}^{\rm SS}(\eta, \Lambda_{1}, \Lambda_{2}) = d_{10} v^{10} + d_{12} v^{12} + \mathcal{O}(v^{13})
+\end{equation}
+Introducing convenient mass-weighted combinations of the tidal parameters, expressed using $X_{1} = m_{1}/(m_{1} + m_{2})$ and $X_{2} = m_{2}/(m_{1} + m_{2})$, \red{[to be checked]}
+\begin{align}
+	\tilde\Lambda &= \frac{16}{13} \frac{(X_1+12X_2)X_1^4\Lambda_1 + (12X_1+X_2)X_2^4\Lambda_2}{(X_1+X_2)^5} \nonumber\\
 	\delta\tilde\Lambda =& \frac{(1319X_1^2 - 7996X_1X_2 - 11005X_2^2)X_1^4\Lambda_1}{1319(X_1+X_2)^6} \\
-& + \frac{(11005X_1^2 + 7996X_1X_2 - 1319X_2^2)X_2^4\Lambda_2}{1319(X_1+X_2)^6}
-\end{align}
-
-$a_5 = -\frac{39}{2}\tilde\Lambda$
-
-$a_6 = -\frac{3115}{64}\tilde\Lambda + \frac{6595}{364}(X_1-X_2)\delta\tilde\Lambda$
+& + \frac{(11005X_1^2 + 7996X_1X_2 - 1319X_2^2)X_2^4\Lambda_2}{1319(X_1+X_2)^6} \,,
+\end{align}
+the coefficients are given by \red{[to be checked]}
+\begin{align}
+	d_{10} &= -\frac{39}{2}\tilde\Lambda \nonumber\\
+	d_{12} &= -\frac{3115}{64}\tilde\Lambda + \frac{6595}{364}(X_1-X_2)\delta\tilde\Lambda \,.
+\end{align}
 

### 2018-11-16 paper/paper.tex
Updated injection PE plots.

@@
 
-\title{Surrogate waveform model using Gaussian process regression for an aligned-spin effective one body model of binary neutron star inspirals}
+%\title{Surrogate waveform model using Gaussian process regression for an aligned-spin effective one body model of binary neutron star inspirals}
+\title{Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression}
 
@@
 \includegraphics[width=0.39\textwidth]{tildelambda.pdf}
-\caption{Marginalized PDFs for two waveform injections with zero noise.
+\caption{Marginalized PDFs for waveform injections with zero noise.
 The parameters are identical except for the distances which are set such that the network SNRs are 30 (blue) or 60 (red).
 The shaded regions give the 90\% credible interval.
-The vertical solid black line represents the injected parameter value.
-The vertical dashed lines represent the parameter value with maximum likelihood.}
+The  dashed black line represents the injected parameter value.}
 \label{fig:pe}
+\end{figure}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.39\textwidth]{chieff_q.pdf}
+\caption{1d and 2d marginalized PDFs.
+The shaded regions for the 1d PDFs give the 90\% credible interval.
+The contours for the 2d PDFs give the 90\% credible regions.
+The dashed black lines and circle represent the injected parameter value.}
+\label{fig:2dpe}
 \end{figure}

### 2018-11-19 paper/paper.tex
Edited PE section.

@@
 
-As a final test of the surrogate accuracy, we perform parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template.
+As a final test of the surrogate accuracy, we perform several parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template.
 We use the design sensitivity PSDs for the two advanced LIGO detectors~\cite{Aasi:2013wya} and the advanced Virgo detector~\cite{virgoPSD}.
 We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$ corresponding to the equation of state named MS1b in~\cite{ReadLackey2009}.
-The sky location and inclination angle are fixed at an arbitrary value, and the distance is chosen such that the optimal network signal-to-noise ratio (SNR) is either 30 or 60.
+The sky location and inclination angle are fixed at an arbitrary value, and the distances is chosen such that the optimal network signal-to-noise ratios (SNR) are either 30, 60, or 120.
 We inject the waveforms into zero noise data so that the recovered parameters do not depend on the specific noise realization~\cite{NissankeHolzHughes2010}.
 
-As discussed in more detail in~\cite{BNSPE}, the posterior $p(\vec\vartheta | d)$ for the parameters $\vec\vartheta$ given the detector data $\vec d$ is given by Bayes' theorem:
+As discussed in more detail in~\cite{BNSPE}, the posterior $p(\vec\vartheta | d)$ for the parameters $\vec\vartheta$ given the detector data $\vec d$ is given by Bayes' theorem
 \begin{equation}
@@
 where $p(\vec\vartheta)$ is the prior and $\mathcal{L}(\vec d | \vec\vartheta)$ is the likelihood. 
-We choose for our priors the sky position and orientation uniform on the unit sphere.
-The distance is uniform in co-moving volume.
-The masses, spins, and tidal parameters are uniform in the ranges $m_{1, 2} \in [0.7, 2]M_\odot$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$.
-The mass ratio is restricted to the range $q\in[0.5, 1]$.
-We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal}.
-As was done in~\cite{BNSPE}, we sample the data and waveform at \red{4096}~Hz, such that the Nyquist frequency is \red{2048}~Hz, and integrate the likelihood function in the interval \red{$[20, 2048]$~Hz}.
-Approximately \red{$10^x$} iterations are needed to produce \red{$\sim 5000$} statistically independent samples.
+For the prior we choose the sky position and orientation to be uniform on the unit sphere; the distance to be uniform in co-moving volume; the masses, spins, and tidal parameters to be uniform in the ranges $m_{1, 2} \in [0.7, 2]M_\odot$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$; and the mass ratio is restricted to the range $q\in[0.5, 1]$.
+As was done in~\cite{BNSPE}, we sample the data and waveform at \red{4096}~Hz such that the Nyquist frequency is \red{2048}~Hz, and integrate the likelihood function in the interval \red{$[20, 2048]$~Hz}.
+We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal} which requires approximately \red{$10^x$} iterations to produce \red{10,000--20,000} statistically independent samples.
+
 We note that the \texttt{SEOBNRv4T} waveform model still has some signal above \red{2048~Hz} (see Fig.~\ref{fig:hoff}).
-To test if this had any effect on the recovered posteriors, we also performed runs with a sampling frequency of \red{8192}~Hz and integrated the likelihood function in the interval \red{$[20, 4096]$}~Hz, but did not see a significant difference in marginalized posteriors.
-
-Marginalized 1-dimensional posteriors for four key parameters are shown in Fig.~\ref{fig:pe}.
-These are the chirp mass $\mathcal{M} = (m_1 m_1)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$.
-We note that the peaks of the marginalized posteriors do not line up with the injected values.
-One reason is that the priors are not flat in these parameters.
-More importantly, because the posterior is bounded by the prior for many parameters, strong correlations cause a shift in other parameters.
-We therefore need another measure to examine the systematic error due to the surrogate.
-
-For data with zero noise, the fit between the injected waveform and the template is best when the likelihood is maximum.
-We therefore use the difference $\Delta\vec\vartheta_{\rm sys} = \vec\vartheta_{\rm max} - \vec\vartheta_{\rm inj}$ between the parameters that maximize the likelihood $\vec\vartheta_{\rm max}$ and the injected parameters $\vec\vartheta_{\rm inj}$ as our estimate for the systematic error due to the surrogate.
-In Fig.~\ref{fig:pe}, the maximum sampled likelihood is dashed, and the injected value is solid.
-This difference is a conservative upper limit because the maximum sampled likelihood does not necessarily correspond to the actual maximum.
-%This is especially true for strongly correlated parameters such as the pair $q$ and $\chi_{\rm eff}$. 
-Finally, we can compare this systematic error to the statistical error which we define as the 90\% credible interval shown by the shaded region.
-We find that the systematic error is significantly less that the statistical error, even for the SNR=60 signal.
-
-%For data with zero noise, the fit between the injected waveform and the template is best when the likelihood is maximum.
-However, as a result of the priors that are bounded and not flat for some parameters, the posterior, and especially the marginalized posteriors, will not be maximized at the injected parameters even with a perfect template.
-We therefore use the difference between the parameters that maximize the likelihood $\vec\vartheta_{\rm max}$ and the injected parameters $\vec\vartheta_{\rm inj}$ as our estimate for the systematic error due to the surrogate.
-%
-%Marginalized 1-dimensional posteriors for four key parameters are shown in Fig.~\ref{fig:pe}.
-These are the chirp mass $\mathcal{M} = (m_1 m_1)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$.
-The maximum value of the likelihood found for each run is dashed, and the injected value is solid.
-Because the sampled parameters do not necessarily contain the parameters for the maximum likelihood, the difference between these parameters $\Delta\vec\vartheta_{\rm sys} = \vec\vartheta_{\rm max} - \vec\vartheta_{\rm inj}$ is an upper bound on the systematic error that results from using the surrogate as the template instead of the original \texttt{SEOBNRv4T} waveform.
-We can compare this systematic error to the statistical error $\Delta\vec\vartheta_{\rm stat}$ from the finite sensitivity of the detectors.
-Here, we will $\Delta\vec\vartheta_{\rm stat}$ as half the width of the 90\% credible interval.
-Table~\ref{tab:syserrors} shows that the systematic error is small compared to the statistical errors even for the injection with SNR=60.
+We therefore also performed runs with a sampling frequency of \red{8192}~Hz and integrated the likelihood function in the interval \red{$[20, 4096]$}~Hz.
+However, this did not have a noticeable effect on the recovered posteriors.
+
+We show marginalized 1-dimensional posteriors for four key parameters in Fig.~\ref{fig:1dpe}.
+These are the chirp mass $\mathcal{M} = (m_1 m_2)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$.
+As the SNR increases, the peaks of the 1-d marginalized posteriors become more closely aligned with the injected value.
+However, the distributions are noticeably asymmetric and are not centered on the injected values.
+This results mainly because prior boundaries on some parameters, such as the mass ratio $q$, shift the marginalized PDFs for correlated parameters.
+This effect is strongest for the interaction between $\chi_{\rm eff}$ and $q$, shown in Fig.~\ref{fig:2dpe}, where the injected value $q=1$ is at the boundary of the prior.
+Although the injected parameter $(\chi_{\rm eff}, q) = (0.1, 1)$ lies on the ridge of maximum density in the marginalized 2-d PDF, the marginalized 1-d PDF for $\chi_{\rm eff}$ is noticeably offset.
+To counteract this effect in Fig.~\ref{fig:1dpe}, we also take a thin cross section containing the injected value of $q=1$ by cutting samples outside the interval $q \in [0.9, 1]$.
+This causes the peaks of the other 1-d distributions to be much closer to the injected values.
+We consider the difference between these peaks and the injected values to be a conservative bound on the size of the statistical errors due to errors in the surrogate model.
 
@@
 \includegraphics[width=0.39\textwidth]{tildelambda.pdf}
-\caption{Marginalized PDFs for waveform injections with zero noise.
-The parameters are identical except for the distances which are set such that the network SNRs are 30 (blue) or 60 (red).
+\caption{Marginalized 1-d PDFs for waveform injections with zero noise.
+The parameters are identical except for the distances which are chosen such that the network SNRs are 30 (blue), 60 (green), or 120 (red).
+The black PDF only includes samples for the SNR=120 injection that satisfy $q>0.9$.
 The shaded regions give the 90\% credible interval.
 The dashed black line represents the injected parameter value.}
-\label{fig:pe}
+\label{fig:1dpe}
 \end{figure}
@@
 \includegraphics[width=0.39\textwidth]{chieff_q.pdf}
-\caption{1d and 2d marginalized PDFs.
-The shaded regions for the 1d PDFs give the 90\% credible interval.
-The contours for the 2d PDFs give the 90\% credible regions.
+\caption{Marginalized 1-d and 2-d PDFs for the parameters $\chi_{\rm eff}$ and $q$ for the injections with SNRs of 30 (blue) and 120 (red).
+The shaded regions for the 1-d PDFs give the 90\% credible interval.
+The contours for the 2-d PDFs give the 90\% credible regions.
 The dashed black lines and circle represent the injected parameter value.}
@@
 \end{figure}
-
-%\begin{table}[htb]
-%\centering    
-%\caption{Estimated systematic errors due to surrogate model and statistical errors from finite detector sensitivity.
-The ratio $\Delta\vec\vartheta_{\rm sys} /\Delta\vec\vartheta_{\rm stat}$ is given for the two injections with network SNRs of 30 and 60.}
-%\begin{tabular}{ccc}        
-%\hline
-%$\vartheta$ & SNR=30 & SNR=60 \\
-%\hline
-%$\mathcal{M}$ &  $<$ & $<$ \\
-%$q$ & $<$  & $<$ \\
-%$\chi_{\rm eff}$ & $<$  & $<$ \\
-%$\tilde\Lambda$ &  $<$ & $<$ \\
-%\hline
-%\end{tabular}
-%\label{tab:syserrors}
-%\end{table}
 

### 2018-11-21 paper/paper.tex
Minor edits to most of the paper.

@@
 
-%\title{Surrogate waveform model using Gaussian process regression for an aligned-spin effective one body model of binary neutron star inspirals}
 \title{Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression}
@@
 
-%Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as the recently discovered event GW170817.
-The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as the effect of dynamical tides, and the formalism is accurate enough to avoid significant biases in the measured parameters.
-Unfortunately, current implementations take several minutes to a day to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations.
-We construct a frequency-domain surrogate of this model that requires less than 0.3~s to evaluate for a waveform with a starting frequency of 10~Hz (0.04~s for a starting frequency of 30~Hz).
-We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter.
-We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with $\sim 1000$ waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods.
-\red{[[Mismatch is ...]] [[It can be used in arbitrary frequency range by matching to analytic TaylorF2.]] [[Improvement to previously published EOB model [cite] by including spin-induced quadrupole moment and prescription for waveform tapering at merger.]] }
-
-Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as the recently discovered event GW170817.
-The effective one body formalism was previously shown to produce waveforms that agree with numerical simulations of aligned-spin BNS systems during the inspiral.
-This model includes aligned-spin effects, the quadrupolar and octopolar adiabatic tides, and the effect of dynamical tides.
-We improve this model by adding the spin-induced quadrupole moment and a prescription for tapering the end of the waveform to qualitatively agree with numerical simulations of the merger.
+Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817.
+We present a completed version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}.
+This model includes the quadrupolar and octopolar adiabatic and dynamical tides, and we improve it by including the spin-induced quadrupole moment effect and a prescription for tapering the end of the waveform to agree with numerical simulations.
 The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters.
-This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters.
+%This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters. 
 Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations.
-We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression to efficiently sample the parameter space.
+We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression.
 The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz.
@@
 
-\red{Choose point-mass or point-particle.}
-
-\red{Use $\chi$ for dimensionless spin instead of $S$.}
-
-\red{Choose notation $\Lambda_2^A$ or $\Lambda_1$.}
-
-\red{only use texttt for actual programs and functions. TaylorF2 should usually be standard text}
-
 \section{Introduction}
 
-%The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors, has demonstrated that the parameters of the binary can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant~\cite{GW170817Hubble}, and understanding the structure of neutron stars (NSs) through tidal interactions~\cite{GW170817}.
-Measuring these parameters accurately will also be necessary to properly interpret the gamma ray burst [[cite]] and optical afterglow [[cite]] that was found in conjunction.
-Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} indicates that, when second generation detectors reach design sensitivity, including KAGRA~\cite{Somiya2012} and possibly LIGO-India~\cite{IyerSouradeepUnnikrishnan2011}, we could eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}.
-This will allow us to estimate the NS mass distribution [[cite]], spin distribution [[cite]], and universal equation of state (EOS) that is common to all NSs\cite{DelPozzoLiAgathos2013, LackeyWade2015}.
-
-%\red{Purpose: Intro. Justify this field of research.}
-
-The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors has demonstrated that tidal interactions during the inspiral can be measured with enough precision to constrain the radius and equation of state~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, BNSPE, EOSPaper} of NSs.
+The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can accurately measure quantities such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}.
 Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further.
-These multiple events could then be stacked to significantly improve measurements of the radius and EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}.
-%\red{[[Mention A+ being funded? How will this improve each parameter?]]}
-
-%These results are obtained using a Bayesian analysis that compares the data to parameterized waveform models.
-To avoid systematic errors highly accurate models are needed, and they must include all relevant physical effects.
-This includes descriptions of the point-particle interactions, spins, and tides.
-Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models.
-These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5PN and 6PN order.
-Unfortunately, the lack of point-particle terms above 3.5PN order leads to systematic errors in the tidal parameters that are as large as the statistical errors~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
-Alternative models exist with more accurate representations of the point particle terms.
-\MP{What are these alternative models besides EOB?} However, they do not always simultaneously include the spin and tidal effects as well.
-Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$~\cite{Favata2014}.
-In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].
-
-%\red{Purpose: Explain the problem. Fast and accurate waveforms are needed}
-
-These results are obtained using Bayesian analyses that compare the data to parameterized waveform models.
+These multiple events could then be stacked to significantly improve measurements of NS radii and the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.
+
+The reliability of these measurements, however, depend on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data.
 Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations.
 Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event.
+To avoid large systematic errors, highly accurate waveform models are needed that include all relevant physical effects such as point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
 Most time-domain waveform implementations involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain.
@@
 
-%In addition, to avoid systematic errors, highly accurate models are needed that include all relevant physical effects.
-This includes descriptions of the point-particle interactions, spins, and tides.
-Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models [[cite some examples]].
-These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5PN and 6PN order.
-Unfortunately, the lack of point-particle terms above 3.5PN order can lead to systematic errors in the tidal parameters that are as large as the statistical errors~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
-\red{[[Describe what was found for GW170817 in PE paper?
-This seemed to be the case for large tidal parameters but not for the actual event.
-Also you didn't compare different PN waveforms there.]]}
-
-%\red{Purpose: Explain what waveforms have been used for GW170817 and other works and why yet another one is needed.}
-
-In addition, to avoid systematic errors that can be as large as statistical errors, highly accurate waveform models are needed that include all relevant physical effects, including point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
-For GW170817, a set of four frequency-domain waveform models were used as templates for the main results.
+For the analysis of GW170817 in Ref.~\cite{BNSPE}, therefore, a set of four frequency-domain waveform models were used as templates for the main results.
 These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein).
 They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT).
-Tidal interactions, on the other hand, were included simply by adding a correction to the phase of each waveform.
+Tidal interactions were then included by simply adding a correction to the phase of each waveform.
 For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added.
-For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was used for the phase correction~\cite{DietrichBernuzziTichy2017}.
-
-%For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates~\cite{PEPaper}.
-The first was the analytic PN waveform model known as \texttt{TaylorF2}~\cite{BuonannoIyerOchsner2009}.
-Because of the known systematic errors that result from PN waveforms, three additional models were produced for comparison: \texttt{IMRPhenomD\_NRTides}, \texttt{SEOBNRv4\_ROM\_NRTides}, and \texttt{IMRPhenomP\_NRTides}.
-These three models used an identical prescription for tides presented in Ref.~\cite{DietrichBernuzziTichy2017, NewNRTidalPaper}.
-At low frequencies, the tidal effect is that of \texttt{TaylorF2}.
-It then transitions to a fit to the tidal effect measured by a small set of numerical simulations with varying mass ratio and aligned-spins.
-The point-particle and spin prescriptions were different.
-For \texttt{IMRPhenomD\_NRTides}, the waveform is based on the phenomenological \texttt{IMRPhenomD} BBH model~\cite{HusaKhanHannam2016, KhanHusaHannam2016}.
-It uses \texttt{TaylorF2} at low frequencies then transitions to fits to the EOB waveform at mid frequencies and fits to NR simulations at high frequencies.
-The \texttt{SEOBNRv4\_ROM\_NRTides} model is based on the fast surrogate model \texttt{SEOBNRv4\_ROM}~\cite{Puerrer2015} for the EOB waveform \texttt{SEOBNRv4} for BBH systems~\cite{Bohe:2016gbl}.
-Finally, the \texttt{IMRPhenomP\_NRTides} is an extension of \texttt{IMRPhenomD\_NRTides}, but also includes a spin-induced quadrupole moment and an approximates precession with a single effective spin parameter~\cite{}.
-These models are improvements to \texttt{TaylorF2}, but only treat matter effects as an additive phase correction to BBH waveform models.
-\red{[[Are all these details needed?]]} [[These waveform models were in broad agreement with each other for $\tilde\Lambda$ (and other parameters?) to 10\% error.]]
-
-%We therefore build a fast surrogate that reproduces as accurately as possible \texttt{SEOBNRv4T} in the frequency domain.
-
-%********************
-
-%Alternative models exist with more accurate representations of the point particle terms.
-\MP{What are these alternative models besides EOB?} However, they do not always simultaneously include the spin and tidal effects as well.
-Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$~\cite{Favata2014}.
-In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].
-
-%The alternative version, \texttt{TEOBResumS}~\cite{Nagar2017}, uses a slightly different point-particle and spin prescription.
-And, in addition, includes the $\ell = 2, 3$, and 4 tidal effects and the spin induced quadrupole moment, but does not include the effect of dynamical tides from the induced $f$-mode resonances.
-It does, however, use a method for re-summing the expressions for tidal interactions.
-
-%In addition to accuracy, useful waveform models must be fast enough to be used in Bayesian parameter estimation algorithms such as Markov chain Monte Carlo (MCMC) and Nested Sampling that currently require $10^7$--$10^8$ sequential waveform evaluations.
-Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event.
-Most time-domain waveform implementations such as the PN approximants and the EOB models involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain.
-For PN waveform models, this can take several seconds for a starting frequency of 10~Hz, while for the \texttt{SEOBNRv4T} model discussed here, the equivalent evaluation time can be a day.
-
-%For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates.
-The first was the stationary phase approximation PN waveform model known as \texttt{TaylorF2}~\cite{BuonannoIyerOchsner2009}.
-Because of the known systematic errors that result from PN waveforms, two additional models were produced for comparison: \texttt{IMRPhenomD\_NRTides} and \texttt{SEOBNRv4\_ROM\_NRTides}.
-Both of these models used an identical prescription for tides presented in Ref.~\cite{DietrichBernuzziTichy2017}.
-At low frequencies, the tidal effect is that of \texttt{TaylorF2}.
-It then transitions to a fit to the tidal effect measured by a small set of numerical simulations with varying mass ratio and aligned-spins.
-The point-particle and spin prescriptions were different.
-For \texttt{IMRPhenomD\_NRTides}, the waveform is based on the phenomenological \texttt{IMRPhenomD} BBH model~\cite{HusaKhanHannam2016, KhanHusaHannam2016}.
-It uses \texttt{TaylorF2} at low frequencies then transitions to fits to the EOB waveform at mid frequencies and fits to NR simulations at high frequencies.
-The \texttt{SEOBNRv4\_ROM\_NRTides} model is based on the fast surrogate model \texttt{SEOBNRv4\_ROM}~\cite{Puerrer2015} for the EOB waveform \texttt{SEOBNRv4} for BBH systems~\cite{Bohe:2016gbl}.
-Both of these models are improvements to \texttt{TaylorF2}, but do not incorporate all known physical effects in a consistent way.
-We therefore build a fast surrogate that reproduces as accurately as possible \texttt{SEOBNRv4T} in the frequency domain.
-
-%\red{[[Purpose: The EOB waveform that's expensive but solves all of humanity's problems.]]}
-
-In this paper we will use a new implementation of the effective one body (EOB) formalism, that treats matter effects consistently with the other parameters.
+For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.
+
+In this paper, we will use an implementation of the effective one body (EOB) formalism that treats matter effects consistently with the other parameters.
 It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations.
@@
 For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions.
-It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances or a tapering to zero amplitude after merger.
-%\red{[[Must mention evaluation time to segue into next 2 paragraphs.]]}
-
-%One such alternative to these waveforms that do treat matter effects consistently with the other parameters is the effective one body (EOB) formalism discussed in section~\ref{sec:eob}.
-It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations.
-Until recently, however, no version of the EOB formalism included the effects of spin and tides in the same model.
-Recently two versions have become available that include both tidal effects and aligned spin and are implemented under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}.
-The \texttt{SEOBNRv4T} model~\cite{Hinderer:2016eia,Steinhoff:2016rfi} used in this paper, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2017}.
-This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}.
-It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations.
-The other model, \texttt{TEOBResumS}~\cite{Nagar2017, NewPaper2018}, uses a slightly different point-particle and spin prescription.
-For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions.
-It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances or a tapering to zero amplitude after merger.
-
-%\red{[[Purpose: Why other attempts for using EOB with parameter estimation have been insufficient.]]}
+It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.
 
 Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible.
-For GW170817~\cite{BNSPE}, an alternative highly parallelized code, \texttt{RapidPE}, was used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}.
-Unfortunately, while the results were broadly consistent with the MCMC results when using the same waveform models, there were noticeable differences.
-In particular the shape of the PDF was different, and the credible intervals for the tidal parameters were shifted downward by $\sim 10$\%~\cite{BNSPE}.
-Another limitation is that RapidPE does not currently marginalize over calibration errors.
+In Ref.~\cite{BNSPE}, an alternative highly parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}.
+Unfortunately, direct comparisons between this code and traditional samplers have not been published for BNS systems using the same waveform models.
+It is therefore not clear how accurate this code is at measuring quantities such as the tidal parameters.
 As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved.
-In addition, different methods of formulating the problem can allow for faster integration~\cite{NagarNettegno2018}.
+Different formulations of the differential equations can also speed up integration~\cite{NagarNettegno2018}.
 However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides.
 As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.
-
-%*********** Part 2 of intro: Surrogates ************
 
@@
 Previous works have focused on efficiently representing the space of all waveforms using a reduced basis of orthonormal functions.
-% whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. 
 The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
-%method that selects a subset of training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
 In this work, instead of using a reduced basis, we will instead use the fact that the Fourier transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model.
@@
 The final step in building a surrogate is to interpolate between waveform parameters $\bx$.
-Previous work focused on parameter spaces with three or fewer dimensions.
+Many previous works focused on parameter spaces with three or fewer dimensions.
 This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}.
@@
 
-This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}.
+This work also takes a different approach from the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform.
 Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation.
 It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
-%The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.
+
+%This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}.
+Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation.
+It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
 
@@
 In section Section V we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817.
-Finally, we discuss applications and future improvements in Section V.
+Finally, we discuss future improvements in Section VI.
+We also give the expressions for the TaylorF2 base model in the Appendix.
 
@@
 two neutron stars, with masses and spins $m_{A}$ and $\bs{S}_{A}$ for $A=1,2$, have spin-induced mass quadrupole moments $Q_{A,\mr{SS}}^{ij}=-\kappa_A S_A^{\langle i}S_A^{j\rangle}/m_A$ \cite{Poisson1998}, with $\kappa_{A}$ the quadrupole-monopole parameter.
-The case of two black holes is recovered for $\kappa_1=\kappa_2=1$, while for a neutron star $\kappa$ can be larger, of order 10 for hard equations of state.
+The case of two black holes is recovered for $\kappa_1=\kappa_2=1$, while for a neutron star $\kappa$ can be larger, of order 10 for hard EOSs.
 
@@
 \ee
-in terms of the tota mass $M$, the reduced mass $\mu = M \eta$, and the spin combination
+in terms of the total mass $M$, the reduced mass $\mu = M \eta$, and the spin combination
 \be	
@@
 For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter.
-At a starting frequency of 20~Hz, \red{$\sim x$--y}~hours are needed to evaluate a single \texttt{SEOBNRv4T} waveform on a standard CPU, so this is at the limit of what is reasonable.
+At a starting frequency of 20~Hz, over an hour is needed to evaluate a single \texttt{SEOBNRv4T} waveform on a standard CPU, so this is at the limit of what is reasonable.
 However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below.
@@
 
-%\red{[[Michael: Add your discussion of resampling the waveform here.]]}
-
-%\MP{Say that we use sparse frequency grids based on power-law $\Delta f$ in the evaluation of the surrogate; perhaps describe in more detail how the model is evaluated in the LAL code.}
-%\MP{Discuss linear vs cubic splines if we decide to use them. Tradeoff in accuracy and speed.}
-
 \subsection{Iterative construction of training set and surrogate}
@@
 This avoids wasting samples when one of the parameters has much less of an influence than the other parameters.
-%As an example, the tidal parameters have a minimal impact on the waveform at low frequencies, so it would not make sense to densely sample a grid of tidal parameters at low frequencies.
-However, during the merger the tidal parameters can be more important than the other parameters.
-One therefore might want more samples for the tidal parameters and less samples for the other parameters.
-However, due to the expense of the waveforms we must use the same set of waveforms for low and high frequencies.
-An LHD with its noncollapsing property bypasses this problem.
-%\MP{Note (perhaps in a footnotes) that it is possible to have separate surrogates for low and high frequency which are then joined smoothly as done for \texttt{SEOBNRv4\_ROM}.}
 
@@
 \caption{Inverse Fourier transformed surrogate (red dashed) for the parameters $\{q, \chi_1, \chi_2, \Lambda_1, \Lambda_2\} = \{0.61, -0.23, -0.33, 180, 4390\}$ that have the largest mismatch with the test-set EOB waveforms (black).
+The $+$ polarizations and amplitudes are shown.
 The two waveforms are aligned in the time domain, before the start of the plot, in the interval $(t-t_{\rm Merger})/M \in [-10^7, -10^6]$.}
@@
 The waveform was sampled with a uniform frequency spacing corresponding to a sampling rate of 4096~Hz.
-At large starting frequencies the evaluation time is limited by the \red{$\sim 10$~ms} needed to calculate the GPR fits at each frequency node.
+At large starting frequencies the evaluation time is limited by the \red{$\sim 0.01$~s} needed to calculate the GPR fits at each frequency node.
 As $f_{\rm min}$ decreases, the waveform becomes longer and a smaller frequency spacing is needed.
 The evaluation time is then dominated by the spline interpolation needed to resample the waveform.
-We also compare the evaluation time to three BNS waveform models used recently in the analysis of GW170817, (\texttt{TaylorF2}, \texttt{SEOBNRv4\_ROM\_NRTidal}, \texttt{IMRPhenomD\_NRTidal})~\cite{BNSPE} as well as the BBH model \texttt{SEOBNRv4\_ROM}.
+We also compare the evaluation time to three BNS waveform models used recently in the analysis of GW170817 (\texttt{TaylorF2}, \texttt{SEOBNRv4\_ROM\_NRTidal}, \texttt{IMRPhenomD\_NRTidal})~\cite{BNSPE} and the BBH model \texttt{SEOBNRv4\_ROM}.
 Only the analytic \texttt{TaylorF2} model is noticeably faster.
@@
 
-As a final test of the surrogate accuracy, we perform several parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template.
+As a final test of the surrogate, we perform several parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template.
 We use the design sensitivity PSDs for the two advanced LIGO detectors~\cite{Aasi:2013wya} and the advanced Virgo detector~\cite{virgoPSD}.
-We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$ corresponding to the equation of state named MS1b in~\cite{ReadLackey2009}.
-The sky location and inclination angle are fixed at an arbitrary value, and the distances is chosen such that the optimal network signal-to-noise ratios (SNR) are either 30, 60, or 120.
+We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$ corresponding to the EOS named MS1b in~\cite{ReadLackey2009}.
+The sky location and inclination angle are fixed at an arbitrary value, and the distances are chosen such that the optimal network signal-to-noise ratios (SNRs) are either 30, 60, or 120.
 We inject the waveforms into zero noise data so that the recovered parameters do not depend on the specific noise realization~\cite{NissankeHolzHughes2010}.
@@
 These are the chirp mass $\mathcal{M} = (m_1 m_2)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$.
-As the SNR increases, the peaks of the 1-d marginalized posteriors become more closely aligned with the injected value.
+As the SNR increases, the peaks of the marginalized 1-d posteriors become more closely aligned with the injected value.
 However, the distributions are noticeably asymmetric and are not centered on the injected values.
@@
 \centering
-\includegraphics[width=0.39\textwidth]{chirpmass.pdf}\\
-\includegraphics[width=0.39\textwidth]{q.pdf}\\
-\includegraphics[width=0.39\textwidth]{chieff.pdf}\\
-\includegraphics[width=0.39\textwidth]{tildelambda.pdf}
+\includegraphics[width=0.49\textwidth]{chirpmass.pdf}\\
+\includegraphics[width=0.49\textwidth]{q.pdf}\\
+\includegraphics[width=0.49\textwidth]{chieff.pdf}\\
+\includegraphics[width=0.49\textwidth]{tildelambda.pdf}
 \caption{Marginalized 1-d PDFs for waveform injections with zero noise.
@@
 \centering
-\includegraphics[width=0.39\textwidth]{chieff_q.pdf}
+\includegraphics[width=0.49\textwidth]{chieff_q.pdf}
 \caption{Marginalized 1-d and 2-d PDFs for the parameters $\chi_{\rm eff}$ and $q$ for the injections with SNRs of 30 (blue) and 120 (red).
@@
 The rest of the time is spent resampling the waveform with spline interpolation.
-For a starting frequency of 10~Hz, this takes \red{0.3~s} when the waveform is matched with data uniformly sampled at 4096Hz.
-For a starting frequency of 30~Hz, the total evaluation time is \red{0.04~s}.
+For a starting frequency of 20~Hz, this takes \red{0.13~s} when the waveform is matched with data uniformly sampled at 4096Hz.
 For current parameter estimation codes, this is sufficient.
@@
 Such a model would have the accuracy of EOB below $\sim 400$~Hz and the accuracy of numerical simulations for the last several cycles before merger.
-Current state of the art NR simulations have phase errors of several tenths of a radian over the last $\sim 20$ gravitational-wave cycles~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}.
+Current state-of-the-art NR simulations have phase errors of several tenths of a radian over the last $\sim 20$ gravitational-wave cycles~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}.
 Using GPR and uncertainty sampling discussed above, one could optimally choose the waveform parameters for the numerical simulations, and run them in parallel to build the training set.
@@
 This approach would provide a fast, accurate model for BNS systems with tides and generic spins.
-We note, however, that including precession using the \texttt{IMRPhenomP\_NRTidal} model did not significantly affect the results for GW170817~\cite{BNSPE}.
+We note, however, that including precession did not noticeably affect the parameter estimation results for GW170817~\cite{BNSPE}.
 
@@
 
-BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the University of Washington for useful discussions at the start of this work.
-The authors were supported by...
-\red{[[Get AEI funding info.]]} \red{[Tanja Hinderer and Justin Vines derived the quadrupole-monopole terms for SEOB]}
+The authors sincerely thank Zoheyr Doctor, Bhooshan Gadre, and Prayush Kumar for reviewing and testing the implementation of \texttt{SEOBNRv4T\_surrogate} in \texttt{LAL}.
+BL also thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the University of Washington for helpful advice at the start of this work.
 

### 2018-11-28 paper/paper_no_event.tex
Adding version of paper without PE for GW170817.

--- 2018-11-21 paper/paper.tex
@@
 The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz.
-With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).
+%With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).
 
@@
 
-%This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}.
-Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation.
-It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
-
 We organize the paper as follows.
-In Section II we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space.
-In Section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model.
-We then compare the accuracy and speed of the surrogate to the original model in Section IV.
-In section Section V we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817.
-Finally, we discuss future improvements in Section VI.
-We also give the expressions for the TaylorF2 base model in the Appendix.
+In Section~\ref{sec:eob} we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space.
+In Section~\ref{sec:surrogate}, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model.
+We then compare the accuracy and speed of the surrogate to the original model in Section~\ref{sec:performance}.
+In Section~\ref{sec:pe} we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817.
+Finally, we discuss future improvements in Section~\ref{sec:discussion} and give the expressions for the TaylorF2 base model in the Appendix.
 
@@
 \section{Surrogate model}
+\label{sec:surrogate}
 
@@
 The algorithm for choosing the $N_{\rm new}$ new points is as follows.
-
-%\begin{algorithm}[H]
-%\caption{Uncertainty Sampling}
-%\label{alg:uc}
-%\begin{algorithmic}[1]
-%\State {\bf Input:} Initial surrogate 
-%\For{$i = 1 \to N_{\rm New}$}
-%
-%\State Construct the GPR error estimator (Eq.~\eqref{eq:errorrms}).
-In practice this can be done by specifying the samples $\bx$, the hyperparameters $\btheta$, and dummy data for $\by$ since Eq.~\eqref{eq:errorrms} does not depend on $\by$.
-%
-%\State Find $({\bm x}_{\rm max}, \epsilon_{\rm rms})$ that maximizes Eq.~\eqref{eq:errorrms} over the parameter space.
-%
-%\State Add ${\bm x}_{\rm max}$ to the list of samples: ${\bm x} = [{\bm x}, {\bm x}_{\rm max}]$.
-%
-%\EndFor
-%\State {\bf Output:} Updated surrogate
-%\end{algorithmic}
-%\end{algorithm}
 
@@
 \section{Performance of the surrogate}
+\label{sec:performance}
 
@@
 
-\subsection{Injection runs}
+%\subsection{Injection runs}
 
@@
 
-\subsection{GW170817}
-
-Having confirmed that the surrogate accurately recovers the injected waveform parameters, we now do parameter estimation for GW170817.
+%\subsection{GW170817}
+%
+%Having confirmed that the surrogate accurately recovers the injected waveform parameters, we now do parameter estimation for GW170817.
 In the PE paper and catalog papers~\cite{}, the original \texttt{SEOBNRv4T} model was used with the parallelized PE code RapidPE.
 However, there were uncertainties about its ability to accurately recover the tidal parameter.
-
-We use the priors discussed in~cite{O2CatalogPaper}.
+%
+%We use the priors discussed in~cite{O2CatalogPaper}.
 
@@
 \section{Discussion and future work}
+\label{sec:discussion}
 

### 2018-11-30 paper/paper_no_event.tex
Checked numbers in paper. Minor edits.

@@
 The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters.
-%This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters. 
-Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations.
+Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require $\mathcal{O}(10^7)$ sequential waveform evaluations.
 We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression.
-The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz.
-%With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).
+The resulting surrogate has a maximum mismatch of $4.5\times 10^{-4}$ for the Advanced LIGO detector, and requires $\sim 0.13$~s to evaluate for a waveform with a starting frequency of 20~Hz.
 
@@
 
-The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can accurately measure quantities such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}.
+The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can measure BNS properties such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}.
 Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further.
-These multiple events could then be stacked to significantly improve measurements of NS radii and the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.
+These multiple events could then be stacked to significantly improve measurements of the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.
 
@@
 For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added.
-For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.
-
-In this paper, we will use an implementation of the effective one body (EOB) formalism that treats matter effects consistently with the other parameters.
+For the other three models, a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.
+
+In this paper, we will use an implementation of the EOB formalism that treats matter effects consistently with the other parameters.
 It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations.
@@
 It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations.
-We also note an alternative model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a slightly different point-particle and spin prescription.
-For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions.
+We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a different point-particle and spin prescription.
+For matter effects, this model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions.
 It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.
@@
 Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible.
-In Ref.~\cite{BNSPE}, an alternative highly parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}.
-Unfortunately, direct comparisons between this code and traditional samplers have not been published for BNS systems using the same waveform models.
-It is therefore not clear how accurate this code is at measuring quantities such as the tidal parameters.
+In Ref.~\cite{BNSPE}, an alternative parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}.
+Unfortunately, direct comparisons of the PDF between this code and traditional samplers using the same waveform models have not been published for BNS systems.
+It is therefore not clear how accurate this code is for measuring quantities such as the tidal parameters.
 As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved.
@@
 The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
-In this work, instead of using a reduced basis, we will instead use the fact that the Fourier transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model.
+In this work, instead of using a reduced basis, we will instead use the fact that the Fourier-transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model.
 We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform.
@@
 In Section~\ref{sec:eob} we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space.
-In Section~\ref{sec:surrogate}, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model.
+In Section~\ref{sec:surrogate} we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model.
 We then compare the accuracy and speed of the surrogate to the original model in Section~\ref{sec:performance}.
-In Section~\ref{sec:pe} we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817.
+In Section~\ref{sec:pe} we verify that the surrogate can correctly extract the parameters of injected waveforms.
 Finally, we discuss future improvements in Section~\ref{sec:discussion} and give the expressions for the TaylorF2 base model in the Appendix.
@@
 
-In this Section we describe how we decompose the EOB waveform into smooth, slowly-varying functions and train a surrogate model for these functions.
+In this section we describe how we decompose the EOB waveform into smooth, slowly-varying functions and train a surrogate model for these functions.
 We will work with the frequency-domain waveform $\tilde h(Mf; \bx)$.
@@
 However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below.
-The method we choose is Gaussian process regression (GPR)~\cite{RasmussenWilliams2006}.
+The method we choose is GPR~\cite{RasmussenWilliams2006}.
 
@@
 \end{align}
-We set these functions to 0 below the first frequency node $MF_0 =  Mf_{{\rm trunc},i}$ so that the waveform transitions to \texttt{TaylorF2} at lower frequencies.
+We set these functions to 0 below the first frequency node $MF_0 =  Mf_{{\rm trunc},i}$ so that the waveform transitions to TaylorF2 at lower frequencies.
 With the analytic expressions for $A_{\rm F2}(Mf;\bx)$ and $\Phi_{\rm F2}(Mf;\bx)$ and the interpolated expressions $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$, the final surrogates for the amplitude and phase are
@@
 The non-smoothness of the curves results because the basin-hopping algorithm does not always find the global maximum of $\epsilon_{\rm RMS}$ at each iteration.
-We find that increasing the number of waveforms from 160 to 560 decreases the global maximum of $\epsilon_{\rm RMS}$ by a factor of \red{$\sim 4$} for the Mat\'{e}rn kernel and a factor of \red{$\sim 10$} for the squared exponential kernel.
+We find that increasing the number of waveforms from 160 to 560 decreases the global maximum of $\epsilon_{\rm RMS}$ by a factor of $\sim 4$ for the Mat\'{e}rn kernel and a factor of $\sim 10$ for the squared exponential kernel.
 In Section~\ref{sec:accuracy} we will examine how well this estimated error agrees with the true error.
@@
 We generate 1000 waveforms with the original parameters sampled uniformly in the ranges $q \in [1/3, 1]$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$.
-In Fig.~\ref{fig:uncsamp}, we show box plots of the RMS phase error of the test set waveforms for each of 8 surrogate models generated.
+In Fig.~\ref{fig:uncsamp}, we show box plots of the RMS phase error of the test set waveforms for each of the 8 surrogate models.
 The first two surrogates are constructed using the initial design with either the Mat\'{e}rn or squared exponential kernel.
@@
 We use the design sensitivity aLIGO PSD~\cite{Aasi:2013wya} and a total mass of $M=2.8M_\odot$.
-Our integration bounds are $Mf_{\rm low} = 0.00021$ and $Mf_{\rm high} = 0.07$, corresponding to physical units of \red{15.2}~Hz and \red{5076}~Hz respectively.
+Our integration bounds are $Mf_{\rm low} = 0.00021$ and $Mf_{\rm high} = 0.07$, corresponding to physical units of 15.2~Hz and 5076~Hz respectively.
 As we increase the number of samples in our training set, the mismatches decrease.
@@
 For our final surrogate, we simply choose the surrogate that has the smallest maximum and median mismatches, and this is the surrogate with the Mat\'{e}rn kernel and all 960 waveforms.
-This surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$}.
+This surrogate has a maximum mismatch of $4.5\times 10^{-4}$.
 
@@
 
-In Fig.~\ref{fig:timing}, we show the surrogate evaluation time as a function of the starting frequency $f_{\rm min}$ for a equal mass $1.4M_\odot$--$1.4M_\odot$ binary evaluated on a 3.5~GHz Intel Xeon processor.
+In Fig.~\ref{fig:timing}, we show the surrogate evaluation time as a function of the starting frequency $f_{\rm min}$ for an equal mass $1.4M_\odot$--$1.4M_\odot$ binary evaluated on a 3.5~GHz Intel Xeon processor.
 The waveform was sampled with a uniform frequency spacing corresponding to a sampling rate of 4096~Hz.
-At large starting frequencies the evaluation time is limited by the \red{$\sim 0.01$~s} needed to calculate the GPR fits at each frequency node.
+At large starting frequencies the evaluation time is limited by the $\sim 0.01$~s needed to calculate the GPR fits at each frequency node.
 As $f_{\rm min}$ decreases, the waveform becomes longer and a smaller frequency spacing is needed.
@@
 
-%\subsection{Injection runs}
-
 As a final test of the surrogate, we perform several parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template.
-We use the design sensitivity PSDs for the two advanced LIGO detectors~\cite{Aasi:2013wya} and the advanced Virgo detector~\cite{virgoPSD}.
+We use the design sensitivity PSDs for the two advanced LIGO detectors and the advanced Virgo detector~\cite{Aasi:2013wya}.
 We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$ corresponding to the EOS named MS1b in~\cite{ReadLackey2009}.
@@
 For the prior we choose the sky position and orientation to be uniform on the unit sphere; the distance to be uniform in co-moving volume; the masses, spins, and tidal parameters to be uniform in the ranges $m_{1, 2} \in [0.7, 2]M_\odot$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$; and the mass ratio is restricted to the range $q\in[0.5, 1]$.
-As was done in~\cite{BNSPE}, we sample the data and waveform at \red{4096}~Hz such that the Nyquist frequency is \red{2048}~Hz, and integrate the likelihood function in the interval \red{$[20, 2048]$~Hz}.
-We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal} which requires approximately \red{$10^x$} iterations to produce \red{10,000--20,000} statistically independent samples.
-
-We note that the \texttt{SEOBNRv4T} waveform model still has some signal above \red{2048~Hz} (see Fig.~\ref{fig:hoff}).
-We therefore also performed runs with a sampling frequency of \red{8192}~Hz and integrated the likelihood function in the interval \red{$[20, 4096]$}~Hz.
+As was done in~\cite{BNSPE}, we sample the data and waveform at 4096~Hz such that the Nyquist frequency is 2048~Hz, and integrate the likelihood function in the interval $[20, 2048]$~Hz.
+We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal} which requires approximately $10^7$ iterations to produce 10,000--20,000 statistically independent samples.
+
+We note that the \texttt{SEOBNRv4T} waveform model still has some signal above 2048~Hz (see Fig.~\ref{fig:hoff}).
+We therefore also performed runs with a sampling frequency of 8192~Hz and integrated the likelihood function in the interval $[20, 4096]$~Hz.
 However, this did not have a noticeable effect on the recovered posteriors.
@@
 This results mainly because prior boundaries on some parameters, such as the mass ratio $q$, shift the marginalized PDFs for correlated parameters.
-This effect is strongest for the interaction between $\chi_{\rm eff}$ and $q$, shown in Fig.~\ref{fig:2dpe}, where the injected value $q=1$ is at the boundary of the prior.
+This effect is strongest for the interaction between $\chi_{\rm eff}$ and $q$ (Fig.~\ref{fig:2dpe}), where the injected value $q=1$ is at the boundary of the prior.
 Although the injected parameter $(\chi_{\rm eff}, q) = (0.1, 1)$ lies on the ridge of maximum density in the marginalized 2-d PDF, the marginalized 1-d PDF for $\chi_{\rm eff}$ is noticeably offset.
@@
 The black PDF only includes samples for the SNR=120 injection that satisfy $q>0.9$.
-The shaded regions give the 90\% credible interval.
+The shaded regions give the 90\% credible intervals.
 The dashed black line represents the injected parameter value.}
@@
 
-%\subsection{GW170817}
-%
-%Having confirmed that the surrogate accurately recovers the injected waveform parameters, we now do parameter estimation for GW170817.
-In the PE paper and catalog papers~\cite{}, the original \texttt{SEOBNRv4T} model was used with the parallelized PE code RapidPE.
-However, there were uncertainties about its ability to accurately recover the tidal parameter.
-%
-%We use the priors discussed in~cite{O2CatalogPaper}.
-
 %% ________________________________________________________
@@
 This aligned-spin model, \texttt{SEOBNRv4T}, incorporates the tidally induced $\ell=2$ and 3 multipole moments as well as the effect of dynamical tides as the excitation approaches the $\ell=2$ and 3 $f$-mode frequencies.
-We have achieved mismatches of no more than \red{$4.5 \times 10^{-4}$} and phase errors of \red{$\lesssim 1$~rad} up to the merger frequency.
+We have achieved mismatches of no more than $4.5 \times 10^{-4}$ and phase errors of $\lesssim 1$~rad up to the merger frequency.
 These are sufficient to not bias results in any of the parameters.
 
-The evaluation time has a flat cost of \red{$\sim 0.02$~s} to perform the GPR interpolation at each node.
+The evaluation time has a flat cost of $\sim 0.01$~s to perform the GPR interpolation at each node.
 The rest of the time is spent resampling the waveform with spline interpolation.
-For a starting frequency of 20~Hz, this takes \red{0.13~s} when the waveform is matched with data uniformly sampled at 4096Hz.
+For a starting frequency of 20~Hz, this takes 0.13~s when the waveform is matched with data uniformly sampled at 4096Hz.
 For current parameter estimation codes, this is sufficient.
 However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, or multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017}.
-Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^4)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.
+Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^3)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.
 

### 2018-12-02 paper/paper_no_event.tex
Paper: made keffdyn and keffdyn,diss explicit in Sec II A

@@
 with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$.
-Here, $\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor (Eq.~(11) of Ref.~\cite{Dietrich:2017feu}), which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.
-This enhancement factor is 1 for low orbital frequencies, and increases to $\sim 2$ as $\ell$ times the orbital frequency approaches the $f$-mode frequency.
-
-Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
+Here, $\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor~\cite{Steinhoff:2016rfi, Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.
+
+Introducing a Keplerian orbital frequency defined from the radius variable as $M\Omega \equiv u^{3/2}$, using the notations (dropping the indices $A,B$)
+\begin{subequations}
+\begin{align}
+	x = \frac{\hat{\omega}_{0\ell}}{m\Omega} \,, \; \epsilon \equiv \frac{256\eta}{5} \left( \frac{\hat{\omega}_{0\ell}}{m} \right)^{5/3} \,, \; \hat{t} \equiv \frac{8}{5\sqrt{\epsilon}} \left[ 1 - x^{5/3} \right] \,,
+\end{align}
+\end{subequations}
+the dynamical enhancement factor  of~\cite{Steinhoff:2016rfi, Dietrich:2017feu} reads
+\be\label{eq:kldyn}
+	\hat{k}_{\ell, \rm{dyn}}(u; \omega_{0\ell}) = a_{\ell} + b_{\ell} \left[ f(x) +\sqrt{\frac{\pi}{3 \epsilon}} x^{2} \mathcal{Q} \right]\,,
+\ee
+with $a_{\ell}$, $b_{\ell}$ constant coefficients with relevant values $\{a_{2}, b_{2}\} = \{1/4, 3/4\}$ and $\{a_{3}, b_{3}\} = \{3/8, 5/8\}$.
+These expressions are used in practice for $\ell= m$.
+Here, we have rewritten the first two resonant terms of Eq.~(11) of~\cite{Dietrich:2017feu} as
+\be\label{eq:deff}
+	f(x) = x^{2} \left[ \frac{1}{x^{2} - 1} + \frac{5}{6} \frac{1}{1 - x^{5/3}} \right] \,.
+\ee
+Although the two terms are individually divergent\footnote{This requires to single out the near-resonance region in a numerical implementation.} at the resonance $x=1$, the function $f(x)$ is actually  regular, taking the value $f(1) = -1/12$.
+The last term in~\eqref{eq:kldyn} above reads
+\begin{align}
+	\mathcal{Q} &= \cos\left( \frac{3}{8} \hat{t}^{2} \right) \left[ 1 + 2 F_{S} \left( \frac{\sqrt{3}}{2\sqrt{\pi}} \hat{t} \right) \right] \nonumber\\
+	& - \sin\left( \frac{3}{8} \hat{t}^{2} \right) \left[ 1 + 2 F_{C} \left( \frac{\sqrt{3}}{2\sqrt{\pi}} \hat{t} \right) \right] \,,
+\end{align}
+with $F_{S}$, $F_{C}$ the Fresnel sine and cosine functions.
+This term is regular at resonance $\hat{t} = 0$.
+The enhancement factor~\eqref{eq:kldyn} is 1 for low orbital frequencies, and increases to $\sim 2$ as $\ell$ times the orbital frequency approaches the $f$-mode frequency.
+
+Tidal effects also enter the radiative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
 \begin{equation}
@@
 (A14)-(A17) of Ref.~\cite{Damour:2012yf}.
-Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
+Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu} as
+\be
+	\hat{k}_{\ell, \rm{dyn}}^{A, \mathrm{diss}}(u; \omega_{0\ell}) = \hat{k}_{\ell, \rm{dyn}}^{A} \frac{4 x + 6 X_{B}}{3(1+2X_{B})} \,.
+\ee
 The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the gravitational-wave flux (Eqs.~(5) and~(6) of Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived.

### 2018-12-02 paper/paper_no_event.tex
Paper: corrections in the Appendix.

@@
 \subsection{Spin-induced quadrupole-monopole terms}
+\label{subsec:QM}
 
@@
 
-The authors sincerely thank Zoheyr Doctor, Bhooshan Gadre, and Prayush Kumar for reviewing and testing the implementation of \texttt{SEOBNRv4T\_surrogate} in \texttt{LAL}.
+The authors sincerely thank Justin Vines and Tanja Hinderer for deriving the quadrupole-monopole corrections needed in the model.
+The authors are grateful to Zoheyr Doctor, Bhooshan Gadre, and Prayush Kumar for reviewing and testing the implementation of \texttt{SEOBNRv4T\_surrogate} in \texttt{LAL}.
 BL also thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the University of Washington for helpful advice at the start of this work.
@@
 
-The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
+The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform,
 \begin{equation}
@@
 \end{equation}
-where $v=(\pi M f)^{1/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{LAL}~\cite{lal} \red{[explain]}. The phase has the schematic form
+where $v=(\pi M f)^{1/3}$ is the standard PN parameter. The phase has the schematic form
 \begin{align}
@@
 We separated point particle terms $\Phi_{\rm F2}^{\rm PP}(\eta)$, spin terms $\Phi_{\rm F2}^{\rm Spin}(\eta, \chi_{1}, \chi_{2})$ and tidal terms $\Phi_{\rm F2}^{\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$.
-We include point particle terms to the highest know order of 3.5PN~\cite{}.
-
-The spin terms can be decomposed into spin-orbit and spin-spin terms (ignoring cubic-in-spin contributions at 3.5PN~\cite{}) as $\Phi_{\rm F2}^{\rm Spin} = \Phi_{\rm F2}^{\rm SO}  + \Phi_{\rm F2}^{\rm SS} $.
-We include spin-orbit terms up to 3.5PN~\cite{}.
-For spin-square terms, as the \texttt{SEOBNRv4T} waveform does not include yet 3PN spin-spin effects (see Sec.~\ref{}), we stop at  the leading 2PN in the \texttt{TaylorF2} waveform and ignore the known 3PN terms.
+We include point particle terms to the highest know order of 3.5PN (see e.g.~\cite{Blanchet:2013haa}).
+
+The spin terms can be decomposed into spin-orbit and spin-spin terms (ignoring cubic-in-spin contributions at 3.5PN) as $\Phi_{\rm F2}^{\rm Spin} = \Phi_{\rm F2}^{\rm SO}  + \Phi_{\rm F2}^{\rm SS} $.
+We include spin-orbit terms up to 3.5PN~\cite{Blanchet:2013haa}.
+For spin-square terms, as the \texttt{SEOBNRv4T} waveform does not include yet 3PN spin-spin effects (see Sec.~\ref{subsec:QM}), we stop at  the leading 2PN in the \texttt{TaylorF2} waveform and ignore the known 3PN terms.
 The main reason for doing this is because including only common physical effects in both \texttt{SEOBNRv4T} and \texttt{TaylorF2} makes the residual $\Delta\Phi(Mf; \bx)$ smaller and have less variation, making it easier to fit at a given error requirement.
@@
 \begin{align}
-	a_{2} &= \frac{55 \nu }{9}+\frac{3715}{756} \nonumber\\
+	a_{2} &= \frac{55 \eta }{9}+\frac{3715}{756} \nonumber\\
 	a_{3} &= -16 \pi \nonumber\\
-	a_{4} &= \frac{3085 \nu ^2}{72}+\frac{27145 \nu }{504}+\frac{15293365}{508032} \nonumber\\
-	a_{5} &= -\frac{65 \pi  \nu }{9}-\frac{65}{3} \pi  \nu   \ln v +\frac{38645}{252} \pi   \ln v +\frac{38645 \pi }{756} \nonumber\\
-	a_{6} &= -\frac{127825 \nu ^3}{1296}+\frac{76055 \nu ^2}{1728}+\frac{2255 \pi ^2 \nu }{12} \nonumber\\ 
-	& -\frac{15737765635 \nu }{3048192}-\frac{6848  \ln v }{21}-\frac{640 \pi ^2}{3}-\frac{6848 \gamma_{E} }{21} \nonumber\\ 
+	a_{4} &= \frac{3085 \eta ^2}{72}+\frac{27145 \eta }{504}+\frac{15293365}{508032} \nonumber\\
+	a_{5} &= -\frac{65 \pi  \eta }{9}-\frac{65}{3} \pi  \eta   \ln v +\frac{38645}{252} \pi   \ln v +\frac{38645 \pi }{756} \nonumber\\
+	a_{6} &= -\frac{127825 \eta ^3}{1296}+\frac{76055 \eta ^2}{1728}+\frac{2255 \pi ^2 \eta }{12} \nonumber\\ 
+	& -\frac{15737765635 \eta }{3048192}-\frac{6848  \ln v }{21}-\frac{640 \pi ^2}{3}-\frac{6848 \gamma_{E} }{21} \nonumber\\ 
 	& +\frac{11583231236531}{4694215680}-\frac{13696 \log (2)}{21} \nonumber\\
-	a_{7} &= -\frac{74045 \pi  \nu ^2}{756}+\frac{378515 \pi  \nu }{1512}+\frac{77096675 \pi }{254016}
+	a_{7} &= -\frac{74045 \pi  \eta ^2}{756}+\frac{378515 \pi  \eta }{1512}+\frac{77096675 \pi }{254016}
 \end{align}
@@
 \begin{align}
-	b_{3} &= \frac{113 \delta   \chi_a }{3}+\left(\frac{113}{3}-\frac{76 \nu }{3}\right)  \chi_s \nonumber\\
-	b_{5} &=  \chi_s  \left(\frac{340 \nu ^2}{9}+\frac{24260 \nu }{81}+\frac{340}{3} \nu ^2  \ln v +\frac{24260}{27} \nu   \ln v -\frac{732985  \ln v }{756}-\frac{732985}{2268}\right) \nonumber\\
-	& \quad + \chi_a  \left(-\frac{140 \delta  \nu }{9}-\frac{732985 \delta }{2268}-\frac{140}{3} \delta  \nu   \ln v -\frac{732985}{756} \delta   \ln v \right)  \nonumber\\
-	b_{6} &= \frac{2270 \pi  \delta   \chi_a }{3}+\left(\frac{2270 \pi }{3}-520 \pi  \nu \right)  \chi_s \nonumber\\
-	b_{7} &=  \chi_a  \left(-\frac{1985 \delta  \nu ^2}{48}+\frac{26804935 \delta  \nu }{6048}-\frac{25150083775 \delta }{3048192}\right)+\left(\frac{5345 \nu ^3}{36}-\frac{1042165 \nu ^2}{3024}+\frac{10566655595 \nu }{762048}-\frac{25150083775}{3048192}\right)  \chi_s \,,
+	b_{3} &= \frac{113 \delta   \chi_a }{3}+\left(\frac{113}{3}-\frac{76 \eta }{3}\right)  \chi_s \nonumber\\
+	b_{5} &=  \chi_s  \left(\frac{340 \eta ^2}{9}+\frac{24260 \eta }{81}+\frac{340}{3} \eta ^2  \ln v +\frac{24260}{27} \eta   \ln v -\frac{732985  \ln v }{756}-\frac{732985}{2268}\right) \nonumber\\
+	& \quad + \chi_a  \left(-\frac{140 \delta  \eta }{9}-\frac{732985 \delta }{2268}-\frac{140}{3} \delta  \eta   \ln v -\frac{732985}{756} \delta   \ln v \right)  \nonumber\\
+	b_{6} &= \frac{2270 \pi  \delta   \chi_a }{3}+\left(\frac{2270 \pi }{3}-520 \pi  \eta \right)  \chi_s \nonumber\\
+	b_{7} &=  \chi_a  \left(-\frac{1985 \delta  \eta ^2}{48}+\frac{26804935 \delta  \eta }{6048}-\frac{25150083775 \delta }{3048192}\right)+\left(\frac{5345 \eta ^3}{36}-\frac{1042165 \eta ^2}{3024}+\frac{10566655595 \eta }{762048}-\frac{25150083775}{3048192}\right)  \chi_s \,,
 \end{align}
@@
 \begin{align}
-	c_{4} &= \chi_s ^2 \left(-50 \delta   \kappa_a +100  \kappa_s  \nu -50  \kappa_s -\frac{195 \nu }{2}-\frac{5}{8}\right) + \chi_a   \chi_s  \left(-100 \delta   \kappa_s -\frac{5 \delta }{4}+200  \kappa_a  \nu -100  \kappa_a \right) \nonumber\\
-	& + \chi_a ^2 \left(-50 \delta   \kappa_a +100  \kappa_s  \nu -50  \kappa_s +100 \nu -\frac{5}{8}\right) \,.
-\end{align}
-As explained above, to be consistent with the fact that this PN information has not yet been incorporated in \texttt{SEOBNRv4T}, we do not include the next-to-leading 3PN term, which for reference is given by
-\begin{align}
-	c_{6} &=  \chi_s ^2 \left(-\frac{1495 \delta   \kappa_a  \nu }{6}+\frac{26015 \delta   \kappa_a }{28}-240  \kappa_s  \nu ^2-\frac{44255  \kappa_s  \nu }{21}+\frac{26015  \kappa_s }{28}+\frac{3415 \nu ^2}{9}+\frac{829705 \nu }{504}-\frac{1344475}{2016}\right) \nonumber\\
-	& + \chi_a   \chi_s  \left(-\frac{1495 \delta   \kappa_s  \nu }{3}+\frac{26015 \delta   \kappa_s }{14}+\frac{745 \delta  \nu }{18}-\frac{1344475 \delta }{1008}-480  \kappa_a  \nu ^2-\frac{88510  \kappa_a  \nu }{21}+\frac{26015  \kappa_a }{14}\right) \nonumber\\
-	& + \chi_a ^2 \left(-\frac{1495 \delta   \kappa_a  \nu }{6}+\frac{26015 \delta   \kappa_a }{28}-240  \kappa_s  \nu ^2-\frac{44255  \kappa_s  \nu }{21}+\frac{26015  \kappa_s }{28}-240 \nu ^2+\frac{267815 \nu }{252}-\frac{1344475}{2016}\right)\,.
+	c_{4} &= \chi_s ^2 \left(-50 \delta   \kappa_a +100  \kappa_s  \eta -50  \kappa_s -\frac{195 \eta }{2}-\frac{5}{8}\right) + \chi_a   \chi_s  \left(-100 \delta   \kappa_s -\frac{5 \delta }{4}+200  \kappa_a  \eta -100  \kappa_a \right) \nonumber\\
+	& + \chi_a ^2 \left(-50 \delta   \kappa_a +100  \kappa_s  \eta -50  \kappa_s +100 \eta -\frac{5}{8}\right) \,.
+\end{align}
+As explained above, to be consistent with the fact that this PN information has not yet been incorporated in \texttt{SEOBNRv4T}, we do not include the next-to-leading 3PN term, which for reference is given by~\cite{Bohe:2015ana}
+\begin{align}
+	c_{6} &=  \chi_s ^2 \left(-\frac{1495 \delta   \kappa_a  \eta }{6}+\frac{26015 \delta   \kappa_a }{28}-240  \kappa_s  \eta ^2-\frac{44255  \kappa_s  \eta }{21}+\frac{26015  \kappa_s }{28}+\frac{3415 \eta ^2}{9}+\frac{829705 \eta }{504}-\frac{1344475}{2016}\right) \nonumber\\
+	& + \chi_a   \chi_s  \left(-\frac{1495 \delta   \kappa_s  \eta }{3}+\frac{26015 \delta   \kappa_s }{14}+\frac{745 \delta  \eta }{18}-\frac{1344475 \delta }{1008}-480  \kappa_a  \eta ^2-\frac{88510  \kappa_a  \eta }{21}+\frac{26015  \kappa_a }{14}\right) \nonumber\\
+	& + \chi_a ^2 \left(-\frac{1495 \delta   \kappa_a  \eta }{6}+\frac{26015 \delta   \kappa_a }{28}-240  \kappa_s  \eta ^2-\frac{44255  \kappa_s  \eta }{21}+\frac{26015  \kappa_s }{28}-240 \eta ^2+\frac{267815 \eta }{252}-\frac{1344475}{2016}\right)\,.
 \end{align}
@@
 
-Finally, the tidal contributions to the phasing are given by~\cite{VinesFlanaganHinderer2011}
-\begin{equation}
-	\Phi_{\rm Tidal}^{\rm SS}(\eta, \Lambda_{1}, \Lambda_{2}) = d_{10} v^{10} + d_{12} v^{12} + \mathcal{O}(v^{13})
-\end{equation}
-Introducing convenient mass-weighted combinations of the tidal parameters, expressed using $X_{1} = m_{1}/(m_{1} + m_{2})$ and $X_{2} = m_{2}/(m_{1} + m_{2})$, \red{[to be checked]}
-\begin{align}
-	\tilde\Lambda &= \frac{16}{13} \frac{(X_1+12X_2)X_1^4\Lambda_1 + (12X_1+X_2)X_2^4\Lambda_2}{(X_1+X_2)^5} \nonumber\\
-	\delta\tilde\Lambda =& \frac{(1319X_1^2 - 7996X_1X_2 - 11005X_2^2)X_1^4\Lambda_1}{1319(X_1+X_2)^6} \\
-& + \frac{(11005X_1^2 + 7996X_1X_2 - 1319X_2^2)X_2^4\Lambda_2}{1319(X_1+X_2)^6} \,,
-\end{align}
-the coefficients are given by \red{[to be checked]}
+Finally, the tidal contributions to the phasing take the form~\cite{VinesFlanaganHinderer2011}
+\begin{equation}
+	\Phi_{\rm F2}^{\rm Tidal}(\eta, \Lambda_{1}, \Lambda_{2}) = d_{10} v^{10} + d_{12} v^{12} + \mathcal{O}(v^{13}) \,.
+\end{equation}
+Introducing convenient mass-weighted combinations of the tidal parameters, expressed using $X_{1} = m_{1}/(m_{1} + m_{2})$ and $X_{2} = m_{2}/(m_{1} + m_{2})$,
+\begin{align}
+	\tilde\Lambda &= \frac{16}{13} \left[(12 - 11 X_1)X_1^4\Lambda_1 + (12 - 11 X_2)X_2^4\Lambda_2 \right] \nonumber\\
+	\delta\tilde\Lambda =& \frac{1}{1319} \left[(-11005 + 14014 X_1 - 1690 X_{1}^{2})X_1^4\Lambda_1 \right. \nonumber\\ 
+	& \left. \quad\quad + (11005 - 14014 X_2 + 1690 X_{2}^{2}) X_2^4\Lambda_2 \right] \,,
+\end{align}
+the coefficients are given by
 \begin{align}

### 2018-12-02 paper/paper_no_event.tex
Paper: small corrections, in particular change sign of A1

@@
 \end{widetext}
-Here, $M_{A,B}$ are the masses of bodies $A$ and $B$, $M = M_A + M_B$ is the total mass, $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
-Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$\footnote{Here, $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.}.
+Here, $m_{A,B}$ are the masses of bodies $A$ and $B$, $M = m_A + m_B$ is the total mass, $X_{A,B}=m_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities.
+Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$\footnote{Here, $\hat{\omega}_{0\ell}^{A,B}=m_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.}.
 In particular, the dimensionless dynamical tidal polarizability reads
@@
 \end{equation}
-with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$.
+with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/m_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$.
 Here, $\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor~\cite{Steinhoff:2016rfi, Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.
@@
 (A14)-(A17) of Ref.~\cite{Damour:2012yf}.
-Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu} as
-\be
-	\hat{k}_{\ell, \rm{dyn}}^{A, \mathrm{diss}}(u; \omega_{0\ell}) = \hat{k}_{\ell, \rm{dyn}}^{A} \frac{4 x + 6 X_{B}}{3(1+2X_{B})} \,.
-\ee
+Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
+%\be
+%	\hat{k}_{\ell, \rm{dyn}}^{A, \mathrm{diss}}(u; \omega_{0\ell}) = \hat{k}_{\ell, \rm{dyn}}^{A} \frac{4 x + 6 X_{B}}{3(1+2X_{B})} \,.
+%\ee
+
 The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the gravitational-wave flux (Eqs.~(5) and~(6) of Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived.
@@
 
-After rescaling with the total mass $M$, the aligned-spin EOB waveform with dynamical tides depends on 13 intrinsic parameters: the mass ratio $q=M_B/M_A\leq1$, the two dimensionless spin components along the orbital angular momentum $\chi_{A,B}$, the two spin-induced quadrupole-monopole parameters $\kappa_{A,B}$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
+After rescaling with the total mass $M$, the aligned-spin EOB waveform with dynamical tides depends on 13 intrinsic parameters: the mass ratio $q=m_B/m_A\leq1$, the two dimensionless spin components along the orbital angular momentum $\chi_{A,B}$, the two spin-induced quadrupole-monopole parameters $\kappa_{A,B}$, the two adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the two adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, and the four $\ell=2,3$ fundamental $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$.
 To reduce the dimensionality of the intrinsic parameter space, we use nearly EOS-independent fits (universal relations) for these parameters in terms of $\Lambda_2$ as discussed below.
@@
 \begin{equation}
-A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} v^{-7/2} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)v^{2} \right],
-\end{equation}
-where $v=(\pi M f)^{1/3}$ is the standard PN parameter. The phase has the schematic form
+A_{\rm F2} = -\sqrt{\frac{5\pi\eta}{24}} v^{-7/2} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)v^{2} \right],
+\end{equation}
+where $v=(\pi M f)^{1/3}$ is the standard PN parameter. The sign is set following the conventions of \texttt{LAL}~\cite{lal}. The phase has the schematic form
 \begin{align}
@@
 We separated point particle terms $\Phi_{\rm F2}^{\rm PP}(\eta)$, spin terms $\Phi_{\rm F2}^{\rm Spin}(\eta, \chi_{1}, \chi_{2})$ and tidal terms $\Phi_{\rm F2}^{\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$.
-We include point particle terms to the highest know order of 3.5PN (see e.g.~\cite{Blanchet:2013haa}).
+We include point particle terms to the highest know order of 3.5PN (see e.g.~\cite{Blanchet2014Review}).
 
 The spin terms can be decomposed into spin-orbit and spin-spin terms (ignoring cubic-in-spin contributions at 3.5PN) as $\Phi_{\rm F2}^{\rm Spin} = \Phi_{\rm F2}^{\rm SO}  + \Phi_{\rm F2}^{\rm SS} $.
-We include spin-orbit terms up to 3.5PN~\cite{Blanchet:2013haa}.
+We include spin-orbit terms up to 3.5PN~\cite{Blanchet2014Review}.
 For spin-square terms, as the \texttt{SEOBNRv4T} waveform does not include yet 3PN spin-spin effects (see Sec.~\ref{subsec:QM}), we stop at  the leading 2PN in the \texttt{TaylorF2} waveform and ignore the known 3PN terms.

### 2018-12-02 paper/paper_no_event.tex
Paper: edits in Sec V

@@
 For current parameter estimation codes, this is sufficient.
-However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, or multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017}.
+However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017} or relative binning~\cite{Zackay:2018qdy}.
 Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^3)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.
@@
 Using GPR and uncertainty sampling discussed above, one could optimally choose the waveform parameters for the numerical simulations, and run them in parallel to build the training set.
-Because the difference between \texttt{SEOBNRv4T} and numerical BNS simulations is exceptionally small, one would not need a high fractional accuracy for a surrogate of the difference, and 10--100 waveforms may be sufficient.
+If the difference between \texttt{SEOBNRv4T} and numerical BNS simulations is small, one would not need a high fractional accuracy for a surrogate of the difference, and 10--100 waveforms may be sufficient.
 
 This model notably does not include precession for non-aligned spins.
-Although none of the EOB models currently available include both tidal effects and precession, there are ways to boost this waveform model to a precessing frame and approximately incorporate precession.
-For example, Chatziioanno {\it et al}.\ have analytically solved the 2PN-accurate precession equations for generic spins~\cite{ChatziioannouKleinCornish2017a, ChatziioannouKleinCornish2017b}, with the exception of transitional precession which is unlikely for low mass-ratio and spin BNS systems.
-Using stationary uniform asymptotics, they can also analytically Fourier-transform the solution.
+Although none of the EOB models currently available include both tidal effects and precession, there are ways to rotate this waveform model to a precessing frame and approximately incorporate precession.
+For example, Chatziioannou {\it et al}.\ have analytically solved the 2PN-accurate precession equations for generic spins~\cite{ChatziioannouKleinCornish2017a, ChatziioannouKleinCornish2017b}, with the exception of transitional precession which is unlikely for low mass-ratio and spin BNS systems.
+Using shifted uniform asymptotics~\cite{Klein:2014bua}, they can also analytically Fourier-transform the solution.
 Importantly, one is free to specify the frequency-domain amplitude and phase evolution, such as the \texttt{SEOBNRv4T\_surrogate} here, for the waveform in the co-precessing frame.

### 2018-12-02 paper/paper_no_event.tex
Paper: 2nd paragraph of intro + small edits.

@@
 To avoid large systematic errors, highly accurate waveform models are needed that include all relevant physical effects such as point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
-Most time-domain waveform implementations involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain.
-Depending on the model, this can take seconds to hours for long BNS waveforms, so analytic frequency-domain approximations are often used instead.
+Most time-domain waveform implementations involve solving a set of ordinary differential equations, which depending on the model can take seconds to hours for long BNS waveforms.
+The results then need to be transformed into the frequency domain where the analysis takes place.
+Because of this, analytic approximations providing waveforms directly in the frequency-domain  are often used instead.
 
@@
 These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein).
-They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT).
+They use either an aligned-spin post-Newtonian (PN) approximant in the Fourier domain (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT).
 Tidal interactions were then included by simply adding a correction to the phase of each waveform.
@@
 It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations.
-We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a different point-particle and spin prescription.
+We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018, TEOBResumSupdate}, that uses a different point-particle and spin prescription.
 For matter effects, this model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions.
@@
 
-\red{[check notations have been introduced]} For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are based on the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins.
+For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are based on the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins.
 It includes consistently spin-orbit interactions up to 3.5PN order and spin-spin interactions up to 2PN order.
@@
 This effect is quadratic in the spins, appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin~\cite{HarryHinderer2018}.
-Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only, see~\cite{}.}, at the moment the effect is only consistently included at the leading 2PN order.
-We leave for future work an extension to higher orders\footnote{\red{complete this} We note that TaylorF2 models do include all spin-square terms at 3PN, see App.~\ref{sec:taylorf2}.
-They are also present in the IMRPhenomPv2+NRTides model, but not in the SEOBNRv4ROM+NRTides model.
-TEOBResumS is also limited to the leading 2PN order.}.
+Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only, see~\cite{Levi:2015ixa}.}, at the moment the effect is only consistently included at the leading 2PN order.
+An extension to higher orders is left for future work\footnote{We note that TaylorF2 models do include all known spin-square terms at 3PN, see App.~\ref{sec:taylorf2}.
+They are also present in the IMRPhenomPv2+NRTides and SEOBNRv4ROM+NRTides models.
+TEOBResumS was recently updated~\cite{TEOBResumSupdate} to include these next-to-leading terms.}.
 
@@
 \end{align}
-In physical units, for an inclination angle $\iota$, the $+$ and $\times$ polarizations of the waveform are 
+In physical units, for an inclination angle $\iota$, the $+$ and $\times$ polarizations of the waveform are (for $f>0$)
 \begin{align}
@@
 This causes the peaks of the other 1-d distributions to be much closer to the injected values.
-We consider the difference between these peaks and the injected values to be a conservative bound on the size of the statistical errors due to errors in the surrogate model.
+We consider the difference between these peaks and the injected values to be a conservative bound on the size of the systematic errors due to errors in the surrogate model.
 

### 2018-12-02 paper/paper_no_event.tex
Paper: changes in II C
- renamed to 'waveform termination'
- effective post-merger rather than post-merger model
- changed the 20M entering the definition of tfreq to 12M

@@
 
-\subsection{Post-merger waveform}
-
-We now look for a suitable time $t_{\rm match}$ to stop the numerical integration and match with a post-merger waveform.
+\subsection{Waveform termination}
+
+We now look for a suitable time $t_{\rm match}$ to stop the numerical integration and match with an effective post-merger waveform.
 In the point-mass model, \texttt{SEOBNRv4}, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks, and $t_{\rm match}$ is chosen to be the time of the peak amplitude $t_{\rm peak}^{\rm amp}$.
-Here, we do not use nonquasicircular corrections tuned to BNS simulations, so we have less control over the behavior of Eq.~(\ref{hlminspplunge}) in the late inspiral.
+Here, although we include nonquasicircular corrections, they are computed as in the BBH case and are not tuned to BNS simulations, so we have less control over the behavior of Eq.~(\ref{hlminspplunge}) in the late inspiral.
 We therefore choose the following definition for $t_{\rm match}$.
@@
 
-In BNS simulations, the waveform rapidly tapers to zero amplitude in 1--2 gravitational-wave cycles after reaching peak amplitude.
+In BNS simulations, the waveform rapidly decreases in amplitude in 1--2 gravitational-wave cycles after reaching peak amplitude.
 The BNS system then either undergoes prompt collapse or post-merger oscillations of the remnant.
-We model this tapering in terms of analytic functions for the amplitude $A^{\textrm{post-mrg}}(t)$ and phase $\phi^{\textrm{post-mrg}}(t)$ such that $h_{22}^{\textrm{post-mrg}}(t) = A^{\textrm{post-mrg}}(t) \exp{[i\phi^{\textrm{post-mrg}}(t)]}$.
+We do not attempt to model post-merger oscillations, and instead model an approximate peak emission and a subsequenct tapering to zero.
+This signal will be represented in terms of analytic functions for the amplitude $A^{\textrm{post-mrg}}(t)$ and phase $\phi^{\textrm{post-mrg}}(t)$ such that $h_{22}^{\textrm{post-mrg}}(t) = A^{\textrm{post-mrg}}(t) \exp{[i\phi^{\textrm{post-mrg}}(t)]}$.
 
@@
 
-For the phase, we smoothly extend the waveform frequency such that it agrees with the inspiral frequency at a time $t_{\rm freq} = t_{\rm match} - 20M$ before the matching time $t_{\rm match}$, but then stretches out the frequency evolution such that it only approaches $\omega_{\rm match}$ asymptotically.
+For the phase, we smoothly extend the waveform frequency such that it agrees with the inspiral frequency at a time $t_{\rm freq} = t_{\rm match} - 12M$ before the matching time $t_{\rm match}$, but then stretches out the frequency evolution such that it only approaches $\omega_{\rm match}$ asymptotically.
 We define this frequency evolution as
 \begin{align}
-\omega(t) = \omega_{\rm match} - \Delta\omega \exp\left[ -(t - t_{\rm freq}) / (20M) \right],
+\omega(t) = \omega_{\rm match} - \Delta\omega \exp\left[ -(t - t_{\rm freq}) / (12M) \right],
 \end{align}
@@
 \phi_{22}^{\textrm{insp-plunge}}(t), & t \leq t_{\rm freq} \,, \\
-\phi_{\rm freq} + \omega_{\rm match}(t - t_{\rm freq}) + 20M \Delta\omega \left\{ \right. & \\
-  \left. \exp\left[ -(t - t_{\rm freq}) / (20M) \right] - 1\right\} \,,& t > t_{\rm freq} \, .
+\phi_{\rm freq} + \omega_{\rm match}(t - t_{\rm freq}) + 12M \Delta\omega \left\{ \right. & \\
+  \left. \exp\left[ -(t - t_{\rm freq}) / (12M) \right] - 1\right\} \,,& t > t_{\rm freq} \, .
 \end{array}\right.
@@
 
-Although this post-merger model has not been fit to NR BNS simulations, it is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}.
+Although this effective post-merger model has not been fit to NR BNS simulations, it is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}.
 A more sophisticated, NR-informed model of the post-merger emission will be part of future investigations.

### 2018-12-02 paper/paper_no_event.tex
Paper: changes in the abstract

@@
 Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817.
-We present a completed version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}.
-This model includes the quadrupolar and octopolar adiabatic and dynamical tides, and we improve it by including the spin-induced quadrupole moment effect and a prescription for tapering the end of the waveform to agree with numerical simulations.
-The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters.
-Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require $\mathcal{O}(10^7)$ sequential waveform evaluations.
-We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression.
+We present a frequency-domain surrogate version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}.
+This model includes the quadrupolar and octopolar adiabatic and dynamical tides.
+The version presented here is improved by the inclusion of the spin-induced quadrupole moment effect, and completed by a prescription for tapering the end of the waveform to reproduce qualitatively numerical simulations.
+The resulting model has 14 intrinsic parameters, so we reduce its dimensionality by using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters.
+The previous implementation of this model could take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require $\mathcal{O}(10^7)$ sequential waveform evaluations.
+We therefore construct a fast and faithful frequency-domain surrogate of this model using Gaussian process regression.
 The resulting surrogate has a maximum mismatch of $4.5\times 10^{-4}$ for the Advanced LIGO detector, and requires $\sim 0.13$~s to evaluate for a waveform with a starting frequency of 20~Hz.

### 2018-12-19 paper/paper_no_event.tex
Paper: fix structure of PP terms in appendix.
Made Michael's suggested changes to paper.

@@
 This model includes the quadrupolar and octopolar adiabatic and dynamical tides.
-The version presented here is improved by the inclusion of the spin-induced quadrupole moment effect, and completed by a prescription for tapering the end of the waveform to reproduce qualitatively numerical simulations.
-The resulting model has 14 intrinsic parameters, so we reduce its dimensionality by using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters.
-The previous implementation of this model could take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require $\mathcal{O}(10^7)$ sequential waveform evaluations.
+The version presented here is improved by the inclusion of the spin-induced quadrupole moment effect, and completed by a prescription for tapering the end of the waveform to qualitatively reproduce numerical relativity simulations.
+The resulting model has 14 intrinsic parameters.
+We reduce its dimensionality by using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters.
+The implementation of the time-domain model can take up to an hour to evaluate using a starting frequency of 20~Hz, and this is too slow for many parameter estimation codes that require $\mathcal{O}(10^7)$ sequential waveform evaluations.
 We therefore construct a fast and faithful frequency-domain surrogate of this model using Gaussian process regression.
 The resulting surrogate has a maximum mismatch of $4.5\times 10^{-4}$ for the Advanced LIGO detector, and requires $\sim 0.13$~s to evaluate for a waveform with a starting frequency of 20~Hz.
+Finally, we perform an end-to-end test of the surrogate with a set of parameter estimation runs, and find that the surrogate accurately recovers the parameters of injected waveforms.
 
@@
 The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can measure BNS properties such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}.
-Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further.
-These multiple events could then be stacked to significantly improve measurements of the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.
-
-The reliability of these measurements, however, depend on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data.
+Furthermore, an estimate of the BNS merger rate of 110--3840 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{O2Catalog} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further.
+These multiple events could then be ``stacked'' to significantly improve measurements of the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.
+
+The reliability of these measurements, however, depends on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data.
 Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations.
@@
 
-For the analysis of GW170817 in Ref.~\cite{BNSPE}, therefore, a set of four frequency-domain waveform models were used as templates for the main results.
+For the analysis of GW170817 in Refs.~\cite{BNSPE, O2Catalog}, therefore, a set of four frequency-domain waveform models were used as templates for the main results.
 These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein).
@@
 
-Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible.
-In Ref.~\cite{BNSPE}, an alternative parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}.
-Unfortunately, direct comparisons of the PDF between this code and traditional samplers using the same waveform models have not been published for BNS systems.
+Although these EOB models are more expensive than frequency-domain models, parameter estimation results are still possible.
+In Refs.~\cite{BNSPE, O2Catalog}, an alternative parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}.
+%Unfortunately, direct comparisons of the PDF between this code and traditional samplers using the same waveform models have not been published for BNS systems.
 It is therefore not clear how accurate this code is for measuring quantities such as the tidal parameters.
+The measured tidal parameter using EOB waveforms and \texttt{RapidPE} were broadly consistent with the results using frequency-domain waveforms and traditional samplers (Fig.
+9 of Ref.~\cite{O2Catalog}).
+However, a direct comparison using identical waveforms has not yet been published for BNS systems.
 As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved.
@@
 Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation.
-It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
+It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas}, multi-banding~\cite{VinciguerraVeitchMandel2017}, and relative binning~\cite{Zackay:2018qdy}.
 
@@
 The black curve represents the estimated error using the Mat\'{e}rn kernel, and the gray curve represents the estimated error using the squared exponential kernel.
+400 samples were added to the initial 160 samples for a total of 560 samples.
 Also shown are box plots of the maximum RMS phase errors at the nodes $MF_j \le 0.03$ for the surrogate compared to the test set of 1000 waveforms (Sec.~\ref{sec:accuracy}).
@@
 As was done in~\cite{BNSPE}, we sample the data and waveform at 4096~Hz such that the Nyquist frequency is 2048~Hz, and integrate the likelihood function in the interval $[20, 2048]$~Hz.
-We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal} which requires approximately $10^7$ iterations to produce 10,000--20,000 statistically independent samples.
+We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal} which requires $\sim 10^7$ iterations to produce 10,000--20,000 statistically independent samples.
 
@@
 
-The point particle terms coefficients up to 3.5PN can be found e.g. in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}. The non-zero coefficients read
-\begin{align}
+The point particle terms coefficients up to 3.5PN can be found e.g. in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, and have the structure
+\be
+	\Phi_{\rm F2}^{\rm PP}(\eta) = \sum_{k=0}^{7} a_{k} v^{k} + \mathcal{O}(v^{8}) \,.
+\ee
+The non-zero coefficients read
+\begin{align}
+	a_{0} &= 1 \nonumber\\
 	a_{2} &= \frac{55 \eta }{9}+\frac{3715}{756} \nonumber\\
@@
 	& +\frac{11583231236531}{4694215680}-\frac{13696 \log (2)}{21} \nonumber\\
-	a_{7} &= -\frac{74045 \pi  \eta ^2}{756}+\frac{378515 \pi  \eta }{1512}+\frac{77096675 \pi }{254016}
-\end{align}
+	a_{7} &= -\frac{74045 \pi  \eta ^2}{756}+\frac{378515 \pi  \eta }{1512}+\frac{77096675 \pi }{254016} \,,
+\end{align}
+where $\gamma_{E}$ is Euler's constant.
 

### 2018-12-20 paper/paper_no_event.tex
Paper: fix nu->eta in Eq. 14
Minor corrections to text. Updated references.

@@
 These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein).
-They use either an aligned-spin post-Newtonian (PN) approximant in the Fourier domain (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT).
+They use either an aligned-spin post-Newtonian (PN) approximant in the Fourier domain (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT), or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT).
 Tidal interactions were then included by simply adding a correction to the phase of each waveform.
@@
 It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations.
-We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018, TEOBResumSupdate}, that uses a different point-particle and spin prescription.
+We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018, NagarMessinaRettegno2018}, that uses a different point-particle and spin prescription.
 For matter effects, this model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions.
@@
 It is therefore not clear how accurate this code is for measuring quantities such as the tidal parameters.
-The measured tidal parameter using EOB waveforms and \texttt{RapidPE} were broadly consistent with the results using frequency-domain waveforms and traditional samplers (Fig.
+The measured tidal parameter using EOB waveforms and \texttt{RapidPE} were broadly consistent with the results using frequency-domain waveforms and traditional samplers (see Fig.
 9 of Ref.~\cite{O2Catalog}).
@@
 This effect is quadratic in the spins, appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin~\cite{HarryHinderer2018}.
-Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only, see~\cite{Levi:2015ixa}.}, at the moment the effect is only consistently included at the leading 2PN order.
-An extension to higher orders is left for future work\footnote{We note that TaylorF2 models do include all known spin-square terms at 3PN, see App.~\ref{sec:taylorf2}.
-They are also present in the IMRPhenomPv2+NRTides and SEOBNRv4ROM+NRTides models.
-TEOBResumS was recently updated~\cite{TEOBResumSupdate} to include these next-to-leading terms.}.
+Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only~\cite{Levi:2015ixa}.}, at the moment the effect is only consistently included at the leading 2PN order.
+An extension to higher orders is left for future work\footnote{We note that TaylorF2 models do include all known spin-square terms at 3PN (see Appendix~\ref{sec:taylorf2}).
+They are also present in the PhenomPNRT and SEOBNRT models.
+TEOBResumS was recently updated to include these next-to-leading terms~\cite{NagarMessinaRettegno2018}.}.
 
@@
 \be
-h_{22}^\tr{LO-SS}=-\frac{8\nu M^3\omega^2}{R}\sqrt{\frac{\pi}{5}}e^{-2i\Phi}\mc A \,,
+h_{22}^\tr{LO-SS}=-\frac{8\eta M^3\omega^2}{R}\sqrt{\frac{\pi}{5}}e^{-2i\Phi}\mc A \,,
 \ee
@@
 When translating the above for the EOB factorized waveform (see e.g.
-(17) in~\cite{Taracchini:2012ig}), spin contributions in the effective source, tail factors and phase contribution would enter only at higher order, so that the only modification to consider is in the $\rho_{\ell m}$ factor for $\ell = 2$, $m = 2$.
+(17) in~\cite{Taracchini:2012ig}), spin contributions in the effective source, tail factors, and phase contribution would enter only at higher order, so that the only modification to consider is in the $\rho_{\ell m}$ factor for $\ell = 2$, $m = 2$.
 We have simply
@@
 For current parameter estimation codes, this is sufficient.
-However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017} or relative binning~\cite{Zackay:2018qdy}.
+However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017}, or relative binning~\cite{Zackay:2018qdy}.
 Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^3)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.
@@
 The hierarchical method presented here, where we begin with the analytic TaylorF2 reference model then make a surrogate of the residual, can be used to further improve the waveform model.
-For example, one could build a surrogate for numerical BNS simulations using \texttt{SEOBNRv4T\_surrogate} as the base model, and constructing a surrogate of the residual.
+For example, one could build a surrogate for numerical BNS simulations using \texttt{SEOBNRv4T\_surrogate} as the base model, then constructing a surrogate of the residual.
 Such a model would have the accuracy of EOB below $\sim 400$~Hz and the accuracy of numerical simulations for the last several cycles before merger.

### 2018-12-20 paper/paper_no_event.tex
Moved timing figures to paper directory. Minor corrections to text.

@@
 \newcommand{\doe}{\partial}
-%\newcommand{\be}{\begin{equation}}
-%\newcommand{\ee}{\end{equation}}
 \newcommand{\bea}{\begin{eqnarray}}
@@
 
-%\def\l{\ell}
-%\def\blambda{{\bm \lambda}}
-%\def\bLambda{{\bm \Lambda}}
-%\def\blam{\bar{\lambda}}
-%\def\Lam{\Lambda}
-%\def\Lamt{\tilde{\Lambda}}
-%\def\dLamt{\tilde{\delta{\Lambda}}}
-%\def\bmu{{\bm \mu}}
-
 \def\bx{\mathbf{x}}
@@
 \newcommand{\MP}[1]{{\textcolor{blue}{\texttt{MP: #1}} }}
-%\newcommand{\bs}[1]{{\textcolor{green}{\texttt{SB: #1}} }}
 \newcommand{\bl}[1]{{\textcolor{blue}{\texttt{BL: #1}} }}
-\newcommand{\cg}[1]{{\textcolor{orange}{{CG: #1}} }}
-\newcommand{\cvdb}[1]{{\textcolor{purple}{\texttt{CVDB: #1}} }}
 \newcommand{\red}[1]{\textcolor{red}{#1}}
@@
 In Refs.~\cite{BNSPE, O2Catalog}, an alternative parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}.
-%Unfortunately, direct comparisons of the PDF between this code and traditional samplers using the same waveform models have not been published for BNS systems.
-It is therefore not clear how accurate this code is for measuring quantities such as the tidal parameters.
 The measured tidal parameter using EOB waveforms and \texttt{RapidPE} were broadly consistent with the results using frequency-domain waveforms and traditional samplers (see Fig.
@@
 Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
-%\be
-%	\hat{k}_{\ell, \rm{dyn}}^{A, \mathrm{diss}}(u; \omega_{0\ell}) = \hat{k}_{\ell, \rm{dyn}}^{A} \frac{4 x + 6 X_{B}}{3(1+2X_{B})} \,.
-%\ee
 
@@
 We only include the dominant $(\ell,m)=(2,2)$ mode in constructing the final waveform, although we use all modes when evaluating the radiation reaction force.
-
-%Let $\Omega$ be the instantaneous orbital frequency, $r$ be orbital separation, $p_r$ be the radial momentum, $f_s$ be the waveform sampling frequency in Hz, $\omega_{\textrm{peak}}^{\textrm{NR}}$ be the NR fit for the (2,2)-mode amplitude peak of a BNS merger defined in Eq.~(2) of Ref.~\cite{Bernuzzi:2015rla}~\footnote{Note that this fit was tuned to NR simulations only for $0 \leq \kappa_2^T \leq 500$, where $\kappa_2^T$ is defined by Eq.~(1) of Ref.~\cite{Bernuzzi:2015rla}.
-In our model, for $\kappa_2^T >500$ we use the value of this fit at $\kappa_2^T=500$.}.
-Then the numerical integration of the orbital dynamics stops when any of the following events happens: (i) $\Omega$ peaks, (ii) $p_r$ becomes positive, (iii) $\dot{r}$ becomes positive, (iv) $\dot{p}_r$ becomes positive, (v) $r$ does not evolve more than a 1000th of the leading-order estimate for $\dot{r}$, that is $\propto 64/5 \nu/r^4$, (vi) $2\Omega$ reaches $\omega_{\textrm{peak}}^{\textrm{NR}}$, (vii) $2M\Omega$ reaches $\pi M f_s$.
-In order to avoid triggering these stopping conditions at low frequency, where residual orbital eccentricity due to imperfect quasicircular initial conditions could be present, we exploit the $\omega_{\textrm{peak}}^{\textrm{NR}}$ fit to define a radius below which the stopping conditions can be triggered; such threshold value is taken to be $1.5 (\omega_{\textrm{peak}}^{\textrm{NR}}/2)^{-2/3}$.
 
@@
 	& \; \cdot \bigg[\frac{\mu}{M}S_*^iS_*^j+(\kappa_1-1)\frac{m_2}{m_1}S_1^i S_1^j+(\kappa_2-1)\frac{m_1}{m_2}S_2^i S_2^j\bigg] \label{eq:HBNSextra}
-%\\\nnm
-%&=\frac{1}{2r^3}(3n_in_j-\delta_{ij})\bigg[2\frac{\mu}{M}S_1^iS_2^j+\left(\kappa_1-1+\frac{m_2^2}{M^2}\right)\frac{m_2}{m_1}S_1^i S_1^j+\left(\kappa_2-1+\frac{m_1^2}{M^2}\right)\frac{m_1}{m_2}S_2^i S_2^j\bigg],
 \end{align}
@@
 
-%In the BBH model, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks at a prescribed time and has prescribed amplitude and frequency, hence $t_{\textrm{match}}^{22}=t_{\textrm{peak}}^{\textrm{amp}}$.
-Here, however, we do not impose such corrections because of limited input from NR BNS simulations, therefore we have little control over the behavior of Eq.~(\ref{hlminspplunge}) in the late inspiral.
-This motivates our choice of $t_{\textrm{match}}^{22}$ in Eq.~(\ref{tmatch}), with the goal of reducing to the BBH prescription in the non-tidal limit and avoiding non-monotonic growth of the GW frequency in the late inspiral.
-To help bracket the stopping time, we exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics.
-
-%Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state.
-The model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state.
-\red{[[To what accuracy?
-Is this paragraph already in the intro and can be deleted?]]}
-
 In BNS simulations, the waveform rapidly decreases in amplitude in 1--2 gravitational-wave cycles after reaching peak amplitude.
@@
 With these relations the waveform only depends on the five intrinsic parameters $\bx=\{q, \chi_A, \chi_B, \Lambda_2^A,\Lambda_2^B\}$. 
-%The aligned component spins are defined as $\chi_i = \hat L_N \cdot \vec S_i / m_i^2$ where $\vec S_i$ are the dimensionfull spin vectors, $\hat L_N$ the Newtonian orbital angular momentum and $m_i$ the component masses.
 We note that this list does not include the total mass $M$ of the system.
@@
 With these optimized hyperparameters, the final interpolating function is given by Eq.~\eqref{eq:mean} and its uncertainty by Eq.~\eqref{eq:var}.
-%We note that once the vector $(K^{-1})_{ij}y_j$ is precomputed, evaluating Eq.~\eqref{eq:mean} is a fast $\mathcal{O}(N)$ operation.
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{../src/TEOB-LAL-implementation/timings/Models_timings_FD_monotonic.pdf}
-\includegraphics[width=0.49\textwidth]{../src/TEOB-LAL-implementation/timings/SEOBNRv4T_surrogate_speedup_deltaF_monontonic.pdf}
+\includegraphics[width=0.49\textwidth]{Models_timings_FD_monotonic.pdf}
+\includegraphics[width=0.49\textwidth]{SEOBNRv4T_surrogate_speedup_deltaF_monontonic.pdf}
 \caption{Waveform evaluation time as a function of the waveform starting frequency $f_{\rm min}$.
@@
 The parameters are identical except for the distances which are chosen such that the network SNRs are 30 (blue), 60 (green), or 120 (red).
-The black PDF only includes samples for the SNR=120 injection that satisfy $q>0.9$.
+The black PDFs only include samples for the SNR=120 injection that satisfy $q>0.9$.
 The shaded regions give the 90\% credible intervals.
@@
 \end{document}
-
-%\begin{widetext}
-%\begin{align}
-%\Lambda_{2,\textrm{dyn}}^{\textrm{(A,B)}}(u) = &\frac{\lambda_{2}^{\textrm{(1,2)}}}{M^5} \left\{\frac{1}{4} + \frac{3}{4} \left(\frac{\omega^{(1,2)}_{f,2}}{2\Omega}\right)^2\left[\frac{4\Omega^2}{(\omega^{(1,2)}_{f,2})^2 - 4\Omega^2} + \frac{10}{3Q}\right.\right.
-\nonumber\\
-%&\left.\left.+\sqrt{\frac{\pi}{3}}\frac{1}{\epsilon_{1,2}}\left[(1 + 2\mathcal{S}{(y_{1,2})})\cos{\left(\frac{\pi}{2}y^2_{1,2}\right)} - (1 + 2\mathcal{C}{(y_{1,2})})\sin{\left(\frac{\pi}{2}y_{1,2}^2\right)}\right]\right]\right\}\,,\\
-%\Lambda_{3,\textrm{eff}}^{\textrm{(1,2)}} (u)= &\frac{\lambda_{3}^{\textrm{(1,2)}}}{M^7} \left\{\frac{3}{8} + \frac{(\omega^{(1,2)}_{f,3})^2}{\Omega^2}\left(\frac{25}{48 O_{1,2}} + \frac{5}{72}\frac{9\Omega^2}{(\omega^{(1,2)}_{f,3})^2-9\Omega^2}\right)\right.
-\nonumber\\
-%&\left.+ O_{1,2}\left[\cos{(w^2)}\left(\frac{1}{2} +\mathcal{S}(sqrt2pi*w)\right) - \sin{(w^2)}\left(\frac{1}{2} + \mathcal{C}(sqrt2pi*w)\right)\right]\right\}\,,
-%\end{align}
-%\end{widetext}
-%where $\lambda_{2}^{\textrm{(1,2)}}$ are the (constant) adiabatic values of the quadrupolar tidal polarizabilities, $\lambda_{3}^{\textrm{(1,2)}}$ are the (constant) adiabatic values of the octupolar tidal polarizabilities, $\omega^{(1,2)}_{f,2}$ are the quadrupolar $f$-mode frequencies, $\omega^{(1,2)}_{f,3}$ are the octupolar $f$-mode frequencies, $\mathcal{S}$ is the Fresnel sine integral, $\mathcal{C}$ is the Fresnel cosine integral, $\epsilon_{1,2} = 2^{19/3}\nu (M\omega^{(1,2)}_{f,2})^{5/3}/5$,   $y_{1,2}=\sqrt{3/(\pi \epsilon)} Q_{1,2}/5$, and $Q_{1,2} = 4- 2^{1/3}(M\omega^{(1,2)}_{f,2})^{5/3}u^{-5/2}$.
-%, $w_{1,2}=O_{1,2}/(4\,3^{2/3}\sqrt{10}(M\omega^{(1,2)}_{f,3})^{5/6}\sqrt{\nu})$, $O_{1,2}=5\sqrt{5\pi}/u^3(M\omega^{(1,2)}_{f,3})^{7/6}/(192\,3^{2/3}\sqrt{\nu})$.
-%

### 2018-12-20 paper/paper.tex
Renamed paper file to paper.tex. Final version submitted to arXiv.

@@
+% tar -cvzf paper.tgz paper.tex paper.bbl hoff.pdf dhofs.pdf trainingset3d.pdf trainingset2d.pdf uncertaintysampling.pdf mismatch.pdf htildemaxerror.pdf hmaxerror.pdf Models_timings_FD_monotonic.pdf SEOBNRv4T_surrogate_speedup_deltaF_monontonic.pdf chirpmass.pdf q.pdf chieff.pdf tildelambda.pdf chieff_q.pdf
+
 \documentclass[prd,aps,letter,twocolumn,floatfix,notitlepage,nofootinbib]{revtex4-1}
@@
 Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817.
-We present a completed version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}.
-This model includes the quadrupolar and octopolar adiabatic and dynamical tides, and we improve it by including the spin-induced quadrupole moment effect and a prescription for tapering the end of the waveform to agree with numerical simulations.
-The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters.
-%This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters. 
-Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations.
-We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression.
-The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz.
-With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).
+We present a frequency-domain surrogate version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}.
+This model includes the quadrupolar and octopolar adiabatic and dynamical tides.
+The version presented here is improved by the inclusion of the spin-induced quadrupole moment effect, and completed by a prescription for tapering the end of the waveform to qualitatively reproduce numerical relativity simulations.
+The resulting model has 14 intrinsic parameters.
+We reduce its dimensionality by using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters.
+The implementation of the time-domain model can take up to an hour to evaluate using a starting frequency of 20~Hz, and this is too slow for many parameter estimation codes that require $\mathcal{O}(10^7)$ sequential waveform evaluations.
+We therefore construct a fast and faithful frequency-domain surrogate of this model using Gaussian process regression.
+The resulting surrogate has a maximum mismatch of $4.5\times 10^{-4}$ for the Advanced LIGO detector, and requires $\sim 0.13$~s to evaluate for a waveform with a starting frequency of 20~Hz.
+Finally, we perform an end-to-end test of the surrogate with a set of parameter estimation runs, and find that the surrogate accurately recovers the parameters of injected waveforms.
 
@@
 
-The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can accurately measure quantities such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}.
-Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further.
-These multiple events could then be stacked to significantly improve measurements of NS radii and the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.
-
-The reliability of these measurements, however, depend on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data.
+The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can measure BNS properties such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}.
+Furthermore, an estimate of the BNS merger rate of 110--3840 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{O2Catalog} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further.
+These multiple events could then be ``stacked'' to significantly improve measurements of the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.
+
+The reliability of these measurements, however, depends on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data.
 Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations.
@@
 To avoid large systematic errors, highly accurate waveform models are needed that include all relevant physical effects such as point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}.
-Most time-domain waveform implementations involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain.
-Depending on the model, this can take seconds to hours for long BNS waveforms, so analytic frequency-domain approximations are often used instead.
-
-For the analysis of GW170817 in Ref.~\cite{BNSPE}, therefore, a set of four frequency-domain waveform models were used as templates for the main results.
+Most time-domain waveform implementations involve solving a set of ordinary differential equations, which depending on the model can take seconds to hours for long BNS waveforms.
+The results then need to be transformed into the frequency domain where the analysis takes place.
+Because of this, analytic approximations providing waveforms directly in the frequency-domain  are often used instead.
+
+For the analysis of GW170817 in Refs.~\cite{BNSPE, O2Catalog}, therefore, a set of four frequency-domain waveform models were used as templates for the main results.
 These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein).
-They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT).
+They use either an aligned-spin post-Newtonian (PN) approximant in the Fourier domain (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT), or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT).
 Tidal interactions were then included by simply adding a correction to the phase of each waveform.
 For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added.
-For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.
-
-In this paper, we will use an implementation of the effective one body (EOB) formalism that treats matter effects consistently with the other parameters.
+For the other three models, a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.
+
+In this paper, we will use an implementation of the EOB formalism that treats matter effects consistently with the other parameters.
 It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations.
@@
 It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations.
-We also note an alternative model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a slightly different point-particle and spin prescription.
-For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions.
+We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018, NagarMessinaRettegno2018}, that uses a different point-particle and spin prescription.
+For matter effects, this model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions.
 It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.
 
-Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible.
-In Ref.~\cite{BNSPE}, an alternative highly parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}.
-Unfortunately, direct comparisons between this code and traditional samplers have not been published for BNS systems using the same waveform models.
-It is therefore not clear how accurate this code is at measuring quantities such as the tidal parameters.
+Although these EOB models are more expensive than frequency-domain models, parameter estimation results are still possible.
+In Refs.~\cite{BNSPE, O2Catalog}, an alternative parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}.
+The measured tidal parameter using EOB waveforms and \texttt{RapidPE} were broadly consistent with the results using frequency-domain waveforms and traditional samplers (see Fig.
+9 of Ref.~\cite{O2Catalog}).
+However, a direct comparison using identical waveforms has not yet been published for BNS systems.
 As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved.
@@
 Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation.
-It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
-
-%This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}.
-Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation.
-It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.
+It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas}, multi-banding~\cite{VinciguerraVeitchMandel2017}, and relative binning~\cite{Zackay:2018qdy}.
 
 We organize the paper as follows.
-In Section II we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space.
-In Section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model.
-We then compare the accuracy and speed of the surrogate to the original model in Section IV.
-In section Section V we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817.
-Finally, we discuss future improvements in Section VI.
-We also give the expressions for the TaylorF2 base model in the Appendix.
+In Section~\ref{sec:eob} we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space.
+In Section~\ref{sec:surrogate} we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model.
+We then compare the accuracy and speed of the surrogate to the original model in Section~\ref{sec:performance}.
+In Section~\ref{sec:pe} we verify that the surrogate can correctly extract the parameters of injected waveforms.
+Finally, we discuss future improvements in Section~\ref{sec:discussion} and give the expressions for the TaylorF2 base model in the Appendix.
 
@@
 \end{equation}
-with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$.
-Here, $\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor (Eq.~(11) of Ref.~\cite{Dietrich:2017feu}), which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.
-This enhancement factor is 1 for low orbital frequencies, and increases to $\sim 2$ as $\ell$ times the orbital frequency approaches the $f$-mode frequency.
-
-Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
+with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/m_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$.
+Here, $\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor~\cite{Steinhoff:2016rfi, Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.
+
+Introducing a Keplerian orbital frequency defined from the radius variable as $M\Omega \equiv u^{3/2}$, using the notations (dropping the indices $A,B$)
+\begin{subequations}
+\begin{align}
+	x = \frac{\hat{\omega}_{0\ell}}{m\Omega} \,, \; \epsilon \equiv \frac{256\eta}{5} \left( \frac{\hat{\omega}_{0\ell}}{m} \right)^{5/3} \,, \; \hat{t} \equiv \frac{8}{5\sqrt{\epsilon}} \left[ 1 - x^{5/3} \right] \,,
+\end{align}
+\end{subequations}
+the dynamical enhancement factor  of~\cite{Steinhoff:2016rfi, Dietrich:2017feu} reads
+\be\label{eq:kldyn}
+	\hat{k}_{\ell, \rm{dyn}}(u; \omega_{0\ell}) = a_{\ell} + b_{\ell} \left[ f(x) +\sqrt{\frac{\pi}{3 \epsilon}} x^{2} \mathcal{Q} \right]\,,
+\ee
+with $a_{\ell}$, $b_{\ell}$ constant coefficients with relevant values $\{a_{2}, b_{2}\} = \{1/4, 3/4\}$ and $\{a_{3}, b_{3}\} = \{3/8, 5/8\}$.
+These expressions are used in practice for $\ell= m$.
+Here, we have rewritten the first two resonant terms of Eq.~(11) of~\cite{Dietrich:2017feu} as
+\be\label{eq:deff}
+	f(x) = x^{2} \left[ \frac{1}{x^{2} - 1} + \frac{5}{6} \frac{1}{1 - x^{5/3}} \right] \,.
+\ee
+Although the two terms are individually divergent\footnote{This requires to single out the near-resonance region in a numerical implementation.} at the resonance $x=1$, the function $f(x)$ is actually  regular, taking the value $f(1) = -1/12$.
+The last term in~\eqref{eq:kldyn} above reads
+\begin{align}
+	\mathcal{Q} &= \cos\left( \frac{3}{8} \hat{t}^{2} \right) \left[ 1 + 2 F_{S} \left( \frac{\sqrt{3}}{2\sqrt{\pi}} \hat{t} \right) \right] \nonumber\\
+	& - \sin\left( \frac{3}{8} \hat{t}^{2} \right) \left[ 1 + 2 F_{C} \left( \frac{\sqrt{3}}{2\sqrt{\pi}} \hat{t} \right) \right] \,,
+\end{align}
+with $F_{S}$, $F_{C}$ the Fresnel sine and cosine functions.
+This term is regular at resonance $\hat{t} = 0$.
+The enhancement factor~\eqref{eq:kldyn} is 1 for low orbital frequencies, and increases to $\sim 2$ as $\ell$ times the orbital frequency approaches the $f$-mode frequency.
+
+Tidal effects also enter the radiative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
 \begin{equation}
@@
 Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
+
 The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the gravitational-wave flux (Eqs.~(5) and~(6) of Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived.
@@
 
-%Let $\Omega$ be the instantaneous orbital frequency, $r$ be orbital separation, $p_r$ be the radial momentum, $f_s$ be the waveform sampling frequency in Hz, $\omega_{\textrm{peak}}^{\textrm{NR}}$ be the NR fit for the (2,2)-mode amplitude peak of a BNS merger defined in Eq.~(2) of Ref.~\cite{Bernuzzi:2015rla}~\footnote{Note that this fit was tuned to NR simulations only for $0 \leq \kappa_2^T \leq 500$, where $\kappa_2^T$ is defined by Eq.~(1) of Ref.~\cite{Bernuzzi:2015rla}.
-In our model, for $\kappa_2^T >500$ we use the value of this fit at $\kappa_2^T=500$.}.
-Then the numerical integration of the orbital dynamics stops when any of the following events happens: (i) $\Omega$ peaks, (ii) $p_r$ becomes positive, (iii) $\dot{r}$ becomes positive, (iv) $\dot{p}_r$ becomes positive, (v) $r$ does not evolve more than a 1000th of the leading-order estimate for $\dot{r}$, that is $\propto 64/5 \nu/r^4$, (vi) $2\Omega$ reaches $\omega_{\textrm{peak}}^{\textrm{NR}}$, (vii) $2M\Omega$ reaches $\pi M f_s$.
-In order to avoid triggering these stopping conditions at low frequency, where residual orbital eccentricity due to imperfect quasicircular initial conditions could be present, we exploit the $\omega_{\textrm{peak}}^{\textrm{NR}}$ fit to define a radius below which the stopping conditions can be triggered; such threshold value is taken to be $1.5 (\omega_{\textrm{peak}}^{\textrm{NR}}/2)^{-2/3}$.
-
 \subsection{Spin-induced quadrupole-monopole terms}
-
-\red{[check notations have been introduced]} For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are based on the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins.
+\label{subsec:QM}
+
+For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are based on the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins.
 It includes consistently spin-orbit interactions up to 3.5PN order and spin-spin interactions up to 2PN order.
@@
 This effect is quadratic in the spins, appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin~\cite{HarryHinderer2018}.
-Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only, see~\cite{}.}, at the moment the effect is only consistently included at the leading 2PN order.
-We leave for future work an extension to higher orders\footnote{\red{complete this} We note that TaylorF2 models do include all spin-square terms at 3PN, see App.~\ref{sec:taylorf2}.
-They are also present in the IMRPhenomPv2+NRTides model, but not in the SEOBNRv4ROM+NRTides model.
-TEOBResumS is also limited to the leading 2PN order.}.
+Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only~\cite{Levi:2015ixa}.}, at the moment the effect is only consistently included at the leading 2PN order.
+An extension to higher orders is left for future work\footnote{We note that TaylorF2 models do include all known spin-square terms at 3PN (see Appendix~\ref{sec:taylorf2}).
+They are also present in the PhenomPNRT and SEOBNRT models.
+TEOBResumS was recently updated to include these next-to-leading terms~\cite{NagarMessinaRettegno2018}.}.
 
@@
 
-%In the BBH model, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks at a prescribed time and has prescribed amplitude and frequency, hence $t_{\textrm{match}}^{22}=t_{\textrm{peak}}^{\textrm{amp}}$.
-Here, however, we do not impose such corrections because of limited input from NR BNS simulations, therefore we have little control over the behavior of Eq.~(\ref{hlminspplunge}) in the late inspiral.
-This motivates our choice of $t_{\textrm{match}}^{22}$ in Eq.~(\ref{tmatch}), with the goal of reducing to the BBH prescription in the non-tidal limit and avoiding non-monotonic growth of the GW frequency in the late inspiral.
-To help bracket the stopping time, we exploit a fit for the (2,2)-mode peak frequency of NR BNS simulations~\cite{Bernuzzi:2015rla} to estimate the final orbital frequency of the EOB inspiral-plunge dynamics.
-
-%Ref.~\cite{Dietrich:2017feu} carried out comprehensive comparisons of this model to state-of-the-art NR simulations of coalescing nonspinning BNSs with different mass ratios and equations of state.
-The model can accurately describe the GW phasing up to the last cycle before merger for systems with soft equation of state, while improvements are still needed for systems with stiff equation of state.
-\red{[[To what accuracy?
-Is this paragraph already in the intro and can be deleted?]]}
-
-In BNS simulations, the waveform rapidly tapers to zero amplitude in 1--2 gravitational-wave cycles after reaching peak amplitude.
+In BNS simulations, the waveform rapidly decreases in amplitude in 1--2 gravitational-wave cycles after reaching peak amplitude.
 The BNS system then either undergoes prompt collapse or post-merger oscillations of the remnant.
-We model this tapering in terms of analytic functions for the amplitude $A^{\textrm{post-mrg}}(t)$ and phase $\phi^{\textrm{post-mrg}}(t)$ such that $h_{22}^{\textrm{post-mrg}}(t) = A^{\textrm{post-mrg}}(t) \exp{[i\phi^{\textrm{post-mrg}}(t)]}$.
+We do not attempt to model post-merger oscillations, and instead model an approximate peak emission and a subsequenct tapering to zero.
+This signal will be represented in terms of analytic functions for the amplitude $A^{\textrm{post-mrg}}(t)$ and phase $\phi^{\textrm{post-mrg}}(t)$ such that $h_{22}^{\textrm{post-mrg}}(t) = A^{\textrm{post-mrg}}(t) \exp{[i\phi^{\textrm{post-mrg}}(t)]}$.
 
@@
 \section{Surrogate model}
-
-In this Section we describe how we decompose the EOB waveform into smooth, slowly-varying functions and train a surrogate model for these functions.
+\label{sec:surrogate}
+
+In this section we describe how we decompose the EOB waveform into smooth, slowly-varying functions and train a surrogate model for these functions.
 We will work with the frequency-domain waveform $\tilde h(Mf; \bx)$.
@@
 
-%\begin{algorithm}[H]
-%\caption{Uncertainty Sampling}
-%\label{alg:uc}
-%\begin{algorithmic}[1]
-%\State {\bf Input:} Initial surrogate 
-%\For{$i = 1 \to N_{\rm New}$}
-%
-%\State Construct the GPR error estimator (Eq.~\eqref{eq:errorrms}).
-In practice this can be done by specifying the samples $\bx$, the hyperparameters $\btheta$, and dummy data for $\by$ since Eq.~\eqref{eq:errorrms} does not depend on $\by$.
-%
-%\State Find $({\bm x}_{\rm max}, \epsilon_{\rm rms})$ that maximizes Eq.~\eqref{eq:errorrms} over the parameter space.
-%
-%\State Add ${\bm x}_{\rm max}$ to the list of samples: ${\bm x} = [{\bm x}, {\bm x}_{\rm max}]$.
-%
-%\EndFor
-%\State {\bf Output:} Updated surrogate
-%\end{algorithmic}
-%\end{algorithm}
-
 For $i = 1, \dots, N_{\rm new}$:
@@
 
-\subsection{Injection runs}
-
 As a final test of the surrogate, we perform several parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template.
-We use the design sensitivity PSDs for the two advanced LIGO detectors~\cite{Aasi:2013wya} and the advanced Virgo detector~\cite{virgoPSD}.
+We use the design sensitivity PSDs for the two advanced LIGO detectors and the advanced Virgo detector~\cite{Aasi:2013wya}.
 We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$ corresponding to the EOS named MS1b in~\cite{ReadLackey2009}.
@@
 For the prior we choose the sky position and orientation to be uniform on the unit sphere; the distance to be uniform in co-moving volume; the masses, spins, and tidal parameters to be uniform in the ranges $m_{1, 2} \in [0.7, 2]M_\odot$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$; and the mass ratio is restricted to the range $q\in[0.5, 1]$.
-As was done in~\cite{BNSPE}, we sample the data and waveform at \red{4096}~Hz such that the Nyquist frequency is \red{2048}~Hz, and integrate the likelihood function in the interval \red{$[20, 2048]$~Hz}.
-We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal} which requires approximately \red{$10^x$} iterations to produce \red{10,000--20,000} statistically independent samples.
-
-We note that the \texttt{SEOBNRv4T} waveform model still has some signal above \red{2048~Hz} (see Fig.~\ref{fig:hoff}).
-We therefore also performed runs with a sampling frequency of \red{8192}~Hz and integrated the likelihood function in the interval \red{$[20, 4096]$}~Hz.
+As was done in~\cite{BNSPE}, we sample the data and waveform at 4096~Hz such that the Nyquist frequency is 2048~Hz, and integrate the likelihood function in the interval $[20, 2048]$~Hz.
+We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal} which requires $\sim 10^7$ iterations to produce 10,000--20,000 statistically independent samples.
+
+We note that the \texttt{SEOBNRv4T} waveform model still has some signal above 2048~Hz (see Fig.~\ref{fig:hoff}).
+We therefore also performed runs with a sampling frequency of 8192~Hz and integrated the likelihood function in the interval $[20, 4096]$~Hz.
 However, this did not have a noticeable effect on the recovered posteriors.
@@
 The parameters are identical except for the distances which are chosen such that the network SNRs are 30 (blue), 60 (green), or 120 (red).
-The black PDF only includes samples for the SNR=120 injection that satisfy $q>0.9$.
-The shaded regions give the 90\% credible interval.
+The black PDFs only include samples for the SNR=120 injection that satisfy $q>0.9$.
+The shaded regions give the 90\% credible intervals.
 The dashed black line represents the injected parameter value.}
@@
 
-\subsection{GW170817}
-
-Having confirmed that the surrogate accurately recovers the injected waveform parameters, we now do parameter estimation for GW170817.
-In the PE paper and catalog papers~\cite{}, the original \texttt{SEOBNRv4T} model was used with the parallelized PE code RapidPE.
-However, there were uncertainties about its ability to accurately recover the tidal parameter.
-
-We use the priors discussed in~cite{O2CatalogPaper}.
-
 %% ________________________________________________________
 \section{Discussion and future work}
+\label{sec:discussion}
 
@@
 This aligned-spin model, \texttt{SEOBNRv4T}, incorporates the tidally induced $\ell=2$ and 3 multipole moments as well as the effect of dynamical tides as the excitation approaches the $\ell=2$ and 3 $f$-mode frequencies.
-We have achieved mismatches of no more than \red{$4.5 \times 10^{-4}$} and phase errors of \red{$\lesssim 1$~rad} up to the merger frequency.
+We have achieved mismatches of no more than $4.5 \times 10^{-4}$ and phase errors of $\lesssim 1$~rad up to the merger frequency.
 These are sufficient to not bias results in any of the parameters.
 
-The evaluation time has a flat cost of \red{$\sim 0.02$~s} to perform the GPR interpolation at each node.
+The evaluation time has a flat cost of $\sim 0.01$~s to perform the GPR interpolation at each node.
 The rest of the time is spent resampling the waveform with spline interpolation.
-For a starting frequency of 20~Hz, this takes \red{0.13~s} when the waveform is matched with data uniformly sampled at 4096Hz.
+For a starting frequency of 20~Hz, this takes 0.13~s when the waveform is matched with data uniformly sampled at 4096Hz.
 For current parameter estimation codes, this is sufficient.
-However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, or multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017}.
-Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^4)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.
+However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017}, or relative binning~\cite{Zackay:2018qdy}.
+Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^3)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.
 
 The hierarchical method presented here, where we begin with the analytic TaylorF2 reference model then make a surrogate of the residual, can be used to further improve the waveform model.
-For example, one could build a surrogate for numerical BNS simulations using \texttt{SEOBNRv4T\_surrogate} as the base model, and constructing a surrogate of the residual.
+For example, one could build a surrogate for numerical BNS simulations using \texttt{SEOBNRv4T\_surrogate} as the base model, then constructing a surrogate of the residual.
 Such a model would have the accuracy of EOB below $\sim 400$~Hz and the accuracy of numerical simulations for the last several cycles before merger.
@@
 
-The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
-\begin{equation}
-A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} v^{-7/2} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)v^{2} \right],
-\end{equation}
-where $v=(\pi M f)^{1/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{LAL}~\cite{lal} \red{[explain]}. The phase has the schematic form
+The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform,
+\begin{equation}
+A_{\rm F2} = -\sqrt{\frac{5\pi\eta}{24}} v^{-7/2} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)v^{2} \right],
+\end{equation}
+where $v=(\pi M f)^{1/3}$ is the standard PN parameter. The sign is set following the conventions of \texttt{LAL}~\cite{lal}. The phase has the schematic form
 \begin{align}
@@
 We separated point particle terms $\Phi_{\rm F2}^{\rm PP}(\eta)$, spin terms $\Phi_{\rm F2}^{\rm Spin}(\eta, \chi_{1}, \chi_{2})$ and tidal terms $\Phi_{\rm F2}^{\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$.
-We include point particle terms to the highest know order of 3.5PN~\cite{}.
-
-The spin terms can be decomposed into spin-orbit and spin-spin terms (ignoring cubic-in-spin contributions at 3.5PN~\cite{}) as $\Phi_{\rm F2}^{\rm Spin} = \Phi_{\rm F2}^{\rm SO}  + \Phi_{\rm F2}^{\rm SS} $.
-We include spin-orbit terms up to 3.5PN~\cite{}.
-For spin-square terms, as the \texttt{SEOBNRv4T} waveform does not include yet 3PN spin-spin effects (see Sec.~\ref{}), we stop at  the leading 2PN in the \texttt{TaylorF2} waveform and ignore the known 3PN terms.
+We include point particle terms to the highest know order of 3.5PN (see e.g.~\cite{Blanchet2014Review}).
+
+The spin terms can be decomposed into spin-orbit and spin-spin terms (ignoring cubic-in-spin contributions at 3.5PN) as $\Phi_{\rm F2}^{\rm Spin} = \Phi_{\rm F2}^{\rm SO}  + \Phi_{\rm F2}^{\rm SS} $.
+We include spin-orbit terms up to 3.5PN~\cite{Blanchet2014Review}.
+For spin-square terms, as the \texttt{SEOBNRv4T} waveform does not include yet 3PN spin-spin effects (see Sec.~\ref{subsec:QM}), we stop at  the leading 2PN in the \texttt{TaylorF2} waveform and ignore the known 3PN terms.
 The main reason for doing this is because including only common physical effects in both \texttt{SEOBNRv4T} and \texttt{TaylorF2} makes the residual $\Delta\Phi(Mf; \bx)$ smaller and have less variation, making it easier to fit at a given error requirement.
@@
 
-The point particle terms coefficients up to 3.5PN can be found e.g. in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}. The non-zero coefficients read
-\begin{align}
-	a_{2} &= \frac{55 \nu }{9}+\frac{3715}{756} \nonumber\\
+The point particle terms coefficients up to 3.5PN can be found e.g. in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, and have the structure
+\be
+	\Phi_{\rm F2}^{\rm PP}(\eta) = \sum_{k=0}^{7} a_{k} v^{k} + \mathcal{O}(v^{8}) \,.
+\ee
+The non-zero coefficients read
+\begin{align}
+	a_{0} &= 1 \nonumber\\
+	a_{2} &= \frac{55 \eta }{9}+\frac{3715}{756} \nonumber\\
 	a_{3} &= -16 \pi \nonumber\\
-	a_{4} &= \frac{3085 \nu ^2}{72}+\frac{27145 \nu }{504}+\frac{15293365}{508032} \nonumber\\
-	a_{5} &= -\frac{65 \pi  \nu }{9}-\frac{65}{3} \pi  \nu   \ln v +\frac{38645}{252} \pi   \ln v +\frac{38645 \pi }{756} \nonumber\\
-	a_{6} &= -\frac{127825 \nu ^3}{1296}+\frac{76055 \nu ^2}{1728}+\frac{2255 \pi ^2 \nu }{12} \nonumber\\ 
-	& -\frac{15737765635 \nu }{3048192}-\frac{6848  \ln v }{21}-\frac{640 \pi ^2}{3}-\frac{6848 \gamma_{E} }{21} \nonumber\\ 
+	a_{4} &= \frac{3085 \eta ^2}{72}+\frac{27145 \eta }{504}+\frac{15293365}{508032} \nonumber\\
+	a_{5} &= -\frac{65 \pi  \eta }{9}-\frac{65}{3} \pi  \eta   \ln v +\frac{38645}{252} \pi   \ln v +\frac{38645 \pi }{756} \nonumber\\
+	a_{6} &= -\frac{127825 \eta ^3}{1296}+\frac{76055 \eta ^2}{1728}+\frac{2255 \pi ^2 \eta }{12} \nonumber\\ 
+	& -\frac{15737765635 \eta }{3048192}-\frac{6848  \ln v }{21}-\frac{640 \pi ^2}{3}-\frac{6848 \gamma_{E} }{21} \nonumber\\ 
 	& +\frac{11583231236531}{4694215680}-\frac{13696 \log (2)}{21} \nonumber\\
-	a_{7} &= -\frac{74045 \pi  \nu ^2}{756}+\frac{378515 \pi  \nu }{1512}+\frac{77096675 \pi }{254016}
-\end{align}
+	a_{7} &= -\frac{74045 \pi  \eta ^2}{756}+\frac{378515 \pi  \eta }{1512}+\frac{77096675 \pi }{254016} \,,
+\end{align}
+where $\gamma_{E}$ is Euler's constant.
 

### 2019-01-30 paper/response_to_referee.txt
Start of response to referee.

Re: DZ12205
   Surrogate model for an aligned-spin effective one body waveform model
   of binary neutron star inspirals using Gaussian process regression
   by Benjamin D. Lackey, Michael Pürrer, Andrea Taracchini, et al.

Dear Dr. Lackey,

The above manuscript has been reviewed by one of our referees.
Comments from the report appear below for your consideration.

When you resubmit your manuscript, please include a summary of the
changes made and a brief response to all recommendations and
criticisms.

Yours sincerely,

Joshua C. Sayre
Associate Editor
Physical Review D
Email: prd@aps.org
https://journals.aps.org/prd/

----------------------------------------------------------------------
Report of the Referee -- DZ12205/Lackey
----------------------------------------------------------------------

In this manuscript, the authors construct a surrogate for SEOBNRv4T
using Gaussian process regression (GPR). The surrogate is intended for
spin-aligned binary neutron star waveforms, but the techniques they
introduce are flexible enough to accommodate precession in the future.

One of the novel aspects of their work is to fit to residuals between
SEOBNRv4T and TaylorF2 thus reducing the requirement for very high
accuracy. They keep their TaylorF2 model as "ignorant" as SEOBNRv4T to
make the phase residual smaller as explained in App. A. The EOB model
includes dynamical tides from the induced f-mode resonances. They use
the Yagi et al. I-Love-Q relations to reduce the dimension of the
parameter space from 14 to 5 for the training sets. They then sample
this 5-D "cube" as a Latin Hyper Cube, thus an initial 128 points + 32
corners suffices to train the surrogate. The "training" over the 5
parameters is done using GPR which is the other novel contribution of
this work to the relevant literature. GPR has the additional nice
property that it "provides a convenient estimate of its own
uncertainty" to use the authors' words. The authors have the option of
retraining the surrogate using 400+400 more points in the 5-D space.
Their final choice for the best surrogate is generated from 960
training sets using a Matern kernel for the covariance. This surrogate
has a mismatch of at worst 4.5E-4 and a phase disagreement of <~ 1
radian at f <~ 2 kHz. The surrogate is finally shown to be good enough
for parameter extraction in LIGO-Virgo.

The article is well written, the figures are of high quality and the
work is original. In my opinion, this article will make a nice
addition to the existing literature on this topic. I recommend it for
publication modulo minor questions/comments/corrections that I list
below.

1) In the abstract, it is written the surrogate takes 0.13 seconds to
evaluate from 20Hz, but given the flat cost of 0.01s of GPR
interpolation at each frequency node and given the fact that there are
roughly 40 frequency according to Sec. III.C, isn't the overall run
time more like 40 x 0.01 + 0.13 = 0.53 second?

 - The confusion is due to bad phrasing on our part. When we wrote 0.01s to evaluate each node, we meant 0.01s to evaluate all 40 nodes. We have changed the wording in Section IV.B (paragraph 2) and Section VI (paragraph 2) to clarify this.

2) In the left column of page 2, the sentence following Refs. [32-36]
contains the word "instead" twice, which seems redundant.

 - We removed the second "instead".

3) In the upperleft part of page 3, it is written that "up to 0.85 for
unequal-mass (up to 0.98 for equal-mass)". Where are these values
obtained from? According to Ref[24] of this manuscript, the equal-mass
case seems to have been extended to possibly 0.994.

4) The authors may wish to add https://arxiv.org/abs/1812.02744 to
their list of papers that accommodate tidal effects to EOB in the
first paragraph of page 3.

 - We have added this reference to the second paragraph of Section II.A.

5) Immediately after Eq.(9), the authors cite Damour-Nagar-Villain
2012. It might be useful to also cite recent work of
https://arxiv.org/abs/1805.07266 that have now extended DNV.

6) Typo in the lowerleft of page 5 where the word "subsequent" has
been written as "subsequenct".

 - We have fixed this typo.

7) Could the authors comment on why they have chosen 12M instead of
15M in Eqs.(22-23)? Is the matching to omega_{match} made to be
asymptotic to ensure smoother Fourier modes?

8) Could the authors comment on the bifurcation of Phi in Fig. 1? It
seems to me that there would be such a difference between the
Lambda_A=Lambda_B=1000 corners of the parameter space vs.
Lambda_A=Lambda_B=0 corners, but the count does not quite add up. Do
these branches each contain 16 waveforms? It might be nice to add a
brief explanation about this to the caption of Fig. 1 for the sake of
the curious, but less familiar, readers.

 - The bifurcation is due to the two extremes in parameter space for q. The lower 16 curves correspond to q=1, and the upper 16 curves correspond to q=1/3. This results because the mass ratio dominates over the other 4 parameters. We added a sentence to the caption of Fig. 1 to clarify this.

9) For Sec IV. B regarding run times, could the authors explicitly
provide an overall run time for the 20Hz non-spinning BNS run
including the flat cost of ~0.01sec per frequency node for computing
the GPR fits? According to my estimation in point 1) above, this would
be 0.53 second which interestingly coincides with the value given by
Table IV of 1812.02744 in their App. B. Looking at that table begs the
question: do the authors have any idea how long their surrogate would
take to run after being trained over a set starting from 10Hz?

 - We apologize for the confusion that also resulted in comment 1 above. The times shown in Fig. 8 are in fact the total evaluation times for the waveform including evaluating the GPR fits at all nodes and resampling the final waveform at uniform frequency spacing (~0.01s for 80Hz, 0.13s for 20Hz, and ~1 s for 10Hz).

 We have clarified that the evaluation times shown are the *total* surrogate evaluation time in Section IV.B. sentence 1 of paragraph 2. In the same paragraph, we also clarified that the 0.01s is the time needed to evaluate all 40 nodes (not just 1) as discussed in comment 1 above.

10) In Sec V. the authors show that their surrogate faithfully
recovers the injected parameters. First of all, I wanted to commend
the authors for sticking with presenting the more pain-in-the-neck q=1
case despite the fact that this value sits on the upper bound of the q
prior. Have the authors repeated their parameter estimation
computations for "friendlier" values of q such as 0.8? If so, do the
PDFs look less asymmetric and more importantly are the centres of the
peaks of the PDFs closer to the injected values?

 - At the moment one of our internal LIGO reviewers is performing an analysis with a less symmetric masses. Those runs, however, are not complete.

### 2019-01-30 paper/paper.tex
Start of response to referee.

@@
 The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
-In this work, instead of using a reduced basis, we will instead use the fact that the Fourier-transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model.
+In this work, instead of using a reduced basis, we will use the fact that the Fourier-transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model.
 We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform.
@@
 
-This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu} for binaries that contain NSs.
+This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu,AkcayBernuzziMessina2018} for binaries that contain NSs.
 The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case.
@@
 The BNS system then either undergoes prompt collapse or post-merger oscillations of the remnant.
-We do not attempt to model post-merger oscillations, and instead model an approximate peak emission and a subsequenct tapering to zero.
+We do not attempt to model post-merger oscillations, and instead model an approximate peak emission and a subsequent tapering to zero.
 This signal will be represented in terms of analytic functions for the amplitude $A^{\textrm{post-mrg}}(t)$ and phase $\phi^{\textrm{post-mrg}}(t)$ such that $h_{22}^{\textrm{post-mrg}}(t) = A^{\textrm{post-mrg}}(t) \exp{[i\phi^{\textrm{post-mrg}}(t)]}$.
@@
 Bottom two panels: the phase $\Phi(Mf; \bx)$ and residual $\Delta\Phi(Mf; \bx)$ relative to TaylorF2 as defined in Eq.~\eqref{eq:phaseres}.
+For the phase, the upper 16 curves correspond to $q=1/3$, and the lower 16 curves correspond to $q=1$.
 Vertical dashed lines represent the frequency nodes $MF_j$ where the residuals are interpolated as functions of $\bx$ using Gaussian process regression.
@@
 
-In Fig.~\ref{fig:timing}, we show the surrogate evaluation time as a function of the starting frequency $f_{\rm min}$ for an equal mass $1.4M_\odot$--$1.4M_\odot$ binary evaluated on a 3.5~GHz Intel Xeon processor.
+In Fig.~\ref{fig:timing}, we show the total surrogate evaluation time as a function of the starting frequency $f_{\rm min}$ for an equal mass $1.4M_\odot$--$1.4M_\odot$ binary evaluated on a 3.5~GHz Intel Xeon processor.
 The waveform was sampled with a uniform frequency spacing corresponding to a sampling rate of 4096~Hz.
-At large starting frequencies the evaluation time is limited by the $\sim 0.01$~s needed to calculate the GPR fits at each frequency node.
+At large starting frequencies the evaluation time is limited by the $\sim 0.01$~s needed to calculate the GPR fits for all 40 frequency nodes.
 As $f_{\rm min}$ decreases, the waveform becomes longer and a smaller frequency spacing is needed.
@@
 
-The evaluation time has a flat cost of $\sim 0.01$~s to perform the GPR interpolation at each node.
+The evaluation time has a flat cost of $\sim 0.01$~s to perform the GPR interpolation for all 40 nodes.
 The rest of the time is spent resampling the waveform with spline interpolation.

### 2019-05-14 paper/paper.tex
Changed AEI address.
Load utf8 package
Address item 3

@@
 
+%%% contains utf-8, see: http://inspirehep.net/info/faq/general#utf8
+\usepackage[utf8]{inputenc}
+
 %% macros for comments
@@
 \affiliation{
-Max Planck Institute for Gravitational Physics, Albert Einstein Institute, D-14476 Golm, Germany
+%Max Planck Institute for Gravitational Physics, Albert Einstein Institute, D-14476 Golm, Germany
+Max Planck Institute for Gravitational Physics (Albert Einstein Institute), Am M\"uhlenberg 1, Potsdam-Golm 14476, Germany
 }
@@
 The EOB approach to the general-relativistic 2-body problem, first described in Ref.~\cite{Buonanno:1998gg}, has proven successful in modeling the dynamics and GW emission of compact binaries.
-State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of NR simulations of aligned-spin BBH systems for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to 0.98 for equal-mass) binaries.
+State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of NR simulations of aligned-spin BBH systems for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to 0.994 for equal-mass) binaries.
 The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}.

### 2019-05-15 paper/paper.tex
Addressing 5) and 7) of the referee response

@@
 where the point-mass piece $h_{\ell m}^{\textrm{pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and the tidal piece $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs.
-(A14)-(A17) of Ref.~\cite{Damour:2012yf}.
+(A14)-(A17) of Ref.~\cite{Damour:2012yf} (see also the recent extension~\cite{Banihashemi:2018xfb}).
 Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.
@@
 It is given by
-\begin{equation}
+\begin{equation}\label{eq:ampwindow}
 W(t) = \frac{1}{1+\exp{[(t- t^{22}_{\textrm{match}}-15M)/\tau]}}\,.
@@
 We define this frequency evolution as
-\begin{align}
+\begin{align}\label{eq:omegatransition}
 \omega(t) = \omega_{\rm match} - \Delta\omega \exp\left[ -(t - t_{\rm freq}) / (12M) \right],
@@
 
+The choice of the constant transition timescales $15M$ in~\eqref{eq:ampwindow} and $12M$ in~\eqref{eq:omegatransition} is partly arbitrary and was made primarily for robustness, keeping in mind that this termination is effective.
 Although this effective post-merger model has not been fit to NR BNS simulations, it is in reasonable qualitative agreement with the post-merger behavior of the two equal-mass nonspinning NR BNS simulations that were analyzed in Ref.~\cite{Hinderer:2016eia}.
